\chapter{On Rudras going towards the heavens on the celestial car}

1-19. Vy\^asa said :-- O mighty armed Kuru! What you have asked me just now, I also asked the same thing to N\^arada, the lord of the Munis and he gave me the following reply :-- O Vy\^asa! What shall I say to you on this point more than this that a doubt occurred to me also in my former days. The question that you have put to me today rose in my mind before; and I went to my father Brahm\^a, of endless energy and asked to him thus :-- O Lord! O Father! Whence is this whole Brahm\^anda born? Have You created it? Or is it Visnu or Mahe\'svara? O all pervading soul! Who is there in this Brahm\^anda fit to be worshipped? O Lord of the world! Who is the top-most Lord ruling over everything? Kindly say. O Brahm\^an! I am plunged in this sea of Maya and perils; my heart is agitated with doubts; hence it is not appeased in any place of pilgrimage ; or in thinking any Deva or in practising any S\^adhan or in any other object. O Sinless one! Give me the answers duly and thus remove my doubts. O Tormentor of foes! Unless the highest truth is not known, peace is not found. This heart, distracted in various ways, cannot rest fixed on one subject. Whom am I to remember? Whom to worship? Where to go? Whom to praise? Who is the Supreme God this Universe? I do not understand these things. O Satyavat\^i's son! Hearing these my serious queries, Brahm\^a, the grand-Sire of beings, replied to me as follows :-- O highly illustrious son! What more shall I say to you than this that even Visnu is unable to answer your questions; so difficult are they indeed! O great intelligent one! Nobody that is attached to the world knows anything about this. Those who are unattached to this world, who are free from any envy, these who are without desires and calm, those highsouled ones know the secret of all this. In former days when all was water, water everywhere and all things, moving and non-moving were destroyed, when five elements were sprung, then I was also born from the lotus navel of Visnu. Then not seeing Moon, Sun, trees, or mountains or anything and sitting on the centre (Karnik\^a) of the lotus thought thus :-- When I am born in this great ocean of waters? Who has created me? Who is now my protector?

And Who will be my Destroyer when this cycle ends? There is no earth distinctly visible anywhere here; on what, then, this mass of water rests? Lotus is termed Pankaja because it springs from mud and dirt; so unless there exists the earth underneath with mud and dirt, how this lotus will come out here! Now let me try and find out where is the root of this lotus, where is the mud and dirt? If this be found, then the earth will be also certainly there. Thus thinking, I dived underneath the water and searched for one thousand years but could not find earth anywhere, when the celestial voice entered my ears ``Practise tapasy\^a (austerities).'' Hearing this celestial voice, I sat on the lotus, my birth place, and practised tapasy\^a for one thousand years.

20-30. Next, the celestial voice came again ``Create.'' Hearing this, I became quite confounded and began to think within myself ``now what am I to create? What to do?'' After this, the two terrible Daityas Madhu and Kaitabha came to me and affrighted me saying ``Fight with us.'' I became quite terrified and holding the stem of the lotus, I got down within the water. There I saw a wonderful person, sleeping on the Ananta serpent. He was of a deep blue colour like a rain-cloud, wearing yellow clothes, four-armed, garlanded with forest flowers, and the Lord of this whole Universe. On the four arms of this Mah\^a Visnu there were conch-shells, disc, club, and lotus and other weapons. I saw this Achyuta Purusa, sleeping on the Ananta serpent bed, motionless and under the influence of Yoga Nidr\^a. I then thought within myself ``What am I to do?'' Not being able to find out any other way, I recollected the Dev\^i who was then of the nature of sleep and began to praise Her. The auspicious Dev\^i Yoga Nidr\^a, whose form could not be determined, immediately left the body of Visnu and decorated with divine ornaments, began to shine in the air. After She left the body of Visnu, Visnu immediately got up. And He fought terribly for five thousand years with the D\^anavas Madhu Kaitabha; then by the grace of the Bhagavat\^i, He extended His own thighs and then, on those thighs, He slew the two demons. Where Visnu and myself were standing, Rudra Deva came also and joined with us. Then we three saw the beautiful Dev\^i in the celestial space.

31-40. We three, then, commenced to chant hymns to Her and She gladdened our hearts by Her gracious look and said :-- ``O Brahm\^a! O Visnu! O Rudra! The two great Daityas are slain. Now forsake your laziness and do your respective works of creating, preserving, and destroying the Universe; create your own abodes, and live in happiness; create by your respective lordly powers, the fourfold beings.'' Hearing the Dev\^i's gentle sweet words, we spoke :-- ``O Mother! There is no wide earth here; all is one mass of infinite ocean. No five elements, no five tanm\^atr\^as,

no sensual organs, no Gunas, nothing exist here; how can we then execute the works of creation, etc. Hearing our words, the Dev\^i smiled. Immediately there came from the sky overhead a beautiful aerial car. The Dev\^i said :-- ``O Brahm\^a! O Visnu! O Rudra! Get in this car without any fear. To-day I will show you one wonderful thing.'' At Her word, we got into the beautiful car without any fear. It was decorated with various gems and jewels, bedecked with pearls, emitting sweet tinkling sounds of bells and looking as the abode of the celestials. Seeing us seated without any fear, She made the car get high up in the sky by Her force.

Thus ends the second chapter of the third skandha on Brahm\^a, Visnu and Rudra's going towards the heavens on the celestial car, given by the Dev\^i in the Mah\^apur\^ana \'Sr\^imad Dev\^i Bh\^agavatam of 18,000 verses by Maharsi Veda Vy\^as.