\chapter{On the questions put by Janamejaya regarding Krisna's incarnation}

1. Janamejaya said :-- O V\^asaveya! the Chief amongst the Munis, the Ocean of all knowledge, O the Sinless One! the Lord and Well Wisher of our families, I come to you with a mind to ask you certain questions.

* V\^asav\^i is the name of the mother of Vy\^asa.

2-3. I heard of yore, but I do not know why was that illustrious son of Sûrasena, that powerful \^Anakadundubhi\footnotemark, the pious and illustrious Vasudeva, the father of \'Sr\^i Krisna, the incarnation of the God Hari Himself, and who was worshipped even by the Gods, thrown into prison by Kamsa?

\footnotetext{\^Anakadundubhi is the epithet of Vasudeva, father of \'Sr\^i Krisna, since at Vasudeva's birth, drums called \^Anakas and Dundubhis were resounded in the sky.}

4-5. What faults had he and his wife Devak\^i committed? Why that Kamsa the descendant of Yay\^ati, killed the six infant sons of Devak\^i? And for what reason did the God \'Sr\^i Hari incarnate Himself as the son of Vasudeva in the prison house of Kamsa?

6-11. How was that Lord of the Universe, \'Sr\^i Bhagav\^an, and the Ruler of the Y\^adava clan, taken to Gokula? Why was He, born of a Ksattriya family, generally recognised as of the Gop\^ala (cow-herd) clan? Why were His father Vasudeva and mother Devak\^i thrown into prison? And why did not \'Sr\^i Krisna, of indomitable prowess, and capable to create and preserve the worlds, could release instantly, father and mother from their imprisoned state? I cannot conceive that the so-called Fate could have any influence on such high souled persons, the father and mother of the Supreme Being, \'Sr\^i Krisna; who were those sons of Vasudeva that were killed by Kamsa? And who was that girl child who when struck by Kamsa on a slab of stone instantly rose up above the sky, assuming the form of Astabhuj\^a, the eight armed Goddess? O Sinless One! Kindly explain unto me how did \'Sr\^i Hari manage to perform the house-holder's duties, when he had married the several wives? and what were those glorious deeds that he did in His this incarnation and how did he finally pass away from his mortal physical coil? My mind sinks into an ocean of confusion, when I hear of the several things done by \'Sr\^i Hari; some times I find the deeds, not capable of being done by any other than the Supreme Being Himself and sometimes I hear of deeds that can be done by  an ordinary man. And, therefore I cannot decide whether Vasudeva was the Incarnation of God or an ordinary being. Be pleased to remove these doubts from my mind and describe the life of V\^asudeva in its true light.

12-14. In days of yore, the two sons of Dharma were the two best amongst the Risis, and were the Devas Nara and N\^ar\^ayana. They were very high souled persons and they performed severe austerities for long extending years. They were born as part incarnations of Visnu; and, for the good of the world, did they, in the Vadarik\^a'srama, perform penances, controlling their six passions, and free from desires.

15. The all-knowing sages N\^arada and others say that the well known Arjuna and \'Sr\^i Krisna of indomitable prowess were the two part incarnations of those two ancient Munis Nara and N\^ar\^ayana.

16. How came those two Devas Nara and N\^ar\^ayana to be born in the two bodies of Krisna and Arjuna, though they did not relinquish their previous bodies!

17. And also when those two Munis had attained liberation, their goal in their Yogas, by performing severe penances, how could they again be born in other bodies!

18-19. If any \'Sûdra dies performing his own religion, he takes up a Vai\'sya body in his next incarnation; if any Vai\'sya died so, he takes up a Ksattriya body and a Ksattriya when adhering to his own rites and ceremonies, dying takes up a Br\^ahman body in his next incarnation. And if a Br\^ahmin be free from desires and resorts to the path of peace, when he dies, he becomes free from incarnations and is saved from this disease of getting into the world.

20-21. Now the reverse seems to take place in the case of Nara and N\^ar\^ayana. In spite of withering up their bodies by hard penances, they took up Ksattriya bodies. Under what influence of Karma, did they take up up birth when they were Yogis? Or might they, the Br\^ahmins, become Ksattriyas owing to some curse? Whatever it may be, kindly remove my doubts, explaining to me their causes.

22. It is heard that the Y\^adava clan suffered destruction through the curse of a Br\^ahman and in spite of \'Sr\^i Krisna being the incarnation of the Supreme Being, his family died of the effect of curse from G\^andh\^ar\^i.

23. How was it that Pradyumna was stolen away by \'Sambara, the lord of the Asuras; and for what purpose when Vasudeva, the Deva of the Devas was present, how was it that his son was stolen away from the lying-in room? This seems impossible.

24. Why did not Vasudeva see, with His inner vision, the stealing away of his son from the impregnable walls of His fortress-like mansion in Dw\^ark\^a? (And could thus have prevented this!)

25-39. O Muni! After V\^asudeva had gone to Heaven, his wives were plundered by the dacoits on the way. I am very much in doubt on this point. Also I cannot understand why did this event take place just after His translation to Heaven? Again how was \'Sr\^i Krisna, who was Visnu's Full incarnation, born into this world to take off the vicious load of this earth, and to destroy the evil-doers, terrified and He fled through the fear of Jar\^asandha, evacuated His kingdom of Mathur\^a and went with His armies and friends to Dw\^ark\^a? V\^asudeva came here to destroy the evil miscreants and to conserve the religion; how then did He not, as He was all knowing, previously kill those dacoits, who afterward stole and plundered His wives? Was it, that He, being Omniscient did not know those dacoits? He protected the P\^andavas, no doubt, who were high souled, righteous and virtuous; but I cannot understand how did He consider the high souled virtuous persons like Bh\^isma, Drona and others as loads of earth and slay them. The devotees of \'Sr\^i Krisna, observing good customs, conducts and practices, Yudhisthira and his brothers performed the R\^ajasuya sacrifice, according to rules, giving various offerings to the Br\^ahmins and depended entirely on V\^asudeva; still, O Muni! they suffered terrible hardships; whither were their virtuous deeds by this time? what horrible sins did they commit that they had to suffer pains and troubles in the assembly. The highly merited Draupad\^i arose from the midst of sacrificial fire and is born of the part of Laksm\^i, pure and devoted to \'Sr\^i Krisna. How did such a glorious woman meet with incomparable terrible pains often and often; how could she be caught hold of by her hair on her head by Duh's\^asan; and carried to the royal assembly, when she was in her menses and extremely terrified and harassed? How did she come to be a slave of Matsya R\^aj in his kingdom Vir\^at, and though actually crying aloud like a female osprey (eagle) she was highly insulted by K\^ichaka! Alas! how could Draupad\^i be stolen away by Jayadratha, though latterly released by the P\^andavas? What evil deeds did the P\^andavas commit in their previous births, that they had to befall under so many telling difficulties?

40. O high minded Muni! My ancestors performed the R\^ajasûya sacrifice and still they fell under so many serious difficulties. Kindly explain the causes of these to me.

41-42. If it be urged, that they suffered so many serious calamities, due to their actions in their former lives, that seems impossible. Because they are born of the Devas; thus arises my doubt; be pleased to explain to me.

Again how the P\^andavas, the sons of Kunti, of good conduct, and knowing the illusory nature of the world, why did they, out of pretence kill Bh\^isma, Drona and others?

43. This appears a riddle to me that these P\^andavas were led to the extermination of their race, being inspired by Hari V\^asudeva, to this horrible act.

44. Rather to live on begging alms and to live on rice, growing wild or without cultivation, or to live as an artisan or artist than to kill the valiant warriors unlawfully in a battle, simply out of voluptuous greed.

45. O Best of the Munis! You have preserved this extirpated race by producing the Goloka sons (i. e. sons born by other persons of women after their husbands are dead) of indomitable prowess.

46. And why did my honoured father, born of Uttara in this respected family , encircle a snake round the neck of a Br\^ahmin ascetic?

No body, born of a Ksattriya family, shews signs of hatred and jealousy towards a Br\^ahmin. Is it that my father shewed such feeling to that ascetic, who took the vow of silence!

O Best of the Munis -- These things and lots of others are troubling my mind with many grave doubts. O merciful saint! You know every thing; be kind enough to quell this the disturbed state of my mind.

Thus ends the First Adhy\^aya in the Fourth Book of \'Sr\^i Mad Dev\^i Bh\^agavata Pur\^ana of 18,000 verses by Maharsi Veda Vy\^asa.