\chapter{On \'Sukr\^a's going to Mah\^adeva to get the Mantra}

1. Vy\^asa said :-- When the Devas retired from the battlefield, \'Sukr\^ach\^arya addressed the Demons thus :-- O best of the Demons! Please hear, what had been told to me by Brahm\^a in days of yore.

2-3. Jan\^ardan Visnu is coming here to slay all the Demons, He killed before Hirany\^aksa, the king of the Asuras, in His Boar incarnation. He killed Hiranyakasipu by assuming His Man-Lion appearance; now too, he will kill all the Daityas, no doubt, with great enthusiasm.

4-5. Now my mantra force will not be of any avail before \'Sr\^i Hari. And without my help you will not be able to conquer the Devas. Therefore, O Demons, wait for some time for me; I will start today to the presence of \'Siva Mah\^adeva to obtain from him the Mantra \'Sakti.

6. I will return, as early as possible, getting the Mantra from \'Sr\^i Mah\^adeva; and with the help of the power of those great Mantras I will be able to protect you thoroughly.

7. The Daityas said :-- O best of the Munis! We are now defeated and our forces are well nigh exhausted; how will we be able to stay on this earth and expect your return for so long a time?

8. Those who were of great strength amongst us, they all are slain; now we are left very few in number. In this crisis it is not advisable and auspicious for us to remain here in this battlefield.

9. \'Sukr\^ach\^arya said :-- Better you all stay here until I return, getting the requisite Mantra; remain peaceful and engaged in asceticism.

10. The heroes apply one or other of the following four measures :-- conciliation, alliance, gifts and bribery, partitions and sowing dissensions and punishment or open attack according to the conditions of time, place strength and circumstances.

11. Intelligent and well wishing persons do serve their enemies even in time of distress; but no sooner they find that their strength and army have increased, than they try at once to kill all their enemies.

12. Now, therefore, pretend your meekness and allegiance and adopt peace and remain in your own place until my arrival.

13. O Demons! I will come back with the Mantra from Mah\^adeva. I will fight again with the Devas, by that veritable mantra power.

14. O king! Thus firmly making up his mind to have the Mantra, \'Sukr\^ach\^arya went to Mah\^adeva.

15. The D\^anavas sent Prahl\^ada to the Devas for conciliation. The Devas regarded Prahl\^ada truthful and they all trusted on him.

16. Prahl\^ada with Asuras addressed thus the Devas with gentleness and humility.

17. O Immortals! We all have abandoned our weapons and armour. Now we desire to wear barks of trees and practise asceticism.

18. The Dev\^as took Prahl\^ada's words to be true and deserted from further fight; were freed of anxiety and felt all delighted.

19. On the Daitya's leaving their weapons, the Gods desisted from battle, went to their respective places and freely gave themselves up to pleasure and amusements.

20. The Daityas controlled themselves and practised asceticism in the Ka\'syapa's hermitage, expecting the return of \'Sukr\^ach\^arya.

21. \'Sukr\^ach\^arya went to Kail\^asa and made respectful obeisance to the God Mah\^adeva. Mah\^adeva enquired of the cause of his coming.

22. \'Sukr\^ach\^arya said :-- I ask for Mantras, O Deva! from you that are not possessed by Brihaspat\^i, for the defeat of the Devas, and the victory of the Asuras.

23. Vy\^asa said :-- On hearing his words, the all-knowing \'Sankara \'Siva began to think what He would do in this matter.

24. Certainly, this is with the revengeful object of attacking the Devas, for their defeat and for the victory of the Asuras that this Muni has come.

25. The Devas ought to be protected by me; thus thinking, \'Sankara Mahe\'svara advised him an entirely difficult tapasy\^a to practise.

26. Full one thousand years he would practise tapasy\^a with feet upwards and head downwards, he will have to inhale the smoke of burnt husk. Then he will get the Mantra and his desires will be fulfilled.

27. That would be done; thus saying to \'Sankara, \'Sukr\^ach\^arya practised that excellent vow, peaceful and inhaling the smoke of husk to get that Mantra.

28-29. The Devas came to know that \'Sukr\^ach\^arya is practising his vow and the Daityas have become arrogant. They then counselled.

30. And came to the conclusion, took up weapons and arms and went to the Daityas, ready to fight.

31. The Daityas, seeing the Devas dressed in armour and holding weapons and coming from all sides, became very much afraid and anxious.

32. The Daityas, seeing this all on a sudden, were attacked with fear and began to address the Devas, proud with their army, in words, full of good meanings and morals.

33. O Devas! We have abandoned our arms; we are now armless; our spiritual guide is in his tapasy\^a, you gave us words of fearlessness; why then you have come now dressed in full armour and with armies to kill us.

34. O Devas! Where is your truth. And where is your religion acccording to \'Sruti? It is stated in \'Sruti never to kill the weaponless, the fear stricken, and the refugees.

35. The Devas said :-- You sent your good \'Sukr\^ach\^arya, out of pretence to acquire the Mantras; your asceticism is veiled under a deceitful object. Therefore we will fight with you certainly.

36. Be ready now and dress yourself with your arms and ammunitions, ``Lo! Whenever you get any loop hole in your enemies, catch hold of it and kill your enemy.'' This is the eternal religion.

37. Vy\^asa said :-- On thus hearing the reply from the Devas, the Daityas after consideration quitted that place and fled away with terror.

38. And they took refuge under the \'Sukr\^ach\^arya's mother. She saw the Daityas very much fearstricken and at once guaranteed to them protection from fear.

39. The mother of K\^avya \'Sukr\^ach\^arya said :-- Don't fear; don't fear; cast away fear. O D\^anavas! In my presence, no fear can overtake you.

40. The Asuras on hearing her words were free from anxiety and pain and remained in that hermitage, in no way now bewildered or agitated, though they had no arms.

41-42. Here the Devas, seeing the Daityas flying away, pursued them and entering the hermitage were ready to kill the Daityas, not taking into account what strength they gained there. The mother of \'Sukr\^a warned the Devas not to kill; but, inspite of her hindrance, they began to slay the Daityas.

43. Seeing the Daityas thus attacked, the mother was furiously irritated and told them she would make all of them overpowered by sleep by her tapas strength, clarified intellectual force.

Note :--Is it by the asphyxiating gas? Or, by poisonous vapours or liquids?

44. So saying she sent the Goddess of sleep who at once overpowered the Gods and made them all lie down on the grounds senseless. Indra with the other Devas lay there dumb, and miserable.

45. On seeing Indra thus stupefied by sleep, the Bhagav\^an Visnu told Indra to enter into His body. He would then carry him to another place and he will be better.

46-47. Indra entered into the Visnu's body; and, under His protection, he became free from sleep and fear. On seeing Indra thus sheltered by Visnu and fearless, the mother of K\^avya spoke.

48. O Indra! I will devour you with Visnu today by my Tapas force. All the Devas will presently see all this and my extraordinary power.

N.B.: Is it by making the earth go down, causing a fissure, thus engulfing the whole army? Or by blowing them up? Or by showering jets of poisonous gas or hot water or vapour from all sides.

49. Vy\^asa said :-- O king! No sooner the mother spoke thus, than both Indra and Visnu were both stupefied under her magical spell, superior thought power, and a thorough learning of the art of warfare.

50. The Devas, seeing them very much overpowered and bewildered, were greatly struck with wonder; they became desperate and began to cry aloud.

51. Indra, on hearing the Devas cry aloud, told Visnu, O Madhusûdana! I am more bewildered in particular than yourself.

52. O M\^adhava! No need of any further consideration. Before this wretch, inflated with pride by her tapasy\^a, burns us, better cause her death as early as possible.

53. When thus requested by Indra who was very much perplexed, Bhagav\^an Visnu quickly remembered his Sudar\'sana disc, casting aside the thought that it is hateful to kill a woman.

54-55. The disc, the ever obedient weapon of Visnu appeared instantly at his remembrance; and Visnu, becoming angry as prompted by Indra held the disc in His hand, and, hurling it off on the \'Sukr\^a's mother, severed off her head quickly. The god Indra became very glad at this.

56. The Devas became free from sorrow, got very much pleased and heartily exclaimed victory to Hari and worshipped Him and began to chant His praises.

57. Indra and Visnu then became free from all troubles; but they began to fear that Bhrigu (\'Sukr\^a's father) would curse them terribly and without fail.

Here ends the Eleventh Chapter in the Fourth Book of \'Sr\^i mad Dev\^i Bh\^agavatam, the Mah\^a Pur\^anam of 18,000 verses by Maharsi Veda Vy\^asa, on \'Sukr\^a's going to Mah\^adeva to get the Mantra.