\chapter{On Bhrigu's curse and the dialogue between \'Sukr\^ach\^arya and the Daityas}

1. Vy\^asa said :-- O king! On seeing Visnu killing his wife, and thus committing a dreadful atrocious act, the Bhagav\^an Bhrigu was very sorry and began to tremble with anger and addressed Madhusûdana thus.

2. Bhrigu said :-- O Visnu! You have done an extremely sinful act. O intelligent one! Knowing it, you have done so; what a great wonder! Nobody dreams of the murder of this Br\^ahman's daughter; and you have committed it in deed, black and white.

3. O Deva! The Mah\^arsis declare you to be S\^attvic, engaged in preservation; Brahm\^a to be R\^ajasic (creator) and \'Sambhu \'Siva to be T\^amasic (destroyer). Why then is the contrary thing visible in this case?

4. Why have you become T\^amasic? Why have you done this heinous crime? O Visnu! The females are never to be killed; this is a known fact; then why have you killed this woman without any fault.

5-6. You have done a very execrable act. What shall I do to you? It is mete that I would curse you. O Great Sinner! You have pained me very much and made me very weary. I will not curse Indra. You always assume a deceitful appearance and behave like a black cruel serpent; your mind is all full of wickedness; I will curse you.

7. O Jan\^ardana. Those Munis who call you S\^attvic are fools; I have seen today that your ways are exceptionally vicious and T\^amasic.

8. O Visnu! I curse you now to take frequent births, suffer very frequently in different wombs, in the earth and thus suffer the pains of remaining in the wombs.

9. O king! Therefore whenever religion subsides in the world, Bhagav\^an Visnu incarnates frequently in this human world, due to the curse of Bhrigu.

10. The king said :-- O best of the Munis! Did that Mah\^atm\^a Bhrigu again marry and lead a householder's life when his wife was thus killed by the lustrous disc Sudar\'sana.

11. Vy\^asa said :-- O king! The sacrificer Bhrigu angrily cursed Hari and next took that severed head and quickly placed it over the body as before and said :--

12-14. O Dev\^i! Visnu has slain you today; I will make you regain your life just now. If I am acquainted with all the Dharmas and if I have practised these in my life and if I have spoken truth always, then dost thou regain your life by my religious merit. Let all the Devas witness my power and strength. If I know the True, if I have studied all the Vedas and if I have realised the Knowledge of the Vedas, then I, sprinkling your body with this cold water, charged with my mantras, will revive you.

15. Vy\^asa said :-- O king! Sprinkled by the water by Bhrigu, his wife regained her life and rose up at once and felt herself glad and smiled.

16. All the persons and living creatures seeing her stand, as if awoken from her sleep, at once exclaimed from all sides ``well done, well done!'' thanked Bhrigu and his wife very much and highly praised them.

17. Thus seeing the fair complexioned wife regain her life through Bhrigu, Indra and all the Devas were very much struck with wonder.

NOTE :-- Thought power and inhalation of some medicines might have revived her.

18. Indra then adressed the Devas :-- Now the mother of \'Sukra has regained her life through Bhrigu; but when \'Sukr\^ach\^arya will receive the mantra after his severe tapasy\^a, we do not know what terrible harm; he would do to us!

19. Vy\^asa said :-- O king! Though the deep sleep that overtook Indra had now left him, yet he became very unhappy, remembering the severe asceticism of \'Sukra's tapasy\^a and his receiving mantras.

20. Thinking over deeply in his mind, Indra called his daughter Jayant\^i and spoke to her with affection.

21-22. Go! Daughter! I have given thee over to the ascetic K\^avya. Do this service for me, go to K\^avya and help him in his asceticism and bring him under your control. By whatever acts the Muni be pleased, dost thou do at once and please him in every way and thus remove my fear.

23. The large-eyed beautiful Jayant\^i, hearing her father's words, went to the hermitage and there saw the Risi inhaling or drinking the smoke.

24. On seeing the body of the ascetic, and remembering her father's words she brought the leaves of the plantain trees and began to fan him.

25. The intelligent Jayant\^i with no excitement used to bring the pure, clear, cool, and well scented water and place it, with great devotion, before the sage for his drinking.

NOTE: Here Indra bribed by giving away his daughter.

26. When the sun was over his head, she used to prepare some protection, with her clothes, from the sun and thus made the shade for him. Thus, in various ways, did she begin to serve the Muni, herself remaining steady in her religion.

27. She brought sweet and ripe fruits proper according to the \'S\^astras and used to place them before him for his repast.

28. In the performance of his daily duties, she used to collect the Ku\'sa grass of the span of thumb and fore finger, and flowers, white and yellow and placed them before him.

29. For the Muni's bedding to lie down and sleep, she used to gather soft, green leaves and with them she used to keep his bed ready; and when the Muni laid himself down, she used to fan him gently.

30. Never did she exhibit for fear of curse, any gesture or posture calculated to disturb his mind.

31. The sweet-tongued, thin lady used to praise \'Sukr\^ach\^arya in pleasing and favourable terms.

32. When the Muni awoke, she placed water before him to wash his face and hands. Thus serving the Muni, did she stay in the hermitage.

33. The fear stricken Indra used to send messengers to fathom the mind of that self controlled Muni.

34. Thus did Jayant\^i serve for good many years the Muni, being void of anger and observing duly her celibacy.

35. In this way one thousand years passed duly, the God Mah\^adeva was very pleased and addressed \'Sukr\^ach\^arya that He had come to give him boons that he desire.

36-37. The God Mah\^adeva said :-- Whatever there exists in this universe, whatever you see with your eyes and whatever cannot be described in words, you will be the lord of all these and the conqueror of all. There is no doubt in this. You cannot be killed by any creature; you will be the ruler of all beings and will be reckoned as the best of the Br\^ahmanas.

38-39. Vy\^asa said :-- The God Mah\^adeva disappeared then and there, after granting him the boons. Then \'Sukr\^ach\^arya saw Jayant\^i and said to her :-- O lady of beautiful hips! Who and whose daughter art thou? What is the desire in your mind? What for you have come here? O one of beautiful thighs! what is your duty?

40. O beautiful eyed! I am very pleased with all that you have done for me. What do you want? O one of good vows! Ask boons from me; I will grant them even if they be difficult.

41. On hearing this, Jayant\^i's face became bright with joy; she said :-- whatever I desire, you know that already by force of your Tapasy\^a.

42. The K\^avya said :-- Your mental desires I know; still mention them yourself particularly; I will do good to you in any way you like; I am pleased with your services.

43-44. Jayant\^i said :-- O Br\^ahmin! I am the daughter of Indra and I am the younger of my brother Jayanta. Father has given me over to you; I now like to live with you. Kindly fulfil my desires now. O blessed one! You better take me and enjoy me according to Dharma. This is my desire.

45. \'Sukr\^ach\^arya said :-- O thou of large and handsome hips! Better enjoy with me for ten years, according to your liking, without being visible to anybody.

46-48. Vy\^asa said :-- Thus saying, \'Sukr\^ach\^arya went to his home and, according to religious observances, married Jayant\^i and lived with her for ten years under the shade of M\^ay\^a and unobserved by anybody. On the other hand, the Daityas, hearing that \'Sukr\^ach\^arya returned home successful getting his desired Mantra, were very glad and went to his house to pay their respects to him. But \'Sukra was then living with Jayant\^i; hence the Asuras could not see him.

49. Then they got very much dejected, sorry and were drooped in spirits; they searched for him again and again.

50. Not being able to see \'Sukra under the shade of M\^ay\^a, the Daityas were very sorry and got afraid and then returned to their own residences.

51. On the other hand, the Devas, knowing that \'Sukra was holding intercourses with Jayant\^i, Indra, the God of Devas, spoke to the Deva Guru

Brihaspat\^i. O Guru! Advise us what to do under the present circumstances and relieve us from the impending difficulties.

52. O Br\^ahmana. Better go today to the D\^anavas and do that by which our honour may be maintained. You better put the D\^anavas under some charm by your magical spell and think and do good to us.

53. On hearing Indra's words and knowing that \'Sukra is now enjoying with Jayant\^i, Brihaspat\^i went to the D\^anavas, putting on the appearance of \'Sukr\^ach\^arya.

54. Going to the Demons, he called on them very politely and sweetly. The Asuras said that \'Sukr\^ach\^arya had come in presence of them.

55. The Demons were very glad and bowed down before \'Sukr\^ach\^arya and could not make out that he was the false \'Sukr\^ach\^arya, under the m\^ayic charm of Brihaspat\^i.

56. Then the false \'Sukr\^ach\^arya asked the welfare of the Daityas and said :-- I went so long for your good; I performed severe tapasy\^a and satisfied the God \'Sambhu and acquired from him the Mantra, the sacred knowledge, and that I will frankly explain to you.

57. On hearing this, the Demons were very pleased and, knowing that the Guru had been successful, were filled with joy.

58. They saluted him with joy and began to live quietly and freely and without pains, having got rid of fear from the Devas.

Here ends the twelfth Chapter in the Fourth Book of \'Sr\^i Mad Dev\^i Bh\^agavatam, the Mah\^apur\^anam of 18,000 verses by Mah\^arsi Veda Vy\^asa on Bhrigu's curse and the dialogue between \'Sukr\^ach\^arya and the Daityas.