\chapter{On the prayer and hymns to the Dev\^i}

1. Vy\^asa said :-- O King! Then Indra and all the Devas became very glad to see the great Mahis\^asura slain; they all began to praise and chant hymns to the World-Mother.

2-33. The Devas said :-- It is by thy Power that Brahm\^a becomes able to create this world, Visnu, to preserve, and Mahe\'svara to destroy during the Pralaya time (the Great Dissolution) of this universe. But when they are bereft of Thy Power, they are quite unable to do such. Therefore, O Dev\^i! Thou art undoubtedly the Prime Cause in the preservation and destruction of this whole Universe. O Dev\^i; Thou art, in this world, Fame, idea and ideal, memory; Thou art the goal, mercy, compassion, faith, constancy, earth; Thou art Kamal\^a, the Mantra Ajap\^a, respiration and perspiration, nourishment, Jay\^a, Vijay\^a (the destroyer of obstacles; a name of Durg\^a); Thou art contentment, correct notion, measure, intellect, Ram\^a (Laksm\^i), (wealth), knowledge, forgiveness, beauty, intelligence; Thou art the \'Sakti (power) of Rudra, Thou art Girij\^a and the Energy of God Um\^a and all other forces in this universe; this is known to everyone in the three worlds. Without any or all of these forces, no one is able to perform any action. Thou art the Supreme Cause of all this world. Therefore everything rests on Thee. If Thou wert not the upholding Power, how could Kurma (in the Tortoise Incarnation) and Ananta have upheld this world? O Mother! Wert Thou not this Earth, could all these world-load of things have rested on the sky? O Mother! Those human beings that worship Brahm\^a, Visnu, Rudra, Moon, Fire, Yama, the God of Death, V\^ayu, Gane\'sa, and the other Devas, they are certainly deluded by Thy M\^ay\^a. Could all those Devas do any action or any favour without Thy Energy? O Mother! Those that offer in any Sacrifice, a profuse quantity of ghee (clarified butter) as oblations to the several Devas, they are certainly conceived to be of very narrow views; Wert Thou not the Sv\^ah\^a, could it have been possible for those Devas to get the offered oblations at that very instant? Certainly, therefore, they are fools and ignorant persons. There is no doubt in this. O Mother! Thou givest the several objects of nourishment and enjoyment to all the beings in this universe by Thy parts (the several transformations of these material things); it is Thou that nourishest the Devas, Thy devotees, as well as the others (the D\^anavas, according to their Karma). O Mother! As the owner of any garden plants, with pleasure, the beautiful trees in his garden for his delight and, finding some of them not to bear any fruits or leaves or of a bitter taste, does not cut them off by their roots, so, O Dev\^i! Thou hast brought into existence these Daityas out of their inferior Karmas and Thou art supporting them. Knowing that the Daityas like to enjoy the celestial nymphs, Thou hast, out of compassion, killed them by their arrows in the battlefield, to afford them facilities in their re-birth in Heavens and thus to enjoy the Deva women which they could not have got in any other possible way. Therefore this Thy dealings with them are to fulfil their intentions and not to kill them. O Mother! It is a great wonder that to kill these Asuras Thou hadst to assume this Divine Body; Thou couldst have done so by Thy mere will. It seems that this act of Thine is but a mere Pastime. There is no other cause for this. O Dev\^i! Those human beings that do not worship Thee in this dreadful age of K\^al\^i, they are certainly deceived by the cunning Pur\^ana makers who have deluded them to worship Hari and Hara, who are Thy creations. Oh! What an amount of evil has befallen to those poor souls! O Dev\^i! Those men know that the Devas, tormented by the Asuras, are Thy devotees, and yet they worship them; certainly such fellows, holding the lighted torches in their hands, plunge deep into the darkest waterless wells. O Mother! Thou art the Vidy\^a (Blissful Intelligence) and Thou grantest pleasure and liberation; Thou art the Avidy\^a, (Great Delusion) and thus Thou causest bondage and pain in this world. O Mother! Thou only destroyest the affliction of the human beings; those that want liberation worship Thee, and those that are ignorant and attached to worldly enjoyments do not worship Thee. What more can be said than this, that Brahm\^a, Visnu, Mahesa and the other Devas incessantly worship Thy adorable lotus-feet; but those men that are of dull intellect and are mistaken, they do not meditate Thy feet and, therefore, they come again and again into this ocean of world. O Chandik\^a! It is through the grace of the dust of Thy lotus-feet that Brahm\^a, Visnu, and Mahe\'svara are creating, preserving and destroying this universe. Therefore, O Goddess! Those men that do not serve Thee, are certainly very unfortunate. O Mother of the Universe! Thou art the Goddess of speech of the Suras and the Asuras; thus if Thou didst not dwell in their mouths, they would not have been able to utter a single word; therefore, O Goddess! How can men speak, when they are thus deprived of Thee! O Mother! It is due to the curse of Bhrigu Muni that Hari takes several incarnations as Fish, Tortoise, Boar, Man-Lion, and deceitful Dwarf Incarnations; all these show clearly the dependence of Hari,

*N.B. - The Devas and the Daityas are the opposite polarities of the same creation.

How, then, can they avoid the fear of death when they serve these dependent incarnations! O Mother! It is well known that the male generative organ of \'Sambhu, the Mah\^adeva fell unto the ground, owing to the curse of Bhrigu Muni, when he went to the hermitage of the Risis. How can, then, happiness come in this world or in the next, to those who worship such a \'Sambhu who wears human skulls on His body! O Dev\^i! Those that worship Gane\'sa, born of the above qualified Mah\^a Deva are awfully mistaken; they are especially quite ignorant of Thee, the Goddess of the Universe, that can be easily worshipped and that can give the fourfold aims of human existence. O Dev\^i! It is out of Thy kindness that Thou hast slain with Thy arrows the enemies and thus hast translated them into Heavens; otherwise they would have certainly gone down to Hell owing to their own Karmic effects. Brahm\^a, Hari, Hara and the other Devas cannot realise Thy greatness; how can, then, ordinary men know Thee, when they are deluded by immeasureably strong S\^attva, R\^ajas and T\^amas qualities. O Mother! Those who do not worship Thy lotus-feet as very hard to be brought within this mind and therefore worship this visible Sun and Fire, they cannot grasp the Essence of the Vedas, demonstrated by hundreds of passages of \'Sruti; they are deluded and simply suffer pains. O Mother! I think that the influences of Thy S\^attva, R\^ajas and T\^amas qualities are widely known in this world; those qualities as taught in various deluding schools of Tantras by various persons, stimulate people to the worship of Visnu, Mahe\'svara, Sun and Gane\'sa and thus detract them from worshipping Thee. O Mother! Those that detract thus the Br\^ahmanas from worshiping Thy lotus-feet and advise them through the \^Agamas, to worship Hari, Hara and others, Thou dost not get angry with them, rather Thou dost shew Thy kindness to them and make them widely celebrated as possessing the occult powers of enchanting, bringing others under their control, or attracting towards them various other persons. In the Satya Yuga, S\^attva Guna was more powerful and therefore the untrue \'S\^astras could not rear their heads; but in this K\^al\^i Age, owing to the S\^attva Guna being not so powerful, the lower qualities have got preponderance; so these so-called clever Pundits instead of worshipping Thee, worship Hari, Hara and the other Devas, the products of their fancy and hide Thee. O Mother! Thou art the Brahm\^a Vidy\^a, the knowledge of the Supreme Consciousness, Thou givest liberation to Thy devotees when they succeed in their Yogas. Therefore the pure S\^attvik Muni meditate on Thee and Thee alone. Those that get themselves diluted in Thee, they are very blessed; what more to speak of them in their praise, they will no longer have to suffer any pains in their mother's wombs! O Mother! Thou art inherent as Chit \'Sakti (the power of consciousness) in the Supreme Spirit and therefore He is become manifest specially as this Great Cosmos and becomes known as the Creator, Preserver, and Destroyer of this world, fashioned out of five elements. O Dev\^i! What male can by his own power work out this Jagat Prapancha, enjoy it and move in this, without the aid of Thy power. O Bhagavat\^i! This universe has been created by Thee; Thou art, therefore, its Mother. The twenty-four Essences or Tattvas are inert; how can they without Thy Chit\'sakti, create this universe? O Dev\^i! Never can these senses and organs, endowed with Guna and Karma, do any work or bestow any fruits without Thy energy.

O Mother! Wert thou not Sv\^ah\^a, the instrumental cause in the sacrifice, how could the Devas have got their shares of the ghee offered in the Yajñ\^as by the Munis! Therefore, O Dev\^i! Thou art, no doubt, preserving this universe. O Bhagavat\^i! It is Thou that hast created this world in the beginning; it is Thou that art preserving the gods Hari, Hara and others; it is Thou that art destroying this universe. Therefore, O Br\^ahman! The Devas cannot know Thy deeds; how can, then, the men who are of dull intellect, know Thee. O Mother! Thou hast now saved the Devas by killing this terrible Mahis\^asura. O Mother! All the Vedas have not been able to know exactly all Thy movements; how can we, of dull intellect, praise Thee! O Mother! Thou has served our cause by killing our enemy, the wicked D\^anava, the inconceivable source of pain to all the world by this act of Thine, Thy fame has spread far and wide in this universe; therefore, O Thou of renowned prowess! Thou art the Mother of this world; save us, and maintain us by Thy mercy.

34-35. Vy\^asa said :-- O King! The Devas having praised the Dev\^i thus, the Dev\^i addressed them gently :-- ``O Devas! Say if you have any other difficult thing for Me to do; remember Me whenever any difficult crisis occurs to you; I will destroy that evil.''

36-42. The Devas said :-- ``O Dev\^i! All our purposes have been served when Thou hast killed lately our enemy Mahis\^asura. Now dost Thou do for us so that we can always recollect Thy lotus feet, and our Bhakti be firm and steadfast towards Thee. It is only the Mother the bears the thousand offences of the son; we, therefore, cannot say why men, knowing this, do not worship the Mother of the Universe. There are two birds always dwelling in this body, Jiv\^atm\^a (human soul) and the Param\^atm\^a (the Supreme Soul). They are so very intimate friends toward each other that they never separate. But there is no other third friend that can bear the faults of these two. Therefore the embodied soul that forsakes Thee, his friend, can never attain any welfare; what more to say on this! That vicious soul is very unlucky amidst the Devas and men, no doubt. He who on attaining this excellent human body, attained with much difficulty, does not remember Thee frequently by words and deeds, is certainly the vilest of men. O Dev\^i! Whether in times of distress or happiness, Thou art our Saviour; therefore dost Thou protect us with Thy best weapons. O Dev\^i! There is no other means of our security than the Grace of the dust of Thy Feet.''

43. Vy\^asa said :-- O King! The Devas having prayed to the Dev\^i thus, the Dev\^i vanished then and there. The Devas, seeing the disappearance of the Dev\^i, were sufficiently struck with surprise.

Here ends the Nineteenth Chapter of the Fifth Book on the prayer and hymns to the Dev\^i in \'Sr\^i Mad Dev\^i Bh\^agavatam, the Mah\^a Pur\^anam of 18,000 verses by Maharsi Veda Vy\^asa.