\chapter{On the Dev\={\i} Yaj\~na by \'Sr\={\i} Vi\d{s}\d{n}u}

1-3. The King spoke :-- O Grandfather! How did Vi\d{s}\d{n}u, the Powerful, the Cause of the world, perform the Dev\={\i} Yaj\~na? Whose help did he receive and what priests with their knowledge of the Vedas did he engage, kindly tell all these to me in special details. I will very attentively hear first this Dev\={\i} Yaj\~na, performed by Vi\d{s}\d{n}u; and then I will make arrangements to do that myself accordingly.

4. Vy\=asa said :-- O noble one! Hear in detail that very wonderful thing, how Vi\d{s}\d{n}u celebrated the Dev\={\i} Yaj\~na comformably to the rules laid down in the \'S\=astras.

5. When Brahm\=a, Vi\d{s}\d{n}u, and Mahe\'sa were each given powers by the Dev\={\i}, the Goddess of the Universe, and when they parted, these three Devas became free from their womanhood, while they were coming back in their aerial cars and turned out men, as they were before.

6. These three Devas, found the great ocean before them. They brought out world from it; and built, in that world, dwelling abodes; and they themselves began to live in some of them.

7. That world became fixed, steady and the supporter of all beings when the Goddess imparted the power of fixture, steadiness, and the power of supporting to the world. The earth, filled with marrow, then, became fixed and the great supporter by Her power.

8-10. O King! The name of this earth is Medin\={\i}, since it was made out of the marrow of the two Asuras Madhu and Kaitabha. This earth is termed Dhar\=a because it supports all; is termed Prithv\={\i} because it is very capacious; and it is called Mah\={\i} because it is great, since it supports so many beings. O King! the Ananta serpent is holding it on her thousand-hoods. To make the earth remain solid and compact, Brahm\=a built at places mountains. As iron nails in a log of wood, so these hills and mountains within this earth made it fixed. Therefore the Pundits call these mountains ``Mah\={\i}dhara,'' holder of the earth.

11. O King! Thus the golden Meru, the great mountain, many Yojanas wide, adorned with many golden mountain peaks was created.

12. Next Mar\={\i}chi, N\=arada, Atri, Pulastya, Pulaha, Kratu, Dak\d{s}a, and Va\'sistha were created by Brahm\=a; these are the Brahm\=a's mental sons (sons created by the sheer power of mind).

13. The son Ka\'syapa was born to Mar\={\i}chi and thirteen daughters were born of Dak\d{s}a. From these daughters and out of the seed of Ka\'syapa, various Devas and Daityas were born.

14. Then human beings, animals, serpents and many other classes were created. This is called the K\=asyap\={\i} \'Sr\={\i}sti or the Ka\'syapa's creation.

15. Next Sv\=ayambhava Manu sprang from the lower half of Brahm\=a; and the daughter named \'Satar\=up\=a came out of the left hand side of the Brahm\=a's body.

16. The two sons Priyavrata and Utt\=anap\=ada were born of Manu in the womb of \'Satar\=up\=a and the three daughters, very beautiful and fair complexioned, were also born of him.

17. Creating then, the Bhagav\=an, the lotus born Brahm\=a built the beautiful Brahma-loka, on the top of the Meru mountain.

18. Then the Bhagav\=an Vi\d{s}\d{n}u built the Vaikuntha city on the top of all the lokas or worlds to dwell with his consort Lak\d{s}m\={\i}.

19. Mah\=adeva, too, built the exceedingly beautiful Kail\=a\'sa and stayed there with his Bh\=utas and played with them at his will.

20. The third Loka termed Heaven was built on the top of Meru, decorated with various precious gems and jewels and stones. It was fixed as the abode of Indra.

21-22. When the great ocean was churned, P\=arij\=ata, the best of all the trees, the elephant Air\=avata with four tusks, the Heavenly cow yielding the milk of all desires, the Uchchai\'srav\=a horse and Apsar\=as, Rambh\=a and others, arose and were taken by Indra. These became the ornaments of Heaven.

23. The Moon and Dhanvantar\={\i}, the great phyisician also came out of the churning of the ocean. These surrounded with many other members began to shine, being situated above the Heavens.

24. O King! Thus the three varieties, human beings, Devas, and Tiryakas (birds, etc.) and their great subvarieties sprang up.

25. The four classes of J\={\i}vas, Andaja (born from the eggs), Svedaja (born out of sweats), Udbhija (plants, etc.) and Jar\=ayuj\=a (men, etc.) were created, being endowed with the fruits of their past auspicious or inauspicious Karmas, as the case may be.

26. Brahm\=a, Vi\d{s}\d{n}u, and Mahe\'sa began to play and walk about at their leisures in their respective spheres, after they finished all their creations.

27. Thus the wheel of creation being started, the Great God Bhagav\=an Vi\d{s}\d{n}u Achyuta remained in sport with Mah\=a Lak\d{s}m\={\i} in His own sphere Vaikuntha.

28. Then Bhagav\=an Vi\d{s}\d{n}u, while sitting one day in Vaikuntha, got in his mind the memory of that beautiful island, called Ma\d{n}i Dv\={\i}pa, adorned with precious stones, gems and jewels.

29-33. O King! In this Ma\d{n}i Dv\={\i}pa, the Bhagav\=an Vi\d{s}\d{n}u got the vision of the Mah\=a M\=ay\=a and the auspicious mantra. He now thought to celebrate a Yaj\~na in honour of the Dev\={\i}, when he recollected the highest Energy, the Great Goddess. He then went out of his abode and invited Brahm\=a, Mahe\'svara, Kuvera, Indra, Varu\d{n}a, Fire, Yama, Va\'sistha, Ka\'syapa, Dak\d{s}a, V\=amadeva, Brihaspati and others and began to collect lots of materials necessary for the Dev\={\i} sacrifice. Selecting a site S\=attvik, beautiful, and possessing great spiritual powers, Vi\d{s}\d{n}u erected, with the help of the great artists and engineers, a wide capacious sacrificial hall, and for the due celebration and fulfilment of it, appointed twenty seven priests; who pledged a solemn vow to complete it duly.

34. When the big altar and chiti (pile of wood for burning, stack of sacrificial bricks) were finished, the Br\=ahma\d{n}as began to recite slowly the Dev\={\i} mantrams with their root mantrams.

Note :-- Chiti, lit pile of wood for burning, is perhaps the M\=ul\=adh\=ara, the sacral plexus, where the fire called Kundalin\={\i} is first kindled by processes of Yoga.

35-37. Then the profuse quantities of ghee were offered as oblations to to the Sacred Fire. Thus when the Homa, (offering oblations of ghee to the fire) ceremony was finished elaborately and conforming to the rules of the \'S\=astras, the sweet and melodious Heavenly voice was heard in the air, addressing Vi\d{s}\d{n}u Bhagav\=an, thus :-- O Vi\d{s}\d{n}u! Let You be the supreme amongst the gods; honour and worship shall be Thine first; and you would be the most powerful of the Devas. Indra together with Brahm\=a, and the other Devas, all will worship You.

38-39. O Achyuta! (O Infallible One!) Those men on the earth that will be devoted to you, will certainly be endowed with power and you shall be the bestower of boons and all their desires. O Vi\d{s}\d{n}u! you will be the Supreme of the Devas and you will be the God of the gods; you will be the first and foremost in all the sacrifices and you will be worshipped by the sacrificers.

40-41. People will worship you; and you will favour them with the boons. O best of the Puru\d{s}as! When the Devas will be troubled by the Asuras, they will come and take refuge in Thee. You will be the Protector of all, there is no doubt in this. In all the Pur\=a\d{n}as and all this vast Vedas, You will be first worshipped.

42. O Ke\'sava! Wherever there will be decay and decline in religion, You will incarnate in your parts and preserve the religion.

43-44. O M\=adhava! Avat\=aras, renowned in all the worlds, will come down on earth as Your part incarnations in all sorts of wombs, in due order, and will be respected by all the high souled personages. O Madhus\=udana! Those Avat\=aras will be the best of all the Avat\=aras and will be famous all the Lokas, the worlds.

45. In all your Avat\=aras, you will get your attendants, the \'Saktis (females) drawing their energies from My parts; and they will serve your purposes.

46-47. V\=ar\=ah\={\i}, N\=arasimh\={\i}, etc., and various other \'Saktis of auspicious appearances, endowed with various weapons and decked with all the ornaments will serve as your attendants; no doubt in this. O Vi\d{s}\d{n}u! Always with their help and under the influence of My favour, you will no doubt be quite competent to serve the purposes of the Devas.

48. You should respect and worship all those powers by all means and with very great attention; never shew the slightest trace of pride to them; never do you thus insult them.

49. These \'Saktis, capable of bestowing all the desires, will be worshipped in Pratim\=as (images of clay, etc.) in the whole of India.

50. O Deva of the Devas! The fame of all these \'Saktis, as well as of You will be spread in the seven worlds and in the whole Universe.

51. O Hari! The human beings on this earth will constantly worship with selfish ends these powers and you, for the fructification of their desires.

52. Having various desires, men will, in your worship, present various offerings, recite the Veda mantrams, and repeat the names of you and the powers.

53. O Vi\d{s}\d{n}u! You will be the God of the Immortals and your glory will be enhanced by the worship offered by the human beings in the world as well as in the heavens.

54. Vy\=asa said :-- O king! The heavenly voice, thus, bestowing boons on Vi\d{s}\d{n}u, ceased. Vi\d{s}\d{n}u Bhagav\=an became very glad to hear this.

55-58. Then Hari, the God of Gods, completed duly the sacrifice and dismissed the Devas and the Munis, the sons of Brahm\=a. Then ascending on Garuda, (His V\=ahana), He went up to Vaikuntha with his followers. The Devas also went to their own respective places. The Munis also retired gladly to their own hermitages, all thunderstruck to see these things, conversing with each other about this sacrifice.

59. O king! All were filled with best devotion towards the Supreme Force, on hearing this clear beautiful, sonorous heavenly voice; then the Dv\={\i}jas, the Munis, and Mun\={\i}ndras began to worship with devotion, according to the Vedas, that Highest Force, the Supreme Goddess, giving all desired objects profusely in all the details.

Here ends the thirteenth chapter on the Dev\={\i} Yaj\~na by \'Sr\={\i} Vi\d{s}\d{n}u in the third Adhyaya in \'Sr\={\i}mad Dev\={\i} Bh\=agavatam; the Mah\=a Pur\=a\d{n}am by Mahar\d{s}i Veda Vy\=asa.



