\chapter{On the Daityas getting back their \'Sukr\=ach\=arya}

1. Vy\=asa said :-- Thus arguing in his mind, \'Sukr\=ach\=arya smilingly said to the Daityas :-- O Daityas! what for have you all been cheated by Brihaspat\={\i}, the Guru of the Devas, in my guise.

2. I am \'Sukr\=ach\=arya; you are my disciples; this man is Brihaspat\={\i}, wanting to serve the Gods. He has cheated you, there is no doubt in this.

3. This vain villainous person has assumed my form; do not put the least faith in his words. O Daityas! You are my disciples, follow unto me; leave this Brihaspat\={\i}, vain and arrogant.

4. The Daityas, on hearing his words were struck with wonder at the resemblance of the two personages and came to the conclusion that ``the person just come is the real \'Sukr\=ach\=arya.''

5. Then Brihaspat\={\i}, the false \'Sukr\=ach\=arya explained to the Daityas in plain and enchanting words that ``the second man just come is the Deva Guru Brihaspat\={\i}; he has come here in my guise. His object is to cheat you.

6. He has come here to cheat you and serve the purpose of the Devas; do not believe in his words.

7. What knowledge I have acquired from the God \'S\=ambhu, I am teaching it to you; I will make you, no doubt, victorious in the battle with the Devas.''

8. Thus the Daityas, hearing the words of the false \'Sukr\=ach\=arya, thought the false one to be real and placed implicit reliance in his words.

9. The real \'Sukr\=ach\=arya, then explained them as much as he could; but the Daityas, owing to the magic of the false \'Sukr\=ach\=arya and to the wonderful effect of time, did not hear his words.

10. The Daityas thus convinced said to the real \'Sukr\=ach\=arya, ``this man before us is our Guru for our welfare and enlightenment, he is the foremost religious \'Sukr\=ach\=arya; for ten years continuously he is giving us advices. You are not our Guru; you appear fictitious and false; you better leave this place and go away.''

11. The dull brained Daityas repeatedly told to \'Sukr\=ach\=arya the reproachful words and bowed down to the false \'Sukr\=ach\=arya, and, saluting, him, gladly welcomed him as their Guru.

12-14. The real \'Sukr\=ach\=arya, seeing the Daityas exceedingly attached to Brihaspat\={\i}, the Guru of the Devas, and deceived by his words, cursed them out of anger and said :--``As you have not taken my words though I have explained everything to you, you would lose your knowledge and get defeat.

15. As you have shown disrespect towards me, you will get the fruits of it at an early date and will then understand the deceitful behaviours of that Deva Guru.''

16. Vy\=asa said :-- Thus saying \'Sukr\=ach\=arya hurriedly went away, infuriated with anger. Brihaspat\={\i} was glad and remained there with his mind calm.

17-19. Brihaspat\={\i} then knowing the Daityas cursed by \'Sukr\=ach\=arya, assumed his real appearance, left that place, and hurriedly returned to Indra. He began to say ``I have undoubtedly succeeded in my undertaking; the Daityas have been cursed and left by me too. They are now helpless; so, O Good Suras! I have made them cursed, you would better now try to fight with them.''

20-21. Indra heard their Guru and became very glad; all the other Devas were glad and worshipped him. They held another cabinet, secret council, how to fight with the Daityas; and, then, all uniting together marched out in battle against the Asuras.

22. Seeing that the powerful Devas are marching towards them ready to fight and knowing their false Guru had fled, the Daityas became very anxious.

23. They told each other :-- Alas! We were enchanted by the Devaguru; the highsouled \'Sukr\=ach\=arya angrily left us; now it is our incumbent duty to satisfy him.

24. That vicious, dirty inside and pure outside, hypocrite Pundit Deva Guru, who used to go to his brother's wife, has really cheated and has quitted us.

25. What are we to do now? where to go? How to satiate now the anger of \'Sukr\=a so that he might be glad and help us.

26. Thus pondering over they all unitedly, shuddering with fear, again went to \'Sukr\=ach\=arya, keeping Prahl\=ada in front of them.

27. They all bowed down at the feet of their Guru; \'Sukr\=a remained silent; then, bursting with anger and with eyes red, told them.

28. You were all warned by me that you were being cheated by the M\=ay\=a of Brihaspat\={\i}; you did not take my word, as worth hearing, though unselfish, pure and leading to your welfare.

29. Rather you were influenced by him and infatuated with vanity, you insulted me; now you will have to bear undoubtedly the effects of that Karma.

30. You are now vitiated from the path of your welfare; go now where that disguised cheat resides for the welfare of the Gods; know me not to be a cheat like him.

31. Vy\=asa said :-- O king! When \'Sukr\=a finished saying his uncertain words, Prahl\=ada clasped his feet and began to say thus :--

32. Prahl\=ada said :-- O our Guru Bh\=argava! Today we have come to you in a very distressed condition! O Omniscient! we are your disciples; we are your good sons; you ought not to quit us.

33. On your departure to get the Mantra, that hypocrite, vicious Brihaspat\={\i} getting the opportunity, assumed your false appearance and cheated us.

34. Peaceful persons do not take any offence committed with ignorance; you know everything; you know very well that our hearts are devoted to thee and to thee alone. There is no need of telling anything further to you.

35. O Highly Intelligent One! By your Tapas, you know our inner minds and relinquish your anger. The sages say that the anger of the saints is not lasting.

36. O Muni! Water is naturally cool; when in contact with fire it gets hot; but, when the heat is removed, it gets cold shortly after.

37. O observer of good vows! Anger is like chand\=ala; sages therefore quit it. Our prayer to you is that you leave your anger and be pleased with us.

38. If you do not quit your anger and if you make us overpowered with grief and sorrow, O blessed one! We, being abandoned by you, will go down to the P\=at\=ala.

39. Vy\=asa said :-- Bh\=argava heard Prahl\=ada's words and, with his intuitioned eye, came to see the proper state of affairs and was pleased and lovingly said.

40. You will not have to fear nor to enter into the P\=at\=ala. You are my Yajam\=anas; I will certainly protect you all by my never failing Mantra power.

41. O knower of religion! What Brahm\=a of yore told me, I am now telling you accordingly. Hear my truthful words, leading to your welfare.

42. Whatever is inevitable, doomed to pass, must come to pass, be it auspicious or inauspicious. No one is able in this world to go against the current of Fate.

43. Under the influence of Time you are now deprived of strength, therefore you will have to suffer defeat at the hands of the Devas and you will have to go once to the P\=at\=ala.

44-45. Brahm\=a said :-- When your time to enjoy the sovereignty of the Trilok\={\i} had come, you enjoyed the kingdoms of the Trilok\={\i} with all its wealth and power. You attacked the Devas and, helped by Time, had been able to trample them under your feet and held your sovereignty for full ten yugas and enjoyed the pleasures without any hitch.

46. You will regain this kingdom in the S\=avar\d{n}ik manvantara. Then Bali will come in your family as the grandson of Prahl\=ada and will conquer the Trilok\={\i} and will get name and fame throughout his kingdom.

47-48. When the Lord of Vaikuntha had incarnated as V\=amana and stolen away the kingdom of Vali, then the Jan\=ardan Vi\d{s}\d{n}u told Bali, the king of the Demons that ``I have taken away your kingdom by pretext to serve the purpose of the Gods; you will become Indra, no doubt, in the coming S\=avar\d{n}ika manvantara.''

49. Bh\=argava said :-- According to the sayings of \'Sr\={\i} Bhagav\=an Hari, your grandson Bali is now invisible to all creatures and is now passing away his time, very much terror stricken.

50-51. Being afraid of Indra, he is now staying in a lonely house as an ass. One day Indra on seeing him enquired of him, in various ways, the cause of his assuming that ass-body.

52. O Lord of the Daityas! You have always enjoyed pleasures of all the world; you are the ruler of the Daityas; you ruled over all the worlds; do you not feel shame now in thus assuming this ass body. The Lord of the Daityas, hearing him, spoke thus.

53-54. O Indra, there is nothing to be sorry in these matters. When the most powerful Vi\d{s}\d{n}u can assume fish and tortoise incarnations, then what wonder is there that I by virtue of the force of Time, am now staying as an ass? When you had murdered a Br\=ahma\d{n}, you, too had hid yourself in the M\=anasa lake in the lotuses; similarly today distressed, I am staying here in this ass body.

55. O Indra! What sorrow or happiness can be to a person who is under the control of Fate. To him everything is alike; for whatever the Time wishes, It can act accordingly.

56. Bh\=argava said to Prahl\=ada! Both persons Bali and Indra got enlightenment at the mutual conversations; and they went away to their places at their own will.

57. O Lord of the Asuras! I have narrated to you this story indicating how powerful is the Time. Know that the Devas and Daityas and all the human beings and this whole universe is under that Great Fate.

Here ends the Fourteenth Chapter in the Fourth Book of \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, the Mah\=a Pur\=a\d{n}am of 18,000 verses on the Daityas getting back their \'Sukr\=ach\=arya, by Mahar\d{s}i Veda Vy\=asa.



