\chapter{On chanting the hymns to the Dev\={\i}}

1. Vy\=asa said :-- The Bhagav\=an Vi\d{s}\d{n}u spoke again unto Praj\=apati :-- Br\=ahma\d{n}! All these beings fascinated by M\=ay\=a, cannot know the Real Essence, the Highest Truth.

2. We, too, are fascinated by that M\=ay\=a; and hence we also, being blinded by that, do not at all remember That Highest Eternal Puru\d{s}a, calm and quiet, the World Teacher, the Highest Self, of the nature of Pure Existence, Intelligence and Bliss.

3. O Brahm\=a! I am Vi\d{s}\d{n}u, I am Brahm\=a, I am Rudra, thus our I-ness ahamk\=ara has blinded our eyesight; and we are made unable to recognise That Eternal Highest Self.

4. As the wooden dolls dance according to the will of the player, the magician, I also am similarly fascinated by the M\=ay\=a and am thus incessantly rolling about like a dependent man.

5-6. O Brahm\=a :-- In the beginning of the Kalpa, Mahe\'svara, You and I saw the wonderful unspeakable form and glory of that Highest Self at the time of R\=asamandala in the M\=ani Dv\={\i}pa where there was the Mand\=ara Tree and the Devas assembled. Then I also saw that wonderful thing a second time in the Sudh\=arnava ocean of nectar, and the most wonderful of it is this, that until we were able to See that Form, we did not hear anything of Her before!

7. Therefore, O Devas! Today do you all remember that Prime Force, the Highest Self, that all beneficent Force that yields all good and auspicious things; That Very Force will fulfill now all your desires.

8. Vy\=asa said :-- O king! No sooner the Bhagav\=an Hari addressed thus, Brahm\=a and the other Devas at once mentally began to meditate on that Eternal Yoga M\=ay\=a, the Goddess of the Universe.

9. Being thus meditated, at once appeared before their eyesight the Dev\={\i}, the Goddess of the Universe, whose colours were like the blood red Jav\=a flower, holding in two of Her hands noose, hook, or goad, while the third hand indicated favour and the fourth hand bade all discard every sort of fear.

10. As the web comes out of a spider and sparks come out of fire, so this whole Universe comes out of that Goddess. We all bow down before Her with our humble hearts, full of devotion.

11. We all salute to that Goddess of the Universe, Bhuvane\'svar\={\i}, under Whose M\=ayic force this whole Universe, moving and unmoving, is created; Who is All-Intelligence and the Ocean of Mercy.

12. This world appears real to him who is unable to realise Her Real Essence; and the world drops off as unreal no sooner he realises Her Presence. We all meditate on that All Intelligent Goddess and we all pray to Her that She may grant unto us more strength to meditate Her and Her alone so.

13. We all want to know Mah\=a Lak\d{s}m\={\i}, we all meditate on the Omnipotent; May the Goddess grant unto us strength to meditate on Her so.

14. O Thou, the Remover of the world's afflictions! Best Thou pleased unto us; O Thou, kind hearted! Do this work for us and promote our good. O Thou, Lord of the Earth! Dost Thou relieve the burden of the earth by killing these Asuras and bring on our welfare.

15. O Thou, Lotus-Eyed! If Thou dost not show Thy mercy towards the gods, they will never be able to strike their enemies with their weapons in the battlefield. O Goddess! Thou didst verify the truth of this before, when Thou didst assume the appearance of a Yak\d{s}a and utter the following sentences ``O Fire! You burn this blade of grass,'' etc. (in the Kena Talavak\=ara Upan\={\i}\d{s}ada.)

16. O Mother! Kamsa, Bhauma, K\=alayavana, Ke\'s\={\i}, Jar\=asandha, the son of Brihadratha, Vaka, P\=utan\=a, Khara, \'S\=alva and other vicious kings respectively are dwelling on the earth. Dost Thou better kill them and relieve the burden of the Earth.

17. O Lotus-Eyed! Thou hadst killed those Asuras without any effort that were not slain by Vi\d{s}\d{n}u or Mahe\'svara and they ended their lives, while they were beholding Thy pleasing countenance.

18. O Goddess! Holding Moon on Her forehead! These Hari, Hara, Brahm\=a and the other gods are unable to move a step without the help of Force; nothing to speak any further, the thousand headed King of Serpents is unable to hold this earth even for a moment, were he not provided with the \'Sakti of upholding this world.

19. Indra said :-- O Bhagavat\={\i}! Would Brahm\=a be ever able to create this world without the aid of Sarasvat\={\i}! Would Vi\d{s}\d{n}u, the Deva of the Devas, be ever able to preserve it without the aid of Lak\d{s}m\={\i} or would Mahe\'svara be ever able, to destroy this world without the help of His Um\=a; no, never! These Devas, the Trinity, are able to take their respective parts in the keeping up of the world simply because they are united with the forces, incarnate in Sarasvat\={\i}, etc., who are again parts of Thee.

20. Vi\d{s}\d{n}u said :-- O Vimale! Were we deprived of Thy Force, the Creator would never have been able to create the world, I would never have been able to preserve it and Mahe\'svara would never have been able to destroy. Therefore O Dev\={\i}! Thou alone art reigning as the Supreme Goddess amidst all the glories in the creation, preservation, and destruction of this Universe.

21-22. Vy\=asa said :-- O king! Thus praised by the Gods, the Goddess then addressed them :-- O Devas! Let all anxieties leave you all; I will do what the Devas have desired, even if it be very difficult to do in this world; now explain what is troubling the Goddess Earth.

23. The Devas replied :-- The wicked kings are troubling very much this Earth. She cannot any longer bear their burden; therefore She came crying unto us.

24. O Bhuvane\'svar\={\i}! Now it is Your duty to remove this load of earth. O Auspicious One! Know this, the desire of the gods.

25-28. O Mother! In days of yore, You have killed the most powerful D\=anava named Mahi\d{s}a, attended by lakhs and lakhs of D\=anavas and Daityas; What to speak! You have slain \'Sumbha, Ni\'sumbha, Raktav\={\i}ja, the very powerful Chanda Munda, Dhumralochana, Durmukha, Duhsana, the strong Kar\=ala and many other cruel D\=anavas.

Now, similarly kill the wicked kings, the enemies of the Gods and save the Earth from their heavy loads.

29. Vy\=asa said :-- Thus addressed by the Devas, the Most Auspicious One, the Dev\={\i}, whose colour was black and who looked askance, said smiling in a tone, grave and deep, like that of the rumbling of a raincloud.

30. O Suras! I have already thought over how the burden of those wicked kings, the parts incarnate of Asuras, can be removed.

31. I will, by My own power, rob them of their powers, of that highly prosperous and affluent J\=arasandha, the king of Magadha, and others, the parts incarnate of the wicked Asuras, that are now shining on the Earth.

32. O Devas! You would all better go down and incarnate yourselves as part incarnations, impregnated with My power, and help thus in the removal of the load of this earth.

33. The Deva Praj\=apati Mahar\d{s}i Ka\'syapa would first of all go with his his wife and incarnate as \=Anaka Dundubhi Vasudeva, in the race of Yadu.

34. The Imperishable Bhagav\=an Vi\d{s}\d{n}u will also incarnate as part, owing to the curse of Bhrigu, as the son of Vasudeva.

35. O Devas! I will be born, then, in the womb of Ya\'sod\=a, in Gokula and will accomplish all the ends of the Gods.

36. I will send Vi\d{s}\d{n}u from the prison to the Gokula; also I will transfer An\=ant\=a Deva from the womb of Devak\={\i} to the womb of Rohi\d{n}\={\i}.

37. They two will grow, day by day, by My Force and at the end of the Dv\=apara Yuga, will no doubt kill the wicked kings.

38-39. Arju\d{n}a too, the part incarnate of Indra, will destroy the armies of those kings. Yudhisthira, the part incarnate of Dharma, Bh\={\i}msena, that of of V\=ayu, Nakula and Sahadeva the incarnate of the two A'svins, and Bh\={\i}sma, the incarnate of Vasu as the son of the Gang\=a, will take their respective births and exhaust their armies.

40. O Devas! Now be calm and quiet and go. Let the Earth be also calm and quiet; be confident that I will certainly remove the load of this earth.

41. I will make the above mentioned Gods as my instruments merely and I will destroy, by My own force, the K\d{s}atriyas in the battlefield of Kuruk\d{s}ettra.

42. Malice, jealousy, foul intention, desire, attachment, vanity, covetousness, desire for victory, lust, and fascination are the vices that will destroy the Y\=adavas.

43. The race of Yadu will end owing to the curse of a Br\=ahma\d{n}a. The Bhagav\=an also will leave His mortal coil due to a curse.

44. Now go and incarnate in your respective parts, with your wives, in Gokula and Mathur\=a, and be helpers in the work of the Lord.

45-46. Vy\=asa said :-- Saying thus, the Dev\={\i} the Goddess of the Universe, the M\=ay\=a Incarnate of the Highest Self disappeared. The Devas and the Earth went to their respective places. O king Janamejaya! the Goddess Earth was pleased with the Dev\={\i}'s words, and being quiet, remained surrounded with various medicinal plants and creepers.

47. Then the subjects felt peace and pleasure, the Br\=ahma\d{n}s also grew stronger in happiness and prosperity, and the Munis became glad and began to perform their religious acts with fresh zeal and alacrity.

Here ends the Nineteenth Chapter in the Fourth Book of \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses, the Mah\=a Pur\=a\d{n}am, on chanting the hymns to the Dev\={\i} by Mahar\d{s}i Veda Vy\=asa.



