\chapter{On the birth of \'Sr\={\i} Kri\d{s}\d{n}a}

1-2. Vy\=asa said :-- The six sons of Devak\={\i} being thus killed by Kamsa and the seventh womb being miscarried, Kamsa began, when Devak\={\i} was pregnant for the eighth time, to think of the cause of his own death as advised by N\=arada and watched carefully, day and night, so that this time there would be born a son and there be no miscarriage.

3. On the other hand the Bhagav\=an Hari came down and rested on the body of V\=asudeva as part incarnate and then entered into the womb of Devak\={\i}, in due order.

4. That very time the Goddess Yogam\=ay\=a, of Her own accord, enters into the womb of Yasod\=a in order to help the workings of the gods.

5. Rohin\={\i}, the wife of V\=asudeva was anxiously dwelling at that time through the fear of Kamsa in the Nanda Gokula; Balar\=ama, the part incarnate of Ananta, took his birth there as Her son.

6. Kamsa then imprisoned Devak\={\i}, who was being worshipped by gods, and engaged servants to take care of her.

7. V\=asudeva, too, entered into the prison with his wife Devak\={\i}, thinking of his issue and also being attracted by his attachment towards his wife.

8. Vi\d{s}\d{n}u, here, the Deva of the Devas, to accomplish the purpose of the gods began to grow regularly in the womb of Devak\={\i}, being incessantly adored and praised by the gods.

9-10. When the tenth month was complete, on the eighth day of the dark fortnight, under the star Rohin\={\i}, Kamsa became much bewildered with fear and called on his D\=anava followers and said ``You must all carefully protect Devak\={\i} in the prison.

11-12. The son born from this eighth womb of Devak\={\i} will turn out my dire enemy; therefore do you all very carefully protect that child, my death incarnate (so that the child may not be transferred anywhere else). O Daityas! I will be able to sleep without any anxiety then, when I will be successful. Kill this eighth son of Devak\={\i} who has become the source of my constant worries and endless troubles.

13. Equipped with swords, shafts and bow and darts do you all incessantly watch with eyes open on all sides, leaving off your sleep and drowsiness.''

14. Vy\=asa said :-- Thus ordering the Demons, the king Kamsa who had become worn out with this anxiety and was bewildered with fear, went quickly to his own palace room; but he could not find any trace of happiness.

15-18. On the other hand, Devak\={\i}, at dead of night, in that prison, told V\=asudeva ``O king! My birth pain has come; I see here many guards, terrific, who are watching and protecting us; what shall I do now? Yasod\=a, the wife of Nanda, promised me before this. O respected one! Your heart is almost burnt up with the fire of grief; therefore do you send to my house your son; I will protect it with utmost care; and especially to incur Kamsa's faith and belief, I will give you another son. O Lord! This is now a very hard moment; what are we to do now?

19. And again how are you to exchange the two sons? However let what may come, come to pass; now the moment of child birth has come; I am possessed by a feeling of shame not liable to be overcome. You better turn away your face; there is no other remedy here.''

20. Thus saying to that blessed V\=asudeva, worshipped by the gods, Devak\={\i}, at dead of night, delivered a wonderful child.

21. The blessed Devak\={\i} looking on that exceedingly beautiful child, was struck with wonder and told her husband, her whole body being filled with joy.

22. O Lord! Look at the face of your new born child, a very rare thing to see such a face as this. Alas! The son of my paternal uncle, Kamsa, will kill my this new born child.

23. ``Kamsa will do so.'' Saying this, V\=asudeva took up that child in his arms and began to look amorously at the face of that child of wonderful deeds.

24. Thus looking, V\=asudeva thought ``What can I do now so as to relieve me from my sorrows owing to the future destruction of this child.''

25-27. While V\=asudeva was thus anxiously pondering over, the Heaven Voice clearly said thus :-- ``O V\=asudeva! Quickly go to Gokula with the child. The keepers and guards I have made dead asleep by My M\=ay\=a. The eight very strong doors are now lying wide open. You better free yourself from your chains and take this child, keep it in the house of Nanda and bring Yoga M\=ay\=a from there and come back here.''

28-29. Thus hearing the Heavenly Voice, he cast his glance at the doors and found them all wide open. O King! Very quickly then he took the child and stepped out of the prison, quite unnoticed by the guards and watchmen. Going to the banks of the Jumn\=a he found there the daughter of Kalindi, flowing with great force and became anxious.

30-31. But the river Jumn\=a instantaneously turned out to be easily fordable just knee deep; then V\=asudeva, guided by Yoga M\=ay\=a, crossed the Jumn\=a and betaking to an unfrequented road arrived at Gokul at dead of night. There, at the doorway of the house of Nanda, he began to see the cows, buffaloes, the wealth and property of Nanda.

32. At that very moment, there was born of Yo\'sod\=a, the Mah\=a Dev\={\i}, the part of the Divine Yoga M\=ay\=a, the Incarnate of the three qualities.

33. Then the Mah\=adev\={\i} Yoga M\=ay\=a, assuming the appearance of a female artisan, taking that divine female child in her arms came there and handed her over to V\=asudeva.

34. V\=asudeva, too, delivered his child over to the lotus like hands of the Dev\={\i} and, taking the female child instead, quickly returned with his heart full of joy.

35. He went unto the prison and kept that female child in the bed of Devak\={\i} and remained aside, fearful and anxious.

36-37. But that child, the moment she was lulled to sleep, began to cry in a sonorous voice; the royal guards at once woke up on that cry, and being bewildered with fear, hurriedly went to their king and said ``O King! Come quickly. Devak\={\i} has brought forth a child.''

38-39. The King of the Bhojas, hearing their words, went there quickly and saw the doors open, called on V\=asudeva ``O blessed one! Let me have the eighth son of Devak\={\i}, my death incarnate; I will at once kill that enemy of mine, born as part incarnate of Hari.''

40. Vy\=asa said :-- O King! Hearing Kamsa's words, V\=asudeva handed over to him the female child crying and very much bewildered with fear.

41. At the sight of the female child, the king was very much astonished and began to think that the Heavenly Voice and the Seer N\=arada's words turned out false.

42. How can V\=asudeva be able to bring about, in this dreadful place, the unnatural act of turning male into a female. Especially my guards are carefully watching the place. There is no doubt in this.

43. How has this female child come here? Where has that eighth born child gone? I ought not to doubt in this matter. For the ways of Time are mysterious!

44-46. Thinking thus, the cruel King Kamsa caught hold of that female child by her legs and, raising her high up in the air, was going to strike her against a stone, when the female child slipped out of his hands and flying into the air assumed a divine appearance and gently spoke to Kamsa, thus :-- ``What will you get by killing me? Your powerful enemy is already born on the earth. O vilest of men! Disgrace to your family! He, the Excellent Human being Who is very difficult to be worshiped will certainly kill you.''

47-49. Saying thus, the auspicious female child, able to go anywhere according to her will, disappeared. Kamsa astonished returned to his own home and, becoming impatient with fear and anger, called all the D\=anavas Baka, Dhenuka, Vatsa and others, and addressed them thus :-- ``O D\=anavas. Go all of you to serve my purpose. Kill anywhere the child whom you see just born.

50. Let P\=utan\=a, expert in killing children go today to Nanda's Gokula. My order is this :-- Any child recently born they must kill.

51. Dhenuka, Vatsaka, Ke\'si, Pralamba, and Vaka, etc., all should remain in Gokula to carry out my order.''

52. The cruel king Kamsa, thus ordering the Demons, went away to his own palace, and, thinking on this matter over and over again became, very much afflicted with fear and his mind got immensely depressed.

Thus ends the 23rd Chapter in the 4th Skandha of \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, the Mah\=a Pur\=a\d{n}am, of 18,000 verses by Mahar\d{s}i Veda Vy\=asa, on the birth of \'Sr\={\i} Kri\d{s}\d{n}a.



