\chapter{On the killing of V\=askala and Durmukha}

1-6. Vy\=asa said :-- O King! The two powerful D\=anavas V\=askala and Durmukha, well-versed in arts of warfare, went out for battle, maddened with their prowess. The two D\=anavas, elated with vanity, went to the battle-field and began to address the Dev\={\i} in voice deep as the rumbling of a cloud. O Beautiful Dev\={\i}! You better choose and worship the Lord of the Daityas, that high-souled Mahi\d{s}\=asura who has conquered all the Devas. He will come before you in privacy in a human shape, with all auspicious signs and adorned with beautiful ornaments. O Sweet smiling One! better place your highest feelings of love on the lovely Mahi\d{s}a as your husband, and you will get all the pleasures of the three worlds as you desire. O Sweet speaking! In short, if you select him as your husband, you will be the mistress of those incomparable worldly happinesses that women always aspire.

7-13. Hearing thus the words of V\=askala and Durmukha, the Dev\={\i} said :-- ``O Stupid! Do you think Me as deluded by passion? Do I not possess strength and intelligence that I will worship that hypocrite Mahi\d{s}a

as husband? See! The ladies of a high family select those persons that are equal in rank as far as family and distinctions, qualifications and propriety of conduct are concerned or those who are superior in beauty, cleverness, intelligence and other qualifications. Then how can a Dev\={\i}, becoming passionate, worship the worst of all beasts, the beast Mahi\d{s}a? O two Asuras! Go you immediately to your King Mahi\d{s}a resembling in his body like an elephant and having a pair of horns and tell him, Go either to P\=at\=ala (the nether regions) or come and fight with Me; the Lord of the Devas will no doubt be happy if the war ensues. O Stupid! My advent here cannot go in vain; I will easily slay you and then depart; knowing this do as you like. O Beast! Without conquering Me, you would get no shelter either in the heavens, or in this earth, or in the caves of mountains?''

14-25. Vy\=asa said :-- Hearing thus, the two powerful Daityas, with eyes reddened with anger, firmly resolved to fight and took bows and arrows in their hands. O Descent of Kuru! The Dev\={\i} then made a terrible noise and fearlessly stood there. The two D\=anavas then began to shoot dreadful arrows at Her. For the victory of the Devas, the Dev\={\i} also begin to hurl arrows after arrows on the two D\=anavas, emitting a sweet sound. V\=askala first came forward with no delay; and Durmukha stood aloof there simply as a witness. The terrible fight then ensued between the Dev\={\i} and V\=askala; arrows, swords and weapons were seen shining in the air and raised terror to those that were dull in intellect. Then the Mother of the Universe seeing V\=askala growing turbulent shot at him five arrows sharpened on stone. The D\=anava, too, cut off the arrows of the Dev\={\i} and hurled seven arrows at Her, seated on a lion. The Dev\={\i} cut off the D\=anava's arrows and shot at that hypocrite, sharpened arrows and began to laugh frequently. She again cut off his arrows with Ardhachandra arrow; V\=askala then pursued the Dev\={\i} with a club in his hands to slay Her. Seeing the arrogant D\=anava with club in his hands, Chandik\=a Dev\={\i} struck him down on the ground with Her own club. The very powerful V\=askala fell down on the ground but rose up within a very short time and hurled again on the Dev\={\i} his club. Seeing him again attacking Her, the Dev\={\i} got angry and pierced him with Her trident; V\=askala fell down, thus pierced, and died.

26-38. V\=askala falling thus dead on the field, the soldiers of the wicked demon routed; whereas the Devas became glad and repeatedly shouted aloud, ``Victory to the Dev\={\i}.'' On this Daitya being slain, Durmukha came forward on the battle-field, filled with anger and accompanied by a stronger army. Mounted on a chariot, shielded all over his

body with a coat of armour, Durmukha came before the Dev\={\i}, shouting all along, ``Wait, wait, O You weak woman!'' and with bows and arrows in his hands. The Dev\={\i} blew Her conchshell and made sounds by stretching Her bow in order to make the D\=anava infuriated with anger. The Asura then began to shoot sharp arrows after arrows like poisonous snakes. The Mah\=am\=ay\=a, by Her own arrows, cut off those of Her enemy and began to shout loudly. The fight then raged furiously, when both parties began to use arrows, \'Saktis, clubs, Mu\d{s}alas, and Tomaras. Blood began to flow in the battle-field in torrents like rivers and on the banks of that river of blood, were seen the severed heads of the dead bodies which looked like so many hollow shells of gourds, as if kept there by the attendant of the god of Death, for their swimming purposes. The battle-field, then, became very dreadful and impassable; at some places dead bodies are lying; wolves are feeding on their flesh; at other places are seen jackals, dogs, herons, crows, vultures, eagles, and other voracious birds and beasts and iron-tipped arrows, eating the dead bodies of those wicked demons. Air began to emit an offensive smell, because of its contact with these corpses; and there were heard the heart-rending sounds of various carnivorous birds and animals. Then the wicked Durmukha began, as if inspired by the god of Death, to address the Dev\={\i} angrily and arrogantly with his right hand raised up before Her. ``Your brain has become perverted; fly away just now or I will send you unto death, or you better accept the proud Mahi\d{s}a, the lord of the Daityas, as your husband.''

39-50. The Dev\={\i} said :-- ``O Villain! I see your death at hand this very day; therefore you are deluded and therefore raving like a mad man. I will kill you today like V\=askala. O Stupid! Better fly away; or if you prefer death, then wait; I will slay you first; then the dull Mahi\d{s}a, the son of a she-buffalo.'' Hearing thus, Durmukha, as if prompted by Death, hurled dreadful arrows on the Dev\={\i}. Instantly the Dev\={\i}, too, cut off all his arrows and, infuriated with anger, pierced the D\=anava by sharpened arrows as Indra had pierced Vritr\=asura before. The fight then turned out very dreadful. O King! Weak persons become very afraid and strong ones become very excited. Instantly the Dev\={\i} cut off the Asura's bow and broke his chariot by five arrows. On seeing his chariot broken, the powerful Durmukha attacked on foot the Dev\={\i} with his club, very hard to overcome. He knocked at the head of the lion with that club with great force; but the powerful lion did not become unsteady, though so very hard hit. Seeing the demon thus standing before Her, the goddess Ambik\=a cut down his head by her sharpened axe. On his head being thus severed, Durmukha fell down dead on the field. The band of Immortals, then, loudly shouted, ``Victory to the Dev\={\i}.'' When Durmukha was slain, the

Immortals from the celestial space began to chant praises and hymns to the Dev\={\i}, showered down flowers on Her head and gave shouts of ``Victory to the Dev\={\i}.'' The \d{R}i\d{s}is, Siddhas, Gandarbhas, Vidy\=adharas, and Kinnaras all became very glad to see the Demon dead on the field.

Here ends the Thirteenth Chapter of the Fifth Book on the killing of V\=askala and Durmukha in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, the Mah\=apur\=a\d{n}am, of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



