\chapter{On the description of the fighting of the goddesses}

1-5. Vy\=asa said :-- O King! The Dev\={\i}, hearing thus, laughed and spoke the following reasonable words, in a voice deep as a rumbling cloud :-- O Dull brained one! Already I told to that messenger before in reply to you; why then do you boast in vain? If there be any such in the three worlds who can stand equal to Me in appearance, strength, and prosperity, I will marry him. Go to \'Sumbha and Ni\'sumbha and inform them that thus I promised before; therefore let any of them conquer Me in battle and then marry Me according to the prescribed rules. You have come here to execute the order of \'Sumbha; so either stand up and fight or fly to P\=at\=ala with your King.

6-11. Vy\=asa said :-- O King! Hearing the Dev\={\i}'s words, the D\=anava was filled with anger and began to shoot dreadful arrows at once on the Lion. Ambik\=a, then, ready-handed, seeing the multitude of arrows coming in the sky like serpents, cut them off in a moment to pieces by Her sharp arrows. The Dev\={\i}, then, drawing Her bow, shot arrows sharpened on stone at the great Asura Raktab\={\i}ja. Then that wicked Demon, thus shot at by the arrows, fell unconscious on the chariot. When he lay thus senseless, a great uproar arose amidst his army and the soldiers began to cry aloud saying, ``Alas! We all are killed.'' Then \'Sumbha, the king of the Asuras, hearing the sound of Boomb\=a (a danger cry by hands and mouth) ordered all the D\=anavas to be ready for the battle.

12. Then \'Sumbha said :-- Let all K\=ambojas go to the battle with all their forces; let other powerful heroes, especially the K\=alakey\=as, too, who are very strong heroes, go up for the battle. (Note: K\=amboja, name of a people and their country. They inhabited the Hindoo Koosh mountain which separates the Giljit valley from Balkh and probably extended up to little Tibet and L\=adak. The K\=alakey\=as may be the Afridis.)

13-33. Vy\=asa said :-- O King! Thus ordered, all the fourfold army of \'Sumbha, viz., cavalry, infantry, elephants and chariots went out, intoxicated for war, to the battle ground where the Dev\={\i} existed. The Dev\={\i} Chandik\=a, seeing the D\=anava forces coming near, made at once terrible sounds frequently. The Ambik\=a Dev\={\i} also made the sound with Her bowstring and blew Her conchshell. K\=al\={\i}, then, shouted aloud opening Her mouth widely. The powerful Lion, the Dev\={\i}'s V\=ahana, hearing the terrible sounds, roared so loudly that the D\=anavas were struck with strange terror. The powerful D\=anavas, then, hearing that sound became

impatient with anger and shot arrows after arrows on the Dev\={\i}. The wonderful horrible battle, then, ensued, causing horripilation and the \'Saktis of Brahm\=a and the other Devas began to come to Chandik\=a Dev\={\i}. The Dev\={\i}s, the wives of the several Devas, then went, to the battlefield in their respective forms with ornaments and V\=ahanas as generally on such occasions. The \'Sakti (wife) of Brahm\=a named Brahm\=a\d{n}i, mounting on the back of Her Swan came there with a string of beads and Kamandalu (an wooden waterpot used by ascetics). The Vai\d{s}\d{n}av\={\i} with Her yellow robes came there mounted on Garuda (the sacred bird of Vi\d{s}\d{n}u, the carrier of Him) with conch, discus, club, and lotus in Her hands. The Dev\={\i} \'Sankar\={\i}, the wife of \'Siva, the Auspicious One, arrived on the back of Her Bull. The emblem of half-moon was on Her forehead while in Her hands She held snake, bracelet, and trident (Tris\=ula) and the sign of fearlessness for Her devotees. The beautiful wife of K\=artika, Kaum\=ari Dev\={\i}, looking like Kartika, came to fight there mounted on a peacock. The fairfaced Indr\=a\d{n}\={\i}, decked with ornaments on Her several limbs, came there to fight, with thunder-bolt in Her hand, mounted on the elephant Air\=avata. The V\=ar\=ah\={\i} Dev\={\i} looking like a female boar, came also, seated on an elevated seat of departed souls (Preta). The N\=arasimh\={\i}, resembling Nrisimha (the Man Lion Incarnation) came there. The wife of Yama, looking fearful like Yama arrived there on the battlefield smiling and with staff in Her hand and mounted on the back of a buffalo. Thus the wives of Kuvera, Varu\d{n}a, and other Devas came there with proper forms, V\=ahanas, ornaments, accompanied by their forces and all excited. Seeing them all, the Dev\={\i} Ambik\=a became glad; the Devas, too, became peaceful and expressed their great joy; the D\=anavas were afraid at the sight of them. \'Sankara, auspicious to all the beings, came there to the battlefield, surrounded by these goddesses and thus said to Chandik\=a :-- Slay quickly all these Asuras, \'Sumbha, Ni\'sumbha and all other D\=anavas to serve the cause of the Devas. Let all the goddesses destroy the D\=anava race and thus free the world from dangers; they may, then, return to their own abodes. Let the Devas get their shares of sacrifices, let the Br\=ahma\d{n}as perform the sacrificial acts, and let all the creatures, moving or not moving, be happy. Let all calamities such as proceed from drought, excessive rain, rats, foreign invasion, locusts and birds, bats, etc., come to an end. Let the clouds bestow rains regularly and let the cultivation yield ample harvests. Note here that rats bring in plague.

34-40. Vy\=asa said :-- O King! When \'Sankara, the Lord of the gods and Auspicious to all, said thus, a wonderful female ensued out of the body of Chandik\=a, very furious, horrible, with hundreds of jackals surrounding and yelling; then that \'Sakti, of dreadful appearance, said

to the five-faced \'Siva, smiling :-- O Deva of the Devas! Go quickly to the lord of the Daityas; do the duty of a messenger for us, O Destroyer of lust! O \'Sankara! Tell that haughty lustful \'Sumbha and Ni\'sumbha leave the Heavens and go to P\=at\=ala. Let the Devas reign in the Heavens; Indra go to his own beautiful throne; let all the Devas go to their appropriate places in Heaven and receive their sacrificial offerings duly. And if the Demons desire to live, let them go quickly to the city of P\=at\=ala where other D\=anavas are dwelling. Else, if they like to die, let them go quickly to the battlefield and let their flesh be eaten up by Her jackals.

41. Vy\=asa said :-- O King! Hearing Her words, \'S\=ulap\=a\d{n}i went quickly to \'Sumbha, the lord of the D\=anavas, seated in the assembly, and said thus :--

42-44. O King! I am Hara, the Destroyer of the Asura Tripur\=a; now I have come to you for your good as a messenger of Ambik\=a Dev\={\i}. Quit the Heavens and Earth and go quickly to P\=at\=ala where reside the powerful Bali and Prahl\=ada, or if you like to court Death, come to fight; I will slay you all in a moment. O King! The Great Queen Ambik\=a Dev\={\i} has sent me with these instructions for your welfare and information.

45-63. Vy\=asa said :-- O King! Thus conveying the nectarlike, beneficial words of the Dev\={\i} to the chief Daityas, \'Siva, the Holder of the trident, returned to his own place. The \'Sakti that sent \'Sambhu as a messenger to the D\=anavas is known in three worlds as \'Siva D\=ut\={\i}. The Daityas, thus hearing the rigorous words of the Dev\={\i}, put on their coats of armour and taking their bows and arrows quickly went out to fight. They came with great force to the battlefield, stretched their bows to their ears and shot piercing arrows, sharpened on stone and tipped with iron, at the Chandik\=a Dev\={\i}. The K\=alik\=a Dev\={\i}, too began to strike some with the trident, some with \'Sakti weapon, and some with clubs and rent them asunder and devoured them all, and began to roam in the field. Brahm\=a\d{n}i began to pour water from Her Kamandalu on the powerful D\=anavas in the battlefield and thus destroyed them. Mahe\'svar\={\i}, mounting on Her Bull gave violent blows by Her trident and thus lay them dead on the ground. Vai\d{s}\d{n}av\={\i}, with the blow of Her club, took away the lives of many Daityas and striking them with Her Discus out off the heads of several others. Indr\=an\={\i} hurled Her thunderbolt on the chief D\=anavas, already struck by the feet of the elephant and lay them dead on the field. N\=arasimh\={\i} tore the strongest D\=anavas with the sharp nails and, devouring them, walked to and fro and made dreadful sounds. \'Siva D\=ut\={\i} began to laugh hoarsely and laid the D\=anavas flat on the field, when they were at once devoured by K\=alik\=a and Chandik\=a. Kaum\=ari, seated on a peacock, by drawing the bowstring to

Her ears hurled arrows sharpened on stone on the enemies and killed them to serve the cause of the gods. V\=aru\d{n}i tied down the D\=anavas by Her P\=asa weapons in a face to face fighting; thus they lay down senseless on the ground. O King! Thus the M\=atrik\=as, the goddesses, crushed the forces. Then the other powerful soldiers fled away terrified. The ``Boomb\=a'' danger cry rose then loudly; on the other hand, the Devas began to shower flowers on the Dev\={\i}s. Hearing the distressful agonies of the Asuras and the shouts of victory of the Devas, Raktab\={\i}ja, the chief of the D\=anavas became very angry. Seeing specially the D\=anavas flying away and the Devas shouting, that powerful Demon came hurriedly to the battlefield with anger. Then with eyes reddened with anger, and with various weapons came before the Dev\={\i}, Raktab\={\i}ja mounting on a chariot, and making sounds unusual with his bowstring.

Here ends the Twenty-eighth Chapter of the Fifth Book on the description of the fighting of the goddesses in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam the Mah\=a Pur\=a\d{n}am, of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



