\chapter{On the Deva defeat and on Vritra's tapasy\=a}

1-3. Vy\=asa said :-- O King! Having the Svastyayana ceremony (a performance of rite to secure welfare or avert calamity) performed by the Br\=ahma\d{n}as versed in the Vedas, the powerful Vritra mounted on his

chariot and started to kill Indra, the King of the Gods. The D\=anavas that were previously defeated by the Devas now knowing Vritr\=asura to be powerful, came up to him to serve his cause. The messengers of Indra, when they saw him ready for battle, hurriedly came to Indra and informed him all about his doings and other matters connected with it.

4-7. The messengers said :-- O Lord! Vi\'svakarm\=a, having been very much grieved his son being slain, got very angry and by Abhich\=ara process (an incantation with a design to injure or magic spells or charms used for a malevolent purpose) has created a son in order to kill you. That indomitable Vritr\=asura is now your powerful enemy; mounting on his chariot he is coming here to fight with you, surrounded by other Asuras. O highly Fortunate One! This enemy of yours is as high as the mountain Meru; he is now coming hurriedly to you, making a terrible noise; guard yourself carefully. O King! While Indra was hearing the messengers, the Devas came there panic-stricken and terrified and said :--

8-16. The Ga\d{n}as said :-- O Lord of the Suras! Ominous signs are being seen in the houses of the Gods; the birds are making sounds, very inauspicious and foreboding a great calamity. Crows, vultures, herons, falcons, and other ugly inauspicious birds are crying and making hoarse sounds on the tops of houses. Other birds are making incessantly harsh sounds like chichi koochy. The carriers of the several Devas are weeping and shedding tears always. O highly Fortunate One! On the tops of houses are heard very loud and very dreadful sounds of the crying R\=ak\d{s}as\={\i}s at dead of night. O Giver of honour! The flags on the chariots are falling to the ground without any trace of wind. Thus ominous signs are being visible on earth and in the air. O King of the Devas! The ugly faced women, wearing black clothes, are roaming from house to house and always repeating ``Leave the house, and go away at once.'' The Deva women while sleeping in their own temples are seeing in their dreams that terrible R\=ak\d{s}as\={\i}s, coming to them are cutting away their hairs on their heads and are frightening them. O Indra of the Devas! The inauspicious signs like these and earthquakes and the falling of the meteors are taking place. The jackals come in the courtyard of houses at night and yell horrible heartrending sounds. Lizards are moving always in the rooms and the several limbs of our bodies are shaking and thus making very inauspicious signs.

17. Vy\=asa said :-- O King! Hearing their words, Indra became very anxious and called Brihaspati, the Deva Guru, and asked him :--

18-20. Indra spoke :-- O Br\=ahma\d{n}a! Very inauspicious signs are being visible; dreadful winds are blowing and stars are falling from the skies what are all these? O Intelligent One! You are very wise and versed in the \'Sastras and the Guru of the Devas; you are omniscient and know very well how to remedy the evils. Therefore perform the rites by which enemies can be killed; do such as our miseries be all averted.

21-31. Brihaspati said :-- ``O Thousand-eyed! What shall I do? You have committed shortly a heinous crime; you killed that innocent Muni and so you have earned a very bad Karmic effect. Very violent sins and good deeds produce their effects very quickly. It is, therefore, highly incumbent on those that desire for their own welfare, to take up any work with great discretion. It is never advisible to do any action that leads to the tormenting of others. Never do they find happiness who give pains to others. O Indra! You have committed Brahmahatty\=a under the influence of greed and delusion; now suddenly has appeared the fruit of that act. O King of the Suras! This Vritra Asura is born invulnerable to all the Devas. That powerful indomitable Asura chief is now coming, mounted on a chariot, to kill you, surrounded by the other D\=anavas and taking with him the Vi\'svakarm\=a-made divine arms and weapons equal to thunderbolt. He is coming like a second K\=ala, as it were, to destroy the whole Universe. There is none in this Triloka, capable kill him; and his death will not also take place. While Brihaspati was thus speaking, a great tumultuous uproar rose at once. The Gandharbas, Kinnaras, Yak\d{s}as, Munis and other Immortals began to fly away from their quarters. Indra seeing the Devas flying away became very anxious and gave orders at once that all subservient to him must be ready at once for battle; they must go and call the Vasus, the Rudras, the twin A\'svins, the \=Adityas, P\=us\=a, Bhaga, V\=ayu, Kuvera, Varu\d{n}a, Yama and the other Devas to come there at once. The enemy is well nigh; so let all the Devas come on their Vim\=anas quickly there.''

32-44. Thus ordering, Indra mounted on the Air\=avata elephant and taking the Sura Guru in front started from his own temple. The other Devas mounted on their respective carriers and, firmly resolved to fight, started with all their arms and weapons. On the other hand, Vritr\=asura surrounded by the Demons, came up to the beautiful mountain, adorned with trees, on the north side of the M\=anasarovara Lake. Indra, too, came there with Brihaspati in front and attended by all the other Devas to that mountain, north of the M\=anasa Lake and began to fight. A dreadful fight, then, ensued between Vritra and Indra with clubs, swords, Parighas, P\=a\'sas, arrows, \'Saktis, Parsus and other weapons. The terrible

fight lasted for full one hundred human years, terrifying to the self-controlled \d{R}i\d{s}is and all the human beings. Varu\d{n}a first turned his back; then V\=ayu, then Yama, the Sun and Moon and then Indra fled from the battle-field. Seeing Indra and the other Devas flying away, Vritr\=asura came to the hermitage and there bowed down to his father who looked very glad; and he said :-- O Father! I have carried out your orders; Indra and all the other Devas are defeated in the battle; as elephants and deer fly away seeing a lion, so the Devas all fled away to their respective abodes. I have taken the Air\=avata, the best of elephants, away from Indra who fled away on foot. O Bhagavan! I have brought the elephant here. Kindly accept it. O Father! It is not advisable to kill a man who is terrified, therefore I did not kill them. Now kindly order anything else that I may fulfil your desires. All the Devas fled away from the battle-field, very much tired and terrified; and what more to say than this that Indra, too, fled, leaving his elephant on the field.

45-54. Vy\=asa said :-- O King! Vi\'svakarm\=a became very glad to hear his son's words and said :-- ``Today I can rightly say that I have got my son and that my life is successful. O son! To-day you have sanctified me; my cares and worries are abated; my mind is also calm to see your wonderful prowess. O Child! Now hear attentively what I say. O highly intelligent One! Now carefully sit in your steady posture (Sthir\=asan) and practise Tapasy\=a. Never trust anybody; Indra is now your enemy, ever ready to find your faults, and clever in sowing dissensions between you and your well-wishers. O Son! Tapasy\=a is not an ordinary thing; Lak\d{s}m\={\i} (prosperity) is obtained thereby; excellent kingdoms, increase of vigour, and victories in battles are obtained. Therefore worship Hira\d{n}yagarbha and get excellent boons from him; then kill this vicious Indra, guilty of the sin Brahmahatty\=a. Worship the auspicious Creator calmly and carefully. The four-faced Brahm\=a then will be pleased and grant you your desired boon. First please the Creator of indomitable prowess, from Whose womb has sprung all this Universe, and get, then, immortality from Him. Then kill that guilty Indra, my enemy. O Son! My feeling of enmity due to the killing of my son reigns always in my mind; I cannot go to sleep peacefully nor do I get peace in any way. The vicious Indra killed my son; O Vritra! What more shall I say to you; I am merged in the ocean of sorrows; save me.''

55-60. Vy\=asa said :-- O King! Thus hearing his father's words, Vritr\=asura became inflamed with anger and, getting his permission, set out gladly to practise Tapasy\=a. He then went to the Gandham\=adan mountain and performed his bath in the holy and auspicious river

Mand\=a Kin\={\i}, and, preparing a Sthir\=asan, took his seat in the Ku\'s\=a grass, to practise the tapasy\=a. He left off gradually taking his food, then subsisted on water only and remained engaged in Yoga; and, seated in Sthir\=asan meditated incessantly on Praj\=apati, the Creator of this Universe. Indra, on the other hand, knowing Vritr\=asura engaged in tapasy\=a became very anxious and sent to him Gandharvas, Yak\d{s}as, Pannagas, Kinnaras, Vidy\=adharas, Apsar\=as and other Deva messengers, all of unbounded vigour to create obstacles in his austerities. These Gandharvas and other Deva Yonies, expert in exercising magical spells, tried many ways and means and various gestures and postures to create disturbances in his penance; but that great ascetic Vritra, the son of Vi\'svakarm\=a did not swerve a bit from his meditative state.

Here ends the Third Chapter of the Sixth Book on the defeat of the Deva army and on Vritra's tapasy\=a in the Mah\=apur\=anam \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



