\chapter{On the origin of the Haihaya Dynasty}

1-5. Janamejaya said :-- How did the Goddess Lak\d{s}m\={\i}, the daughter of the ocean, come to be born as a mare, when cursed by the Bhagav\=an in His moment of anger, and what did Revanta do at that time? In what country was the Dev\={\i} born as a mare and how did She pass Her time alone like one whose husband had gone abroad. O Muni! How long and in what forest unfrequented by persons did she pass her time, thus deprived of the companion of her husband and what did she do at that time? When was she reconciled with her husband V\=asudeva? and how did she get a son, when she lived in a state of separation from her husband. O best of \=Aryas! I am very curious to hear this excellent story. So describe this in full details to me.

6. S\=uta said :-- O \d{R}i\d{s}is! Thus questioned by Janamejaya, the Dvaip\=ayan Muni began to recite the story in its full details.

7-24. Vy\=asa said :--O King! I will now describe to you the pleasing story of the Pur\=a\d{n}as in a clear distinct language; hear. Revanta, the son of the Sun, became terrified to see V\=asudeva, the Deva of the Devas cursing the Lak\d{s}m\={\i} Dev\={\i} and, after bowing down to Jan\=ardan, the Lord of the world, went off. Seeing the anger of Vi\d{s}\d{n}u, the Lord of the world, he went quickly to his father and informed him of the curse delivered by N\=ar\=aya\d{n}a to the Goddess Lak\d{s}m\={\i}. And the Lak\d{s}m\={\i} Dev\={\i}, the lotus-eyed, thus cursed, got the permission of N\=ar\=aya\d{n}a and with a grieved heart bowed down to him and came down to the world of mortals. She took the form of a mare and went to the spot where the wife of the Sun (named Chch\=ay\=a) practised her asceticism in ancient times. The place was the confluence of the river K\=alindi and the Tamas\=a, and decorated with lovely forests and trees situated north of the mountain named Supar\d{n}\=ak\d{s}a, yielding all desires. There she meditated with her whole heart the auspicious Mah\=adeva \'Sankara, the Giver of all desires, thus :-- That Mah\=adeva is holding the Tri\'s\=ula (the trident) on his arms; His forehead is adorned with beautiful cooling semi-Moon; He has five faces, each face having three eyes; His throat is coloured blue; He has ten arms; His body is white like camphor; He wears a tiger's skin; His upper garment is of elephant's skin; and snakes are his holy thread; He is holding the one-half of the body of Gaur\={\i} and his neck is adorned with garlands of human skulls. The Goddess Lak\d{s}m\={\i}, the daughter of the ocean, assuming the form of a mare, thus practised severe asceticism in that

place of pilgrimage. O King! With a feeling of intense dispassion (Vair\=agyam) towards the worldly things, She spent the divine one thousand years in the meditation of Mah\=adeva, the God of the Gods. After that period, the Highest Lord Mah\=adeva, mounting on His bull, came there with His consort P\=arvat\={\i} and appeared before the Lak\d{s}m\={\i} Dev\={\i}, perceptible by his eyes. Appearing thus with His host of His own persons, He then spoke to Lak\d{s}m\={\i}, dear to Hari, now practising ascetism in the form of a mare, ``O Auspicious One! You are the Mother of this whole Universe and your Husband is the Creator of these worlds and is capable to give all desires. Why are You, then, practising asceticism, when He is present? What is the cause of this? O Dev\={\i}! Why are You praising hymns to me, instead of to V\=asudeva \'Sr\={\i} Hari, Who is capable to yield enjoyments and final liberation, and Who is the Preserver and the Lord of this world. O Dev\={\i}! Work should be done according to the authority of the Vedas; it is stated in the Vedas that the husband is the lord of a woman; therefore it is never advisable to fix one's mind entirely on another person. The eternal Dharma of women is to serve their husbands; whether the husband be a saint or a sinner, the woman, desirous of her welfare, should serve her husband in every way. O Daughter of the Ocean! Your husband N\=ar\=aya\d{n}a is fit to be served by all and He is capable to yield all desires. Why are you then worshipping Me, and leaving the Lord of the Goloka, the Deva of the Devas.''

25-32. Lak\d{s}m\={\i} said :-- O Deva of the Devas! O Seat of Auspiciousness! I know that You are soon pleased with Your servant. My husband has cursed me. O Ocean of mercy! Kindly save me from this curse. O \'Sambhu! When I informed my husband of my mental agonies, graciously and mercifully He then pointed out how I might be freed of this curse thus :-- ``O Kamale! When Your son will be born, then You will be freed of this curse and will, no doubt, come back and live in this Heaven of Vaikuntha.'' Thus spoken, I have come in this hermitage to make tapas and to worship Thee, knowing that Thou art the Bhagav\=an, the Lord of Bhavan\={\i}, the Lord of all and the Giver of all desires. O Lord of the Devas! How can I get a son without the intercourse of My husband? Though I am guiltless, my husband has forsaken me and is living in Vaikuntha. O Mahe\'svara! Thou art doing good to all persons; and if Thou art pleased with me, then grant me a boon. O Lord! I know full well that there is no distinction between Him and Thee. O Lord of Gir\={\i}j\=a! This truth I have come to learn from my husband. O Hara! You are the same thing that He is and what is He is the same as You; there is not the least doubt in this. O Thou, full of auspiciousness! Recognising the Sameness without any distinction between Him and Thee,

I am meditating on Thee. Had it been otherwise, then I would certainly have been guilty when I take Thy refuge and meditate on Thee.

33-36. \'Sankara said :-- ``O Dev\={\i}, the daughter of the Ocean! Tell truly before me how you have been able to realise the identity between Him and Me. The Devas, Munis and the Mahar\d{s}is, versed in the Vedas, get their understandings baffled by wrong argumentations and never realise the identity without any difference between us. Almost everywhere you will perceive that many of my devotees blame me. Specially in this K\=al\={\i} Yuga due to the influence of Time, this happens to a very great extent in many cases. O Auspicious One! Let that go! How have you come to know this matter, which is difficult even for the liberal-minded persons to comprehend. Know that this perception of the identity between me and Hari is very rare.''

37-38. Vy\=asa said :-- O King! When Mah\=adeva asked thus with great pleasure, the Dev\={\i} Kamal\=a, the darling of Hari, gladly replied the essence of the matter to Mah\=adeva.

39-43. Lak\d{s}m\={\i} said :-- ``O Deva of the Devas! One day Bhagav\=an Vi\d{s}\d{n}u, seated in the Padm\=asana posture, was immersed in deep meditation. I was very much astonished at this. When His meditation was over and when He was in a pleasant mood, I asked Him with great humility, O Deva of the Devas! I know that You are the Lord of the world and Master of this whole Universe; when Brahm\=a and the other Devas were united and churned the great ocean, I came out of the waters and looked all around to know who is the superior one whom I can select as my husband and then, thinking You as the superior to all the Devas, I accepted You as my husband. Now whose meditation You are doing again? A great doubt has thus occurred in my mind. O Lord! You are my most Beloved; now disclose to me your innermost desire and thought.''

44-49. Vi\d{s}\d{n}u said :-- ``O Beloved! Hear now, whom I am meditating. I am meditating in the lotus of My heart that Mah\=adeva Mahe\'svara, the Highest of all the Devas. Mah\=adeva, the Deva of the Devas, of indomitable prowess, sometimes meditates on Me and sometimes I meditate on the Lord of the Deva, \'Sankara, the Destroyer of Tripur\=a. I am dear to \'Siva as his life is dear to him and \'Sankara is similarly dear to me. The hearts of us both are attached to each other in the most secret way possible; therefore there is not the least difference betwixt us both. O large-eyed One! Those men who being my devotees hate \'Sankara, certainly go to hell. I speak this very truly unto you.'' O Mahe\'svara! When I asked him this question when he was all alone, that Deva of the Devas, the Highest Vi\d{s}\d{n}u thus said to me. Therefore I am meditating on You, knowing that You are His beloved. O Mahe\'sa! Now find out means by which I can mix with My husband.

50-59. Vy\=asa said :-- O King! Mah\=adeva, skilled in speech, hearing thus the words of Lak\d{s}m\={\i}, consoled Her with sweet words and said :-- ``O Beautiful! Be peaceful; I am pleased with Your tapasy\=a; soon You will come in contact with Your husband. There is not the least doubt in this. When I will send the Bhagav\=an, the Lord of the world, He will come before you in the shape of a horse, to satisfy your desires. I will send the Madhus\=udana, the Deva of the Devas, in such a manner, as he will come in the form of a horse, passionately attached to you. O One of good eye brows! Thus you will get a son equivalent to N\=ar\=aya\d{n}a; and the son will be the King on this earth and will be undoubtedly worshipped by all. O fortunate One! After you get your son, you will go to Vaikuntha with N\=ar\=aya\d{n}a and will reside there as His Beloved. Your son will be famed by the name of Ekav\={\i}ra; and from him will propagate the Haihaya dynasty on the earth. O Kamale! You were blind by prosperity, and, becoming passionate, you forgot the Dev\={\i} Parame\'svar\={\i}, residing in your heart. Therefore you have experienced such a result. Therefore, to expiate that sin, take Her refuge by all means. O Dev\={\i}! If your heart remained attached to the Highest Dev\={\i} the blissful Bhagavat\={\i}, your heart would never have got attached to the Uchchai\'srav\=a horse. Vy\=asa said :-- O King! Thus granting boons to Lak\d{s}m\={\i} Dev\={\i}, He with his consort Um\=a vanished away in her presence.

60-62. Kamal\=a Dev\={\i}, lovely in all respects, whose toe nails are always rubbed by the gems on the coronets of the Devas, began to meditate on the lotus-feet of Ambik\=a and in expectation of his beloved Hari, in the shape of a horse, praised and chanted hymns frequently to the Highest Goddess, in words choked with feelings of intense love.

Here ends the Eighteenth Chapter of the Sixth Book on the origin of the Haihaya Dynasty in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



