\chapter{On the glory of M\=ay\=a}

18-22. O Muni! When my Father learnt the cause of my cares, he smiled and spoke to me in sweet words :-- ``O Child! The Devas, the high-souled Munis, the wise ascetics and the Yogis subsisting on air only are not able to conquer this M\=ay\=a. O N\=arada! The power of M\=ay\=a is so very great that I, Vi\d{s}\d{n}u and \'Sambhu, the Lord of Um\=a, none are able to know Her power.

That Mah\=am\=ay\=a is creating, preserving and dissolving this world by Time, Karma, and Nature and other efficient causes. O Child! Know Her to be inconceivable and unapproachable. O Intelligent One! Do not be sorry nor should you be surprised about M\=ay\=a's great strength, for we all are deluded by Her.

23-25. O Dvaip\=ayan! Thus advised by my Father, my wonder disappeared. I then asked permission of my Father Padma Yoni (Lotus-born) and went out on tour round the sacred places of pilgrimages and on my way, seeing by and by the chief T\={\i}rthas, I have now come here. Therefore, O Muni! Dost thou relinquish your sorrows for the extinction of the Kuru's family and remain here and pass your time in great joy and happiness. One must bear the fruits of one's Karma, good or bad; knowing this fully roam at your will wherever you like.

26-40. Vy\=asa said :-- O King! Mahar\d{s}i N\=arada thus kindling knowledge in me, went away; I also thought over his words. On the banks of the river Sarasvat\={\i}, I composed this Dev\={\i} Bh\=agavat to pass away my time during the excellent period of S\=arasvata Kalpa. This Pur\=a\d{n}am is excellent; it is composed on the authority of the the Vedas; all doubts are removed by it; many nice events are narrated here. Therefore, O King! Not the least doubt should be entertained. As a magician makes the wooden dolls dance in his hands at his will, so this world-enchanting M\=ay\=a is making this world, moving and non-moving, dance from Brahm\=a down to the blades of grass and all human beings. O King! Know M\=ay\=a's triple Gu\d{n}as to be the cause of this mind consisting of five organs of senses, that follows the Chitta (mind, buddhi and Ahamk\=ara). Actions arise from the causes thereof; there is no doubt in this; what doubt, then, there can arise that all these creatures of different temperaments will come out of the different Gu\d{n}as of M\=ay\=a. Peaceful, terrible and stupid become the persons in contact with the Mayic Gu\d{n}as. How, then, can they exist, bereft of them? As the cloth cannot exist without threads, so the embodied beings cannot exist in the world without the triple Gu\d{n}as of M\=ay\=a. There is no doubt in this. As a pot cannot be made without clay, so these bodies, Devas, human or birds, cannot be created without the Gu\d{n}as. Brahm\=a, Vi\d{s}\d{n}u

and \'Siva, too, are possessed of those three Gu\d{n}as and therefore they become sometimes happy and satisfied, sometimes unhappy and dissatisfied and sometimes they become sad and remorseful as they are then under the influence of one Gu\d{n}a or the other. Brahm\=a happens at times to be full of wisdom and knowledge, his temper peaceful, sweet and pleasant and his soul rapt in Sam\=adhi, when he becomes possessed of S\=attva Gu\d{n}a; again when he is void of S\=attva and filled with R\=ajo Gu\d{n}a, His temper becomes unpleasant and his appearance gets dark and awful everywhere; and when he becomes grossly T\=amasic, He becomes sorrowful and bereft entirely of intelligence.

41-51. Vi\d{s}\d{n}u, when resting in S\=attva, becomes peaceful, sweet-tempered, and full of knowledge; when R\=ajo Gu\d{n}a preponderates in Him, He becomes void of sweetness and becomes awful to all the beings. Rudra becomes, too, peaceful and pleasant under the S\=attva Gu\d{n}a, awful and void of sweetness under the R\=ajo Gu\d{n}a, and becomes sad and stupid under the T\=amo Gu\d{n}a. O King! When Brahm\=a, Vi\d{s}\d{n}u, Mahe\'svara and the solar and lunar Kings, the fourteen lords of Manvantaras, Manu and others are under the control of the mayic Gunas, what to speak of other ordinary mortals, men and the other J\={\i}vas. The whole world is under the control of M\=ay\=a; the Devas, men and all other beings. None should doubt on this point. All the embodied beings labour under the directions of M\=ay\=a; never can they work independently. This M\=ay\=a is again always residing in the Highest Essence, the Samvit or the Universal Pure Consciousness. Thus M\=ay\=a is dependent on the Highest Goddess, Who is of the nature of Samvit, and, stimulated by Her, resides in the hearts of all the J\={\i}vas. Therefore one ought to meditate, worship and bow down before the Bhagavat\={\i}, the Creatrix of M\=ay\=a and Who is of the nature of Samvit, Pure Existence, Intelligence and Bliss. Thus She becomes gracious and merciful and liberates the J\={\i}vas, giving them Her realisation and drawing together Her own M\=ay\=a away from them. This whole cosmos is nothing but M\=ay\=a and the Consciousness (Samvit) of the nature of Br\=ahma\d{n} is the Lord of M\=ay\=a. For this reason that Beautiful One in the triple worlds, the Dev\={\i} Bhagavat\={\i} is known by the name Bhuvane\'svar\={\i}, the Great Lady of the worlds.

52-60. O King! If the J\={\i}vas can fix their hearts on that Samvit, then M\=ay\=a, born of the real and unreal, is quite unable to do any harm to them. No other Deva than the Bhuvane\'svar\={\i}, of the nature of pure existence, intelligence and bliss is able to remove this M\=ay\=a. O King! Darkness cannot destroy darkness; the Sun, Moon, Lightning or Fire can destroy it. Therefore it is highly incumbent on us to worship the

Lady of M\=ay\=a, the Samvit, the Mother with a cheerful heart to remove the M\=ay\=a and Her Gu\d{n}as. O King! Now I have narrated to you all the events concerning the killing of Vritr\=asura that you asked. What more do you want to hear now. O One devoted to vows! I have now described the first half of that Pur\=a\d{n}a, which describes in detail the glory of \'Sr\={\i} Dev\={\i} Bhagavat\={\i}. This Pur\=a\d{n}a, the secret of this Mother of the whole Universe, is not to be disclosed indiscriminately to anybody. Those that are peaceful, self-restrained, devoted, and possessed of Bhakti to the Dev\={\i}, the disciples that are devoted to their Gurus and the eldest son, those are the fit recipients thereof. Whoever reads or hears with greatest devotion this Mah\=apur\=a\d{n}am, equivalent to the Vedas, fraught with sound proofs and the essence of all talks, becomes, in this world, possessed of great wealth, becomes wise and passes his time in the greatest happiness. There is no doubt in this.

Here ends the Thirty-first Chapter on the Sixth Book on the glory of M\=ay\=a in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.

[The Sixth Book Finished.]



