\chapter{On the twin A\'svins drinking the Soma Cup}

1-2. Vy\=asa said :-- O King! When the vessel filled with the Soma juice was given to the two A\'svins, Indra became very angry and showing his strength, spoke thus to the Muni Chyavana. O Br\=ahma\d{n}a! Never will you be able to endow him with such a high honour. When you have shewn towards me your enmity, I will kill you, no doubt, exactly like Vi\'svar\=upa.

3-4. Chyavana said :-- ``O Indra! Do not insult the two highsouled A\'svins. They have given me beauty, youth and lustre and made me look like a second Deva. O Lord of the Devas! Just as the other Devas can take the vessel of Soma excluding you, so the two powerful A\'svins can do the same towards you.''

5. Indra said :-- ``They are the physicians; so they cannot in any way have the right to accept the Soma cup. I will just now sever your head.''

6-29. Vy\=asa said :-- O Ornament to the race of Bharata! At these words of Indra, the Muni made the A\'svins accept the Soma cup, thus highly insulting, as it were, Indra and not taking any notice of his words. When the two A\'svins accepted the cup with a desire to drink the Soma thereof, the powerful Indra saw it and said :-- ``If you make them drink Soma out of your own necessity, I will hurl thunderbolt on your head exactly in the same way as I did towards Vi\'svar\=upa.''

The Muni became violently angry at this and made the A\'svins drink the Soma according to due rites and ceremonies. Indra, too, angrily hurled thunderbolt on him in the presence of all the Devas. The weapon shed lustre like million Suns. Seeing the thunderbolt hurled on him, the powerful Muni made the Indra's thunderbolt stand stock still by virtue of his Tapas. The powerful Muni then adopted black magic to kill Indra and offered oblations of clarified butter and grains, purified by Mantrams, in the Fire. By the Tapas of Chyavana, of unbounded lustre, there sprang from the sacrificial hearth Krity\=a (A female deity to whom sacrifices are offered for destructive and magical purposes). And out of Krity\=a originated a very strong person, very cruel and of huge body, a great Demon. The horrible Demon, named Mada, was terrifying to all the beings. His body was huge like a mountain, teeth very sharp and terrible. Four teeth were hundred Yojanas long each, and his other teeth were ten Yojanas long. His arms were like mountains, very long and fierce; tongue, horrible, rough and so very long that it reached up to the heavens. His throat was like a mountain peak hard and of a furious appearance; nails resembled tiger's, hairs horrible. His body was jet black like lamp black; face very terrible, eyes bright like the conflagration fire and awful. One of his jaws touched the ground and the other touched the heaven. Thus was born the Demon, named Mada, of huge form. Looking at him, the Devas became suddenly frightened; Indra, too, got very much terrified at the sight of him and did not want to have any more fight. The Demon swallowed Indra's thunderbolt, looked at the sky, and stood up as if to swallow at once the whole Universe. He became infuriated with rage and pursued Indra to devour him. Seeing this, the Devas cried aloud :-- ``Alas! We are slain.'' Indra had his arms disabled by Mantrams and so he could not hurl his thunderbolt though he wanted to do so. The Lord of the Devas, then, with thunderbolt in his hand, looked on the Demon as Death personified and remembered his Guru, skilled in the knowledge which is the proper time to perform a certain thing. The liberal-minded Brihaspati knowing the time of imminent danger, and that he is remembered by Indra, at once came to the spot. He then judged what to do in the present crisis and told Indra :-- ``O Indra! This cannot be averted even by Mantrams; what to speak of thunderbolt! This powerful Asura Mada has arisen from the sacrificial hearth by virtue of the Tapas of the Chyavana Muni. The Muni's power is especially felt here. O Lord of the Devas! Nobody, You, I, nor any other Deva will be able to resist him. Know this. Even Brahm\=a cannot thwart the anger of one who is devoted to the \'Sakti, the Highest Force; Chyavana is the Bhakta of the Highest \'Sakti. So no

other body is able to defeat him. He is the man himself to take back the Krity\=a that he has originated. There is no doubt in this. So it is better for you now to take the shelter of the Muni.''

30-52. Vy\=asa said :-- O King! Hearing thus from his Guru, Indra went to the Muni and bowed down shuddering, before him. ``O Muni! Forgive me and stop the Asura from his intention to slay the Devas. O All knowing One! Be pleased, I will keep your words. O Bh\=argava! The two A\'svins will, from this day, have the right to drink the Some juice. This I speak out to you in truth. O Br\=ahma\d{n}a! Be graciously pleased unto me. O Ascetic! Your intention will never be baffled. Especially I know you to be a knower of Dharma; so, you will never be able to make your word swerve from truth. The two A\'svins will, by your grace, be able to drink always the Soma cup; and the King \'Sary\=ati's fame will also know no bounds. O Muni! Know that what I have done is simply to test your prowess in Tapas. O Br\=ahma\d{n}a! Now do this favour to me and take back your, this Demon Mada, sprung from your sacrificial hearth and thus do good to all the Devas. There is no doubt in this.'' Thus spoken piteously by Indra, Chyavana, who knew the Highest Reality, drew back within himself the anger arising from the enmity with Indra. Then the Mahar\d{s}i Chyavana consoled the Devas that were very much perplexed and anxious out of terror of the Demon named Mada and divided the Asura into four parts (1) female sex, (2) drinking, (3) gambling and (4) hunting animals. When Mada was thus divided into four parts, the terror stricken Devas felt themselves relieved and saved and got consoled. Chyavana then placed the Devas in their respective stations and completed the sacrifice. As last, the religious Bh\=argava made first Indra and then the two A\'svins drink the Soma Cups. O King! Thus Chyavana had the A\'svins their Soma Cups by virtue of his power of Tapas. Thence the tank with the sacrificial post Y\=upa became famous and the Muni's \=A\'srama also was renowned and honoured in all respects. The King \'Sary\=ati, too, became very glad at this sacrifice and completing the sacrifice returned with his ministers to his city. The Manu's son, the powerful religious King \'Sary\=ati governed his kingdom, free from any obstacles or other enemies. He had one son named \=Anarta; and \=Anarta had a son named Revata born to him.

Revata built the city of Ku\'sasthal\={\i} in the midst of the ocean and began to live there. He enjoyed all the things in the countries named \=Anarta and others. Revata had one hundred sons of whom Kakudmi was the eldest and of pure character. He had one daughter very beautiful named Revat\={\i}, endowed with all auspicious qualities. When the daughter reached a marriageable age, the King began to think where he

could get a prince of a good noble family. That powerful King began to govern his people \=Anartas, with his headquarters at the Raivata hill. ``Whom to betroth this daughter,'' the King thought and settled that he would go to Brahm\=a and ask him, the venerable omniscient Praj\=apati, worshipped by the Devas. Thus the King went with his daughter Revat\={\i} to the Brahmaloka. There the Devas, Yaj\~nas, Vedas, mountains, oceans and rivers all were shining with luminous bodies. There the eternal \d{R}i\d{s}is, Siddhas, Gandharbas, Pannagas and Ch\=aranas were singing hymns to Brahm\=a, standing with folded hands.

Here ends the Seventh Chapter of the Seventh Book on the twin A\'svins drinking the Soma Cup in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, the Mah\=a Pur\=a\d{n}am of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



