\chapter{On the King Revata and the Solar Dynasty}

1-5. Janamejaya said :-- ``O Br\=ahma\d{n}\=a! The King was a K\d{s}attriya; how could he go himself with his daughter Revat\={\i} to the Brahm\=aloka (the abode of Brahm\=a)? I entertain a great doubt on this point. I heard of yore while conversing about matters connected with the Br\=ahmi\d{n}s, that the Br\=ahmi\d{n} only who was self-restrained and the knower of Brahm\=a\d{n} could alone go to the Brahm\=aloka. The Satyaloka is very hard for the worldly people to go; so I doubt how the king could have gone with Revat\={\i} from the Bh\=urloka to the Satyaloka. Man, when he discards his body, can go to the Heavens. So is recited in all the \'S\=astras. How then, people, while, in their human bodies can go to the Brahm\=aloka. So cut asunder my doubts how the King Revata could go to the Brahm\=aloka to ask the Praj\=apati on certain matters.''

6-16. Vy\=asa said :-- ``O King! On the top of the mountain Sumeru, are located the Indra's heavens called Amar\=avati (the abode of the Immortals) the Samyaman\={\i} city of Yama, the Satyaloka, the Vahniloka, the Kail\=asa, Vaikuntha the abode of Vi\d{s}\d{n}u, and others. The great archer Arju\d{n}a, the son of Prith\=a, went to the Indraloka and spent five years there. In ancient times, the Kings Kakutstha and others went to Indraloka, in their human bodies. Even the powerful Daityas used to conquer the Indraloka and resided there at their will and pleasure. In ancient times, when the sovereign of the earth, the King Mah\=abhi\d{s}a went to the Brahm\=aloka, the most beautiful Gang\=a also was coming to the Brahm\=aloka. On the way the King saw Her. O King! Accidentally at that time her clothings were cast aside by the wind; the King saw her partly in her naked state, and, overpowered with lust, smiled; Gang\=a also

smiled. Seeing the states of them, Brahm\=a instantly cursed them; and they had to come in this world and take their births. All the Devas, when oppressed by the D\=anavas, went to Vaikuntha and sang hymns to Hari, the Lord of Kamal\=a. O King! Men can go to all the Lokas; in fact those high-souled men that perform Yaj\~nas or severe asceticism and thus have acquired great merits, those performers of Sacrifices and ascetics surely go to the Heavens. O King! It is only the abundance of good merits that is the only cause of going to Heavens. So you ought not to entertain any doubts on this.''

17-18. Janamejaya said :-- ``O Muni! The King Revata went with his daughter Revat\={\i} to the Brahm\=aloka; but what did he do when he went there? What did Brahm\=a order him? And to whom did the King betroth his daughter, when ordered by Brahm\=a? O Br\=ahma\d{n}a! Speak out all these in details to me now.''

19-21. Vy\=asa said :-- ``O King! Hear. When the King went to Brahm\=aloka to ask about the proper bridegroom of his daughter, there was going on singing and music; so he waited a while to find an opportunity when the assembly would have a leisure; but he was so very pleased with music that he could not desist from hearing it till the end. When the music was finished, the King bowed down to Brahm\=a and shewed him his daughter and informed Him of his intention.''

22-26. The King said :-- ``O Deva! This good daughter is mine; now kindly say who will be her bridegroom. O Brahm\=a! To whom shall I betroth this daughter? I have come to you to ask on this point I have searched for many princes and seen also a good many of them and none of them is to my liking and so my mind is not at rest. O Lord of the Devas! Therefore I have come to you. Kindly select one bridegroom for her. He is to be a Kul\={\i}n (of good family), powerful, religious, liberal, and a prince endowed with all auspicious qualities. This is my prayer.'' Vy\=asa said :-- O King! Brahm\=a, the lotus-born, the Creator of the world, hearing these words, laughed, thinking that a very long interval had passed away. He then said :--

27-43. O King! The princes that you thought would become the bridegroom of your daughter, all died; their sons and grandsons and their friends even have all passed away. The twenty-seventh Manvantara of the Dv\=apara Yuga is now going on; so none of the princes of your family are now existing. The Daityas sacked your city. Now Ugrasena, the King of Mathur\=a, is reigning in that place. He belongs to the illustrious lunar family of Yay\=ati. His son, the powerful Ka\d{n}sa,

born of a D\=anava, began to do injuries always to the Devas; he threw his own father to the prison. Becoming very haughty, he began to govern himself the countries of other kings and began to tyrranise over the subjects. O King. The Earth became so much troubled by the armies of the wicked Demon Kings, that She became quite unable to bear further loads. So She went to seek refuge to Brahm\=a. Brahm\=a and the other Devas then began to say :-- ``O Earth! To remove your burden the lotus-eyed N\=ar\=aya\d{n}\=a will incarnate Himself as part incarnation in the form of \'Sr\={\i} Kri\d{s}\d{n}a. He who is N\=ar\=aya\d{n}\=a practised in ancient times, as the son of Dharma, a very severe asceticism, in company with his brother Nara in the sacred hermitage of Badari. Now this very Deva is born in the great family of Yadu in the womb of Devak\={\i} by Vasudeva and is now celebrated by the name of V\=asudeva. O King! He has slain that vicious wicked Kan\d{s}a and has installed Ugrasena in his place. The very powerful Jar\=asandha, the vicious King of Magadha, is the father-in-law of Kan\d{s}a. On hearing the death of his son-in-law, he became infuriated with rage, came to Mathur\=a, and raged a terrible war. V\=asudeva defeated in a battle that Jar\=asandha, proud of his mighty valour. Though defeated, Jar\=asandha sent K\=alayavana with his host of army to fight again with Kri\d{s}\d{n}a. Bh\=agavan V\=asudeva, when he heard that the King of Yavanas arrived, sent away all the members of his family and the Y\=adavas to Dv\=ark\=a and began to wait with his brother Balar\=ama for the Yavana King. Then he went alone to the camp of Yavana and led him away to a mountain cave where was sleeping the King Muchu Kunda and had then the Yavana King slain by Muchukunda. Kri\d{s}\d{n}a then went to Dv\=ark\=a. The city of Dv\=ark\=a was then a dilapidated condition. Kri\d{s}\d{n}a brought together the artists and got built exquisitely the beautiful palaces, forts, and markets and stalls, etc., and so added to the beauty of the place. That V\=asudeva, of mighty prowess, thus improving the city, made Ugrasena the King of that place; and Kri\d{s}\d{n}a is now waiting there with his friends. His elder brother Baladeva, the carrier of the plough, is celebrated. Thus he with Musala in his hands is a great warrior and the part incarnation of Ananta Deva. He is the fit bridegroom of your daughter. So give your daughter in marriage, without any delay, according to the rules of the marriage ceremony to Sankar\d{s}a\d{n}a Balabhadra. After giving your daughter marriage, go to the hermitage of Badari and practise tapasy\=a. That sacred retreat is the (park) recreation ground of the Devas, holy and yielding to human beings the objects of their desires.

44-46. Vy\=asa said :-- ``O King! Thus ordered by the lotus-born Brahm\=a, the King went to Dv\=ark\=a with his daughter. Reaching there he gave over his all auspicious daughter in marriage to Bala Deva

duly according to the rules and regulations. At last, according to Brahm\=a's injunction, he became engaged in severe austerities in the Badarik\=a\'sram and, when the time of death arrived, left off his mortal coil on the banks of the river and went to the world of Gods.''

47-48. Janamejaya said :-- ``O Bhagav\=an! You have uttered one wonderful thing. One hundred and eight Yugas passed away when the King of Revata with his daughter was deeply absorbed in hearing music in the Brahm\=aloka yet neither the good King nor the daughter did get sufficiently old. How could this be! How could they have lived so long! Were their longevities ordained to be such a long period!''

49-56. Vy\=asa said :-- ``O King! The Brahm\=aloka is not touched by any vice nor sin; old age, hunger, thirst or fear of death nothing exists there; nor is there any other cause by which weariness comes. So what doubt there can be that the people there will be long-lived, free from old age and death! When the King \'Sary\=ati went up to the Heavens, his sons were all destroyed by the R\=aksasas; those that remained, they, terrified left Ku\'sasthal\={\i} and fled on all sides. Vaivasvata Manu sneezed; owing to that, came out of his nose one powerful son; his name was Iks\=aku. He spread the Solar dynasty and became celebrated. Getting excellent initiation from the Mahar\d{s}i N\=arada, he began to meditate the Dev\={\i} constantly and practised severe tapasy\=a for the spread of his race. O King! Iks\=aku had one hundred sons; Vikuksi was the eldest; he was powerful and endowed with great strength. Iks\=aku became king and lived in Ayodhy\=a. He sent his fifty sons, the powerful \'Sakuni and others to Uttar\=apatha (Eastern) provinces for governing those countries. That high-souled monarch sent also other eight sons to govern the countries in the South. (Western). O King! He kept the remaining two sons by his side for his own service.''

Here ends the Eighth Chapter of the Seventh Book on the story of the King Revata and the spread of the Solar dynasty in the Mah\=apura\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



