\chapter{On the getting of youth by Chyavana Muni}

1-6. Vy\=asa said :-- O King! Hearing their words, the princess began to tremble; but holding on patience she spoke to them in reserved terms thus :-- You are the sons of S\=urya and you are the acknowledged deities amongst the gods; specially you know everything, I am a chaste virtuous woman. You ought not to speak to me in the above mannner. O Twin Devas! Father has betrothed me to the Muni practising the Yoga Dharma; besides I am chaste; how can I behave like a prostitute! This Sun is the Witness of the actions good or bad of all people; He is therefore looking on our actions also. Besides you both are born in the family of the high-souled Ka\'syapa. Thus it is utterly wrong for you to utter such (irreligious and infamous) words. You know well the course of Dharma, what is religious and what is irreligious in this world which has got nothing substantial; how can a family woman leave her husband and serve another? Go wherever you like, O Sinless Devas! I am the daughter Sukany\=a of the King \'Sary\=ati, devoted to my husband. Otherwise I will curse you.

7-11. Vy\=asa said :-- O Bharata! Hearing these words, the A\'svins were greatly surprised, and, afraid of the Muni, spoke again :-- ``O Princess! We are very much pleased to see your chastity; therefore, O Beautiful Woman! Ask boon from us; we will grant it for your welfare. O honoured Woman! We, the physicians of the gods, will certainly make your husband exceedingly beautiful and young, O Smart and Intelligent One! When we three will be exactly the same in figure, age and lustre! you better can make one of us your husband.'' Hearing their words Sukany\=a became greatly amazed and went to her own husband and spoke everything what they, the God's physicians, said.

12-17. Sukany\=a said :-- ``O Husband! The A\'svins, the S\=urya's sons, have come close to our \=A\'srama. I have seen the two Devas and their bright bodies. Seeing me beautiful they were overpowered with passion and told me, ``We will, be sure, make your blind husband, young, bright and give him his two eyes again; but you will have to make one condition. Hear it :--Your husband will be exactly like us and you will

have to select your husband amongst three of us.'' O Holy One! Hearing this as strange, I now come to inform you. Now judge and say what I am to do now. The Deva's Maya is very difficult to comprehend; the more so, when I do not know their intentions! O Omniscient One! I will act as you desire.''

18-19. Chyavana said :-- ``O Beloved! Go just now, at my word, to the A\'svins and bring them, O auspicious One! before me. What more shall I say than this :-- Go and observe, what they say, as early as possible. There is no need to think over this matter.''

20-25. Vy\=asa said :-- O King! Thus getting the permission from her husband, Sukany\=a went immediately to them and said :-- ``O A\'svins! You are the chief gods; now do as you say. I agree to observe what you desire.'' Hearing her words, the two Devas then went to the Muni's \=A\'srama and told the princess :-- ``Let your husband enter in the midst of the water.'' The aged Chyavana Muni went down quickly in the midst of waters to attain a good form. Next the two A\'svins entered into the water of that tank. A few minutes after, the three persons came out of the tank. All were equally bright, equally beautiful, equally young and their limbs were decorated equally with earrings and various other ornaments. They all spoke simultaneously :--``O Auspicious One! There is no other woman beautiful like you, especially your face is very clean and fair; therefore select any one of us three as your husband. O Fair One! Whomever you love most, choose him.''

26-30. Vy\=asa said :-- O King! Sukany\=a then saw their bodies are equally bright and beautiful; not the least difference is to be found in beauty, age, voice and dress. She became doubtful on seeing their equal appearances. The princess, not being able to distinguish her husband, became very anxious and thought :-- ``What am I to do now? Whom to choose? They are exactly the same. I cannot distinguish who is my real husband? This may be the magic set up by the two A\'svins. However, I am put to a great crisis. I won't ever select another who is not my husband. Therefore my death is well nigh; what to do now? The third form seen now may be also a Deva's son.'' Thus cogitating, she resolved to meditate on the Highest Prakriti, the Lady of the Universe, the most Auspicious One. Then the thin-bellied princess began to sing the hymns of the Bhagavat\={\i}.

31-38. Sukany\=a said :-- ``O World-Mother! Under most painful circumstances I take refuge unto Thee; preserve my chastity; I bow

down to Thy feet. O Dev\={\i}! Salutations to Thee, born of lotus. O Thou, the dear consort of \'Sa\d{n}kara! Salutations to Thee. O Thou favourite to Vi\d{s}\d{n}u, O Mother of the Vedas! O Sarasvat\={\i}! Salutations to Thee, Thou hast created the world, moving and unmoving; Thou art preserving it without being least excited; again Thou art swallowing it for the peace and well-being of all. What more, Thou art the Most Worshipful Mother of Brahm\=a, Vi\d{s}\d{n}u, and Mahe\'sa. Thou always illuminest the understanding of the illiterate and Thou always grantest liberation to the J\~n\=anins. Thou art the Prime Prakriti in fullness and the Beloved of the Prime Puru\d{s}a. Thou grantest Bhukti (enjoyment) and Mukti (freedom) to the souls that are cleansed and pure; Thou givest pain to those that are entirely void of knowledge and Thou grantest happiness to those that are endowed with Sattva Gu\d{n}a. O Mother! Thou bestowest Siddhi (the success, the eight supernatural powers), fame, and victory to the Yogis! Being merged in an ocean of bewilderment, I come now to take refuge unto Thee. O Mother! The two Devas are playing hypocrisy with me; thus puzzled I can't fix my mind whom to select; therefore I am merged in an ocean of sorrow. Save me by showing my real husband. O Omniscient One! Knowing my vow of chastity dost Thou enlighten me so that I can know my husband.''

39-58. Vy\=asa said :-- O King! Thus pleased by the Sukany\=a's prayers, the Dev\={\i} Tripur\=a Sundar\={\i} then imparted to her the pleasant Sattva J\~n\=an (knowledge pertaining to Sattva Gu\d{n}a). She then looked again at the three personages, and though they were similar in appearance and beauty, instantly she recognised mentally her husband and chose him. When Sukany\=a selected the Muni Chyavana, the two Devas became greatly pleased to see that. The two Devas were pleased by the grace of Bh\=agavat\={\i}; they were further pleased to see the Dharma of chastity and granted her the boon. They then bade good-bye to Chyavana and were ready to start to their own place when Chyavana being very much pleased to get through their grace, his beauty, youth and wife, interrupted them, saying ``O high-minded Devas! You have done much good to me. I used to feel pain every day, in spite of my having this wife having good hairs! But owing to your mercy, I cannot describe how happy I am now in this world of woes and troubles. I was very aged and blind and was without any enjoyment but it is you that coming to this forest have brought to me eyes, youth and exquisite beauty. Therefore, O twin Devas! I desire to do something good to you in return. Fie on him, who does not return anything for the good that he has received from a beneficent friend. That man remains indebted

for ever in this world; therefore I am desirous to give you two now whatever you want. O twin Devas! Even if the Devas or the Asuras find it difficult to attain, I will give that to you to free myself of the debt I owe to you. I am greatly pleased at your good deed; therefore be kind enough to speak out whatever you desire.'' They began to consult with each other, and spoke to the Muni Chyavana who was seated with Sukanya beside him :-- ``O Mahar\d{s}i! We have got all our desires by the grace of Father! Still it is difficult for us to drink some libation along with the other Devas and we thirst after that very strongly. At the great sacrifice of Brahm\=a in the Golden Mountain (Kanak\=achala), Indra, the King of the Devas, ordered us not to drink Soma, as we were physicians. Therefore, O Knower of Dharma! O Ascetic! You will certainly do us a great favour if you can make us drink the Soma juice; we would be very glad and have our desires satisfied.'' Thus hearing the words of the A\'svins, Mahar\d{s}i Chyavana gladly spoke to them the following gentle words, O Twin Devas! I was blind and aged; but now I am become a young and beautiful man, and it is by your grace that I have got back my wife. Therefore I speak this truly before you that at the great sacrifice of the lustrous King \'Sary\=ati, I will gladly make you drink the Soma in the presence of Indra, the King of the Devas. Hearing these words of the Muni, the twin A\'svins were greatly pleased and went back to the world of the Devas. And the Muni Chyavana, too, returned to his own A\'srama with his wife Sukany\=a.

Here ends the Fifth Chapter of the Seventh Book on the getting of youth by Chyavana Muni in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam the Mah\=a Pur\=a\d{n}am of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



