\chapter{On the stay of Hari\'schandra in the burning ground}

1. \'Saunaka said :-- ``O S\=uta! Now describe as quickly as you can in detail what the King Hari\'schandra did afterwards in the house of the Ch\=and\=ala.

2-14. S\=uta said :-- When Vi\'sv\=amitra went away, the mind of the Ch\=and\=ala was filled with joy. He already gave to Vi\'sv\=amitra that amount of jewels; so he tied now the King and, telling him, ``Do you now stand on the path of falsehood?'' began to beat him with sticks. The King was already very much tired of the bereavements from his dear ones; now being beaten by the Ch\=and\=ala, his senses were lost. In this state the Ch\=and\=ala took him to his house and fastened him with a chain. Then the Ch\=and\=ala's troubles were over and he fell asleep. The King lived in the Ch\=and\=ala's house in that state fettered by a chain; but he did not take any food there. Incessantly he wept for his wife and son and others. ``Alas! That thin lady, seeing the sad face of

her son is now remembering me with a morose face. She is now perhaps thinking, with a grieved heart, that whenever the King will get the money, he will pay off the promised money to the Br\=ahmi\d{n} and then will free us from this yoke of slavery. Alas! When will that day come when will he see me and this crying child and speak with us. When the son will cry, saying, `I will go to my father; father!' When will he come and speak with the child? That fawn eyed gentle woman does not know that I am now placed under a Ch\=and\=ala. Alas! I am deprived of my kingdom, friends; and I have sold away my wife and son; now I am bound in the chain of slavery of a Ch\=and\=ala. Alas! So many miseries have fallen on me all one after another.'' Thus thinking incessantly about his dear consort and son, the King passed his days in the house of that Ch\=and\=ala. Four days passed; and on the fifth day the Ch\=and\=ala came there and rebuked the King with very harsh words and freed him from his fastenings and said, ``Go to the burial ground and collect the clothings of the dead bodies. There is a wide \'Smas\=an (burial ground) on the southern part of K\=a\'s\={\i}; go and protect that and whatever is due to you, justly take that; do not leave it. Take this Jarjara club and go there quickly. Say to all that you are the messenger of V\={\i}rav\=ahu and this staff is his.''

15-33. S\=uta said :-- O \d{R}i\d{s}is! Thus Hari\'schandra became a Ch\=and\=ala's servant and was engaged in collecting the cloths of the dead persons. Thus ordered by the Ch\=and\=ala, whose duty was to collect the rags of the dead bodies, the King went to the burial ground. To the south of the city K\=a\'s\={\i}, was situated the dreadful \'Smas\=ana, scattered over with the garlands of the dead, bad odours were emitting on all sides and it was covered all over with smoke. Hundreds of jackals were yelling there and the ground was being reverberated by their yells. Vultures, jackals and dogs were at many places dragging the dead bodies. At other places were scattered heaps of bones; the whole ground was covered with the putrid smell of the dead. At some places it seemed that from within the funeral pyre, the half-burnt dead bodies were laughing wildly with their teeth wide open from their mouths. Thus the dead bodies looked terrible when being placed under fire. Lots of dead bodies were brought there and there was a great tumultuous uproar made by the cries of their friends and relatives. Oh! My son! My friend! My relative! My brother! My child! My dear wife! Oh! My cousin! Oh! My grandfather! Oh! My father! My grandson! My acquaintance! Where hast thou gone leaving me here! Come once and let me have a sight of thee! With such dreadful sounds as these, the burial ground was being echoed. Flesh, marrow, fat all were being burnt in the

fire and a peculiar sound So\d{n}, So\d{n} was being produced there and creating voidness in the minds of the people. The fire was burning with a crackling noise. Thus the \'Smas\=ana looked very terrible as if the universe was being destroyed at the end of a Kalpa. The King Hari\'schandra arrived there; and, with extreme pain, he began to give vent to his sorrows. ``My ministers, servants! Where are you all now? Where is kingdom that I got by a succession of inheritance! O my Son! O my dear wife! Where are you staying now, at what a long distance, leaving me here out of the Br\=ahmi\d{n}'s anger. Without Dharma man can never get auspicious fruits. So men should carefully earn Dharma.'' The King, whose body was covered with dust and dirt, thought thus repeatedly; and at last, remembering the Ch\=and\=ala's words, went out in quest of the dead. Out of this eventful cares and anxieties, his body became lean like a stick; still he ran, to and fro, and calculated thus :-- ``This dead body will fetch for its price one hundred gold mohurs; out of this, this belongs to the King; this to me, and this to the Ch\=and\=ala.'' So he thought constantly and his state became awful. His face, arm, belly and feet and the other parts of body were all covered over with ashes and dust; the King wore a ragged cloth where hundred places were sewn over; his toes were all besmeared with all sorts of flesh, marrow, fat and other things. He began to satisfy his hunger out of the food that was prepared for all sorts of dead bodies; and, taking their garlands, he encircled his head with them. Day and night he did not sleep and always he sighed and sighed, crying, Alas! Alas! Thus one year passed away, as if it had been three hundred years.

Here ends the Twenty-fourth Chapter of the Seventh Book on the stay of Hari\'schandra in the burning ground in the Mah\=a Pur\=a\d{n}am, \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



