\chapter{On the glory of the \'Satak\d{s}i Dev\={\i}}

1-3. Janamejaya said :-- ``O \d{R}i\d{s}i! Wonderful is the story of the religious R\=ajar\d{s}i Hari\'schandra that you have described, the great Bhakta of \'Sat\=ak\d{s}\={\i} Dev\={\i}! Why is that auspicious \'Siv\=a, the wife of \'Siva, called \'Sat\=ak\d{s}\={\i}? Explain to me, the cause of it, O Muni! And thus make my birth full of use and success. Who is there amongst the clear-minded that gets fully satisfied, when he hears the good deeds of the Dev\={\i}? Each sentence, describing the good deeds of the Dev\={\i}, gives the undecaying fruits of A\'svamedha Sacrifice.''

4-45. Vy\=asa said :-- O King. Hear; I am describing the story of \'Sat\=ak\d{s}\={\i} Dev\={\i}. You are the great devotee of the Dev\={\i}; so I have nothing that I cannot say to you. In olden times, there was a great

D\=anava named Durgama: he was very cruel. He, the the son of Ruru, was born in the family of Hira\d{n}y\=ak\d{s}a. Once be thought within himself thus :-- ``The Munis offer oblations by Mantras as ordained in the Vedas. And the Devas, eating the clarified butter (ghee) of these oblations, get
nurtured and strengthened. The Vedas is the strength of the Devas; if the Vedas be destroyed, the Devas also would be destroyed. Thus it is advisable to destroy the Vedas. (There is no other easy way.)'' Thus thinking, he went to the Him\=alay\=as to perform tapasy\=a. He began to meditate Brahm\=a in the space of his heart, and, taking air only, passed away his time. [Mark here that all the Devas reside in space, a magnitude of the Fourth Dimension.] He practised hard tapasy\=a for one thousand years and the Devas and the Asuras and all the Lokas were agitated by the power of his Tejas (fiery lustre). Then the Bhagav\=an, the four-faced Brahm\=a, became pleased with him and mounting on his carrier, the Swan came up there to grant him the boon. Brahm\=a told clearly the Demon, sitting in Sam\=adhi with his eyes closed, ``Let all be well with you; now ask what you desire? Satisfied with your tapasy\=a, I have come to grant you the boon.'' Hearing thus, the Demon got up from his Sam\=adhi and worshipping Him duly, said :-- ``O Lord of the Devas! Give me all theVedas. O Mahe\'svara! Let all the Vedic Mantrams, that are found in the three worlds, with the Br\=ahma\d{n}as and the Devas, come to me and give me such strength as would enable me to conquer the Devas.'' Hearing this, the God Brahm\=a, the author of the four Vedas, replied, ``Let it be as you wish,'' and went away. From that time, the Br\=ahma\d{n}as forgot all about the Vedas. So bathing, Sandhy\=a, daily Homas, \'Sr\=addha, sacrifice, and Japam and other rites and performances, all became extinct. Then a cry of universal distress arose on the surface of this wide earth; the Br\=ahmi\d{n}s began to say to each other :-- ``How has this happened! How has this come to pass! Now what are we to do? Where the Vedas have disappeared.'' Thus when great calamities befell on the earth, the Devas became gradually weaker and weaker, not getting their share of the sacrificial Havis. At this time, that Demon invested the city of Amar\=avat\={\i}. And the Devas, not being able to fight with the Asura, of a thunder-like body fled to various directions. They took refuge in the caves of the mountain Sumeru and the inaccessible passes of the mountain and began to meditate on the Highest Force, the Great Goddess. O King! When oblations of clarified butter are offered to the Fire, those get transferred to the Sun (S\=uryaloka) and get transformed as rains. So when the Homa ceremonies disappeared, there was the scarcity of rain. The earth became quite dry and not a drop of water was found anywhere. The wells, tanks, pools, rivers all were dried up. And this state of ``no rain\'s' lasted one hundred years. Countless people, hundreds and thou-

sands of cows, buffaloes and other beasts went to the jaws of death. The dead bodies of persons remained in heaps in every house; persons would not be found to perform their burning ceremonies. When such calamities were seen, the calm and quiet body of the Br\=ahma\d{n}s, in their earnestness to worship the Supreme Goddess, went to the Him\=alay\=as. They with their whole heart and without taking any food began to worship the Dev\={\i} daily with their Sam\=adhi, meditation and worship. O Mahe\'s\=an\={\i}! Shew mercy on us. O Mother! It's not praiseworthy to Thee to manifest Thy such anger on us, the low persons and guilty of all sins. So, O Deve\'s\={\i}! Forgive us. If Thou art angry on us for our faults, even then we may be excused, for Thou art the Internal Ruler within us all and we do whatever Thou impellest us to do. (The other Devas become pleased and give fruits when they are worshipped by Japam, and other Homa ceremonies; but that is not even possible due to the disappearance of the Vedic Mantrams from amongst us. But You are kind as mothers are towards their children whenever they remember.) So without Thee, there is no other rescue for these people. O Mahe\'svar\={\i}! Whatever Thou willest, Thou canst do that; so what art Thou seeing again and again? O Mahe\'sar\={\i}! How can we live without Water, what is called the Life. Now rescue us from this great difficulty. O Mother of the Worlds! O Mahe\'svar\={\i}! Be pleased. Oh the Ruler of the endless crores of Brahm\=andas! Obeisance to Thee! We bow down to Thee, the Unchangeable, of the nature of Intelligence. We again and again make obeisance to Thee, the Lady of the Universe and realisable by the Ved\=anta words (not this, not this). All the sayings of the Ved\=anta declare Thee, by negating (not this, not this) other transient objects as the Cause of all this Universe. We with all our hearts bow down to the Dev\={\i}. When the body of the Br\=ahma\d{n}as thus praised and chanted the hymns of Mahe\'svar\={\i}, She created innumerable eyes within Her body and became visible. Her colour was dark-blue (colour of the fourth dimension, space) like heaps of collyrium (eye-paint); eyes like the blue lotuses and expanded; breasts hard, regularly elevated round and so fleshy that they touched each other; four handed; with Her right hand, holding arrows; on the under hand holding lotus; on the upper-left hand holding a great bow and on the lower hand, carrying vegetables, fruits, flower and roots with abundance of juice, destroying hunger, thirst and fever. She was the Essence of all Beauty, lovely, luminous like the thousand Suns, and the ocean of mercy. That Upholder of the Universe, showed Her form and began to shed waters from Her eyes. For nine nights continuously, the heavy rains poured down out of the waters flowing from Her eyes. Seeing the misery of all the people, out of

pity, She showered incessantly tears from Her eyes; and all the people and medicines were satisfied. What more than this, out of those tears, the rivers began to flow. The Devas that remained hidden in the mountain caves, now came out. Then the Br\=ahmi\d{n}s, united with the Devas, began to praise and sing hymns to the Dev\={\i}. Thou art known by the Ved\=anta Mah\=av\=akyas. We bow down to Thee. Thou ordained everything to all the worlds by Thy M\=ay\=a; so again and again we bow down to Thee. Our Obeisance to Thee! Who art a Kalpa tree to the Bhaktas yielding all their desires! Thou assumest the body for the Bhaktas! Thou art always satisfied; without any equal; the Lord of the Universe! We bow down to Thee. As Thou, O Dev\={\i}! hast innumerable eyes only for our welfare and peace, therefore Thou wilt be called henceforth by the name ``\'Sat\=ak\d{s}\={\i}.'' O Mother! We are very much hungry; so we have no power to chant hymns to Thee; therefore, O Mahe\'svar\={\i}? Shew mercy on us and deliver to us our Vedas.

46-68. Vy\=asa said :-- O King! Hearing these words of the Devas and the Br\=ahmi\d{n}s, the Auspicious One gave them the vegetables, delicious fruits and roots to them that were on Her Hand, for their eating. After She was prayed, She gave to men sufficient quantity of various articles of juicy food and to the beasts, grass, etc., until new crops came out. O King, from that day She became famous by the name of \'S\=akambhar\={\i} (because She nourished all by vegetables, etc.) Great tumult arose and the Demon Durgama heard all from the emissaries and started out to fight with his weapons and army. He took one thousand Ak\d{s}auhi\d{n}\={\i} armies with him (one Ak\d{s}auhi\d{n}\={\i} army equals large army consisting of 21,870 chariots, as many elephants, 65,610 horses, and 109,350 foot) and, shooting arrows, he came quickly before the Dev\={\i} and invested Her and the Deva army and the Br\=ahmi\d{n}s. At this, a great tumultuous uproar arose and the Devas and the Br\=ahmi\d{n}s united exclaimed :-- ``O Dev\={\i}! Save us; save us.'' The Auspicious Dev\={\i}, then, for the safety of the Devas and the Dv\={\i}jas created round them a luminous circle and She Herself remained outside. The terrible fight, then, ensued between the Dev\={\i} and the D\=anavas. The Sun was covered with their incessant hurling of arrows; and the shooters could not shoot accurately on account of the darkness that then prevailed. Then by the collision of the arrows of both the parties, the arrows caught fire and the battlefield again became filled with light. The quarters on all aides resounded with harsh bow sounds and nothing could be heard. At this moment, came, out of the body of the Dev\={\i}, the principal \'Saktis (forces incarnate) K\=alik\=a, T\=ari\d{n}\={\i}, Soda\'s\={\i}, Tripur\=a, Bhairab\={\i}, Kamal\=a, Bagal\=a, M\=atang\={\i}, Tripur\=a Sundar\={\i}, K\=am\=aks\={\i}, Tulaj\=a Dev\={\i}, Jambhin\={\i}, Mohin\={\i}, Chchinnama\d{s}t\=a,

and ten thousand armed Guhya K\=al\={\i}s and others. Thirty-two \'Saktis, sixty-four \'Saktis, and then innumerable \'Saktis, all armed, came out of the Dev\={\i} successively. When the \'Saktis destroyed one hundred Ak\d{s}auhi\d{n}\={\i} forces, Mridangas, conch-shells, lutes and other musical instruments were sounded in the battle-field. At this time, the enemy of the Devas, Durgama, came in front and first fought with the \'Saktis. The fight grew to such a terrible extent that, within ten days, all the Ak\d{s}auhi\d{n}\={\i} troops were destroyed. So much so as the blood of the dead soldiers began to flow in torrents like rivers. When the fatal eleventh day arrive the D\=anava, wearing red clothes on his waist, red garlands on his neck and annointing his body all over with red sandal paste, celebrated a very grand festivity and mounted on his chariot and went out to fight. With the strenuous effort, he defeated all the \'Saktis and placed his chariot before the Dev\={\i}. Then a terrible fight ensued for two Praharas (six hours). The hearts of all shivered with horror. At this time, the Dev\={\i} shot fifteen very awful arrows at the D\=anava. His four horses (V\=ahanas) were pierced by Her four arrows; the charioteer was pierced by one arrow; his two eyes were pierced by two arrows; his arms by two arrows, his flag by one arrow and his heart was pierced by five arrows. He then left his body before the Dev\={\i}, vomitting blood. The vital spirit, the luminous counterpart, emitting from his body, merged in the space-like body of the Dev\={\i}. The three worlds, then, assumed a peaceful appearance when that greatly powerful D\=anava was killed. Then Hari, Hara, Brahm\=a and the other Devas began to praise and chant hymns to the World Mother with great devotion and in voices, choked with feelings.

69-73. The Devas said :-- ``O Auspicious One! Thou art the only Cause of this Illusion of this world, presenting an unreal appearance (while Brahm\=a is the Only Reality). So Thou art the Lady of all the beings (otherwise why it would be that Thou hast nourished all the beings with vegetables, etc). So, Obeisance to Thee, the \'S\=akambhar\={\i}! Hundred-eyed! O Auspicious One! Thou art sung in all the Upani\d{s}adas; The Destroyer of the Durgama Asura! We bow down to Thee, the Lord of M\=ay\=a, the Dweller in the five sheaths Anna, Rasa, etc. We meditate upon Thee, the Lady of the universe, as demonstrated by Pra\d{n}ava Aum, whom the chief Munis meditate with their Nirvikalpa hearts (hearts free from any Vikalpa, doubts or ignorance). Thou art the Mother of the endless crores of universe! Thou assumest the Divine Bodies at times for our welfare! Thou art the Mother of Brahm\=a, Vi\d{s}\d{n}u and others; we bow down to Thee with all our heart.

Thou art the Mother of all; so, out of mercy, Thou hast shed tears from the hundred eyes, to remove the miseries of the low humble persons. Thou art the Ruler of all!''

74-80. Vy\=asa said :-- O King! Thus when Brahm\=a, Vi\d{s}\d{n}u, Hara and the other Devas praised and chanted various hymns to the Dev\={\i} and worshipped Her with various excellent articles, She became instantly pleased. Then the Dev\={\i}, graciously pleased, handed over the Vedas to the Br\=ahma\d{n}as. At last, She, the Cuckoo-voiced, made a special address to them. ``These Vedas are the excellent parts of My body. So preserve these with your greatest care. The more so, when you all have seen with your own eyes what a great calamity befell on you when these Vedas went away out of your hands! You should all worship and serve Me (the Controller of the Space) always; there is no other thing higher than this that I can advise you for your welfare. Read always these My excellent glorious deeds. I will be pleased thereby and will destroy all your bad calamities and misfortunes. My name is Durg\=a, because I have killed this demon Durgama; so he, who will take My name Durg\=a and \'Sat\=ak\d{s}\={\i}, he will be able to unveil my M\=ay\=a and walk freely. No use in telling more than this that I tell you now, O Devas, the Essence of all essences :-- Both the Suras and the Asuras would always serve Me and Me, alone.''

81-83. Vy\=asa said :-- O King! Thus giving pleasures to the Devas by these words, the Dev\={\i} of the nature of Existence, Intelligence and Bliss disappeared before them. O King! This Grand Mystery I have described to you in detail; but this is the source of good to all; so keep it secret with every care. The person that hears daily with great devotion this Chapter, gets all that he wants and at last gets the worship in the Dev\={\i} Loka.

Here ends the Twenty-eighth Chapter of the Seventh Book on the glory of the \'Satak\d{s}i Dev\={\i} in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



