\chapter{On the King Hari\'schandra's acknowledging of the slavery of the Ch\=and\=ala}

1-5. Vy\=asa said :-- O King! Speaking these harsh and cruel words to the King, the Muni took that money and went away. When Vi\'sv\=amitra went away, the King Hari\'schandra became very much perplexed with sorrows and sighed frequently. He then began to say with his face bent downwards. ``Suffering from constant pain and troubles, I am now turned into a Preta; if anybody finds me serviceable, he may purchase me with value in gold as proper; but he should do this quickly before the sun sets.'' Dharma, then, assuming the form of a heartless

Ch\=and\=ala, came there quickly to test Hari\'schandra. The body of that low mean person was of a black colour, his air looking ferocious, his belly elongated, body emitting stench odours, teeth very long, and his face, covered with beards. He had one bamboo in his hand; in his neck, the bones of the dead were hanging and his chest was very distorted.

6. The Ch\=and\=ala said :-- ``I am in urgent need of a servant; I will keep you as my slave; say, then, quickly what is your price?''

7. Vy\=asa said :-- O King! When the cruel, extremely ferocious and heartless Ch\=and\=ala said thus, the King Hari\'schandra was surprised to see his appearance and said :-- ``Who are you?''

8-12. The Ch\=and\=ala said :-- ``O King! I am the famous Ch\=and\=ala, Prav\={\i}ra; you will have to remain always subject to me and to collect the clothes of the dead persons.'' Hearing his word, the King said, ``I want to be purchased by a Br\=ahmi\d{n} or a K\d{s}attriya. See! The sages say, that the Dharma of good people is excellent; the Dharma of the persons intermediate is middling; and the Dharma of the mean is depressing. You belong to the low and mean class. So my Dharma cannot be observed if I remain in your house.'' The Ch\=and\=ala said :-- ``O King! This is the Dharma of yours now mentioned by you; then why did you mention that anybody can purchase you; without any previous consideration, you spoke before me. He who speaks with preconsideration attains his desired object; but, O Sinless One! You did not consider and you spoke that ordinarily. However, if I take your words that you spoke first to be true, then you are no doubt, purchased by me.''

13. Hari\'schandra said :-- The villain that speaks untruth, goes downright to a terrible hell; so to become a Ch\=and\=ala is far better for me than to use an untrue word.

14-15. Vy\=asa said :-- O King! When the King was speaking thus, the ascetic Vi\'sv\=amitra arrived there out of anger and impatience; he rolled his eyes and said :-- This Ch\=and\=ala is come to give you your desired money; why, then, are you not giving me the remnant of my Dak\d{s}i\d{n}\=a!

16. Hari\'schandra said :-- ``O Kau\'sika! Nothing is unknown to you. My this body is born for the Solar Line; how then can I accept this slavery of a Ch\=and\=ala!''

17-20. Vi\'sv\=amitra said :-- If you do not sell yourself to a Ch\=and\=ala, be certain that I will just now put you under my curse. Give me immediately

my Dak\d{s}i\d{n}\=a, be it whether from a Ch\=and\=ala or from a Br\=ahma\d{n}a. There is no other purchaser at present than this Ch\=and\=ala. But know this as certain that I won't go back until I get my money. O King! If you do not give me money just now, then when half the Ghatik\=a of the day is remaining, I will burn you up by my fire of anger.

2l. Vy\=asa said :-- O King! Hearing these words of Vi\'sv\=amitra, the King became almost dead; bewildered with fear, then he clasped the feet of the \d{R}i\d{s}i and said, ``Be friendly, please.''

22-23. Hari\'schandra said :-- ``O Viprar\d{s}i! I am now very humiliated and have become very afflicted and distressed. Especially I am your Bhakta, I am your servant; so be graciously pleased and free me from this painful companion of a Ch\=and\=ala. O Muni! In lieu of my remnant Dak\d{s}i\d{n}\=a, I will be your obedient slave; I will do your work and follow your commands.''

24. Visvamitra said :-- ``O King! You are then my slave, you will obey always my commands.''

25-26. Vy\=asa said :-- O King! When Vi\'sv\=amitra said so, the King, out of joy, thought that he regained his life and said to Kau\'sika. Always I will obey your words; now order me what work I will have to do.

27-28. Vi\'sv\=amitra, then addressed the Ch\=and\=ala and said :-- ``O Ch\=and\=ala! Come to me and give me the price for this slave. I am now handing this slave over to you; give me the price and take him. I want money; I have no need for a servant.''

29. Vy\=asa said :-- O King! When Vi\'sv\=amitra spoke thus, the Ch\=and\=ala, overflowed with joy, came immediately to the \d{R}i\d{s}i Vi\'sv\=amitra and said :--

30. O Dv\={\i}ja! The relief that you have given me by selling this servant, for that I will give you the ten Yoyanas wide land of Pray\=aga Mandalam, covered over with jewels.

31-36. Vy\=asa said :-- O King! The Ch\=and\=ala then gave one thousand gems, one thousand jewels, one thousand pearls and one thousand gold Mohurs and Vi\'sv\=amitra took them. No signs of distraction nor unpleasantness were visible on the face of the King Hari\'schandra. Rather he laid hold of his patience and thought within himself, ``Vi\'sv\=amitra is now my master; I will do any work that he puts

me in.'' At this time, the incorporeal voice, the voice of the fourth dimensional space, sounded from the Heavens :-- ``O Fortunate One! You are freed from the Dak\d{s}i\d{n}\=a, the debt before that you promised to give me.'' A shower of flowers fell on the head of the King from the Heavens. At this time the powerful Indra and the other hosts of the Devas praised the King, saying :-- ``S\=adhu! S\=adhu! Well-done, Well-done.'' The heart of the King was then filled with intense joy and the King then said to Kau\'sika :--

37-38. O Intelligent One! You are a greater benefactor to me than my father, mother and friend as you have freed me in a moment from my debts. So, O mighty armed one! Your words are beneficial to me. Now order what am I to do.

39. When the King said so, Vi\'sv\=amitra then said :-- Go and observe from today the words of the Ch\=and\=ala. Let good befall on you! Thus saying, the Mahar\d{s}i Vi\'sv\=amitra took the money given by the Ch\=and\=ala and went away to his own place.

Here ends the Twenty-third Chapter of the Seventh Book on the King Hari\'schandra's acknowledging of the slavery of the Ch\=and\=ala in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



