\chapter{On the selling of Hari\'schandra's wife}

1-6. Vy\=asa said :-- O King! When the Queen Madhav\={\i} requested repeatedly the King, He said :-- ``O Good Auspicious One! When you have not met with any scruple to utter clearly these harsh and cruel words, I will do that act now which the most ruthless persons do not dare to do.'' Saying this, the King went with his wife, very distressed, to the city. Placing her on the public road, the King cried out in a voice choked with feelings and eyes full of tears :-- ``O Citizens! Hear you all. Do any one of you require any maidservant? This lady is dearer to me than my life. If any of you be able to offer price of her as I will declare, then let him give it out quickly.'' The Pundits then said :-- ``Who are you? Why are you come here to sell your wife?''

7. The King said :-- ``Are you asking me of my introduction? Hear then; I am a heartless brute and not fit to be called a man; or I am a R\=ak\d{s}asa; nay, I am more than that; I am prepared to do this sinful act.''

8-11. Vy\=asa said :-- O King! Hearing this, Kau\'sika suddenly assumed the form of an old man and came out and spoke to Hari\'schandra :-- I am master of boundless wealth; so I am able to give you the money you want; I am ready to purchase the maidservant by giving an equivalent wealth. Better give me the maidservant. My wife is exceedingly delicate; she is unable to do all the household work; so let me have the maid. But say quickly what value am I to pay? When the Br\=ahmi\d{n} spoke this, Hari\'schandra felt his heart, as it were, torn asunder; so he could not for the moment speak anything.

12-15. The Br\=ahma\d{n}a said :-- Take an equivalent amount of money according to the age, beauty, qualifications and capabilities of your wife and hand her over to me. Hear about the prices of the male and female servants as written in the Dharma \'S\=astras :-- The price of a female servant clever, good, well-qualified and possessing thirty-two auspicious qualities is one Koti gold mohurs; and the male servant similarly qualified fetches one Arbuda gold mohurs. Hari\'schandra became very much pained to hear the Br\=ahmi\d{n} speaking thus; but he could not say anything. The Br\=ahmi\d{n} then placed in front of the King the money over a bark and caught hold of the hair of the Queen and was ready to drag Her.

16-21. The Queen said :-- ``O \=Arya! Let me see once the lotus-face of my son; leave me once. O Br\=ahmi\d{n}! Please see that it will be hard for me again to see this boy. O Son! Behold! Your mother is now a slave. So, O Prince, do not touch me. I am not fit now to be touched by you.'' The boy, then, seeing the mother suddenly snatched away, cried out, ``O Mother! O Mother!'' and followed her with tears in his eyes. That boy tumbled at every step still he caught hold of the mother's clothing by his hand and began to accompany her. The Br\=ahmi\d{n}, seeing this behaviour of the boy, became impatient with anger and began to beat him. Still the boy wept, saying, ``Mother! Mother!'' and never quitted the hold of his mother. The Queen said :-- ``O Lord! Have mercy on me and purchase this boy also. Though you are purchasing me, yet without this boy I will not be able to do your work. My fate is bad; therefore this calamity has happened. Shew this favour to me.''

22-24. The Br\=ahmi\d{n} said :-- Take this money and give me the boy too. For the Wise in the Dharma \'S\=astras fix such to be the prices of a female and a male. The other Pundits make differences in the prices, e.g., one hundred, one thousand, one lakh, one crore and so on, according to the different qualifications. But for the female, who is skilled in all actions, modest, of good behaviour, and well qualified and, on whose body the thirty-two auspicious signs are seen, her price is one Koti gold mohurs and for a man qualified, one Arbuda gold mohurs.

25-35. S\=uta said :-- O King! The Br\=ahmi\d{n} then gave over the price of the boy as decided, in gold mohurs in front of the King over a bark and then tied both the mother and son. He, then, gladly and without any delay, carried them to his home. At the time of departure, the Queen circumambulated the King and, kneeling down, bowed down to him and, in that state of humility, began to speak :-- If ever I have done any charities, if ever I have poured oblations on the Fire, if ever I have satisfied the Br\=ahmi\d{n}s, then, by that virtue, Hari\'schandra will again be my husband. Seeing his wife, dearer then his life, fallen on his feet, the King became very distracted and lamented, crying, ``Alas! Alas! The shadow of a tree never leaves the tree; but you being verily modest and endowed with all qualifications, are now separated from me.'' Speaking thus reasonably with his wife, the King said to his son :-- ``O Child! Where will you go, leaving me here? Where shall I go now? and who will stop my miseries?'' The King, then, spoke to the Br\=ahmi\d{n} :-- ``O Br\=ahmi\d{n}! The pain that I experience in the separation from my son, I did not feel on the occasion of quitting my kingdom or on my being exiled in a forest.

O Auspicious One! The husband, good natured in this world, nourishes always his wife and keeps her always in comfort and happiness. But I am such a bad husband of yours, as I have left you and made you float in the sea of sorrows. Born in the Ik\d{s}\=aku family, I inherited the kingdom and its pleasures; but, Alas! Your getting such a husband has now been reduced to slavery! O Dev\={\i}! I am merged in this ocean of sorrows and troubles. Who will rescue me, by narrating this story of the Pur\=a\d{n}as!''

36-40. S\=uta said :-- O King! The Br\=ahmi\d{n}, then, began to take away the queen and the boy, whipping them, in the face of the King. Seeing his wife and son being dragged away in that state, the King's pain knew no bounds and be frequently sighed and sighed and bitterly wept aloud. Alas! My dear wife, whom the Moon, the Sun, Wind or any other body could not see ere this, has become now reduced to slavery today! Oh! How beautiful and gentle are the fingers of my child? He has been sold off today, being born in the Solar Dynasty? Alas! Fie on my foolish understanding! Oh my Dear! Oh my child Rohit\=a\'sva! Your this wretched condition is due to my An\=arya irrespectable bad maxims! Oh! Through the mockery of the Daiva, I have got this distress! Fie on Me!

41-42 Vy\=asa said :-- The King was lamenting thus when the Br\=ahmi\d{n} disappeared with them, in the very tall trees and walls of palatial buildings. At this time the cruel fiendish Muni, endowed with great power of asceticism came there quickly, accompanied by his disciples.

43. Vi\'sv\=amitra said :-- ``O One of mighty arm! If you think it your duty to respect Truth, then pay me the Dak\d{s}i\d{n}\=a of R\=ajas\=uya sacrifice that you promised before.''

44. Hari\'schandra said :-- ``O R\=ajar\d{s}i! I bow down to Thee. O Sinless One! Now take the Dak\d{s}i\d{n}\=a of the R\=ajas\=uya Sacrifice that I promised to pay you before.''

45. Vi\'sv\=amitra said :-- ``O King! Whence have you collected these gold Mohurs that you are now paying me as my Dak\d{s}i\d{n}\=a. How have you earned this? Say.''

46. Hari\'schandra said :-- ``O Dv\={\i}ja! O Sinless One! What use is there in telling this to you. It will increase agony by hearing. O One of good vows!''

47. Vi\'sv\=amitra said :-- ``I won't accept money earned not rightly.

Give what you have acquired by rightful means. Say truly how you have acquired it.''

48. Hari\'schandra spoke :-- ``O Br\=ahmi\d{n}! I have sold my wife the Dev\={\i} Madhav\={\i} for one Koti Gold Mohurs and my son for ten Kotis of gold Mohurs. So take this eleven Koti Gold Mohurs from me.''

49. S\=uta said :-- Seeing the gold collected out of the sale of wife and son very small, and seeing the King overpowered with pain and sorrow, Kau\'sika angrily spoke :--

50-52. O King! The Dak\d{s}i\d{n}\=a of the R\=ajas\=uya Sacrifice cannot be so small; so collect quickly other money to complete it. O Vilest of K\d{s}attriyas! If you think this much to be proper for me, see first the enormous power of mine that I possess of my tapasy\=a, practised duly, of my pure Br\=ahma\d{n}yahood, of my violent power and of my chaste study and then you can pay my Dak\d{s}i\d{n}\=a.

53. Hari\'schandra said :-- ``O Bhagav\=an! I have sold just now my wife; and so wait for some time and I will collect more gold and will pay that to you.''

54. Vi\'sv\=amitra said :-- ``O King! The fourth part of the day is now remaining; I will wait till then. After this you won't expect any other reply from me.''

Here ends the Twenty second Chapter of the Seventh Book on the selling of Hari\'schandra's wife in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



