\chapter{On the birth of the Bhagavat\={\i} in the house of Dak\d{s}a}

1-19. Vy\=asa said :-- O King! Thus I have described the glory of the Dev\={\i}. Now I will narrate, as far as I can, the excellent lives of the kings of the the Solar and the Lunar dynasties respectively. They

all attained their excellent glories, simply because they were favoured by the Grace of the Highest \'Sakti; they were all the great devotees of the Supreme Deity. All their prowess, bravery, prosperity and all their glory, know that those all were derived from the mere parts of the Par\=a \'Sakti. O King! Those Kings and others as well were able to out off the Tree of this World by the Axe of their Knowledge, simply because they were the devotees of the Par\=a \'Sakti. So with all the care possible, the Lady of the Universe is to be worshipped and served. Men should avoid worshipping any other gods, as people avoid the husk to get the grain inside. O King! By churning the ocean of the Vedas, I have got the jewel as the lotus-feet of the Par\=a \'Sakti; and I think that I have discharged all my duties and think myself satisfied and successful. Brahm\=a, Vi\d{s}\d{n}u Rudra, and I\'svara are the four feet and Sad\=a \'Siva is the plank overhead; thus these five form the seat on which the Dev\={\i} is seated. There is no other deity superior to Her. To show this (to the ordinary ignorant people) the Mah\=a Dev\={\i} has taken this seat composed of the five Brahm\=a, Vi\d{s}\d{n}u, Rudra, I\'svara and Sad\=a \'Siva. Superior to these five, what is stated in the Vedas as Vyaktam and in which all this Universe is sewn, as it were, crosswise and lengthwise, lying in and through, that is Bhuvane\'svar\={\i}, the Goddess of the Universe.

[Note: Brahm\=a, Vi\d{s}\d{n}u, Rudra, I\'svara and Sad\=a \'Siva are the Regents or the presiding Deities of earth, water, fire, air and \=Ak\=a\'sa]. No man can be free unless he until the Goddess. When men will be able to encircle the \=Ak\=a\'sa, of the fourth dimension, as if it were an antelope skin, then they will be able to root out the miseries of the world, without knowing the nature of the Dev\={\i} (i.e., impossible). Thus the \'Svet\=a\'svataropani\d{s}ada says :-- ``Those that were engaged in meditation, Dhy\=ana Yoga, they saw the Dev\={\i} covered by the Gu\d{n}as S\=attva, R\=ajas and T\=amas and the forces incarnate respectively of the several Devas.'' So to make the human birth a success, first avoid all companies, be it out of shame, or fear, or devotion, or out of love; then bring the mind and keep it steady in your heart and then be devoted to Her and consider Her as the Supreme. This is the Ved\=anta Dindima (the declaration of the Ved\=anta). Whoever takes the name of the Dev\={\i}, either in sleeping, going or resting or in any other condition, he is certainly freed from the bondage of the world, no doubt. O King! So worship the M\=ahe\'svar\={\i} with all the care that you can. Go on step by step; first worship Her Vir\=at R\=upa (cosmic form); then S\=uk\d{s}ma R\=upa (subtle form) and then her Antary\=am\={\i} R\=upa (inner form, ruling within). Thus when your heart is purified, worship the Par\=a \'Sakti, of the nature of Brahm\=a, beyond this M\=ay\=a, this Prapa\d{n}cha Ull\=asa, of the nature of Existence,

Intelligence and Bliss. When the Chitta (heart) melts in Par\=a \'Sakti, then comes the real \=Ar\=adhan\=a (the real worship). So dilute your heart in Her. O King! Thus I have described to you the sanctifying deeds of the extremely devoted kings of the Par\=a \'Sakti, who were noble minded and religious. One who will hear this will acquire fame, dharma, intelligence, good end, and merits that have no equal. Now what else do you like to hear?

20-22. Janamejaya said :-- ``O Bhagav\=an! In olden times, the World mother Par\=a \'Sakti handed over Gaur\={\i} to Hara, Lak\d{s}m\={\i} to Hari, and Sarasvat\={\i} to Brahm\=a, born of the lotus from the navel of Hari. Now I hear that Gaur\={\i} is the daughter of Him\=alay\=a as well of Dak\d{s}a; and Mah\=a Lak\d{s}m\={\i} is the daughter of the K\d{s}iroda ocean (ocean of milk). They were all originated from the Prime Dev\={\i}; how, then, Gaur\={\i} and Lak\d{s}m\={\i} came to be the daughters of others? O great Muni! This is next to impossible; so my doubt arises. O Bhagav\=an! You are quite competent to cut off all my doubts; so by your axe of knowledge, cut off my present doubt.''

23-44. Veda Vy\=asa said :-- O King! Hear. I am telling you this wonderful secret. You are greatly devoted to the Dev\={\i}; so nothing there can be that I cannot disclose to you. Since the time the Great Mother gave over to Hara, Hari and Brahm\=a, Gaur\={\i}, Lak\d{s}m\={\i} and and Sarasvat\={\i}, respectively, these three Devas, Hara, etc. were performing their tasks, preserving, etc. O King! Once on a time, certain D\=anavas, named Hal\=ahalas were born. In time, they became very powerful and in a short time conquered the three worlds. What more than this, that they being elated with the boon granted to them by Brahm\=a, took their forces and invested the Mount Kail\=a\'sa and the Vaikuntha regions!

Seeing this, Mah\=a Deva and Vi\d{s}\d{n}u both made preparations for war. A terrible fight ensued between both the parties. For sixty thousand years the battle lasted incessantly but the result was stalemate. Gradually there was a great cry of consternation in the two parties. When \'Siva and Vi\d{s}\d{n}u with great effort destroyed the D\=anavas. O King! \'Siva and Vi\d{s}\d{n}u then returned to their own houses and began to brag of their powers before their own \'Saktis Gaur\={\i} and Lak\d{s}m\={\i}; whereas the Demons were killed on account of the \'Saktis of Gaur\={\i} and Lak\d{s}m\={\i}. Seeing them boast, Gaur\={\i} and Lak\d{s}m\={\i} laughed not sincerely whereon the two gods were very much angry. They under the magic spell of the Prime M\=ay\=a insulted them and even used offensive languages. Gaur\={\i} and Lak\d{s}m\={\i} quitted them and disappeared. A great uproar then arose in the worlds.

Both Hari and Hara became lustreless due to their insulting the two \'Saktis. They become powerless and unconscious and turned out mad. Seeing this Brahm\=a became very anxious. Hari and Hara are the two chief Deities; how then these two have become unable to perform the actions of the world! What is the cause? Why this calamity has sprung up out of season? Will there be a Pralaya (a general dissolution) of the world out of some offence, when no actions are being done! I know nothing about this. So how can I find a remedy! Being thus very distressed, he began to meditate with his eyes closed in the fourth dimensional space in the heart. O King! The Lotus born Brahm\=a then found out by his meditation that this calamity was brought about by the great wrath of the Par\=a \'Sakti. He then tried to find out the remedy until Hari and Hara did not regain the former natural position. Brahm\=a began by his own \'Sakti to carry on the functions of them both, viz., that of preservation and destruction for some time. The religious-minded Praj\=apati quickly called his son Manu and Sanaka, etc., the \d{R}i\d{s}is, for bringing peace on the two great Gods! When they came to him, the great ascetic four-faced Brahm\=a told them :-- ``I am now busy with many more works; so I am unable carry on my tapasy\=a? By the wrath of the Highest Force, Hari and Hara have become somewhat distracted; so for the satisfaction of the Par\=a \'Sakti I am performing the three functions, i.e., those of Creation, Preservation and Destruction. So you both practise this hard tapasy\=a with the greatest devotion and bring about Her satisfaction. O sons! Do such as Hari and Hara gain their former states and then be united with their own \'Saktis respectively. Your fame will increase thereby, no doubt. Rather that family where the two \'Saktis will take their birth, will purify the whole world and that man himself will be crowned with success.''

45. Vy\=asa said :-- O King! The pure-hearted Dak\d{s}a and other mind-born sons of Brahm\=a, hearing the words of the Grandsire expressed their desire to worship the Par\=a \'Sakti and went to the forest.

Here ends the Twenty-ninth Chapter of the Seventh Book on birth of the Bhagavat\={\i} in the house of Dak\d{s}a in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



