\chapter{1-2. The Him\=alay\=as said :-- "O Dev\={\i}! Describe the places on this earth that are prominent, sacred, and worth visiting and which Thou likest best. O Mother! Also sanctify us by describing the vows and utsabs that are pleasing to Thee, and by performing which, men become blessed and get themselves satisfied.}
3-10. The Dev\={\i} spoke :-- "O Himavan! All the places that are on this earth are all Mine and all should be visited. And every moment is fit for taking vows and utsabs. For I am of the nature pervading every moment; so whatever actions are performed at any moment are all equal to taking My vows and utsabs. O King of Mountains! Still I am now telling something out of My affection to My Bhakta\d{s}. Hear. There is a great place of pilgrimage named Kolh\=apura in the southern country. Here the Dev\={\i} Lak\d{s}mi always dwells. The second place is M\=atripura in the Sahy\=adr\={\i} mountain; here the Dev\={\i} Re\d{n}uk\=a dwells. The third place is Tulaj\=apur; next is the place Sapta\'sringa, the great places of Hingul\=a and Jv\=al\=a Mukh\={\i}. Then the great places of S\=akambhar\={\i}, Bhr\=am\=ari, \'Sr\={\i}raktadantik\=a and D\=urg\=a. The best of all places is that of Vindhy\=achala V\=asin\={\i}, the great places of Annapurn\=a and the excellent K\=anchipur (Conjiverum). Next come the places of Bh\={\i}m\=a Dev\={\i}, Vimal\=a Dev\={\i}, \'Sr\={\i} Cha\d{n}dral\=a Dev\={\i} of Kar\d{n}\=at, and the place of Kau\'sik\={\i}. Then the great place of N\={\i}l\=amb\=a on the top of the N\={\i}l\=aparvata, the place of J\=amb\=unade\'svar\={\i}, and the beautiful \'Sr\={\i}nagara.
11-20. The great place of \'Sr\={\i} Guhya K\=al\={\i}, well established in Nepal, and that of \'Sr\={\i} M\={\i}n\=ak\d{s}\={\i} Dev\={\i} established in Ch\={\i}damvaram. The great

place named Ved\=ara\d{n}ya where the Sundar\={\i} Dev\={\i} is residing; then the place named Ek\=amvaram, and the place Bhuvane\'svara near Puru\d{s}ottama where I always dwell as Par\=a \'Sakti Bhuva\d{n}e\'svar\={\i}. The famous place of Mah\=alas\=a, known in the south by the name Mall\=ari; the place of Yoge\'svar\={\i} Var\=at, and the widely known place of N\={\i}la \'Sarasvat\={\i} in China. The excellent place of Bagal\=a in Baidyan\=ath, the supreme place Ma\d{n}idv\={\i}pa of \'Sr\={\i}mat\={\i} Bhuvane\'svar\={\i} where I always reside. The Yonimandala K\=am\=akhy\=a, the place of \'Srimat\={\i} Tripur\=a Bhariav\={\i}, the excellent of all the places in this earth, where the Dev\={\i} Mah\=a M\=ay\=a always dwells. There is no other place better than this on the earth. Here the Dev\={\i} becomes every month in Her course of menstruation and where the virtuous men are seen. Here all the Devas remain in the form of mountains and where on the mountains the excellent Devas inhabit. The sages say: That all the places there are of the nature of the Dev\={\i}; there is no better place than this K\=am\=akhy\=a Yonimandala. Pu\d{s}kara, the sacred place, is the seat of G\=ayatr\={\i}; the place of Chandik\=a in Amare\'sa: and the excellent place of Pu\d{s}karek\d{s}i\d{n}\={\i} in Frabh\=asa. The place of Linga-dh\=ari\d{n}\={\i} Dev\={\i} in Naimi\d{s}\=aranya, and the place of Purubut\=a in Pu\d{s}kar\=ak\d{s}a; Rati dwells in \=As\=adh\={\i}.
21-30. Dandin\={\i} Parame\'svari dwells in Chandamund\={\i}. Bh\=uti dwells in Bh\=arabh\=uti; and Nakule \'Svar\={\i} dwells in N\=akula. Chandrik\=a dwells in Hari\'schandra; \'S\=ankari in \'Sr\={\i}giri; Tri\'s\=ul\=a in Jape\'svara; and Suk\d{s}m\=a in \=Amr\=ata Ke\'svara. \'S\=ankar\={\i} dwells in Ujjain, \'Sarv\=a\d{n}\={\i} in the place Madhyam\=a, and M\=arga D\=ayini dwells in the holy K\d{s}etra Ked\=ara. The celebrated Bhairav\={\i} dwells in the place named Bhairava; Mangal\=a in Gay\=a K\d{s}ettra; Sth\=anupriy\=a in Kuruk\d{s}etra; and Sv\=ayambhuv\={\i} Dev\={\i} dwells in N\=akula; Ugr\=a dwells in Kankhal; Vi\'sves\=a dwells in Vimale\'svara, Mah\=anand\=a in Attah\=asa and Mah\=antak\=a in Mahendra. Bhime\'svar\={\i} dwells in Bh\={\i}ma; the Bhav\=an\={\i} \'Sankar\={\i} dwells in Vastr\=apadma; and Rudr\=a\d{n}\={\i} in Ardha Kot\={\i}. Vi\'sal\=ak\d{s}\={\i} dwells in Avimukta; Mah\=abh\=ag\=a dwells in Mah\=alaya; Bhadrakar\d{n}\={\i} in Gokar\d{n}a; and Bhadr\=a resides in Bhadrakar\d{n}ak; Utpal\=ak\d{s}\={\i} dwells in Suvarn\=ak\d{s}a; Sth\=a\d{n}v\={\i}\'s\=a in Sth\=a\d{n}u; Kamal\=a in Kamal\=alay\=a; Chand\=a in Chhagalandaka, situated in the south near the sea coast. Trisandhy\=a dwells in Kurundala; Mukute\'svar\={\i} in M\=akota; \'S\=andak\={\i} in Mandale\'sa; K\=al\={\i} in K\=alanjara; Dhvani in \'Sankukar\d{n}a; Sth\=ul\=a in Sth\=ulake\'svara; and Parame\'svar\={\i} Hrillekh\=a dwells in the heart lotuses of the J\~nanins.
31-34. The places mentioned above are all dearest to the Dev\={\i}. First the merits of these places are to be heard; next the Dev\={\i} is to be

worshipped by the rites and ceremonies according to these rules. Or, O Mountain! All the holy places of pilgrimages exist in K\=a\'s\={\i}. The Dev\={\i} always dwells there. Persons, devoted to the Dev\={\i}, see these places and if they make Japam and meditate on the lotus-feet of the Dev\={\i}, they will certainly be freed from the bonds of Sams\=ara; there is no doubt in this. If anybody, getting up in the morning, recite the names of these places, all his sins would instantly be burnt away.
35-40. And if one reads, in the time of \'Sr\=addha, before the Brahmi\d{n}s, these holy names of the Dev\={\i}, his Pitris will be purified of their sins in the Mah\=ak\=a\'sa by the Mah\=a Pr\=a\d{n}a and will get their highest goal. O One of good vows! I will now describe to you the vows that are to be carefully observed by men and women; hear. Ananta Trit\={\i}y\=akhya Vrata (vow), Rasakaly\=an\={\i} Vrata, and \=Ardr\=anandakara Vrata, these three Vratas are to be observed in the Trit\={\i}y\=a (third) tithi. The next come the Friday vow, the Kri\d{s}\d{n}a Chaturda\'si vows, the Tuesday vow, and the evening twilight vow. In this twilight vow, Mah\=a Deva placed the Dev\={\i} in the evening on an \=Asana; and He, along with the other Devas, began to dance before Her. Fasting is enjoined in this vow; and then in the evening one must worship the Dev\={\i}, the Giver of all auspicious things. Especially in every fortnight, if the Dev\={\i} be worshipped, She gets extremely pleased.
41. O Best of Mountains! The Monday vow is very agreeable to Me; the worship of the Dev\={\i} should be done and then in the night one must take one's food.
42-43. The two nine nights vow called Navar\=atra are to be observed, one in the autumn and the other in the spring season. These are very dear to Me. He is certainly My devotee and very dear who for My satisfaction performs these and the other Nitya Naimittik vows, free from any pride and jealousy. He certainly gets the S\=ajujya Mukti with Me.
44-46. O Nagar\=aja! The Holy (Dol) festival in the month of Chait on tha third day of the white fortnight is very pleasing to Me and should be observed by all. My devotees perform the \'Sayanotsava in the Paur\d{n}\=am\=as\={\i} in the month of \=As\=adha; the J\=agara\d{n}otsava in the Paur\d{n}am\=as\={\i} in the month of K\=artik, the Ratha J\=atr\=a in the 3rd of the white fortnight in \=As\=adha; the Damanotsava in Chaitra. And my dear festivals in the month of \'Sr\=avana and various other festivals.

47-49. In all these festivals one should feast well with gladness all My devotees, and the Kum\=aris (virgins), well clothed and dressed, and the boys, thinking them all to be of My very nature. No miserliness is to be entertained and I should be worshipped with flowers, etc. He is blessed and attains his goal and is dear to Me who carefully and devotedly observes every year all these festivals. O Nagendra! Thus I have described to you in brief all the vows that are pleasing to Me. These instructions are not to be given who is not a disciple nor to one who is not My devotee.
Here ends the Thirty-eighth Chapter of the Seventh Book on the vows and the sacred places of the Dev\={\i} in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



