\chapter{1-19. The Dev\={\i} said :-- "O Girir\=aja! This whole universe, moving and unmoving, is created by My M\=ay\=a \'Sakt\={\i}. This M\=ay\=a is conceived in Me. It is not, in reality, different or separate from Me. So I am the only Chit, Intelligence. There is no other intelligence than Me. Viewed practically, it is known variously as M\=ay\=a, Vidy\=a; but viewed really from the point of Brahma\d{n}, there is no such thing as M\=ay\=a; only one Brahman exists, I am that Brahma, of the nature of Intelligence.}

I create this whole world on this Unchangeable Eternal (Mountain-like) Brahma, (composed of Avidy\=a, Karma, and various Samsk\=aras) and enter first as Pr\=a\d{n}a (vital breath) within it in the form of Chid\=abh\=a\d{s}a. O Mountain! Unless I enter as Breath, how can this birth and death and leaving and retaking bodies after bodies be accounted for! As one great \=Ak\=a\'s\=a is denominated variously Ghat\=ak\=a\'sa (\=Ak\=a\'s\=a in the air), Pat\=ak\=a\'sa (\=Ak\=a\'s\=a in cloth or picture), so I too appear variously by acknowledging this Pr\=a\d{n}a in various places due to Avidy\=a and various Antahkara\d{n}as. As the Sun's rays are never defiled when they illumine various objects on earth, so I, too, am not defiled in entering thus into various high and low Antahkara\d{n}as (hearts). The ignorant people attach Buddhi and other things of activity on Me and say that \=Atman is the Doer; the intelligent people do not say that. I remain as the Witness in the hearts of all men, not as the Doer. O Achalendra! There are many J\={\i}vas and many \=I\'svaras due to the varieties in Avidy\=a and Vidy\=a. Really it is M\=ay\=a that differentiates into men, beasts and various other J\={\i}vas; and it is M\=ay\=a that differentiates into Brahma, Vi\d{s}\d{n}u and other \=I\'svaras. As the one pervading sky (\=Ak\=a\'sa) is called Mah\=ak\=a\'sa Ghat\=aka\'sa (being enclosed by jars), so the One All pervading Param\=atm\=a is called Param\=atm\=a, J\={\i}v\=atm\=a (being enclosed within J\={\i}vas). As the J\={\i}vas are conceived many by M\=ay\=a, not in reality; so \=I\'svaras also are conceived many by M\=ay\=a; not in essence. O Mountain! This Avidy\=a and nothing else, is the cause of the difference in J\={\i}vas, by creating differences in their bodies, indriyas (organs) and minds. Again, due to the varieties in the three Gu\d{n}as and their wants (due to the differences between S\=attvik, R\=ajasik and T\=amasik desires), M\=ay\=a also appears various. And their differences are the causes of different \=I\'svaras, Brahma, Vi\d{s}\d{n}u and others. O Mountain! This whole world is interwoven in Me; It is I that am the \=I\'svara that resides in causal bodies; I am the "Sutr\=atman, Hira\d{n}yagarbha that resides in subtle bodies and it is I that am the Vir\=at, residing in the gross bodies. I am Brahm\=a, Vi\d{s}\d{n}u, and Mahe\'svara; I am the Br\=ahm\=a, Vai\d{s}\d{n}avi and Raudr\={\i} \'Saktis. I am the Sun, I am the Moon, I am the Stars; I am beast, birds, Chand\=alas and I am the Thief, I am the cruel hunter; I am the virtuous high-souled persons and I am the female, male, and hermaphrodite. There is no doubt in this. O Mountain! Wherever there is anything, seen or heard, I alway exist there, within and without, There is nothing moving or unmoving, that can exist without Me. If there be such, that is like the son of a barren woman. Just as one rope is mistaken for a snake or a garland, so I am the One Brahma and appears as \=I\'svara, etc. There is no doubt in this. This world cannot appear without a substratum.

And That Substratum is My Existence. There can be nothing else.
20. The Him\=alay\=as said :-- "O Dev\={\i}! If Thou art merciful on me, I desire, then, to see Thy Vir\=at form in the Fourth Dimensional Space.
This sight is developed when the mind resides in the heart centre or in the centre of the eye-brows. A proper teacher is necessary.
21-41. Vy\=asa said :-- "O King! Hearing the words of Girir\=aja, Vi\d{s}\d{n}u and all the other Devas gladly seconded him. Then the Dev\={\i}, the Goddess of the Universe, knowing the desires of the Devas, showed Her Own Form that fulfils the desires of the Bhaktas, that is auspicious and that is like the Kalpa Vrik\d{s}a towards the Bhaktas. They saw Her Highest Vir\=at Form. The Satyaloka is situated on the topmost part and is Her head; the Sun and Moon are Her eyes; the quarters, Her ears; the Vedas are Her words; the Universe is Her heart; the earth is Her loins; the Bhuvarloka is Her navel; the asterisms are Her Thighs; the Maharloka is Her neck; the Janarloka is Her Face; the Taparloka is Her head, situated below the \'Satyaloka; Indra and the Devas and the Svarloka is Her arms; the sound is the organ of Her ears; the A\'svin twins, Her nose; the smell is the organ of smell; the fire is within Her face; day and night are like Her two wings. The four-faced Brahm\=a is Her eyebrows; water is Her palate; the juice thereof is Her organ of taste; Yama, the God of Death, is Her large teeth; the affection is Her small teeth; M\=ay\=a is Her smile; the creation of Universe is Her sidelooks; modesty is Her upper lip; covetousness is Her lower lip; unrighteousness is Her back. The Praj\=apati is Her organ of generation; the oceans are Her bowels; the mountains are Her bones; the rivers are Her veins; and the trees are the hairs of Her body. O King! Youth, virginity, and old age are Her best gaits, positions or ways (courses) paths, the clouds are Her handsome hairs; the two twilights are Her clothings; the Moon is the mind of the Mother of the Universe; Hari is Her Vijn\=ana \'S\=akti (the knowledge power); and Rudra is Her all-destroying power. The horses and other animals are Her loins; the lower regions Atala, etc., are Her lower regions from Her hip to Her feet. The Devas began to behold Her this Cosmic (Vir\=ata) appearance with eyes, wide awake, with wonder. Thousands of fiery rays emitted from Her form; She began to lick the whole universe with Her lips; the two rows of teeth began to make horrible sounds; fires came out from Her eyes; various weapons were seen in Her hands; and the Br\=ahma\d{n}as and K\d{s}attriyas are become the food of that Awful Deity. Thousands of heads, eyes and feet were seen in that form. Crores of Suns, crores of lightnings flashes, mingled there. Horrible, Awful, That appearance looked terrific to the eyes, heart and mind. The Devas thus beheld and began to utter cries of horror and consternation; their hearts trembled and they were

caught with immoveable senselessness. "Here is the Dev\={\i}, our Mother and Preserver." This idea vanished away at once from their minds.
At this moment the Vedas that were on the four sides of the Dev\={\i}, removed the swoon of the Devas and made them conscious. The Immortals got, then, the excellent Vedas; and, having patience, began to praise and chant hymns in words choked with feelings and with tears flowing flowing from their eyes.
42-53. The Devas said :-- "O Mother! Forgive our faults. Protect us, the miserable, that are born of Thee. O Protectress of the Devas! Withhold Thy anger; we are very much terrified at the sight of Thy this form. "O Dev\={\i}! We are inferior immortals; what prayers can we offer to Thee! Thou Thyself canst not measure Thy powers; how then can we, who are born later, know of Thy greatness! Obeisance to Thee, the Lady of the Universe! Obeisance to Thee of the nature of the Pra\d{n}ava Om; Thou art the One that is proved in all the Ved\=antas. Obeisance to Thee, of the form of Hr\={\i}m! Obeisance to Thee, the Self of all, whence has originated the Fire, the Sun, and the Moon and whence have sprung all the medicinal plants. Obeisance to the Dev\={\i}, the Cosmic Deity, the Self in all whence have sprung all the Devas, S\=adhyas, the beasts, birds, and men! We bow down again and again to the Great Form, M\=ah\=a M\=ay\=a, the Self of all, whence have sprung the vital breath Pra\d{n}a, Ap\=ana, grains and wheats, and Who is the source of asceticism, faith, truth, continence and the rules what to do and what not to do under the present circumstances. The seven Pr\=a\d{n}as, the seven Lokas, the seven Flames, the seven Samidhs, the seven Oblations to Fire, have sprung from Thee! Obeisance to Thee, the Great Self in all! Obeisance to the Universal form of the Deity of the Universe whence have sprung all the oceans, all the mountains, all the rivers, all the medicinal plants and all the Rasas (the tastes of all things). We bow down to that Vir\=at Form, the Great Self, the Mah\=a M\=ay\=a, whence have originated the sacrifices, the sacrifical post (to which the victim about to be immolated is bound) and Dak\d{s}i\d{n}\=as (the sacrificial fees) and the Rik, the Yajus, and the S\=ama Vedas. O Mother! O Mah\=a M\=ay\=a! We bow down to Thy front, to Thy back, to Thy both the sides, to Thy top, to Thy bottom and on all sides of Thee. O Dev\={\i}! Be kind enough to withhold this Extraordinary Terrific Form of Thine, and shew us Thy Beautiful Lovely Form.
54-56. Vy\=asa said :-- "O King! The World Mother, the Ocean of mercy, seeing the Devas terrified, withheld Her Fearful Cosmic Form and showed Her very beautiful appearance, pleasing to the whole world. Her body became soft and gentle. In one hand She held the noose, and in another She held the goad. The two other hands made signs to dispel all their fears

and ready to grant the boons. Her eyes emitted rays of kindness; Her face was adorned with beautiful smiles. The Devas became glad at this and bowed down to Her in a peaceful mind and then spoke with great joy.
Here ends the Thirty-third Chapter of the Seventh Book on the Dev\={\i}'s Vir\=at R\=upa in the Mah\=a Par\=a\d{n}am, Sri Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



