\chapter{1-5. The Dev\={\i} said:--Getting up from the bed early in the morning, one is to meditate on the thousand petalled lotus, bright, of the colour of camphor, in the top part his brain on the head. On this he should remember his \'Sr\={\i} Guru, very gracious looking, well decorated with ornaments, with His Consort \'Sakti and bow down to Him and within Him he should meditate the Kundalin\={\i} Dev\={\i} thus --"I take refuge unto that Highest \'Sakti Kundalin\={\i}, of the nature of the Supreme Consciousness, Who is manifest as Chaityana while up-going to the Brahmarandhra (the aperture supposed to be at the crown of the head, through which the soul takes its flight at death) and Who is of the nature of nectar while returning back in the Su\d{s}umn\=a canal. After meditating thus, he should meditate on the Blissful Form of Mine within the Kundalin\={\i} Fire situated in the Mul\=adh\=ara Lotus (coccygeal lotus). Then he should rise up to go for the calls of nature, etc., and complete Sandhy\=a Bandanams and other duties. The best of the Br\=ahmi\d{n}s, then, should for My satisfaction perform the Agnihotra Homa and sitting in his \=Asana make Sankalap (determination) to do My P\=uj\=a (Worship).}
N. B.--The brain has three divisions, the lower, the middle and the higher, or top-most part which is very pure.
6-10. Next he is to make Bh\=uta \'Suddhi (purification of elements of the body by respiratory attraction and replacement) and then the M\=atrik\=a Ny\=asa, Then he should arrange the letters of the root Mantra of M\=ay\=a and execute the Hr\={\i}llekh\=a M\=atrik\=a Ny\=asa. In this he is to place the letter "Ha" in the Mul\=adh\=ara, the letter "Ra" in his heart and the vowel "\={\i}" in the middle of his brows aud, Hr\={\i}m on the top part of his head. Finishing

then all the other Ny\=asas according to that Mantra, he should think within his body Dharma, J\~n\=ana, Vair\=agyam, and Prosperity as the four legs of the seat and Adharma, Aj\~n\=ana, Avair\=agyam and non-prosperity, these four as the body of the seat on the four quarters East, South, West and North. Then he should meditate on the Great Dev\={\i} in the lotus of his heart blewn by Pr\=a\d{n}\=ay\=ama, situated on the five seats of the Pretas. O Mountain! Brahm\=a, Vi\d{s}\d{n}u, Rudra, Sad\=a\'siva and \=I\'svara are the five Pretas situated under My feet.
11. These are of the nature of earth, water, fire, air, and ether, the five elements and also of the nature of J\=agrat (waking), Svapna (dreaming) Su\d{s}upti (deep sleep state) Turiy\=a (the fourth state) and At\={\i}ta R\=upa, the (the fifth state) excluding the 4 states, corresponding to the five states. But I, who am of the nature of Brahma, am over and above the five elements and the five states; therefore My Seat is always on the top of these five forces.
12. Meditating on Me thus and worshipping Me with his mind concentrated, he is next to make Japam (reciting My name slowly). After Japam he is to make over the fruits of Japam to Me. He should then place the Arghya for the external worship.
13. Then the worshipper is to sprinkle with the Astra mantra 'Phat', all the articles of worship that are placed in front of him and purify them.
14-18. He should close the ten quarters with the Chhotik\=a Mudr\=a and bow down to his Guru. Taking his permission, he should meditate on the outside seat, the beauiful divine form of his heart lotus and invoke the Deity outside and place Her on the seat by Pr\=a\d{n}a Pratisth\=a and perforn \=Av\=ahana, and present to Her Arghya (an offer of green grass, rice, etc., made in worshipping a god), P\=adya (water for washing legs and feet), \=Achaman, water for bath, a couple of clothes, all sorts of ornaments, or scents, flowers and the necessary articles with due devotion and he should worship the attendant deities of the Yantra. If one be unable to worship daily the attendant deities, one must worship them on Friday.
19. Of the attendant deities, one must meditate the principal deity of the nature of Prabh\=a (illumination) and think that by Her rays the three worlds are pervaded.
20. Next he should worship again the Bhuvane\'svari Dev\={\i}, the Chief Deity along with other attendant deities with scent, good smelling flowers; and Naivedya and various other tasteful dishes.

21-22. He should then recite the Sahasran\=ama (thousand names) stotra and the Dev\={\i} S\=ukta Mantra "Aham Rudrebhih, etc.," and "Sarve vai Dev\=a Dev\={\i} mupatasthuh, etc.," the Dev\={\i} Atharva \'Siro Mantra and the Upani\d{s}ad\'s Mantra of Bhuvane\'svar\={\i}, the famous mantras, repeatedly and thus bring My satisfaction.
23-24. With hearts filled with love and with hairs standing on their ends all should satisfy Me frequently with tears of love flowing from their eyes and with voice choked with feelings and with dancing music and singing and with his whole body filled with joy.
25. My glory is well established in the Veda P\=ar\=ayana and in all the Pur\=a\d{n}as. So for My satisfaction, one should offer daily to Me one's everything with one's body and recite the readings from the Vedas.
26-27. Next, after completing the Homa offerings, he should feed the Br\=ahma\d{n}as, tha young virgins well clothed, the boys and the public and the poor, thinking all of them to be so many forms of the Dev\={\i}. Than he should bow before the Dev\={\i} that resides in his heart and finally by Samh\=ara Mudr\=a take leave of the Deity invoked.
28. O One of good vows! The Hr\={\i}llekh\=a Mantra (Hr\={\i}m) is the chief of all mantrams; so My worship and all other actions ought to be performed with this Hr\={\i}llekh\=a Mantram.
29. I am always reflected in this Mirror of Hr\={\i}llekh\=a form; so anything offered in this Hr\={\i}llekh\=a Mantra of Mine is offered as it were with all the Mantras. Then one should worship the Guru with ornaments, etc., and think oneself blessed.
30-31. O Himavan! Nothing remains at any time unavailable to him who worships thus the Bhuvane\'svar\={\i} Dev\={\i}. After quitting his body, he goes to theMa\d{n}i Dv\={\i}pa, My Place. He gets the form of the Dev\={\i}; and the Devas constantly bow down to him.
32-45. O Mah\={\i}dhara! Thus I have described to you the rules of worshipping the Great Dev\={\i}; consider this in all the aspects and worship Me according to your Adhik\=ara (claim) and you will attain your Goal. There is no doubt in this. O Best of mountains! This \'S\=astra Devig\={\i}t\=a you are not to tell to those who are not the devotees, to those who are enemies, and to those who are cunning. If one gives out this secret of G\={\i}t\=a, it is like taking off the covering from the breast of the mother; so carefully keep it secret and think that this is very necessary. This Dev\={\i} G\={\i}t\=a ought to be given to a disciple, a Bhakta, the eldest son, and

to one who is good natured, and well dressed and devoted to the Dev\={\i}. O Mountain! In the time of \'Sr\=addha (solemn obsequies performed in honour of the manes of deceased ancestors) he gets the highest place of the Pitris who reads this Dev\={\i} G\={\i}t\=a before the Br\=ahma\d{n}as. Vy\=asa said:-- The Dev\={\i} vanished there after describing all these. The Devas were glad and considered themselves blessed by the sight of the Dev\={\i}. O Janamejaya! The Haimavat\={\i} next took Her birth in the house of the Him\=alay\=a and was known by the name of Gaur\={\i}. \'S\=a\d{n}kara, the Deva of the Devas, marricd Her. Sad\=anana (K\=artika) was born of them. He killed the T\=adak\=a Asura. O King! In ancient times, when the ocean was churned, many gems were obtained. At that time the Devas chanted hymns to the Dev\={\i} with a concentrated mind to get Lak\d{s}m\={\i} Dev\={\i}. To show favour to the Devas, Ram\=a Dev\={\i} got out of the ocean. The Devas gave Laksmi to Vi\d{s}\d{n}u, the Lord of the Vaikuntha. Vi\d{s}\d{n}u was very glad at this. O King! Thus I have described to you the Greatness of the Dev\={\i} and the birth of Gaur\={\i} and Lak\d{s}m\={\i}. One's desires are all fulfilled when one hears this. O King! This secret I have described to you. Take care not to divulge it to any other body. This is the secret of the G\={\i}t\=a; so carefully conceal it. O One of pure heart! I have told to you this Divine and Sin-destroying narration, that you asked. What more do you want to hear? Say.
Here ends the Fortieth Chapter of the Seventh Book on the External Worship of the Dev\={\i} in the Mah\=apur\=a\d{n}am, \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.
[The. Seventh Book Completed.]



