\chapter{On the uplifting of the Earth by the Sacrificial Boar}

1-10. N\=ar\=aya\d{n}a said :-- O Child! When you have got your senses controlled and have overcome lust, anger, etc., and other enemies, then you are certainly entitled to hear this secret. While Brahm\=a was discussing on this subject with Mar\={\i}chi and the other Brahm\=ar\d{s}is and with Sv\=ayambhuva Manu, suddenly there came out of the nose of the meditative Brahm\=a, one young boar (the child of a boar) of the dimension of one finger only. That little boar, that was just in the air, soon became, while they were looking at it, enlarged to a very big elephant, in an instant; a very strange sight, indeed! Seeing this, the Kum\=aras Sanaka, etc., Mar\={\i}chi and the other Seven \d{R}i\d{s}is, and the Creator Brahm\=a were struck with wonder and Brahm\=a began to discuss thus :-- Oh! This young boar has come out suddenly of My nose; though it is a very small one, it is certainly, come under a disguise and it has all made us merge

in an ocean of wonder. Is this something Divine, of a S\=attvic quality under the guise of this boar? It was of the size of a thumb and it is now like the Him\=alay\=a Mountain? Oh! Is this the Bhagav\=an, the Yaj\~na Puru\d{s}a? While they were thus discussing, the Bhagav\=an, in the shape of the boar, began to make loud sounds, like the rolling of the mountain clouds at the time of the Pralaya resounding all the quarters. At this Brahm\=a and all the \d{R}i\d{s}is assembled there became very glad. Hearing the ghurghura noise of the Boar, their troubles ended and the people of Janarloka, Taparloka, Satyaloka and all the Devas being very much gladdened began to chant sweet hymns with Chha\d{n}das of Rik, Yajus, S\=ama and Atharva Vedas, to that \=Adi Puru\d{s}a, the Bhagav\=an, from all sides.

11-20. Hearing their praises, the Bhagav\=an Hari graciously looked on them and immediately merged Himself in the ocean. The ocean was very much agitated with the fierce striking of the hairs of the Boar, when He entered into the waters, and spoke thus :-- ``O Deva ! O Thou, the destroyer of the afflictions of your refuges! Protect me.'' Hearing the words of the ocean, the God Hari tore asunder all the aquatic animals and went down beneath the waters. Searching violently hither and thither, He knew the earth by her smell. At once the Bhagav\=an Hari, the Lord of all, went to Her and rescued the earth by raising Her up on his big teeth, the Earth that was the abode of all the beings. When the Bhagav\=an, the Lord of all the sacrifices, came up with the Earth on the top of His teeth, He looked beautiful like the elephant of a quarter taking out by the root and holding a thousand petalled lotus upon his tusks. When the Bhagav\=an rescued thus the Earth, raising Her up on His tusks, Brahm\=a and Indra, the Lord of the Devas, the powerful Manu began praise to Him with sweet words :-- Brahm\=a said :-- O Lotus-eyed! O Bhagav\=an! Thou art victorious everywhere; O Thou, the Destroyer of the sufferings of the Bhaktas! Thou hast, by Thy own prowess, humiliated the Heaven, the abode of the gods upto Satyaloka! Thou, the Giver all the desires! O Deva! This earth shines so beautiful on Thy teeth as the thousand petalled lotus, taken up by the root by a mad elephant, shines on his two tusks. O Bhagav\=an! This Form, i. e., that of Thy sacrificial Boar, with earth on Thee looks so very beautiful as the lotus looks beautiful on the tusks of an elephant.

21-22. O Lord! We bow down to Thee, the Creator and the Destroyer of everything; Thou assumest many forms for the purpose of destroying the D\=anavas; Thou dost do many acts that redound to Thy Glory, We bow down to Thee, to Thy Front and to Thy Back. (Really thou hast no front nor back. Thou art everywhere.) Thou art the Upholder of all the Celestials and the Immortals! Thy eyes shine equally everywhere.

23-25. O Deva! By Thy power I am brought up and engaged in the work of creation and by Thy order I create in every Kalpa this Universe and destroy it. O Lord of Immortals! In the ancient times the Devas united churned the vast ocean by Thy help and got according to their merits their due shares. O Hari! Indra, the Lord of the Devas is enjoying the vast kingdom of Heaven, the Trilokas, by Thy Will and appointment. All the Devas worship him.

26-27. So the God of Fire has got his burning power and is residing in the bellies of the Devas, the Asuras, men and all other beings and penetrating their bellies, is satisfying all. Yama, the King of Dharma, by Thy appointment, is the Lord of the southern quarters, is presiding over the Pitris and being the witness of all the actions of the Jiv\=as, is awarding duly to them the fruits thereof.

28-33. Nairrita, the Lord of the Rak\d{s}asas, the witness of all the actions of all the beings, by Thy commandment, is the destroyer of all the obstacles of the devotees that take refuge unto thee, though he is a Yak\d{s}a. The Varu\d{n}a Deva, by Thy order, has become duly the Lord of the waters, and the Regent of the Loka (Dikp\=ala). The V\=ayu, the Life of all, the carrier of the smell, has become by Thy order Lok\=ap\=ala and the Guru of the universe. Kuvera, obedient to Thy order, has become the Lord of the Yak\d{s}as and the Kinnaras and is holding a respectful position like other Lokap\=alas. \=I\'s\=ana, Who is the destroyer of all the J\={\i}vas, has got his Lordship over a quarter by Thy order and is being praised by all the Rudras, the Devas, Gandharbas, Yak\d{s}as, Kinnaras, the men and all the beings. O Bhagav\=an! We bow down to Thee, the Lord of the Universe; the innumerable Devas that are seen are merely the small fractions of Thy powers.

34-38. Naraya\d{n}a said :-- ``O Child N\=arada! When Brahm\=a, the Creator and the Grandsire of the Lokas, praised thus the \=Adi Puru\d{s}a Bhagav\=an, He cast a side long glance at them, offering His Grace. When the Bhagav\=an, the Sacrificial Boar, was coming up with earth, rescued and placed on his teeth, the awful Hira\d{n}yaksa, the chief of the Daityas came before Him and obstructed His passage when He killed him by one violent stroke of his club. He was besmeared all over his body with the blood of the Daitya; and thus He came up from the Ras\=atala and placed the earth on the waters. He then went away to His Vaikuntha abode.

O Child N\=arada! He who hears or reads devotedly this glorious deed of the Bhagav\=an and the deliverance of the Earth, will certainly be freed of all his sins and goes to the highest holiest place of Vi\d{s}\d{n}u, the Lord of all. There is no doubt in this.''

Here ends the Second Chapter of the Eighth Book on the uplifting of the Earth by the Sacrificial Boar in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



