\chapter{On the Ganges and the Var\d{s}as}

1-37. N\=ar\=aya\d{n}a said :-- Jathara and Devak\=uta are the two mountains situated on the east of Sumeru; their dimensions towards the north are eighteen thousand Yoyanas wide and two thousand Yoyanas high. On the west of Meru, are situated the two famous lofty mountains Pavam\=ana and P\=ariy\=atra; their length and height are well known. On the south of Meru are situated the two lofty mountains Kail\=asa and Karavira. On the north, again, of the Sumeru mountain, are situated the mountains \'Sri\d{n}gagiri and Makaragiri. Thus the golden Sumeru mountain, shines like the Sun, surrounded by these eight mountains. In the centre of the Sumeru, there is a Divine city built of the Creator Brahm\=a, ten thousand Yoyanas in dimensions. The learned sages that know everything from the highest to the lowest, describe that as square in dimensions and all made of gold. On the top of the Sumeru, are situated the world known eight golden cities, subservient to the Brahmapur\={\i}, for the eight Lokap\=alas. The Lords of the four directions, east, west, north and south and of the four corners, north-east, north-west, south-west, south-east, occupy these. The dimensions of each of these eight cities are two thousand Yoyanas and a half. In fact, there are nine cities there including the Brahmapur\={\i}. Now hear the names of these nine cities in due order. The First is Manovat\={\i}, the second is Amar\=avat\={\i}, the third is Tejovat\={\i}, next come in order Samyaman\={\i}, Kri\d{s}\d{n}\=angan\=a, \'Sraddh\=avat\={\i}, Gandhavat\={\i}, and Mahoday\=a, the ninth is Ya\'sovat\={\i}. The Lords of the Pur\={\i}s are Brahm\=a, Indra, Fire and the other Dikp\=alas in due order. When Vi\d{s}\d{n}u Bhagav\=an assumed the Tri Vikrama Form in His Dwarf Incarnation and went at the sacrifice, to get back the Kingdom of the Heavens from the demon Vali, there was created one hole or cavity over this Brahm\=anda Kat\=aha (skull) by the nails of the toes of that foot that went up towards his left; and through that cavity, the famous river Bhagavat\={\i} Gang\=a flowed on the top of those heavenly Kingdoms, the clear waters of which are ever ready to destroy the sins of the whole people. For this reason, She is known in the three Lokas as the Vi\d{s}\d{n}upad\={\i} manifested. This took place long long ago, many thousand Yugas ago, a period difficult to determine. The Ganges fell first on the top of the Indra's Heavens, near the place called Vi\d{s}\d{n}udh\=ama, known in the three worlds. Here the pure-souled Dhruva, the son of Utt\=anap\=ada, observed within his heart the lotus-feet

of Sr\={\i} Vi\d{s}\d{n}u Bhagav\=an and exists still there taking refuge on that immoveable position. There the high-souled Seven \d{R}i\d{s}is, knowing the highest purifying influence of the Ganges, circumambulate the river, wishing for the welfare of all the Lokas. This is the great place where one gets success, and final liberation, the success-giving place of the ascetics and where the Munis with clots of hairs on their heads daily take their dip in the Ganges with the greatest pleasure and admiration. The Ganges flows thence from the Dhruva Mandala, the abode of Vi\d{s}\d{n}u, in Kotis and Kotis of Divine Channels, interspersed with many Vim\=anas or carriages, deluges the Chandra Mandala (the Moon Sphere), comes gradually to the Brahm\=a Loka. Here She is divided into the four channels, S\={\i}t\=a, Alakanand\=a, Bhadr\=a, and Chaturbhadr\=a and irrigating many countries, mountains, and forests ultimately falls in the oceans. The S\={\i}t\=a, the famous Dh\=ar\=a, purifying all, while falling down from the Brahm\=aloka, passes round the mountains that form as it were the filaments of the flower-like Sumeru mountain, and falls on the top of the Gandham\=adana range. Thence She, worshipped by the Devas, irrigates the Bhadra\'svavar\d{s}a and falls eastward into the salt ocean. The second Dh\=ar\=a (stream), named Chak\d{s}u, getting out of the Malyav\=an range, gradually gains strength and assumes great force and flows by Ketum\=alavar\d{s}a to the western ocean. The third Dh\=ar\=a (stream), the very pure Alakananda, getting out of the Brahm\=a Loka, passes through the Girik\=uta mountain and other forests, falls to Hemak\=uta; next She flows through the Bh\=arata Var\d{s}a and meets with the southern ocean. No words can describe the glory and the purifying effect of this river; suffice it to say that those who march out to bathe in this river get at each of their steps, the fruits of doing the great sacrifices, R\=ajas\=uya and A\'svamedha, etc. The fourth Dh\=ar\=a of the Trailokyapavan\={\i} Ganga Devi, named Bhadr\=a, falling from the mountain \'Sri\d{n}gav\=an, become swift and capacious, flows by the Northern Kuru country and satisfying the people there flows down into the ocean. Many other rivers, getting out from the mountains, Meru, Mandara and others, flow through the various Var\d{s}as or continents, but of all the Var\d{s}as, continents, this Bh\=arata Var\d{s}a is called the Karma K\d{s}ettra (or the field of actions). The other eight Var\d{s}as, though on earth, give the pleasures of the Heavens. The reason for this is that when the period of the person\'s enjoyments in the Heavens cease, they come and take their birth in the one or other of these eight Var\d{s}as. The people here live for ten thousand years; their body is hard like thunderbolt and all are endowed with powers of Ayuta elephants. No one is satisfied with a little of the pleasures of sexes; so all men live happily with their wives, etc. It is not that the males alone are so happy; the females also are

full of youth throughout their lives and become pregnant, when they are less than one year old. In short, the inhabitants of the Var\d{s}as enjoy for ever as the persons of Tret\=a Yuga do.

Hear ends the Seventh Chapter of the Eighth Book on the story of the Ganges and the continents in Sr\={\i} Mad Dev\={\i} Bh\=agavatam, the Mah\=a Pur\=anam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



