\chapter{On the description of the family of Manu}

1-23. N\=ar\=aya\d{n}a said :-- O N\=arada! When the Bhagav\=an went away to Vaikuntha, establishing the Earth in due position and equilibrium, Brahm\=a spoke thus to his son :-- ``O Powerful Son of mine, O Sv\=ayamhhuva! The best of those that are filled with Teja (energy) and Tapas! Now go on with your work of creation, as you think proper, on this earth, the Upholdress of all the J\={\i}vas. And worship the Puru\d{s}a, the Lord of Sacrifices, according to the division of place and time, and with all the necessary materials of various kinds, high and low, and those that will be useful in performing your sacrifices. Do Dharma according to the \'S\=astras, and according to the Var\d{n}a(the different castes) and \=Asrama (Brahm\=acharya, etc.); thus by gradually going on step by step in the path of Yoga, your progeny will be multiplied. Beget lovely sons and daughters, of good fame, culture, modesty and ornamented with various other good qualifications, like yourself; then marry your daughters when they will come to the marriageable age, to proper persons of good qualifications and then fix your mind thoroughly on the Excellent Puru\d{s}a that is the very Best. O Child! Now go and serve the Bhagav\=an with devotion as I have advised you; and you will certainly attain that which is difficult to be worshipped and obtained. Advising thus his son Sv\=ayambhuva Manu, and starting him in his work; of creation, the Lotus-born, the Lord of all the subjects, Brahm\=a went away to His own abode. When Brahm\=a went away, having ordered his son to create progeny and subjects, Manu, took that seriously in his heart and began to do that work. In due time, he had two powerful sons named Priyavrata and Utt\=anap\=ada and three lovely beautiful daughters endowed with various good qualities. Hear the names of the three daughters. The first daughter, the purifier of the world, was named \=Ak\=uti; the second was named Devah\=uti and the third was Pras\=uti. The first daughter \=Ak\=uti was married to the Mahar\d{s}i Ruchi; the second was married to the Prajapati Karddama; and the third was married to the Prajapati Dak\d{s}a. And know that all the beings in this world had their origin from this last daughter. Now hear the progeny of these three daughters and the Mahar\d{s}is respectively :-- By Mahar\d{s}i Ruchi was born one son named Yaj\~na, in the womb of \=Ak\=uti; he is the part of

the Bhagav\=an \=Adi Purusa Vi\d{s}\d{n}u; by Mahar\d{s}i Karddama, in the womb of Devah\=uti was born the Bhagav\=an Kapila Deva, the famous author of the \'S\=ankhya \'S\=astra; and in the womb of Pras\=uti, by the Praj\=apati Dak\d{s}a were born some daughters only; know that the Devas, men, beasts and birds were all created by this Praj\=apati Dak\d{s}a. These offsprings were the first promulgators in the work of creation. In the Sv\=ayambhuva Manvantara, the powerful Bhagav\=an Yaj\~na, by the help of the Deva named Y\=ama, saved his mother's Father Manu from the attacks of the Rak\d{s}asas; and the great Lord of the Yogis, the Bhagav\=an Kapila, remained for a while in his \=Asrama and gave spiritual instructions to his mother Devah\=uti, by which Avidy\=a could be at once destroyed, and disclosed his \'S\=ankhya \'S\=astra, the great work on the Spiritual Philosophy and the special Dhy\=ana Yoga in all its details and finally went for Sam\=adhi to the \=Asrama of Pulaha where the great Deva S\=amkhy\=a ch\=arya lives up to today. Oh! I bow down to the great Yog\=ach\=arya, the Bhagav\=an Kapila Deva, the Fructifier of all desires, the more remembrance of Whose Name makes easily the Yogi realise the meaning of the S\=amkhya J\~n\=ana. The sins are immediately destroyed of those that hear or read the holy anecdote of the progeny of the daughters of Manu. O Child! Now I describe to you the progeny of the sons of Sv\=ayambhuva Manu. Hear attentively. The hearing of which will enable one to enter into the highest place. Now is being described the history about the progeny of those who formed those Dvipas (islands) Var\d{s}as(countries) and oceans for the welfare and happiness of all the creation and for the use of them. Hear.

Here ends the Third Chapter of the Eighth Book on the description of the family of Manu in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, the Mah\=a Pur\=a\d{n}am, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



