\chapter{On the description of the remaining Dv\={\i}pas}

1. N\=arada said :-- Thou, the Knower of everything! Describe about the remaining Dv\={\i}pas, knowing which we will be highly delighted.

2-36. N\=ar\=aya\d{n}a spoke thus :-- The very vast Ghrita S\=agara (the ocean of clarified butter) is encircling the Ku\'sa Dv\={\i}pa. Next to it is the Krau\d{n}cha Dv\={\i}pa. It is twice as large as Ku\'sa. The K\d{s}\={\i}ra S\=agara (the ocean of milk) is surrounding this Dv\={\i}pa. The Krau\d{n}cha mountain is standing here. The name of this Dv\={\i}pa is derived from this mountain. In days gone by, the highly intelligent K\=artikeya burst this mountain by his own prowess. This Dv\={\i}pa is washed by the K\d{s}\={\i}ra S\=agara; and Varu\d{n}a is its Regent. The son of Priyavrata, Ghritapristha, respected by all and whose prosperity knows no end, is the Lord of this Dv\={\i}pa. He divided this Dv\={\i}pa into the seven parts and distributed them to his sons and named the Var\d{s}as after the names of his sons. He made his sons the rulers of those places and he himself took the refuge of the Bhagav\=an N\=ar\=aya\d{n}a The names of the seven Var\d{s}as are respectively :-- \=Ama, Madhuruha, Meghapristha, Sudh\=amaka, Bhr\=ajistha, Lohit\=ar\d{n}a, and Vanaspati. O N\=arada! The seven mountains and the rivers there are very celebrated throughout the worlds. The names of the mountains are :-- \'S\=ukla, Vardham\=ana, Bhojana, Upavarha\d{n}a, Nanda, Nandana, and Sarvatobhadra. The names of the rivers are :-- Abhay\=a, Amritaugh\=a, \=Aryak\=a, T\={\i}rthavat\={\i}, Vrittir\=upavat\={\i}, \'Sukl\=a, and Pavitravatik\=a. The inhabitants there drink

the highly pure water of these rivers. The people there are divided into four colours Puru\d{s}a, Ri\d{s}abha, Dravi\d{n}a, and Vedaka and they worship the Bhagav\=an Varu\d{n}a, of the form of water. Then they become very discriminative, and, with great devotion, and holding full within their folded palms the water, repeat the following mantra :-- ``O Water! Thou art the essence V\={\i}rya of the Person Bhagav\=ana and Thou sanctifiest the Bh\=urloka, Bhuvarloka, and Svarloka. Thou destroyest the sins of all. We all are touching it; purify our bodies.'' After finishing their mantrams, they sing various hymns to Varu\d{n}a. Next to the K\d{s}iroda S\=agara, is the \'Saka Dv\={\i}pa, thirty two lakh yoyanas wide, surrounded by the Dadhi S\=agara (the ocean of curds), of similar dimensions. Here the most excellent tree named the \'Saka tree exists. O N\=arada! The Dv\={\i}pa is named so after the tree. Medh\=atithi, the son of Priyavrata is the Lord of this Dv\={\i}pa. He divided this land into the seven Var\d{s}as and distributed each to his seven sons respectively; and ultimately he took refuge to the path of Yoga. The names of the seven Var\d{s}as are Purojava, Manojava, Pavam\=anaka, Dh\=umr\=an\={\i}ka, Chitrarepha, Bahur\=upa, and Vi\'svadhrik. In these Var\d{s}as there are seven mountains, one in each Var\d{s}a, as forming their boundaries; and there are seven rivers also. The names of the mountains are :-- \=I\'sana, \=Uru\'sri\d{n}ga, Valabhadra, \'Sata Ke\'sara, Sahasra-srotaka, Devap\=ala, and Mah\=asana; the names of the rivers are :-- Anagh\=a, \=Ayurd\=a, Ubhayaspristi, Apar\=ajit\=a, Pa\~nchapad\={\i}, and Sahasra\'sruti and Nijadhriti. These seven rivers are all very big and resplendent with lustre. The people are divided into four classes :-- Var\d{s}as, Satyavrata, Kratuvrata, D\=anavrata, and A\d{n}uvrata. They all take the Pr\=a\d{n}ay\=ama exercise and thereby bring the R\=ajas, and Tamo Gu\d{n}a under their subjection and they worship Hari, of the nature of Pr\=a\d{n}a Vayu, Higher than the Highest. Their mantra is this :-- ``He has entered into all the living beings and nourishes them by the Pr\=a\d{n}a and other faculties; He is the Internal Ruler of all and the Supreme Controller; this Universe is under His control; let Him protect and nourish us.'' O N\=arada! Next to this Dadhi S\=agara is Pu\d{s}kara Dv\={\i}pa; it is twice as large as \'S\=aka Dv\={\i}pa. It is surrounded by the Dudha S\=agar (the ocean of milk) all twice as large. The leaves of Pu\d{s}kara tree that shines in the Pu\d{s}kara Dv\={\i}pa, are fiery like golden flames; they are as clean and pure. Crores and crores of leaves, golden in colour ornament this Tree. V\=asudeva, the Guru of all the Lokas, has created this Pu\d{s}kara Dv\={\i}pa as the seat of Paramesth\={\i} Brahm\=a, possessed of six extraordinary powers, for the purpose of creation. There is one mountain in this Dv\={\i}pa; it is divided into two parts, named Arv\=ach\={\i}na and Par\=ach\={\i}na. These form the boundaries of the two Var\d{s}as. The

mountain is one Ayuta Yoyana high and one Ayuta Yoyana wide. There are four cities on the four sides. Indra and the three other Lokap\=alas are the lords of these cities. The Sun God comes out from their top and circumambulating Meru, goes there again. The whole year is his Chakram, circle of circuit; His path is Uttr\=aya\d{n}am and Dak\d{s}i\d{n}\=aya\d{n}am. V\={\i}tihotra, the son of Priyavrata is the lord of this island. He distributed the two Var\d{s}as amongst his two sons, Rama\d{n}a and Dh\=atak\={\i}. They rule over the two Var\d{s}as named also after them. Like the inhabitants of the above Var\d{s}as, the people also get powers of themselves and worship devotedly the God seated on the lotus and follow such path of the Yoga as leads them to the Brahm\=a S\=aloky\=a, etc. The mantra runs thus :-- ``We bow down to that One God, without a Second, of the nature of Peace, Who is the Fruit of all the Karmas, Who is the seat of illumination of Brahm\=a, Who is established in Unity, and Who is worshipped by all the Lokas.''

Here ends the Thirteenth Chapter of the Eighth Book on the description of the remaining Dv\={\i}pas in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



