\chapter{On the description of the continents and of Bh\=aratavar\d{s}a}

1. N\=ar\=aya\d{n}a said :-- In this Bh\=aratavar\d{s}a (India) I am present, as the First of all, in the shape of the Man, at present existing before you. Let you praise Me incessantly thus:--

2-28. N\=arada said :-- Thou art the Bhagav\=an. Obeisance to Thee. Thou art completely free from attachment and envy and Thou art the incarnate of Dharma, J\~n\=ana and Vair\=agyam (Dispassion). No trace of Ahamkara (egoism) is in Thee. Obeisance to Thee. Thou hast nothing of wealth; Thou art the foremost of the family of \d{R}i\d{s}is; Thou art Nara N\=ar\=aya\d{n}a. Thou art the Parama Hamsa (the highest class of ascetics and renouncers of the world). Thou art the Highest Guru; Thou art happy with Thyself and the Leader of all; I bow down to Thee. Thou art the Lord of all, yet not mixed in the work of creation, etc. Though Thou residest in every embodied being, yet no hunger and thirst can press Thee; though Thou art the Witness, yet Thy sight is not at all disturbed by the proximity. Thou art not at all connected and no desires can come to Thee; Thou art the Witness. So I bow down to Thee. This path of Yoga had

come out of Thee and it is established in Thee. The Bhagav\=an Hira\d{n}yagarbha has given instructions on the cleverness in Yoga, thus :-- That leaving aside the egoism of this body, the seat of all evils, the people at last will take resort to the path of Bhakti and set their minds on Thee, Who art beyond the pale of the three Gu\d{n}as. If, like the ignorant persons very much attached to this world as well as the next, the wise pass their time in the thoughts of their sons, wives, and riches and ultimately become very much pained on the destruction of this ugly body, persons, versed in J\~n\=ana and Vij\~n\=ana be afraid in the end of their separation fron the body, then their study of the \'S\=astras and all their labours become labours merely, without any effect. When such a thing occurs to the learned, then O Thou, the Subduer of the knowledge of senses! Thou Thyself givest us instructions that we may easily think of Thee. Then this terrible clinging to this ugly body that Thy M\=ay\=a brings about and which it is very difficult to discard by other means, will leave us quickly. N\=arada, the seer of all, well qualified with the knowledge of all the Tattvas, the foremost of the Munis, thus worships it and praises N\=ar\=aya\d{n}a Who is unaffected by any of the \=Up\=adhis and who corresponds to Nitya and Leel\=a (both Permanent and the Pastime-making). O Devar\d{s}i! I now describe to you the rivers and mountains in Bh\=aratvar\d{s}a. Hear attentively. Malaya, Mangalaprastha, Main\=aka, Chitrak\=uta, Ri\d{s}abha, K\=utaka, Kolla, Sahya, Devagiri, Ri\d{s}yam\=uka, \'Sri\'saila, Vyankata, Mahendra, V\=aridh\=ara, Vindhya, \'Suktim\=an, Rik\d{s}a, P\=ariy\=atra, Dro\d{n}a, Chitrak\=uta, Gobardhana, Raivataka, Kakubha, N\={\i}la, Gaurmukha, Indrak\={\i}la, K\=amagiri and many other mountains, that cannot be numbered. Great merits accrue on seeing these mountains. Hundreds and thousands of rivers issue from these mountains. Drinking their waters, bathing in them and visiting them and singing their praises completely destroy the sins of mind, word and body. The names of the rivers are :-- T\=amrapar\d{n}\={\i}, Chandrava\'s\=a, Kritam\=al\=a, Vatodak\=a, Vaih\=ayas\={\i}, K\=aver\={\i}, Ve\d{n}\=a, Payasvini, Tungabhadr\=a, Kri\d{s}\d{n}aven\=a, \'Sarkar\=a, Vartak\=a, God\=avar\={\i}, Bh\={\i}marath\={\i}, Nirbindhy\=a, Payo\d{s}\d{n}ik\=a, T\=ap\={\i}, Rev\=a, Suras\=a, Narmad\=a, Sarasvat\={\i}, Charma\d{n}vat\={\i}, and the Indus, Andha, and Sone, \d{R}i\d{s}ikuly\=a, Tris\=am\=a, Vedasmriti, Mah\=anad\={\i}, Kau\'sik\={\i}, Yamun\=a, Mand\=akini, Drisadvat\={\i}, Gomat\={\i}, Saray\=u, Oghavat\={\i}, Saptavat\={\i}, Susam\=a, \'Satadru, Chandrabh\=ag\=a, Marudbridh\=a, Vitast\=a, Asikn\={\i} and Vi\'sv\=a and many other rivers. Those persons that take their births in this Var\d{s}a enjoy the (Divya) Divine, human, and the lower enjoyments, as the effects of their Karmas, in accordance with their S\=attvik, R\=ajasik, and T\=amasik qualities. All the inhabitants of this Var\d{s}a take their Sanny\=asa, V\=anaprasthas, etc., in accordance with the rules of their Var\d{n}as (castes) respectively and enjoy

beatitudes as dictated (in their \'S\=astras). The Vedav\=adis (the propounders of the Vedas), the \d{R}i\d{s}is, and the Devas declare that this Var\d{s}a is superior to all the other Var\d{s}as as the Grace of God is easily obtained here. They say: Oh! What an amount of good works, the inhabitants of Bh\=arata Var\d{s}a did in their previous lives, that by their influence, the Bhagav\=an Hari has become pleased with them without their having any S\=adhan\=as (good Yogic practices). Therefore we also desire to have our births there, for there if we be born amidst the persons there, we would be able serve Hari, Mukunda in every way. What will severe austerities, gifts, sacrifices, and practising vows avail us? Even if we go to Heaven, what benefit we will derive there? Then we would never be able to bring our mind to the lotus-feet of the Bhagav\=an N\=ar\=aya\d{n}a. Our tastes for sense enjoyments would increase and we would be wholly deprived of Bhagav\=an. Far better it is to be born in this Bh\=arata Var\d{s}a as short-lived persons, than to acquire other places where one can live up to the period of Kalpa and then be reborn. For the intelligent people of Bh\=arata, getting this mortal coil can within a short period devote themselves to the Bhagav\=an Hari and be free from rebirth. That place is never to be visited and served, even if that be Svarga Loka where the nectar-like ocean of the sayings of Vaikuntha are not heard; where there are no assemblages of saints who take refuge of the Lotus feet of Bhagav\=an; where the great sacrifices and festivities of the Bhagav\=an Vi\d{s}\d{n}u are not celebrated. Those persons, that do not try for Mok\d{s}a, when they get human births and when they possess J\~n\=ana, Kriy\=a (methods of doing sacrifices and other works for the help of J\~n\=ana) and Dravya (materials for those sacrifices), are certainly like beasts and get themselves repeatedly entanged into bondages. The people of Bh\=arata Var\d{s}a offer oblations to Indra and other deities, invoking their names separately, with proper rules (Vidhis), mantras, and purod\=asas (a sacrificial oblation made of ground rice and offered in Kap\=alas or vessels); but the All sufficient Bhagav\=an Hari, the Giver of blessings accepts those very gladly.

True that He grants unto the people what they ask from Him; but He hardly gives anybody the Param\=artha, the Highest Goal. The reason being that the people ask again and again after receiving what they had asked. So He gives of Himself His own lotus-feet to those who, having abandoned all their desires, worship Him and Him only with the idea of love. (1) Thus :-- We are living very happily in the Heavens as the fruit of the Ist\=ap\=urta that we performed fully; yet we want that we may also be born in Bh\=arata Var\d{s}a with our memory all renewed with the thought of the Bhagav\=an Hari. The Bhagav\=an residing in this Bh\=aratvar\d{s}a, awards the highest welfare to its inhabitants.

29-32. N\=ar\=aya\d{n}a said :-- O Devar\d{s}i! Thus the Devas of the Heavens, the Siddhas, the Highest \d{R}i\d{s}is, sing the excellent merits of this Bh\=aratavar\d{s}a. There are the eight Upadv\={\i}pas of this Jambudv\={\i}pa :-- When the sons of the King Sagara were searching for the footprints of the stolen horse (for A\'svamedha sacrifice), they found these Upadv\={\i}pas; so it is definitely stated. The names of out and made these are :-- Svar\d{n}aprastha, Chandra\'sukra, \=Avartana, Ram\=a\d{n}aka, Mandarop\=akhya, Hari\d{n}a, P\=anchajanya, and Ceylon. The length and breadth of Jambudv\={\i}pa have been described; now the other six Dv\={\i}pas, Plak\d{s}a and others will be described. Hear.

Here ends the Eleventh Chapter of the Eighth Book on the description of the continents and of Bh\=aratavar\d{s}a in the Mah\=apur\=a\d{n}am, \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



