\chapter{On the narration of Plak\d{s}a, \'S\=almala and Ku\'sa Dv\={\i}pas}

1-37. N\=ar\=aya\d{n}a said :-- The Jambudv\={\i}pa has been described, how it is and what is its width. It is surrounded on all sides by the salt ocean. As Meru is surrounded by Jambudv\={\i}pa, so the salt ocean is surrounded by Plak\d{s}adv\={\i}pa, twice its size. As the ditch is surrounded by gardens, so the salt ocean is surrounded by gardens. As the Jambu tree exists in Jambudv\={\i}pa, so the Plak\d{s}a tree exists in the Plak\d{s}a Dv\={\i}pa and of the same size. The name Plak\d{s}a Dv\={\i}pa is derived from the name of this Plak\d{s}a tree. This tree is of a golden colour. Fire exists at its bottom with form incarnate. This is, named Saptajihva. The Ruler of this island is Idhmajihva, the son of Priyavrata. He divided his island into the seven Var\d{s}as and distributed them, to each of his seven sons and he himself took refuge of the path of Yoga, so much liked by the Knowers of Self and he got the Bhagav\=an V\=asudeva. The names of those seven Dv\={\i}pas are :-- \'Siva, Yavas, Subhadra, S\=anti, K\d{s}ema, Amrita, and Abhaya. Seven rivers and seven mountains exist respectively in the seven islands. The rivers are :-- Aru\d{n}\=a, Nrimn\=a, Angiras\={\i}, S\=avitr\={\i}, Suprabh\=atik\=a, Ritambhar\=a, and Satyambhar\=a. The names of the mountains are Ma\d{n}ik\=uta, Vajrak\=uta, Indrasena, Jyoti\d{s}m\=ana, Supar\d{n}a, Hira\d{n}yasth\={\i}va, and Megham\=ala. The seeing and drinking the waters of these rivers take away all sins and all darkness due to ignorance. The four castes live here, Hamsa, Patanga, \=Urdh\=ayana, and Saty\=anga, corresponding to the four castes Br\=ahma\d{n}a, etc. The inhabitants of this Plak\d{s}a Dv\={\i}pa live for one thousand years and all are of variegated wonderful appearances. They follow the customs

and usages dictated by the Vedas and worship the God Sun for the attainment of Heaven. The mantra by which the worship is done is this :-- We take refuge unto that Sun, Who is the Body Manifest of the Ancient Person Vi\d{s}\d{n}u and Who is the Ordainer of Satya (Truth), Rit\=a (Straight forwardness), Brahm\=a, Amrita (Immortality) and Mrityu (Death); O N\=arada! All the persons here live long, of vigorous senses, energetic, powerful, intelligent, enthusiasts, and valorous; everyone gets extraordinary powers of themselves. Next to this Plak\d{s}a Dv\={\i}pa is the Ik\d{s}hu Ocean. This Ik\d{s}u S\=agara surrounds the Plak\d{s}a Dv\={\i}pa. Next comes \'Salmala Dv\={\i}pa. It is twice as large as Plak\d{s}a. This Dv\={\i}pa is surrounded by Sur\=as\=agara (the ocean of wine). There is a tree named \'Salmal\={\i} in this island, which is as large as the Plak\d{s}a tree. The high-souled Garuda resides on that tree, Yaj\~nav\=ahu is the Ruler of this place. He was born of Priyavrata; he divided his Var\d{s}a into the seven parts and distributed each of them to his seven sons respectively. Now hear the names of these Var\d{s}as :-- Surochana, Saumanasya, Rama\d{n}a, Deva Var\d{s}a, P\=aribhadra, \=Apy\=ayana, and Vij\~n\=ata. Seven mountains and seven rivers exist there respectively. The names of the mountains are :-- Sarasa, \'Sata\'sringa, V\=ama Deva, Kandaka, Kumuda, Pu\d{s}pavar\d{s}a, and Sahasra-\'Sruti. Now hear the names of the rivers. Anumati, Sin\={\i}b\=al\={\i}, Sarasvat\={\i}, K\=uhu, Rajan\={\i}, Nand\=a, and R\=ak\=a; these are the seven rivers. The people are divided into the four castes :-- \'Srutadhara, V\={\i}ryadhara, Vasundhara, I\d{s}undhara. These correspond to the Br\=ahma\d{n}as, etc. They worship the Bhagav\=an Moon, the Controller of all and the Creator of all the Vedas. They offer food duly in the black and white fortnight to their Pitris. The mantra for their worship is :-- ``Let Soma, the King of all, be pleased.'' O N\=arada! Next to Sur\=as\=agara is Ku\'sadv\={\i}pa, surrounded by Ghritas\=agara (the ocean of clarified butter). Its dimensions are twice as large. Here are blades of Ku\'sa grass, of a very resplendent colour. The name of the Dv\={\i}pa is from this Ku\'sastamba. This bundle of Ku\'sa, illumines all the quarters with their gentle rays. The Ruler of the Dv\={\i}pa is Hira\d{n}yaret\=a, the son of Priyavrata. He divided the Dv\={\i}pa into seven parts and distributed each of them to each of his seven sons respectively. The names of the seven sons are :-- Vasu, Vasud\=ana, Dhridharuchi, N\=abhigupta, Stutyavrata, Vivikta, and Bh\=amadevaka. There are seven mountains forming the seven boundaries and so are the seven rivers. Hear the names of these. The names of the mountains are :-- Chakra, Chatuh\'sringa, Kapila, Chitre Dev\=an\={\i}ka, K\=uta, \=Urdharom\=a and Dravi\d{n}a. The names of the rivers are :-- Rasakuly\=a, Madhukuly\=a, Mitravind\=a, \'Srutavind\=a, Devagarbh\=a, Ghrit\=achyut, and Mantramalik\=a. The inhabitants of the Ku\'sadv\={\i}pa drink the waters of these

rivers. There are the four castes here Ku\'sala, Kovida, Abhiyukta, and Kulaka corresponding to the Br\=ahma\d{n}as, etc. They are all powerful like Indra and the other chief Devas; and all of them are omniscient. They worship the Fire God and perform various good works in honour of Him. Their mantra is this :-- O Fire! Thou carriest the oblations direct to Para Brahm\=a. In the sacrifices of the Devas, Thou worshippest that Personal God and offerest to Him whatever are given in oblations, mentioning the respective limbs of His Body. Thus the inhabitants of that Dv\={\i}pa worship the Fire God.

Here ends the Twelfth Chapter of the Eighth Book on the narration of Plak\d{s}a, \'S\=almala, and Ku\'sa Dv\={\i}pas in the Mah\=apur\=anam, \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



