\chapter{On the worship of the Dev\={\i}}

1-2. N\=arada said :-- ``O Bhagav\=an! Of what sort is the Dharma, i.e., the worship of the Dev\={\i}? In what way shall we worship Her, so that She gives us the Highest Place? What are the methods and forms of Her Worship? Where and when shall we worship Her! So that the Durg\=a Dev\={\i} saves us from the above mentioned hells.''

3-20. N\=ar\=aya\d{n}a said :-- O Devar\d{s}i! You are the foremost of the Knowers of Truth. I will therefore tell you how the Dev\={\i} gets pleased and how Her worship is conducted according to the Dharma. Hear attentively. O N\=arada! I will also describe the nature of Sva Dharma. Hear that also. When one worships, with due rites and ceremonies, the Dev\={\i} in this beginningless world, She Herself removes all the terrible dangers and difficulties. Hear the rules how the people worship the Dev\={\i}. When the Prat\={\i}pat Tithi (the first day after the Full or New Moon) comes, one should worship the Dev\={\i} with a present of rice, etc., with ghee (clarified butter) and give that to the Br\=ahmi\d{n}s. Then one becomes completely free from any disease. On the second day (Dv\={\i}tiy\=a Tithi), one must serve the Mother of the Universe with sugar and give that to the Br\=ahmins; he then becomes long-lived. On the third (tithi) on commencing with the Pooj\=a, the worshipper must give milk to the

Dev\={\i} and give that to a best Br\=ahmi\d{n}; he is then freed from all his troubles and ailings. On the fourth (tithi) the worshipper is to offer a cake of flour to the Dev\={\i} and then give that to a Br\=ahmi\d{n}; no obstacles come to that man. On the fifth tithi, the worshipper is to offer plantains to the Dev\={\i} and then to give that to the Br\=ahmi\d{n}s; he thus becomes intelligent. On the sixth tithi, the worshipper gives honey to the Dev\={\i} and then that to a Br\=ahmi\d{n}; he gets thus the beauty of his body. On the seventh tithi, the Br\=ahmi\d{n} gives to the Dev\={\i} the Naivedya (an offering of rice, etc.) with Gur (sugarcandy) and then that to the Br\=ahmi\d{n}s; he then becomes freed from his mental sorrows. On the eighth day, if one gives cocoanut, one is freed of one's remorse, etc.; on the ninth, if anybody gives fried rice (L\=aj), he will have his happiness increased both in this world and in the next. O Muni! If on the tenth tithi, one offers to the Dev\={\i} black Til (sesamum) and then to the Br\=ahmi\d{n}, he becomes free from the fear of death. If, on the eleventh tithi, (ek\=ada\'s\={\i}) one gives the curd to the Dev\={\i} and then to the Br\=ahmi\d{n}, one becomes a great favourite of the Dev\={\i}. If on the twelfth day, one offers to the Dev\={\i} and to the Br\=ahmi\d{n} the Chipitaka rice or grain (well parched and flattened) one becomes a favourite of the Dev\={\i}. If, on the thirteenth day one gives to the Bhagavat\={\i} grains and then that to a Br\=ahmi\d{n}, one gets progeny. If, on the fourteenth day, one gives to the Dev\={\i} the flour of fried barley or other grains (\'Saktu) and then that to a Br\=ahmi\d{n}, one becomes a favourite of \'Siva. If on the Full Moon day, one offers to the Dev\={\i} P\=ayasa and then that to a Br\=ahmi\d{n}, then one's Pitris are uplifted to the higher regions.

21-42. O Muni! On the above tithis, if one forms daily Homas, as stated in the P\=uj\=a Chapter, the Dev\={\i} becomes very pleased. The articles corresponding to the tithi as mentioned above destroy all the evils and inauspicious omens.

On Sunday, it is a rule to give an offering of P\=ayasam (a food prepared of rice, milk, and sugar). On Monday, the milk; on Tuesday, the nice plantains; on Wednesday, the fresh butter; on Thursday, the gud or sugarcandy, Friday, the white sugar, and on Saturday, it is a rule to give the clarified butter of cow's milk. Now hear what should be offered on the Nak\d{s}attras. The following are the Naivedyas given to each of the Nak\d{s}attras, in due order, from A\'svin\={\i} :-- Clarified butter (ghee), sesamum (Til), sugar, curd, milk, Kil\=atak (M\=al\=ai, milk), Dadhik\=urchi (M\=al\=ai Curd), Modaka (a kind of sweetmeat, a confection) Phe\d{n}ik\=a, Ghrita Mandaka, a sort of sweet meat of wheaten flour and gur, Vatapattra, Ghritapura (Ghior), Vataka, Kharjura juice (of the datepalm), a sort of sweet meat of Gur and gram, honey, \'S\=ura\d{n}a, Gur

Prithuka, grapes, datepalms, Ch\=arak\=as, Ap\=upa, Navan\={\i}ta (fresh butter), mudga, modaka, and M\=atulinga. Now hear what are given in the Vi\d{s}kambha and the other Yogas. The World Mother becomes very much pleased when one offers to Her the following things :-- Gur, honey, ghee, milk, curd, Takra, ap\=upa, fresh butter, Karkat\={\i}, Ku\d{s}m\=anda, Modaka, Panasa, plantain, Jambu (rose-apple), mangoe, sesamum, oranges, D\=adima, (pomegranate) Vadar\={\i} (Jujube) the Dh\=atr\={\i} (\=Amalaki) fruit, P\=ayasa, Prithuka, gram, cocoanut, Jamb\={\i}ra. Kaseru, and \'S\=ura\d{n}a. The auspicious events occur when these are offered. The intelligent persons have thus decided to give the above on Vi\d{s}kambha and the other Yogas. Now hear :-- I will describe the things that are offered on the respective Kara\d{n}as :-- Kams\=ara, Mandaka, Phen\={\i}, Modaka, Vatapattraka, Ladduka, Ghritap\=ura, Til (Sesamun), curd, ghee, and honey. These are to be offered devotedly to the Dev\={\i} on the respective Kara\d{n}as. Now I will describe to you the other offerings very pleasing to the Dev\={\i}. Hear. O N\=arada! Hear it with great attention and love. On the third tithi of the bright fortnight, in the month of Chaitra, one is to worship duly the Madh\=uka tree and offer Pa\~ncha Kh\=adya (the five sorts of food). So hear what articles are to be offered according to the rules that are laid down on the third day of the white fortnight of the other months. The Gur, in the month of Vai\'s\=akh; the honey, in Jyaistha; the fresh butter, in \=A\d{s}\=adha; the curd, in \'Sr\=ava\d{n}a; the \'Sarkar\=a, in Bh\=adra; the P\=aya\'sa, in \=A\'svin; the pure milk, in K\=artik; the Phen\={\i}, in Agrah\=aya\d{n}a; the Dadhi K\=urch\={\i}k\=a in Pau\d{s}a; the clarified butter of cow's milk, in M\=agha, and the cocoanut offerings, in the month of Ph\=alguna. Thus with these twelve sorts of offerings, one is to worship the Dev\={\i} in the twelve months respectively.

43-69. One should worship the Dev\={\i} in the Madh\=uka tree with these names :-- Mangal\=a, Vai\d{s}\d{n}av\={\i}, M\=ay\=a, K\=ala-r\=atri, Duratyay\=a, Mah\=am\=ay\=a, M\=atang\={\i}, K\=al\={\i}, Kamalav\=asin\={\i}, \'Siv\=a, Sahasrachara\d{n}\=a, and Sarva mangalar\=upi\d{n}\={\i} (One name for each of the 12 months). Finally, to bring the vow to a good issue and to have one's desires fulfilled with greater success, one is to sing stotras (hymns) to the M\=ahe\'svar\={\i}, the Controller of all the gods, in that Madh\=uka tree, thus :-- Thou art lotus-eyed; obeisance to Thee! Thou art Jagaddh\=atr\={\i}, the Upholdress of the Universe, I bow down to Thee; Thou art Mahe\'svar\={\i}, Mah\=a Dev\={\i}, and Mah\=amangalar\=upi\d{n}\={\i} (Thou art the great Dev\={\i}, and Thou dost great good to all). Thou destroyest the sins, Thou givest Mok\d{s}a or final liberation. Thou art Parame\'svar\={\i}, Thou art the World Mother and Thou art of the nature of the Highest Brahm\=a. Thou art Madad\=atr\={\i} (the giver of Mada, the Supreme Felicity and rapture or excessive

delight), Thou art maddened with Mada the (Excessive Joy); Thou canst be reached when Thou art given proper veneration; Thou art the the Most High. Thou art Intelligent; Thou art meditated upon by the Munis; and Thou dwellest in the Sun. Thou art the Lord of the several Lokas (worlds); Thou art endowed with the Highest Knowledge; and Thou art of the colour of water at the time of Pralaya (the Universal Dissolution). Thou art worshipped by the Gods and the Asuras for the destruction of the Great Moha. So Great Victory to Thee! Thou art the Rescuer of one from the abode of Death; Thou art worshipped by Yama, Thou art the elder of Yama, Thou art the Controller of Yama and Thou art worshipped by all. Obeisance to Thee! Thou art impartial; Thou controllest all; Thou art perfectly unattached; Thou destroyest the people's worldly attachments; Thou art The One to whom all look for the fructification of their desires; and Thou art the Compassion Incarnate. Thou art, worshipped by the names :-- Ka\d{n}k\=ala Kr\=ura, K\=am\=ak\d{s}\={\i}; M\={\i}nak\d{s}\={\i} Marma bhedin\={\i}, M\=adh\=uryar\=upa\'s\=alin\={\i}; and Thou art worshipped with the Pra\d{n}ava Om prefixed to all the Stotras and the Mantras. Thou art of the nature of the Seed M\=ay\=a (m\=ay\=av\={\i}ja); Thou canst be realised by repeating the mantra and Thou canst be pleased by the deep concentration (Nididy\=asana) on Thee. Thou canst be reached by all men through their minds and Thou dost things that are pleasing to the Mah\=a Deva. Thou dwellest in the trees A\'svattha, Vata (Peepul tree), Neem, Mangoe, Kapittha (wood apple tree), and the tree Kul (Jujube) trees. Thou art the Palms (Jack) tree, Thou art Arka, Kar\={\i}ra and K\d{s}\={\i}ra trees. Thou residest in Dugdha vall\={\i} (the milky juice of plants); Thou art the Compassion Incarnate; and fit to shew mercy. Thou art sincerity and kindness and Thou art the Consort of the Omniscient. So Victory to Thee! O N\=arada! After the worship, if one performs the stotra above described, to the Dev\={\i}, the worshipper derives all sorts of Pu\d{n}yams (merits). He who reads daily the Stotra, pleasing to the Dev\={\i}, becomes freed of all sorts of diseases, pain and freed of his fear of passions, so hard to conquer! What more than this that he who wants money, gets money; who wants Dharma, gets Dharma; he who wants K\=ama, gets his K\=ama (objects of desires); and he who wants Mok\d{s}a, gets Mok\d{s}a. The Dev\={\i} is the Awarder of the Chatur Varga fourfold fruits.

If this Stotra be read, the Br\=ahmi\d{n} becomes Vedavit, the knower of the Vedas; the K\d{s}attriya gets the victory; the Vai\'sya gets wealth and the \'S\=udra gets happiness. If this Stotra be read with devotion and attention, the Pitris get undying satisfaction, lasting till Pralaya (the time of universal dissolution). Thus I have described to you the method of

worshipping the Dev\={\i}. The Devas consider it with great attention. He gets the Dev\={\i} Loka, who performs the worship of the Dev\={\i}, as above described, with devotion. O Br\=ahma\d{n}a! When the Dev\={\i} is thus worshipped, all the desires are fulfilled; all sins are destroyed; and, in the end, one's mind becomes pure and the worshipper is respected and worshipped everywhere. O Son of Brahm\=a! His fear of going into hell is destroyed by the Grace of the Dev\={\i}; even in dreams, he does not fear anything. By the Grace of Mah\=a M\=ay\=a, his sons and his grandsons, riches and grains multiply and multiply. He becomes a great and steadfast devotee of the Dev\={\i}; there is no doubt in this. Now I have described to you completely the rules of worship of the Dev\={\i}. When one performs this, one becomes freed of the Narakas; and all sorts of good things come to him. O Muni! The Madh\=uka worship and the monthly worship have been described also. He who performs this Madh\=uka worship fully, never meets with any disease or obstacles. Now I will describe to you the other five aspects of the Great Dev\={\i} of the nature of Prakriti. Her Name, Form and Origin give pleasure to all the worlds. O Muni! Now hear this Prakriti Pa\~nchaka, its narrative and the greatness thereof. Know that this is as curious as it gives liberation.

Here ends the Twenty-fourth Chapter of the Eighth Book on the worship of the Dev\={\i} in the Mah\=a Pur\=a\d{n}am, \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.

[The Eighth Book completed.]

Here ends the Second Volume.



