\chapter{On the narrative of the sins leading to hells}

1. N\=arada said :-- ``O Everlasting One! O Muni! Now describe the various actions that lead to these tormenting hells. I like to hear about them in detail.''

2-52. N\=ar\=aya\d{n}a said :-- O Devar\d{s}i! He who steals other's sons, wives, and riches, the wicked fellow is taken to Yama by His messengers. Tied down hard and fast by the terrible messengers of Yama, by the K\=ala rope (rope of time), he is taken to the T\=amisra hell, the place of many torments. There the Yama's attendants punish him, beat him and threaten him; and he becomes stupefied, and feels himself very weak, distressed and ultimately faints. He who deceives another's husband and enjoys his wives; the Yama's servants drag him down to Andha T\=amisra hell. There he suffers any amount of pain and suffering. He loses instantly his eyesight and his brain gets upset. His state resembles that of a tree when its trunk is broken. For this reason the ancient sages called this Andha T\=amisra. He who being subject to ``My'' and ``Mine'' quarrels with others and being very much attached, maintains his family, leaves his families here and goes with his bad tendencies to the Raurava hell, very horrible to all. The animals whom he injured and killed before in this world, assume the form of Ruru animal and torment him in the next. For this reason, the intelligent knowers of Pur\=a\d{n}a, call this Raurava. The ancients say, that Ruru is more cruel and ferocious than snakes. These animals live in that hell; hence it is termed Mah\=araurava. He who torments others, goes to this hell and these Rurus, the flesh-eaters, spring on his body and bite and eat his flesh. He who cooks other animals and birds, that cruel and ferocious, thus deluded, is cooked in return on the hot oil in the Kumbh\={\i}p\=aka hell by the Yama D\=utas for thousand years. He who quarrels with his Pitris and the Br\=ahma\d{n}as, is taken by the Yama D\=utas to the K\=alas\=utra hell and there be burnt by the fire and Sun. There that hellish person, being troubled very much, inside and outside by hunger and thirst, sometimes sits, sometimes sleeps, sometimes walks, and sometimes runs hither and thither. O Devar\d{s}i! He who transgresses the path of the Vedas in times other than those of calamity and danger and follows other paths

even to a trifling distance, that sinner is taken by Yamad\=utas to Asipatra K\=anana and there whipped severely. Not being able to bear that, he runs wildly to and fro and is pierced by the sharpedged Asi leaves on both his sides. His whole body being cut asunder, he cries ``Oh! I am killed!''and faints away. Then feeling himself pained very much, he tumbles down at every step. Thus the sinner suffers for violating the path of the Vedas. The King or that royal personage who gives orders of punishment, not approved by Dharma, and hurts or punishes the body of a Br\=ahma\d{n}a, the Yama's servants throw him down into the \'S\=ukaramukha Hell and grind down his body with great force as a sugarcane is ground down. He then cries aloud bitterly, he gets fainted and becomes stupefied. He is crushed by them and suffers all sorts of pains and miseries. Again he who knows the feelings of others when they are tormented, pains those insects who live on other's blood as bugs, etc., and who does not realise other's pains, goes, as a punishment for that fault to the Andhak\=upa Hell. There he is pained by the beasts, birds, deers, reptiles, mosquitoes, bugs, louses, flies and Danda\'s\=ukas and various other cruel animals. There he lives in his ugly body and roams like a beast. The man who, getting even a trifling wealth and food-stuff, does not perform the five Mah\=a Yaj\~n\=as and gives not a share of that to the Devas and feeds his own belly with that like a crow, is taken by the ferocious Yamad\=utas to the worst Krimibhojana Naraka for his sinful deeds. This hell is one lakh Yoyanas wide and is the reservoir of worms. It causes terror to the inhabitants of the hell. That sinner assumes the form of an insect and is eaten up on return by the insects and thus passes his time there. When a man does not give any share to the Atithis or the guests and does not offer oblations to the Fire and eats his food, he, too, goes to the above hell. When a man in times other than that of great danger and distress, follows the livelihood of a thief and robs gold and jewels forcibly of a Br\=ahmi\d{n} or any other person, he is taken to this hell and the Yama's servants pierce and cut off his skin by a fiery hot iron cutter. When a man goes for an illicit cohabitation to one who is not fit to be approached and when a woman goes similarly to an unapproachable man, both of them are taken, being whipped, to this hell by the servants of Yama! Where the man is obliged to embrace a fiery hot iron figure of woman and vice versa. When a man goes to all sorts of wombs for unnatural crimes, he is taken to Vajra Kantaka Naraka and placed on the top of an iron \'Salmal\={\i} wood. When a King or any royal personage, subject to the P\=a\d{s}anda Dharma (i.e., the unrighteous path) breaks the boundary of a law, he for that sin goes to Vaitara\d{n}\={\i}, the ditch round that hell. There the aquatic animals eat

his body all around. O N\=arada! Yet neither his life nor his body parts with him. He is thrown, for his Karma, into the rivers filled with faeces, urine, puss, blood, hairs, bones, nails, flesh, marrow, fat, etc., and he becomes very much troubled. Those who are the husbands of Vri\d{s}al\={\i}s (girls under twelve years of age, who have attained menstruation; or the barren women), void of any \'Saucha (cleanliness) or shame and without any \=Ach\=ara Vyavah\=ara (the following of one's natural customs and rites) and those who follow Pa\'sv\=ach\=ara (like beasts), they meet with a very hard fate and are thrown into this hell filled with faeces, urine, cough, blood and other impurities and when they feel hungry, are forced to eat the above things. When those persons that are twice born, maintain dogs and asses, etc., and when they are addicted to hunting, and kill daily, for nothing, beasts, birds and deer, those evildoers are specially watched by the servants of Yama and when they retire, they tear them asunder by shooting bows at them. He who kills animals, engaged vainly in a sacrifice and addicted to haughty tempers and habits, is thrown into this hell by the Yama's servants and whipped very severely. The twice-born that copulates blindly with a savar\d{n}\=a wife, is taken by the Yama's messengers into the hell filled with semen and he is made to drink that. Those who are addicted to robbery, who put fire to other's houses, who make others drink poison, those that are treacherous, and who destroy the interest of the villagers and other persons, those kings or the royal personages are taken after their death by the Yamad\=utas to the S\=aramey\=adana hell. There seven hundred and twenty dogs, very wonderful to behold, come furiously and with great force and energy, over them and feed on them. O N\=arada! This hell is denominated as S\=aramey\=adana Naraka and it is very horrible. Now I will describe to you the other hells Av\={\i}chi and others.

Here ends the Twenty-second Chapter of the Eighth Book on the narrative of the sins leading to hells in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, of 18,000 verses, by Mahar\d{s}i Veda Vy\=asa.



