\chapter{On the origin of the \'Sakti of the Earth}

1-4. \'Sr\={\i} N\=arada said :-- In the twinkling of an eye of the Dev\={\i}, the Pralaya takes place; and in that very time also the Brahm\=anda (cosmos) is dissolved, which is called the Pralaya of Prakriti. During this Pralaya, the Dev\={\i} Vasundhar\=a (Earth) disappears; the whole world is deluged with water and all this appearance of five elements called Prapa\~ncha vanishes in the body of Prakriti. Now where does Vasundhar\=a (Earth), thus vanished, reside? And how does She again appear at the beginning of the creation! What is the cause of her being so much blessed, honoured and capable to hold all and victorious. So tell about Her birth, the source of all welfare.

5-23. \'Sr\={\i} N\=ar\=aya\d{n}a said :-- O N\=arada! So it is heard that the Earth appears at the very outset of the creation. Her appearance and disappearance so occur in all the Pralayas. (This earth) the manifestation of the great \'Sakti, sometimes becomes manifest in Her and sometimes remains latent in Her (the \'Sakti). It is all the will of that Great \'Sakti. Now hear the anecdote of appearance (birth) of the earth, the cause of all good, the source of destruction of all calamities, the destroyer of sin and the cause of furtherance of one's religious merits. Some say that this earth has come out of the marrow of the Daityas, Madhu and Kaitabha; but that is not the fact. Hear now the real fact. Those two Daityas were greatly pleased with Vi\d{s}\d{n}u's valour and prowess in the fight between them and Vi\d{s}\d{n}u; and they said :-- ``Kill us on that part of the earth which is not under water.'' From their words it is evident that the earth was existent during their lifetime but she was not visible. After their death, the marrow came out after their bones. Now hear how the name ``Medin\={\i}'' came to be applied to the earth. She was taken out of the water, and the marrow came to be mixed with the earth. It is on account of this mixing that she is called Medin\={\i}, Now I will tell you what I heard before in Pu\d{s}kara, the sacred place of pilgrimage, from the mouth of Dharma Deva, about the origin of earth, approved by the \'Srutis, consistent, and good. Hear. When the mind of Mah\=a Vir\=at, merged in water, expanded all over his body, it entered into every pore of his body. Next the Mah\=aprithv\={\i} or the Great Earth appeared at the time of Pa\~nch\={\i} Kara\d{n}a (mixing of one-half of each of the elements with one-eighth of each of the other four elements). This Mah\=aprithv\={\i} was broken into pieces and placed in every pore. It is this differentiated earth that appears during creation and disappears during Pralaya. From this mind, concentrated in every pore of the body

of Mah\=a Vir\=at, is born this earth, after a long interval. In every pore in the skin of this Vir\=at Puru\d{s}a there is one earth. She gets manifested and she disappears. This occurs again and again. When she appears, she floats on the water; and when she disappears, she gets merged in the water. There is this earth (world) in every universe; and along with her, there are mountains, forests, seven oceans, seven islands, Sumeru mountain, the Moon, the Sun and other planets, Brahm\=aloka, Visnuloka (the abode of Vi\d{s}\d{n}u) \'Sivaloka and the regions of the other Devas, sacred places of pilgrimage, the holy land of Bh\=aratavar\d{s}a, the K\=anchan\={\i} Bh\=umi, seven heavens, seven P\=at\=alas or nether regions, on the above Brahm\=aloka, and Dhruvaloka. This law holds good in every world in every universe. So every universe is the work of M\=ay\=a and thus it is transient. At the dissolution of Prakriti, Brahm\=a falls. Again when creation takes place, the Mah\=a Vir\=at appears from \'Sr\={\i} Kri\d{s}\d{n}a, the Supreme Spirit. Eternal is this flow of creation, preservation and destruction; eternal is this flow of time, Kasth\=a; eternal is this flow of Brahm\=a, Vi\d{s}\d{n}u and Mahe\'sa, etc. And eternal is this flow of Vasundhar\=a who is worshipped in the V\=ar\=aha Kalpa by the Suras, Munis, Vipras, Gandarbhas, etc. The \'Sruti says that the Presiding Deity of this eternal earth is the wife of Vi\d{s}\d{n}u in His boar-form. Mangala (Mars) is the son of that earth and Mangala's son is Ghate\'sa.

24-26. N\=arada said :-- In what form was the Earth worshipped by the Devas in V\=ar\=aha Kalpa. The V\=ar\=ahi, the receptacle of all things, moving and non-moving, how did she appear, by what method of Pa\~nch\={\i} Kara\d{n}a, from the M\=ulaprakriti? What is the method of her worship in this Bh\=urloka and in the Heavens (Svarloka). Also tell me, O Lord! in detail, the auspicious birth of Mangala (Mars).

27-34. N\=ar\=aya\d{n}a spoke :-- In ancient days, in the V\=ar\=aha Kalpa, Var\=aha Deva (the boar incarnation) when entreated and praised by Brahm\=a, killed the Daitya Hira\d{n}y\=ak\d{s}a and rescued the earth from the nether regions Ras\=atala. He then placed the earth on the waters where she floated as the lotus leaf floats on water. In the meantime Brahm\=a began to fashion the wonderful creation on the surface of the earth. Bhagav\=an Hari, in His boar form and brilliant like ten million suns saw the beautiful and lovely appearance of the presiding deity of the earth, possessed of amorous sentiments. He then assumed a very beautiful form, fit for amorous embraces. They then held their sexual intercourse and it lasted day and night for one Deva year. The beautiful Earth, in the pleasant amorous plays, fainted away; for the intercourse of the lover with the beloved is exceedingly pleasant. And Vi\d{s}\d{n}u, too, at the same time was very much exhausted by the pleasant touch of the

body of the Earth. He did not become conscious even how days and nights passed away. When full one year passed away, they got back their senses and the amorous man then left his hold of the loved. He assumed easily his former Boar form and worshipped Her as the incarnate of the Dev\={\i}, with incense, lights, offerings of food, with vermilion (Sindur, red-powder), sandal-paste, garments, flowers and various other offerings of food, etc. He then said :--

35-37. O Auspicious One! Let Thou beest the receptacle of all things. All the Munis, Manus, Devas, Siddhas, and D\=anavas, etc., will worship Thee with pleasure and willingness. On the day the Ambuv\=achi ceremony closes, on the day when the house construction, i.e., the foundation is laid, on the day when the first entry is made into the newly built houses, when the digging of the well or tank commences, and on the day when tilling the ground commences, all will worship Thee. Those stupid fools that will not observe this, will certainly go to hell.

38-41 The Earth spoke :-- ``O Lord! By Thy command I will assume the form of V\=ar\=ah\={\i} (female boar) and support easily on my back this whole world of moving and non-moving things, but the following things, pearls, small shells, \'S\=alagr\=am, (a black stone, usually round, found in the river Gandak\={\i}, and worshipped as a type of Vi\d{s}\d{n}u), the phallus or emblem of \'Siva, the images of the goddesses, conch-shells, lamps (lights), the Yantras, gems, diamonds, the sacred upanayana threads, flowers, books, the Tulas\={\i} leaves, the bead (Japa m\=al\=a), the garland of flowers, gold, camphor, Gorochan\=a (bright yellow pigment prepared from the urine or bile of a cow), Sandal, and the water after washing the \'S\=alagr\=ama stone, I will not be able to bear. I will be very much pained in case I were to bear these on Me.''

42-45. \'Sr\={\i} Bhagav\=an said :-- ``O Fair One! The fools that will place the above articles on Thy back will go to the K\=alasutra hell for one hundred divine years.'' O N\=arada! Thus saying, the Bhagav\=an N\=ar\=aya\d{n}a remained silent. Now the Earth became pregnant and the powerful planet Mars was born. By the command of \'Sr\={\i} Hari, all began to meditate on Earth according to what is mentioned in K\=a\d{n}va\'s\=akh\=a and began to praise Her. Offerings of food were given, uttering the root Mantra. Thus became extant all over the three worlds Her worship and praise.

46. N\=arada said :-- O Bhagav\=an! Very sacred is the meditation, hymn and the root Mantra of the Earth. I am very anxious to hear them. Kindly describe it in detail.

47-48. N\=ar\=aya\d{n}a said :-- The Earth was first worshipped by Var\=aha Deva; next She was worshipped by Brahm\=a. Next She was worshipped by all the Munis, Devas, Manus and men. O N\=arada! Now hear the Dhy\=an, praise and Mantra of the Dev\={\i} Earth.

49-51. The Earth was first worshipped by Bhagav\=an Vi\d{s}\d{n}u with this root Mantra (m\=ul mantra) ``On Hr\={\i}m \'Sr\={\i}m Kl\={\i}m Vasundhar\=ayai Sv\=ah\=a.'' Next He said :-- O Dev\={\i} Earth! O Thou Smiling One! I worship Thee, who art worshipped by the three worlds, whose colour is white like white lotus, whose face is beautiful like the autumnal moon, who art the Store-house of all gems and jewels, and in whose womb all the precious stones and pearls are inbedded, and who has put on a raiment purified by fire. All then began to worship Her with this Mantra.

52-63. \'Sr\={\i} N\=ar\=aya\d{n}a said :-- Now hear the hymn sung before Her according to K\=a\d{n}va \'S\=akh\=a :-- O Thou, the Giver of Victory! Holder of water! Endowed with water, full of victory; Consort of the Boar Incarnation, Carrier of victory! Bestow victory on me. O Thou Auspicious One! The Store-house of all good, O Thou incarnate of all auspiciousness! Bestower of good, Thou, the Source of all good to bestow all sorts of welfare! Bestow all things that are good and auspicious to me in this world.

O Thou! The Receptacle of all, the Knower of all, all powerful, the Bestower of all desires, O Dev\={\i} Earth! Give me the fruits that I desire.

O Thou! Who art all merits, Thou, the Seed of all religious merits, O Thou, the Eternal, the receptacle of all religious merits, the home of all religious persons, Thou bestowest merits to all.

O Thou! The Store-house of all grains, enriched with all sorts of corns, Thou bestowest harvests to all; Thou takest away all the grains in this world and again, Thou producest all corns of various kinds here. O Earth! Thou art all-in-all to the landlords, the Best Source of refuge and happiness. O Bestower of lands! Give me lands. The above hymn yields great religious merits. He becomes the sovereign of the whole earth for millions and millions of births who rising early in the morning reads this stotra. Men who read this acquire merits due to giving away lands as gifts. People become certainly freed of their sins, if they read this stotra, who take back the lands after making them as gifts, who dig earth on the day of Ambuv\=ach\={\i}, who dig wells without permission on another's well, who steal other's lands, who throw their semen on earth, who place lamps on the earth. Religious merits, equivalent to one hundred horse sacrifices accrue from reading this stotra. There is no doubt in this. This stotra of the great Dev\={\i} is the source of all sorts of welfare and auspiciousness.

Here ends the Ninth Chapter of the Ninth Book on the origin of the \'Sakti of the earth in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



