\chapter{On the greatness of Kali}

1-110. N\=ar\=aya\d{n}a said :-- A part of Sarasvat\={\i} descended in this Bh\=arata Pu\d{n}ya Bh\=umi (land of merits), owing to the curse of Gang\=a; and She remained in full in Vi\d{s}\d{n}u's region, the abode of Vaikuntha. She is named Bh\=arat\={\i}, on account of Her coming to Bh\=arata; she is called Br\=ahm\={\i} because she is dear to Brahm\=a; and She is called V\=an\={\i} as She presides over Speech. Hari is seen everywhere, in tanks, in wells, in running streams (i.e., in Saras). Because He resides in Saras, therefore He is called Sarasv\=an. V\=an\={\i} is the \'Sakti of that Sarasv\=an; therefore She is denominated Sarasvat\={\i}. The river Sarasvat\={\i} is a very sacred T\={\i}rtha. She is the burning fire to the fuel of sins, of sinners. O N\=arada! Through the curse of Sarasvat\={\i}, the Dev\={\i} Gang\=a also assumed the form of a river in part. She was brought down to this earth at the request of Bhag\={\i}ratha. Hence she is called Bh\=agirath\={\i}. While Gang\=a was rushing down to the earth \'Siva capable to bear the great rush of Her, held Her on His head at the request of the Mother Earth. Lak\d{s}m\={\i} also, through the curse of Sarasvat\={\i} came in part of parts Bh\=arata as the river Padm\=avat\={\i}. But She remained in full with Hari. Lak\d{s}m\={\i} appeared also in Her other part as the well-known daughter Tulas\={\i} of the king Dharmadhvaja in India. Last of all, through Bh\=arat\={\i}'s curse and by the command of \'Sr\={\i} Hari, she turned into the Tulas\={\i} tree, purifying the whole world. Remaining for five thousand years of Kali, all them will quit their river appearances and go back to Hari. By the

command of \'Sr\={\i} Hari, all the T\={\i}rthas save K\=as\={\i} and Bindr\=aban will go along with them to Vaikuntha. Next at the expiry of the ten thousand years of Kali, \'S\=alagr\=ama \'Sil\=a (the stone piece worshipped as N\=ar\=aya\d{n}a) \'Siva, and \'Siva \'Sakti and Puru\d{s}ottama Jagann\=atha will leave the soil of Bh\=arata and go to their respective places, (i.e. the M\=ah\=atmyas of those will be extinct from Bh\=arata). There will then cease to be the saints of \'Siva \'S\=akta, G\=a\d{n}apatya and Vai\d{s}\d{n}ava sects, (eighteen) Pur\=a\d{n}as, the blowing of conch shells (auspicious signs), \'Sr\=addhas, Tarpa\d{n}as, and all the rites and ceremonies dictated by the Vedas. The worship and glorification of the Gods, the recitation of their praises, their names will be extinct. The Vedas with their Amgas will no longer be heard of. All these will disappear with them. The assembly of the S\=adhus, the true Dharma, the four Vedas, the village Devas and Dev\={\i}s, the Vratas (vows) the practising of the austerities, fasting, all will disappear. All will be addicted to the V\=am\=ach\=ara ritual (the left-hand ritual T\=antrik form of worship; sarcastically used in the sense of drinking wine and eating flesh, etc.) They will speak falsehood and be deceitful. If anybody worships, his worship will be void of Tulas\={\i} leaves. Almost all will be deceitful, cruel, vain, egoistic, thievish and mischievous. Men will be at variance with one another; women will be at variance with one another; no fear will exist in marriage ties. Properties will be only of those that will make them (i. e. there will cease to be any inheritance from father to son and so forth ). Husbands will be obedient to their wives; unchaste women will be in every house. Wives will rebuke their husbands by incessant noisings and chidings. Wives will be the sole mistresses of houses and husbands will stand before them as servants with folded palms. Fathers-in-law and mothers-in-law will be their servants. The brothers of wives, and their friends will be the managers of the household affairs. But there will be no friendship with one's own class fellows. The brothers and friends of the house owners (masters of the house) will appear quite strangers as if they are new-comers. Without the command of the house-wives, the masters of the houses will be unable to do anything. The divisions of caste (Br\=ahma\d{n}\=a, K\d{s}attriya, Vai\d{s}ya, and \'S\=udra) will entirely disappear. Far from practising Sandhy\=a Bandanam and other daily practices, the Br\=ahma\d{n}as will cease to hold the holy threads even on their bodies. The four colour-classes will practise the doings of the Mlechchas, read the \'S\=astras of the Mlechchas and forsake their own \'S\=astras. The Br\=ahmi\d{n}s, K\d{s}attriyas, and Vai\'syas will become the slaves of \'S\=udras, will become their cooks, runners and carriers of buffaloes. Every one will be devoid of truth. Earth will not yield any grains;

trees will not yield any fruits and women will be issueless. The cows will not yield milk; even if there be a little milk, ghee will not come out of it. The affection between husband and wife will die out and the families will be devoid of truth. The King will not wield any power; the subjects will be over burdened with taxes. The ever flowing big rivers, the petty streams, the caves of mountains all will gradually have very little water in them. The Four Var\d{n}as will be devoid of Dharma and Pu\d{n}ya (merit, virtue). One in a lakh may be virtuous. Afterwards that too will cease. Men, women, boys, all will be ugly and deformed. They will utter bad words and vile sounds. Some villages and towns will be completely deserted by men and will look terrible; at some others few cottages with few inhabitants will be seen. Villages and towns will be jungles and jungles will become filled with men. The inhabitants of the forests will become heavily taxed and disconsolate. The beds of rivers and lakes will become dry owing to want of rains and will be cultivated. The Kul\={\i}nas of high families will become very low. The whole earth will be filled with liars, untruthful cheats and hypocrites. The lands, though cultivated well, will yield grains in name. Those who are well known as the millionaires, they will become poor and those who are devoted to the Devas will be atheists. The towns folk will have no trace of mercy; rather they will hate and envy their neighbours and turn out murderers of men. In the Kali age, males and females will be everywhere, of a dwarfish stature, diseased, shortlived, and of very little youthful virility. The hairs will turn out grey no sooner the people reach their sixteenth year. And they will be very old when they become twenty years old. The girls of eight years will have menstruation and will become pregnant. They will deliver every year. Old age will attack them when they become sixteen years old. Some women will have their husbands and children living. Otherwise almost all will be barren, childless. The four Var\d{n}as will sell their daughters. The paramours of the mothers, wives, son's wives, daughters, and sisters will be the source of support to them all. No one will be able, without money, to collect the merits by repeating the name of Hari. Persons will make gifts for name and fame and ultimately will take back what they had made as gifts. If there be any gifts made by one's own self or by one's forefathers or for a Deva purpose or for Br\=ahmi\d{n}s or for the families of the Gurus, there will not be found wanting attempts to take back those gifts. Some will go to daughters, some to mothers-in-law, some to the wives of sons, some to sisters, some to mothers of co-wives, some will go to the brother's wives. In every house, those who are unfit to be mixed will be mixed with, excepting one's mother. In Kali Yuga

who is whose wife? And who is whose husband? There will be no certainty; who is whose subject and what village is to whom? There will be no surety that such a property belongs to such and such a man. All will turn out to be liars, licentious, thieves, envious of other's wives, and murderers of men. In the houses of the Br\=ahmi\d{n}s, K\d{s}attriyas, and Vai\'syas, the three higher castes, the current of sin will flow. They will live by selling lac, iron, and salts prohibited by the \'S\=astras. The Br\=ahmi\d{n}s will drive buffaloes, burn the dead bodies of the \'S\=udras, eat the food of the \'S\=udras and go to unchaste women. There will be no more faith existing in the five \d{R}i\d{s}i Yaj\~nas. Almost every Br\=ahmi\d{n} will not observe the vows of Am\=avasy\=a Ni\'sip\=alana. The holy threads will be cast away and the Sandhy\=a Bandanam and cleanliness and good practices will cease altogether. The unchaste women who deal in giving loans, etc., and live on interests and the procuresses during menstruation will cook in Br\=ahmi\d{n} families. There will be no distinction of food, no distinction of wombs, no distinction of \=A\'sramas, and no distinction of persons. All will turn out Mlechchas. O N\=arada! Thus, when the Kali will have its full play, the whole world will be filled with Mlechchas, the trees will be one hand high and the men will be of the size of a thumb. Then the most powerful Bhagav\=an N\=ar\=aya\d{n}a will incarnate in His part in the house of a Br\=ahmi\d{n} named Vi\d{s}\d{n}uja\'s\=a as his son. Mounted on a long horse, holding a long sword He will make the world free of the Mlechchas in three nights. Then he will disappear from the face of the Earth and She will be without any sovereign and be filled with robbers. There will be incessant rain, for six nights and it will rain and rain and the whole earth will be deluged; no traces of men, houses, and trees. After this the Twelve Suns will rise simultaneously and by their rays the whole water will be dried up and the earth will become level. Thus the dreadful Kali will pass away when the Satya Yuga will come back, Tapasy\=a and the true religion and Sattva Gu\d{n}a will prevail again. The Br\=ahmi\d{n}s will practise Tapasy\=a, they will be devoted to Dharma and the Vedas. The women will be chaste and religious in every house. Again the wise and intelligent K\d{s}attriyas devoted to the Br\=ahma\d{n}as will occupy the royal thrones and their might, devotion to Dharma and love for good deeds will increase. The Vai\'syas will again go on with their trades and their devotion to their trade and the Br\=ahmi\d{n}s will be reestablished. The \'S\=udras, too, will be again virtuous, and serve the Br\=ahmi\d{n}s. Again the Br\=ahmi\d{n}s, K\d{s}attriyas, and Vai\'syas and their families will have Bhakti towards the Dev\={\i}, be initiated in Dev\={\i} Mantras and all will meditate on the Dev\={\i}. Again there will be spread the knowledge of the Vedas, the Smritis,

and the Pur\=a\d{n}as, all will go to their wives in menstruation periods. No Adharma (unrighteousness) will exist and the Dharma will reign in full, with all the parts (Kal\=as) complete. When the Tret\=a Yuga comes, the Dharma will be three footed; when the Dv\=apara Yuga will come; the Dharma will be two-footed and when Kali will begin, the Dharma will be one-footed, and when Kali will reign supreme, no Dharma will exist, even in name. (O N\=arada! Now I will speak of time.) The seven days of the week, Sunday, \&c., the sixteen tithis, Pratipada \&c., the twelve months Vai\'s\=akha \&c., the six seasons Summer, \&c., the two fortnights (dark and bright) and the two Aya\d{n}as (Northern and Southern) are rendered in vogue. One day consists of four Praharas, one night consists of four Praharas; a day and a night constitute one so-called day. Thirty such days make one month. In the computation of time, five kinds of years (Var\d{s}as) were already mentioned (in the 8th Skandha). As the Satya, Tret\=a, Dv\=apara and Kali roll on turn by turn, so the days, months and years also roll on in turn. One day, according to the Devas, is equal to one year, according to men; three hundred and sixty human Yugas equal to one Deva Yuga. Seventy-one Deva Yugas make one Manvantara. The life period of Indra, the Lord of \'Sach\={\i}, is one Manvantara. Twenty-eight Indra\'s lives equal to one day of Hira\d{n}yagarbha (the golden wombed) Brahm\=a. One hundred and eight such years equal to the life of one Brahm\=a. When this Brahm\=a dies, there is the Pr\=akrita Pralaya. The earth is not visible then. (The dissolution of Prakriti takes place.) The whole Br\=ahmanda is deluged by water; Brahm\=a, Vi\d{s}\d{n}u, Mahe\'svara and the other wise \d{R}i\d{s}is get diluted in Para Brahm\=a whose substance is all truth and consciousness. That time, the Prakriti Dev\={\i}, too, gets merged in Para Brahm\=a. The fall of Brahm\=a and the dissolution of Prakriti are called the Pr\=akrita Pralaya. The duration of this Pralaya is one Nime\d{s}a of the Para Brahm\=a M\=ula Prakriti united with M\=ay\=a. All the Br\=ahmandas (universes) are destroyed at this time. When this Nime\d{s}a expires, the creation begins again in due order. So one cannot count the endless numbers of times when this creation and dissolution work are going on. So who can tell how many kalpas had past away, or how many Kalpas will come, how many Br\=ahmandas were created or how many Br\=ahmandas will be created. Who will be able to count how many Brahm\=as, how many Vi\d{s}\d{n}us or how many Mahe\'svaras there have been. But One and Only One Para Brahm\=a Parame\'svara (the Great God) is The Supreme Lord of these countless Br\=ahmandas. The Parame\'svara of the nature of Existence, Consciousness and Bliss is the Highest Spirit of all. All others, Brahm\=a, Vi\d{s}\d{n}u, Mahe\'svara the

Great Vir\=at, the Smaller Vir\=at, all are His parts. This Brahm\=a is M\=ul\=a Prakriti and from That has appeared \'Sr\={\i} Kri\d{s}\d{n}a, the Lord of his left half which is woman (Ardha N\=ar\={\i}\'svara). It is She that divided Herself into two forms; in Her one form, She resides as the two armed Kri\d{s}\d{n}a in the region of Goloka; and as the four-armed N\=ar\=aya\d{n}a in Vaikuntha. All the things from Brahm\=a, the Highest, to the mere grass the lowest, all are originated from Prakriti. And all the Prakriti-born things are transient. Thus the True, Eternal Para Brahm\=a, beyond the three gu\d{n}as, the Source of all creation, Whose substance is All-Will is the Only Substance beyond the region of Prakriti. He is without Up\=adhis (conditions, as time, space, causation and attributes); He is without any form; and the forms that He assumes, they are for shewing His Grace to the devotees only. The Lotus-born Brahm\=a is able by His Power of Knowledge to create the Br\=ahmanda. It is by His Grace that \'Siva, the Lord of the yogis is named Mrityumjaya (the Conqueror of Death), the Destoyer of all, and the Knower of all Tattvas. By His Tapas, \'Siva has realised Para Brahm\=a and therefore has become the Lord of all, All-knowing, endowed with great Vibh\=utis (lordly powers), the seer of all, omnipresent, the protector of all, the bestower of all prosperities. The devotion and service towards Para Brahm\=a have alone made \'Sr\={\i} Vi\d{s}\d{n}u the Lord of all; and it is through the power of Para Brahm\=a, that Mah\=am\=ay\=a Prakriti Dev\={\i} has become omnipotent and the Goddess of all. Bhagavat\={\i} Durg\=a has got His Grace by Her devotion and service to Him and has become M\=ula Prakriti of the nature of Being, Consciousness and Bliss. And so has the Dev\={\i} S\=avitr\={\i}, the Mother of the Vedas, become the presiding Deity of the Vedas and She is worshipped by the Br\=ahma\d{n}as and the Knowers of the Vedas. That She presides over all the branches of knowledge, is worshipped by all the learned assemblies and by the whole Universe is the result only of worshipping the Prakriti Dev\={\i}. That Lak\d{s}m\={\i} has become the bestower of all wealth and the presiding Deity of all the villages and the mistress of all, worshipped by all and the bestower of sons to all is also the result of worshipping Her. Thus it is through the worship of Prakriti that Durg\=a, the Destroyer of all calamities and troubles has appeared from the left side of \'Sr\={\i} Kri\d{s}\d{n}a; and R\=adh\=a has become the presiding Deity of His Pr\=a\d{n}a (vital airs), and She is worshipped by all and possessed of all knowledge. It is by the worship of \'Sakti that R\=adhik\=a has so much excelled in love, has become the presiding Deity of the pr\=a\d{n}a of Kri\d{s}\d{n}a, has got love and respect, has been placed on His breast and is exceedingly beautiful. With the object of getting Kri\d{s}\d{n}a for her husband, She

practised severe austerities for one thousand Deva years on the mountain \'Sata\'sringa in Bh\=arata to get the M\=ula Prakriti's Grace. And when the \'Sakti M\=ula Prakriti became graciously pleased towards Her, \'Sr\={\i} Kri\d{s}\d{n}a seeing R\=adhik\=a increasing in beauty like the Crescent Moon took Her to His breast and out of tenderness wept and granted Her highest boons so very rare to others and said :-- O Beautiful One! You better remain always in My breast and devoted to Me amongs all my wives; let you be superior to them all in good fortune, respect, love and glory. From today you are my greatest best wife. I will love you as the best amongst them all. O Dear! Always I will be submissive to you and fulfil what you say. Thus saying, \'Sr\={\i} Kri\d{s}\d{n}a selected her as his wife without any co-wives and made Her dear to His Heart. The other Dev\={\i}s besides the five Prakritis, already mentioned, also derived superiorities by serving M\=ula Prakriti. O Muni! What shall I say, everyone reaps the fruits as he practises Tapasy\=a. Bhagavati Durg\=a practised on the Him\=aly\=as tapasy\=a for one thousand Deva years and meditated on the lotus-feet of M\=ula Prakriti and so has come to be worshipped by all. The Dev\={\i} Sarasvat\={\i} practised Tapasy\=a for one lakh Deva years and is come to be respected by all. The Dev\={\i} Lak\d{s}m\={\i} practised Tapasy\=a at Pu\d{s}kara for one hundred Divine Yugas and, by the Grace of M\=ula Prakriti has become the bestower of wealth to all. The Dev\={\i} Savitr\={\i} worshipped \'Sakti for sixty thousand divine years in the Malaya mountain and is respected and worshipped by all. O Bibhu! Brahm\=a, Vi\d{s}\d{n}u, and Mahe\'svara worshipped \'Sakti for one hundred Manvantaras and so have become the Preservers, etc., of this world. \'Sr\={\i} Kri\d{s}\d{n}a practised for ten Manvantaras terrible austerities and therefore obtained his position in the region of Goloka and is remaining there today in greatest bliss. Dharma Deva worshipped \'Sakti with devotion for ten Manvantaras and has become the lives of all, worshipped by all, and the receptacle of all. O Muni! Thus all, whether the Dev\={\i}s, Devas, Munis, Kings, Br\=ahma\d{n}as, all have got their respect in this world by the worship of \'Sakti. O Devar\d{s}i! I have thus described to you all that I heard from the mouth of my Guru, in accordance with the rules of the Vedas. What more do you want to hear?

Here ends the Eighth Chapter of the Ninth Book on the Greatness of Kali in the Mah\=a Pur\=a\d{n}am Sr\={\i}mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



