\chapter{On the hymn, worship and Kavacha of Sarasvat\={\i} Dev\={\i}}

1-3. N\=arada said :-- By your Grace I have heard everything very sweet like nectar, of the origin of things. Now may I ask you which Dev\={\i} of these five Prakritis has been worshipped by what Mantra? and by whom? How has each of them been praised? and by whom? How has the worship of their M\=urtis (form) become prevalent in this world? What are the Stotram (hymn of praise), the Dhy\=an (meditation) glory and life of these? Also what sort of boon do each of the Dev\={\i}s grant? and to whom? Kindly describe all these in detail.

4-29. N\=ar\=aya\d{n}a said :-- O Child! Durg\=a, the mother of Ga\d{n}e\'sa, R\=adh\=a, Lak\d{s}m\={\i}, Sarasvat\={\i} and S\=avitr\={\i}, these are the five Prakritis sprung directly from M\=ula Prakriti. The methods of their worship, wonderful glorious acts, excellent stotrams, and their lives, inculcating good to all, and sweet like nectar are all widely written in the Vedas, Pur\=a\d{n}as, Tantras and other \'S\=astras. So there is no need to describe them here again. Now I am describing in detail the auspicious character of these that are sprung from the parts and Kal\=as of the Prakriti. Hear attentively. K\=al\={\i}, Vasundhar\=a, Gang\=a, Sasth\={\i}, Mangal Chandik\=a, Tulas\={\i}, Manas\=a, Nidr\=a, Svadh\=a, Sv\=ah\=a, and Dak\d{s}i\d{n}\=a, these are the parts of Prakriti. By and by I will describe, briefly, the merit-giving characters and pleasant to hear. Along with these I will describe the Karmas of the J\={\i}vas, and the great exalted lives of Durg\=a and R\=adh\=a. I am now describing Sarasvat\={\i}'s character. Hear, O Muni! \'Sr\={\i} Kri\d{s}\d{n}a introduced first in this Bh\=arata, the worship of the Dev\={\i} Sarasvat\={\i}, the holder of V\={\i}\d{n}\=a in Her hands, under whose influence, the hearts of illiterate stupid persons become illumined with knowledge. The amorous Dev\={\i} Sarasvat\={\i} sprang from the end of the lips of R\=adh\=a and so she desired to marry Kri\d{s}\d{n}a out of amorous feelings. \'Sri Kri\d{s}\d{n}a, the controller of the hearts of all, knew it instantly and addressed the Mother of the people in true words proper to Her and beneficial to Her in the end. O Chaste One! The four-armed N\=ar\=aya\d{n}a is born from My parts; He is young, of good

features and endowed with all qualifications; so much so, he is like Me. He is a Knower of amorous sentiments of women and He fulfils those desires; what to speak of His beauty, ten millions of the God of love are playing in His body. O Beloved! And if you desire to marry and remain with Me, that will not be of any good to you. For R\=adh\=a is near to Me; She is more powerful than you. If a man be stronger than another, he can rescue one who takes his shelter; but if he be weaker, how can he then, himself weak, protect his dependant from others. Though I am the Lord of all, and rule all, yet I cannot control R\=adh\=a. For She is equal to me in power, in beauty, in qualifications, equal to Me in every respect. Again it is impossible for Me to quit R\=adh\=a for She is the presiding Deity of My life. Who can relinquish life? Though a son is very dear to his father, still it may be questioned, is he dearer than his father's life? So, O Auspicious One! Go to the abode Vaikuntha; you will get your desires fulfilled there. You will get for your husband the Lord of Vaikuntha and you will live ever in peace and enjoy happiness. Though Lak\d{s}m\={\i} is residing there yet like you she is not under the control of lust, anger, greed, delusion and vanity. She is also equal to you in beauty, qualities, and power. So you will live with her in great delight and Hari, the Lord of Vaikuntha, will treat both of you equally. Moreover, I say this in particular that in every universe, on the fifth day of the bright fortnight of the month of M\=agha, every year, the day when the learning is commenced, a great festival will be held and men, Manus, Devas, and the Munis desirous of liberation, Vasus, Yogis, N\=agas, Siddhas, Gandarbhas, R\=ak\d{s}asas, all will perform your worship with devotion in every Kalpa till the time of Mah\=a Pralaya comes. All are required to be Jitendriya (having their senses under control) and Samyam\={\i} (concentrating his mind, and with a religious vow) and they will invoke Thee on a jar or on books and then meditate according to what is stated in the Ka\d{n}va \'S\=akh\=a of Yajurveda and then worship and sing hymns to Thee. Thy Kavacha (an armour; a mystical syllable ** considered as a preservative like armour) is written on the bark of the Bh\=urja tree and then with eight kinds of scents mingled with it is placed within a golden nut or ring named M\=aduli and then held on the neck or on the right arm. The learned should recite Thy Stotras during worship. Thus saying, the Pura\d{n} Brahm\=a \'Sr\={\i} Kri\d{s}\d{n}a Himself worshipped the Dev\={\i} Sarasvat\={\i}. Since then, Brahm\=a, Vi\d{s}\d{n}u, Mahe\'sa, Ananta Deva, Dharma, Sanaka and other Mun\={\i}ndras, all the Devas, Munis, all the kings and all the human beings are worshipping the Dev\={\i} Sarasvat\={\i}. O N\=arada! Thus the worship of the Eternal Dev\={\i} is made extant in the three worlds.

30-31. N\=arada said :-- O Chief of the Knowers of the Vedas! Now describe to me the methods of worship, Dhy\=an, Kavacham, hymns, the

appropriate offerings of the P\=uj\=a flowers, sandalpaste and other good things necessary in these worships and which are so sweet to hear. I am ever very eager and anxious to hear these.

32-59. N\=ar\=aya\d{n}a said :-- O N\=arada! I am now stating the method of worship of the Dev\={\i} Sarasvat\={\i}, the Mother of the Worlds, according to Ka\d{n}va \'S\=akh\=a of the Yajurveda. Hear. On the day previous to the fifth day of the bright fortnight of the month of M\=agh or the day of commencement of education, the devotee should control his senses, concentrate his mind and take his bath. Then he is to perform his daily duties and install the jar (Ghata) with devotion and according to the Mantras of the Ka\d{n}va \'S\=akh\=a or the Tantra, as the case may be. He is to worship first on that Ghata (jar) Ga\d{n}apati (Gane\'sa), then meditate the Dev\={\i} Sarasvat\={\i} as described below, invoke Her and again read the Dhy\=an and then worship with Soda\'sopach\=ara (sixteen good articles offered in the worship). O Good One! Now I am speaking, according to my knowledge, about the offerings as ordained in the Vedas or Tantras. Hear. Fresh butter, curd, thickened milk, rice freed from the husk by frying, sweetmeats (Til Laddu) prepared of Til, sugar cane, sugarcane juice, nice Gud (molasses), honey, svastik, sugar, rice (not broken) out of white Dh\=an, chipitak of table rice (\=Alo ch\=al), white Modak, Harbi\d{s}y\=anna prepared of boiled rice with clarified butter and salt, Pistaka of jao\=a or wheaten flour, Param\=anna with ghee, nectar like sweetmeats, cocoanut, cocoanut water, Svastik Pistaka, Svastik and ripe plantain Pistaka, Kaseru (root), M\=ul\=a, ginger, ripe plantains, excellent Bel fruit, the jujube fruit, and other appropriate white purified fruits of the season and peculiar to the place are to be offered in the Pooj\=a. O N\=arada! White flowers of good scent, white sandalpaste of good scent, new white clothes, nice conchshell, nice garlands of white flowers, nice white necklaces, and beautiful ornament are to be given to the Dev\={\i}. I say now the Dhy\=anam sweet to hear, of the Dev\={\i} Sarasvat\={\i} according to the Vedas, capable to remove errors! Hear. I hereby bow down to the Dev\={\i} Sarasvat\={\i}, of a white colour, of a smiling countenance and exceedingly beautiful, the lustre of whose body overpowers that of the ten millions of Moons, whose garment is purified by fire, in whose hands there are V\={\i}\d{n}\=a and books, who is decorated with new excellent ornaments of jewels and pearls and whom Brahm\=a, Vi\d{s}\d{n}u, Mahe\'svara and the other Devas, Munis, Manus and men constantly worship. Thus meditating the Dev\={\i}, the intelligent persons should offer all articles, after pronouncing the root Mantra. Then he is to hymn and hold Kavacha and make S\=ast\=a\d{n}ga pra\d{n}\=ams before the Dev\={\i}. O Muni! Those whose Dev\={\i} Sarasvat\={\i} is the presiding Deity, are not to be spoken of at all (i.e., they

will naturally do all these things and with a greater fervour). Besides all should worship the Dev\={\i} Sarasvat\={\i} on the day of commencement of education and every year on the \'S\=ukla Pancham\={\i} day of the month of M\=agh. The eight-lettered Mantra, as mentioned in the Vedas is the root Mantra of Sarasvat\={\i}. (Aim Kl\={\i}m Sarasvatyai namah). Or the Mantra to which each worshipper is initiated is his M\=ulmantra (not Mantra). Or uttering the Mantra ``\'Sr\={\i}m Hr\={\i}m Sarasvatyai Sv\=ah\=a'' one is to offer everything to the Dev\={\i} Sarasvat\={\i}. This Mantra is the Kalpa Vrik\d{s}a (i. e., the tree which yields all desires). N\=ar\=aya\d{n}a, the ocean of mercy, gave in ancient times, this very Mantra to V\=almik\={\i} in the holy land Bh\=arata Var\d{s}a on the banks of the Ganges; next Bhr\={\i}gu gave this Mantra on the occasion of solar eclipse to Mahar\d{s}i Sukr\=acharya on the Pu\d{s}kara T\={\i}rtha; M\=ar\={\i}cha gave to Brihaspati on a lunar eclipse; Brahm\=a gave to Bhr\={\i}gu in the Vadarik\=a \=A\'srama; Jaratkar\=a gave to \=Astika on the shore of the K\d{s}iroda ocean; Bibh\=andaka gave this to the intelligent Ri\d{s}yasri\d{n}ga on the Sumeru mountain, \'Siva gave this to Ka\d{n}\=ada and Gotama, S\=urya gave to Y\=aj\~navalkya and K\=aty\=ayana, Ananta Deva gave to P\=a\d{n}ini, to the intelligent Bh\=aradv\=aja and to \'S\=akat\=ayana in Bali's assembly in the P\=at\=ala. If this Mantra be repeated four lakhs of times, all men attain success. And when they become Siddhas with this Mantra, they become powerful like Brihaspati. In past times, the Creator Brahm\=a gave a Kavacha named Vi\'svajaya to Bhr\={\i}gu on the Gandham\=adana Mountain. I now speak of that. Hear.

60-61. Once on a time Bhr\={\i}gu asked Brahm\=a the Lord of all, and adored by all, thus :-- ``O Brahm\=an! Thou art the foremost of those that know the Vedas; there is none equal to thee in matters regarding the knowledge of the Vedas (so much so that there is nothing that is not known to thee; for all these have sprung from thee). Now say about the Vi\'svajaya Kavacha of the Dev\={\i} Sarasvat\={\i}, that is excellent, without any faults and embodying in it all the properties of all the Mantras.''

62-91. Brahm\=a said :-- ``O Child! What you have asked about the Kavacha of Sarasvat\={\i} that is sweet to hear, ordained and worshipped by the Vedas, and the giver of all desired fruits, now hear of that. In the very beginning, the all-pervading \'Sr\={\i} Kri\d{s}\d{n}a, the Lord of the R\=asa circle, mentioned this Kavacha to me in the holy Brind\=avana forest in the abode Goloka at the time of R\=asa in R\=asa Mandala. This is very secret; it is full of holy unheard, wonderful Mantras. Reading this Kavacha and holding it (on one's arm) Brihaspati has become foremost in matters of intelligence; by the force of this Kavacha \'Sukr\=ach\=arya

has got his ascendancy over the Daityas; the foremost Muni V\=almik\={\i} has become eloquent and skilled in language and has become Kav\={\i}ndra and Sv\=ayambhuva Manu; holding this Kavacha he has become honoured everywhere. Ka\d{n}\=ada, Gotama, Ka\d{n}va, Pa\d{n}ini, \'Sakat\=ayana, Dak\d{s}a, and K\=aty\=ayana all have become great authors by virtue of this Kavacha; Kri\d{s}\d{n}a Dvaip\=ayana Veda Vy\=asa made the classification of the Vedas and composed the eighteen Pur\=a\d{n}as. \'S\=at\=atapa, Samvarta, Va\'sistha, Par\=asara and Y\=aj\~navalkya had become authors by holding and reading this Sarasvat\={\i} Kavacha. Ri\d{s}ya\'sri\d{n}ga, Bh\=aradv\=aja, \=Astika, Devala, Jaigi\d{s}avya, and Y\=ay\=ati all were honoured everywhere by virtue of this Kavacha. O Dvija! The Praj\=apati Himself is the \d{R}i\d{s}i of this Kavacha; Brihat\={\i} is its Chhanda; and \'S\=arad\=a Ambik\=a is its presiding Deity. Its application (Viniyoga) is in the acquisition of spiritual knowlege, in the fruition of any desires or necessities, in composing poems or anywhere wheresoever success is required. May \'Sr\={\i}m Hr\={\i}m Sarasvatyai Sv\=ah\=a protect fully my head; \'Sr\={\i}m V\=agdevat\=ayai Sv\=ah\=a, my forehead; Om Hr\={\i}m Sarasvatyai Sv\=ah\=a, my ears always; Om \'Sr\={\i}m Hr\={\i}m Bhagabatyai Sarasvatyai Sv\=ah\=a always my eyes; Aim Hr\={\i}m V\=agv\=adinyai Sv\=ah\=a, always my nose; Om Hr\={\i}m Vidyadhisth\=atr\={\i} Devyai Sv\=ah\=a, my lips always; Om \'Sr\={\i}m Hr\={\i}m Brahmyai Sv\=ah\=a my rows of teeth; Aim, this single letter protect my neck; Om \'Sr\={\i}m Hr\={\i}m my throat; \'Sr\={\i}m, my shoulders, Om Hr\={\i}m Vidy\=adhisth\=atr\={\i} Devyai Sv\=ah\=a, always my chest; Om Hr\={\i}m Vidyadhisvar\=up\=ayai Sv\=ah\=a my navel; Om Hr\={\i}m Kl\={\i}m V\=a\d{n}yai Sv\=ah\=a my hands; Om Svarva v\=arnatm\={\i} K\=ayai Sv\=ah\=a my feet; and let Om V\=agadhisth\=atridevyai Sv\=ah\=a protect all my body. Let Om Sarvakanthav\=asinyai Sv\=ah\=a protect my east; Let Om Svarvajibh\=agra v\=asinyai Sv\=ah\=a, the South-east; Om Aim Hr\={\i}m \'Sr\={\i}m Kl\={\i}m Sarasvatyai budhajananyai Sv\=ah\=a, my South; Aim Hr\={\i}m \'Sr\={\i}m, this three lettered Mantra my South-west; Om Aim Jhibhagrav\=asinyai Sv\=ah\=a, my West; Om Svarvam bik\=ayai Sv\=ah\=a, my North west; Om Aim \'Sr\={\i}m Kl\={\i}m Gadyav\=asinyai Sv\=ah\=a my North; Aim Sarva\'s\=astra vasinyai Sv\=ah\=a, my North-east; Om Hr\={\i}m Sarvapujitayai Sv\=ah\=a, my top; Hr\={\i}m Pustakavasinyai Sv\=ah\=a my below and let Om Grantha v\={\i}jasvarup\=ayai Sv\=ah\=a protect all my sides. O N\=arada! This Vi\'svajaya Kavacha of the nature of Brahm\=a and its embodied Mantra I have now spoken to you. I heard this before from the mouth of Dharma Deva in the Ga\d{n}dh\=am\=adana mountain. Now I speak this to you out of my great affection for you. But never divulge this to anybody. One is worship one's spiritual Teacher (Guru Deva) according to due rites and ceremonies with clothings, ornaments, and sandalpaste and then fall down prostrate to him and then hold this Kavacha. Repeating this

five lakhs of times, one gets success and becomes a Siddha. The holder of this Kavacha becomes intelligent like Brihaspati, eloquent, Kav\={\i}ndras, and the conqueror of the three worlds, no sooner one becomes a Siddlha in this. In fact, he can conquer everything by virtue of this Kavacha. O Muni! Thus I have described to you this Kavacha according to K\=a\d{n}va \'S\=akh\=a. Now I am speaking about the method of worship, Dhy\=ana and the praise of this Kavacha. Hear.

Here ends the Fourth Chapter of the Ninth Book on the hymn, worship and Kavacha of Sarasvat\={\i} Dev\={\i} in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



