\chapter{On the curses of Gang\=a, Sarasvat\={\i} and Lak\d{s}m\={\i}}

1-2. N\=ar\=aya\d{n}a said :-- O N\=arada! Thus saying, the Lord of the World stopped. And Lak\d{s}m\={\i}, Gang\=a and Sarasvat\={\i} wept bitterly, embracing one another. All of them then looked to \'Sr\={\i} Kri\d{s}\d{n}a, and gave vent to their feelings one by one with tears in their eyes, and with their hearts throbbing with fears and sorrows.

3-4. Sarasvat\={\i} said :-- ``O Lord! What is now, the way out of this curse, so severe and paining since our births? How long can helpless women live, separated from their husbands? O Lord! I certainly say that I will sacrifice my body when I go to Bh\=arata, by taking recourse to yoga. The Mah\=atm\=as always protect all the persons without fail.''

5-6. Gang\=a said :-- ``O Lord of the Universe! Why have I been abandoned by You. What fault have I committed? I will quit my body. And You will have to partake of the sin due to the killing of an innocent woman. He is surely to go to hell even if he be the Lord of all, who forsakes in this world an innocent wife.''

7-15. Padm\=a said :-- ``O Lord! Thou art of the nature of Sattva Gu\d{n}a in fullness; what wonder, then, how Thou hast become angry! However let Thou be pleased now with Sarasvat\={\i} and Gang\=a. Forgiveness is the best quality of a good husband. I am ready just now to go to Bh\=arata when Sarasvat\={\i} has cursed me. But tell me, how long I will have to stay there? After how many days I shall be able to see again Thy lotus-feet? The sinners will wash away their dirts of sins in my waters by their constant baths and ablutions? By what means shall I be freed again and get back to Thy lotus-feet. How long shall I have to remain in my part, the daughter of Dharma Dhvaja, at the expiry of which I will be able to see Thee again? How long shall I have to assume the form of Tulas\={\i} tree, the abode of Thine. O Thou, the Ocean of mercy! Say, when wilt Thou deliver me? And if Gang\=a have to go to Bh\=arata, by the curse of Bh\=arata, when shall She be freed of the curse and sin and when shall She see back Thy feet? Again if Sarasvat\={\i} have to go to Bh\=arata out of Gang\=a's curse, when will that period of curse expire? How many days after shall She be able to come back to Thy feet? Now, be pleased to cancel Thy order for them to go to Brahm\=a and \'Siva respectively.'' O N\=arada! Thus speaking to Jagann\=atha, the Dev\={\i} Kamal\=a bowed down

at His feet and embracing them by Her own hairs of the bead, cried frequently.

16-37. Now the lotus-navelled Hari, always eager to shew favour to the devotees, smiled and with a gracious heart took up Padm\=a on His breast and said :-- ``O Sure\'svar\={\i}! I will keep my own word, also I will act according as you like. O Lotus-eyed! Hear. How the two ends can be made to meet. Let Sarasvat\={\i} go in her one part to have the form of a river and in her one half part to Brahm\=a and remain with me in Vaikuntha in Her full parts. Gang\=a will have to go in one part to Bh\=arata - to purify the three worlds as she will be urged eagerly to do so by Bhagiratha. And She will remain in her one part in the matted hair of Chandra \'Sekhara (the Mah\=adeva with Moon on his forehead), obtained with a great difficulty, and so will remain there purer than her natural pure state. And let her remain with me in full parts. O Padme! O Lovely-eyed One! You are most innocent; so part of your part will go to Bh\=arata and be the Padm\=avat\={\i} river and you will be the Tulas\={\i} tree. After the expiry of five thousand years of Kali Yuga, your curse will expire. Again you all will come to My abode. O Padme! Calamities are the causes of the happiness of the embodied beings. Without dangers no one can understand the true nature of happiness. The saint worshippers of My mantra who will perform their ablutions in your water will free you all of your curse by touch and sight. O Fair One! By the sight and touch (Dar\d{s}an, Spar\'san) of My bhaktas (devotees), all the sacred places of pilgrimages in the world will be purified. For uplifting and sanctifying the holy earth, My mantrop\=asakas, i.e., \'Saivas, \'Saktas, G\=a\d{n}apatyas, etc., that are devoted to Brahm\=a all are residing in Bh\=arata. Where My Bhaktas reside and wash their feet, that place is undoubtedly reckoned as the holy places of pilgrimages. So much so that by the sight and touch of My devotees, the murderer of a woman, of a cow, of a Br\=ahmin, the treacherous and even the stealer of the wife of one's Guru will be sanctified and liberated while living. Those who do not perform the vow of Ek\=ada\'s\={\i}, who do not perform Sandhy\=as, who are N\=astikas (atheists), the murderers, all are free of their sins by the sight and touch of My devotees. By the sight and touch of My devotees, those who live on their swords, pens, and the royal officials, the beggars in a village and the Br\=ahma\d{n}as who carry (deal in) bullocks are also freed of their sins. The traitors, the mischief makers of their friends, those who give false evidence, those that steal other's trust properties, are also freed of their sins by the sight and touch of My devotees. Those who are foul-mouthed, bastards, the husbands and sons of unchaste women are all purified by the sight and touch of My Bhaktas. The Br\=ahmi\d{n} cooks of \'S\=udras, Br\=ahmi\d{n}s of an inferior order (who subsist

upon the offerings made to the images which he attends), the village mendicants, those who are not initiated by their Gurus, these all are purified by the sight and touch of My devotees. O Fair One! The sins of those persons who do not maintain their fathers, mothers, brothers, wives, sons, daughters, sisters, the blind, friends, the families of the Gurus, the fathers-in-law, the mothers-in-law are also removed by the sight and touch of My devotees. Those that cut the A\'svattha trees, that slander My devotees, and the Br\=ahmi\d{n}s that eat the food of \'S\=udras, are also freed of their sins. Those who steal the Deva's articles, the Br\=ahma\d{n}a's articles, those that sell lac, iron, and daughters, those who commit Mah\=a P\=atakas (Brahm\=ahaty\=a, Sur\=ap\=anam, Steyam, Gurbangan\=aganah, Mah\=anti p\=atak\=any\=ahuh, tatsam-sargahseha Pa\~nchamam) and those that burn the \'S\=udr\=a's dead bodies, these also are freed of their respective sins by the sight and touch of My devotees.''

38-42. Mah\=a Lak\d{s}m\={\i} said :-- ``O Thou gracious to faithful attendants! What are the characteristics, the marks of those Bhaktas of Thine that Thou hast spoken of just now whose sight and touch destroy instantly the Mah\=ap\=atakas (five great sins), that are destroyed after a long time by the water of the T\={\i}rthas and the earthen and stone images of the Gods. The sins of the vilest of men, devoid of Hari bhakti, vain and egoistic, cheats, hvpocrites, slanderers of saints, vicious souls are destroyed by your Bhaktas, whose touch and ablations sanctify the sacred places of pilgrimages; by the touch of the dust and water of whose feet, the earth is purified; whom the Bhaktas of Bh\=arata always pray to see; and there is nothing higher than the meeting of those Bhaktas.'' S\=uta said :-- ``O Great \d{R}i\d{s}i! Thus hearing the words of Mah\=a Lak\d{s}m\={\i}, the Lord smiled and began to speak about the secret things or the marks of the Bhaktas.''

43-54. O Lak\d{s}m\={\i}! The marks of the Bhaktas are all mentioned very hiddenly in \'Srutis and Pur\=a\d{n}as. These are very sanctifying; destructive of sins, giving happiness, devotion, and liberation. These are never to be described to deceitful persons; these are the essences and to be kept hidden. But you are very simple and like my life. I therefore speak to you. Hear. O Fair One! All the Vedas declare him to be holy and the best of men, in whose ears are pronounced from the mouth of a Guru, the Vi\d{s}\d{n}u mantra. At the very moment of his being born into this world, one hundred generations back of that person, whether they be at that time in heaven or hell, get instantaneous liberation and if any of them happen to be born then as J\={\i}vas, they become liberated at once while living and finally get Vi\d{s}\d{n}upadam (the place of Vi\d{s}\d{n}u). That mortal is My Bhakta (devotee) who is full of devotion to Me, who always repeats My glories and acts according to My directions, who hears with all his heart My topics, and hearing which, whose mind dances with joy, whose voice gets choked and

tears incessantly flow out from whose eyes, who loses his outward consciousness. Such a man is indeed, My Bhakta. My Bhaktas do not long for happiness or Mukti, or the four states S\=alokya, S\=ayujya, S\=am\={\i}pya and S\=arst\={\i}, nor the Brahm\=ahood, nor the Devahood (the state of immortality); they want only to do Sev\=a (service) to Me and they are solely intent on doing this. Even in dreams they do not desire the Indraship, Manuship, the state of Brahm\=a so very difficult to be had; nor do they want the enjoyment of kingdoms and heavens. My Bhaktas roam in Bh\=arata, eager to hear My glories, and always very glad to recite My sweet glorious deeds. The birth of such Bhaktas in Bh\=arata is very rare. They purify the world and go ultimately to My abode, the best of all T\={\i}rthas (sacred places). Thus I have spoken O Padme! all that you wanted to hear. Now do as you like. Then Gang\=a and others all went to obey the order of Sr\={\i} Hari, Who went to His own abode.

Here ends the Seventh Chapter of the Ninth Book on the curses Gang\=a, Sarasvat\={\i}, and Lak\d{s}m\={\i} and the way to freedom thereof in the Mah\=apur\=a\d{n}am \'Sr\={\i}mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



