\chapter{On the origin of Brahm\=a, Vi\d{s}\d{n}u, Mahe\'sa and others}

1-34. N\=ar\=aya\d{n}a said:-- ``O Devar\d{s}i! The egg (born of M\=ula Prakriti) that was floating in the waters for a period equal to the life period of Brahm\=a, now in the fullness of time separated into two parts. Within that egg there was a powerful Child, lustrous like one thousand millions of suns. This child could not suck mother's milk, as it was forsaken by Her. So being tired of hunger, the child for a moment cried repeatedly. The child that will become the Lord of countless Brahm\=andas (universes), now an orphan having no father nor mother began to look upwards from the waters. This boy came to be denominated afterwards by the name of Mah\=a Vir\=at, when he became gross and

grosser. As there is nothing finer than radium so there is nothing grosser than Mah\=a Vir\=at. The power of this Mah\=a Vir\=at is one-sixteenth of that of \'Sri Kri\d{s}\d{n}a, the Highest Self. But this boy, (born of the Prakriti R\=adh\=a) is the Sole Stay of all this Universe and he is denominated by the name ``Mah\=a-Vi\d{s}\d{n}u''. In his every pore countless universes are existing. So much so that even \'Sri Kri\d{s}\d{n}a could not count them. If it were possible to count the number of dust particles, it is impossible to count the number of universes. So there are endless Brahm\=as, Vi\d{s}\d{n}us, and Mahe\'svaras. In every Brahm\=anda, there is Brahm\=a, Vi\d{s}\d{n}u, and Mahe\'sa. Each Brahm\=anda extends from P\=at\=ala to the Brahm\=aloka. The abode of Vaikuntha is higher than that (i. e. it is situated outside of Brahm\=anda), again the abode of Goloka is fifty koti yojanas (50 x 10 x 4 x 2 million miles) higher than Vaikuntha. This Goloka Dh\=ama is eternal and real as \'Sri Kri\d{s}\d{n}a is eternal and real. This world composed of the seven islands is surrounded by the seven oceans. Forty-nine Upa Dv\={\i}pas (smaller islands adjacent to them) are existing here. Besides there are countless mountains and forests. Higher than this earth is the Brahm\=aloka with seven heavens and below this are the seven P\=at\=alas. This is the bounding limit of Brahm\=anda. Just above this earth there is the Bh\=urloka; above is Bhuvarloka; then Svarloka, then Janarloka, then Taparloka, then Satyaloka, and above that is Brahm\=aloka. The splendour of Brahm\=aloka is like that of molten gold. But all the substances whether outside or inside this Brahm\=aloka, are transient. When this Brahm\=anda (cosmos) dissolves, everything dissolves and is destroyed. All are temporary like bubbles of water. Only Goloka and Vaikuntha are eternal. In every pore of this Mah\=a Vir\=at is existing one Brahm\=anda (cosmos). What to speak of others when even Kri\d{s}\d{n}a cannot count the number of these Brahm\=andas. In every Brahm\=anda there is Brahm\=a, Vi\d{s}\d{n}u and Mahe\'sa. O Child N\=arada! In every Brahm\=anda, the number of the gods is three kotis or 30 millions. Some of them are the Dikpatis (the Regents of the quarters); some are the Dikp\=alas (the Rulers of the quarters), some are asterisms, and some planets. In the Bh\=urloka, there are four Var\d{n}as (Br\=ahmi\d{n}s, etc.,) and in the P\=at\=alas there are N\=agas. Thus the Universe exists composed of moveable and non-moveable things (This is Brahm\=anda Vivriti). O N\=arada! Now the Vir\=at Puru\d{s}a began to look up to the skies again and again but He could not see anything within that egg except the void. Then distressed with hunger He cried out repeatedly and became merged in anxiety. Next moment getting back His consciousness, He began to think of Kri\d{s}\d{n}a, the Highest Person and saw there at once the eternal light of Brahm\=a. He saw there His form as deep blue like new rain-cloud;

with two hands, garment of a yellow colour, sweet smile on His face, flute in His hand and He seemed to be very anxious to show His Grace to Devotees. Looking at the Lord, His Father, the boy became glad and smiled. The Lord, the Bestower of boon granted him boons appropriate for that moment ``O Child! Let you possess knowledge like Me; let your hunger and thirst vanish; let you be the holder of innumerable Brahm\=andas till the time of Pralaya (the universal dissolution). Be without any selfishness, be fearless and the bestower of boons to all. Let not old age, death, disease, sorrow nor any other ailings afflict thee. Thus saying He repeated thrice on his ear the six-lettered great Mantra ``Om Kri\d{s}\d{n}\=aya Sv\=ah\=a'' worshipped by the Vedas with their Amgas, the Giver of desires and the destroyer of all troubles and calamities. O Brahm\=a's Son! Thus giving the mantra, \'Sri Kri\d{s}\d{n}a arranged for his feeding thus:-- In every universe, whatever offerings will be given to \'Sri Kri\d{s}\d{n}a, one sixteenth of that will go to N\=ar\=ayana, the Lord of Vaikuntha and fifteen-sixteenth is to go to this boy, the Vir\=at. \'Sri Kri\d{s}\d{n}a did not allot any share for Himself. Himself transcending all the Gunas, and Full, He is always satisfied with Himself. What necessity is there for any further offerings? Whatever the people offer with devotion, the Lord of Laksm\={\i}, the Vir\=at eats all these. Bhagav\=an \'Sri Kri\d{s}\d{n}a giving thus to the Vir\=at the boon and the Mantra said:-- ``O Child! Say what more you desire; I will give you that instantly. The Vir\=at boy, hearing thus the words of \'Sri Kri\d{s}\d{n}a, spoke:-- ``O Thou Omnipresent! I have got no desires whatsoever, save this that as long as I live, whether for a short time or for a long time, let me have pure Bhakti towards Thy lotus feet.

35-41. In this world he is J\={\i}vanmukta (liberated whilst living) who is your Bhakta; and that bewildered fool is dead while living who is devoid of any Bhakti to Thee. What needs he to perform Japam, asceticism, sacrifice, worship, holding fasts and observances, going to sacred places of pilgrimages and other virtuous acts if he be without any bhakti to \'Sri Kri\d{s}\d{n}a? Vain is his life who is devoid of any devotion to \'Sri Kri\d{s}\d{n}a, under Whose Grace he has obtained his life and Whom he does not now pay homage and worship. He is endowed with Sakti as long as \=Atma (Self) resides in his body; no sooner the \=Atma departs from his body all the Saktis accompany him. O Great One! And thou art the Universal \=Atman (soul) who transcends Prakriti, Who is All will, the Primeval Person and of the nature of the Highest Light. O Child! Thus saying, the Vir\=at boy remained silent. \'Sri Kri\d{s}\d{n}a then, spoke in sweet words:-- ``O Child! Let you remain as fresh as ever like Me. You will not have any fall even if innumerable Brahm\=as pass away.

42-57. Let you divide yourself in parts and turn into smaller Vir\=ats in every universe. Brahm\=a will spring from your navel and will create

the cosmos. From the forehead of that Brahm\=a will spring eleven Rudras for the destruction of the creation. But they will all be parts of \'Siva. The Rudra named K\=al\=agni, of these eleven Rudras, will be the destroyer of all this Vi\'svas (cosmos). Besides, from each of your sub-divisions, the Vi\'snu will originate and that Bhagav\=an Vi\d{s}\d{n}u will be the Preserver of this Vi\'svu world. I say that under my favour you will always be full of Bhakti towards Me and no sooner you meditate on Me, you will be able to see My lovely form. There is no doubt in this; and your Mother, Who resides in My breast, will not be difficult for you to see. Let you remain here in ease and comfort. I now go to Goloka. Saying thus \'Sri Kri\d{s}\d{n}a, the Lord of the world disappeared. Going to His own abode He spoke instantly to Brahm\=a and \'Sankara, skilled in the works of creation and destruction:-- ``O Child Brahm\=a! Go quickly and be born in parts from the navels of each of the smaller Vir\=ats that will arise from the pores of the Great Vir\=at. O Child Mah\=adeva! Go and be born in parts from the forehead of each Brahm\=a in every universe for the destruction of the creation; (but be careful that you not forget) and perform austerities for a long, long time. O son of the Creator Brahm\=a! Thus saying, the Lord of the Universe remained silent. Brahm\=a and \'Siva, the auspicious, bowing to the Lord, went to their own duties. On the other hand, the great Vir\=at that lay floating in the waters of the Brahm\=anda sphere, created from his every pore each smaller Vir\=at. That youth Jan\=ardan of the form of the Great Cosmos, wearing yellow garment of the bluish-green colour of the Durba grass, lay sleeping pervading everywhere. Brahm\=a took his birth in His navel. He, then, after his birth, began to travel in that navel-lotus and in the stem of the lotus for one lakh yugas. But he could not find out the place whence the lotus or its stem had sprung up. O N\=arada! Then your father became very anxious and came back to his former place and began to meditate on the lotus feet of \'Sri Kri\d{s}\d{n}a. Then, in meditation, with his introspective eye, he first saw the small Vir\=at, then the endless great Vir\=at lying on the watery bed, in whose pores the universes are existing and then he saw the God \'Sri Kri\d{s}\d{n}a in Goloka with Gopas and Gopis. He then began to praise the Lord of Goloka when He granted boons to your father, and he began to do the work of creation.

58-62. From the mind of your father, were born first \'Sanaka and other brothers and then from his forehead eleven Rudras sprang. Then from the left side of that small Vir\=at lying on the bed of waters, the four-handed Vi\d{s}\d{n}u Bhagav\=an, the Preserver of the Universe, came. He went to \'Svetadv\={\i}pa, where he remained. Then your father became engaged in creating this Universe, moveable and non-moveable, composed

of three worlds, heaven, earth and P\=at\=ala, in the navel of that small Vir\=at Purusa. O N\=arada! Thus from the pores of that great Vir\=at each universe has sprung and in every universe there is one small Vir\=at, one Brahm\=a, one Vi\d{s}\d{n}u and one \'Siva and \'Sanaka and others. O Best of twice born! Thus I have described the glories of Kri\d{s}\d{n}a, that give exceeding pleasure and Mok\d{s}a. Now say what more you want to hear?

Here ends the Third Chapter of the Ninth Book on the Origin of Brahm\=a, Vi\d{s}\d{n}u and Mahe\'sa and others in the Mah\=a Pur\=anam \'Sri Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=arsi Veda Vy\=asa.



