\chapter{On the incarnation of Mah\=a Lak\d{s}m\={\i} in the house of Ku\'sadhvaja}

1-30. \'Sr\={\i} N\=ar\=aya\d{n}a said :-- O Muni! Dharmadhvaja and Ku\'sadhvaja practised severe tapasy\=as and worshipped Lak\d{s}m\={\i}. They then got separately their desired boons. By the boon of Mah\=a Lak\d{s}m\={\i}, they

became again the rulers of the earth. They acquired great religious merits and they also had their children. The wife of Ku\'sadhvaja was named M\=al\=avat\={\i}. After a long time, the chaste wife delivered one daughter, born of the parts of Kamal\=a. The daughter, on being born, became full of wisdom. On being born, the baby began to sing clearly the Vedic mantrams from the lying-in-chamber. Therefore She was named Vedavat\={\i} by the Pundits. She bathed after her birth and became ready to go to the forest to practise severe tapas. Everyone then, tried earnestly to dissuade her, devoted to N\=ar\=aya\d{n}a, from the enterprise. But she did not listen to anybody. She went to Pu\d{s}kara and practised hard tapasy\=a for one Manvantara. Yet her body did not get lean a bit; rather she grew more plumpy and fatter. By degrees her youth began to show signs in her body; one day she heard an incorporeal voice from the air above, ``O Fair One! In your next birth \'Sr\={\i} Hari, adored by Brahm\=a and other gods, will be your husband.'' Hearing this, her joy knew no bounds. She went to the solitary caves in the Gandham\=adan mountain to practise tapas again. When a long time passed away in this tapasy\=a, one day the irresistible R\=avana came there as guest. No sooner Vedavat\={\i} saw the guest, then she gave him, out of devotion to the guest, water to wash feet, delicious fruits, and cool water for his drink. The villain accept the hospitality and sitting there, began to ask :-- ``O Auspicious One! Who are you?'' Seeing the fair smiling lady, with beautiful teeth, her face blooming like the autumnal lotus, of heavy loins, and of full breast, that villain became passionate. He lost entirely all consciousness and became ready to make violence on Her. Seeing this, the chaste Vedavat\={\i}, became angry and out of her tapas influence, astounded him and made him insensible to move. He remained motionless like an inanimate body. He could not move his hands nor feet nor could he speak. That wicked fellow then mentally recited praises to her. And the praise of the Higher \'Sakti can never go futile. She became pleased and granted him religious merits in the next world. But she also pronounced this curse :-- ``That when you have touch my body out of passion, then you will be ruined with your whole family for my sake. Now see my power.'' O N\=arada! Thus saying to R\=avana, Vedavat\={\i} left her body by her yogic power. Then R\=avana took her body and delivered it to the Ganges and he then returned to his own home. But R\=avana thought over the matter repeatedly and exclaimed, ``What wonder have I seen! Oh! What a miracle this lady has wrought!'' R\=avana thus lamented. This Vedavat\={\i}, of pure character, took her birth afterwards as S\={\i}t\=a, the daughter of

Janaka. For the sake of this S\={\i}t\=a, R\=avana was ruined with his whole family. By the religious merits of her previous birth, the ascetic lady got Bhagav\=an Hari \'Sr\={\i} R\=ama Chandra, the Fullest of the Full, for her husband and remained for a long time in great enjoyment with the Lord of the world; a thing very difficult to be attained! Though she was a J\=atismar\=a (one who knows all about her past lives), she did not feel any pain due to her practising severe austerities in her previous birth; for when the pains end in success, the pains are not then felt at all. S\={\i}t\=a, in Her fresh youth enjoyed various pleasures in the company of her husband, handsome, peaceful, humorous and witty, the chief of the Devas, loved by the female sex, well-qualified, and just what she desired. But the all-powerful Time is irresistible; the truthful R\=amachandra, the scion of the Raghu's family, had to keep up the promise made by his father and so he had to go to the forest, ordained by Time. He remained with S\={\i}t\=a and Lak\d{s}ma\d{n}a near the sea. Once the God Fire appeared to Him in the form of a Br\=ahma\d{n}a. Fire, in a Br\=ahmin-form, saw R\=ama Chandra morose and became himself mortified. Then the Truthful Fire addressed the truthful R\=amachandra :-- ``O Bhagav\=an R\=amachandra! I now speak to you how time is now coming to you. Now has come the time when your S\={\i}t\=a will be stolen.

31-48. The course of Destiny is irresistible; none else is more powerful than Time, Fate. So give over your S\={\i}t\=a, the World Mother to me and keep with you this Chh\=ay\=a S\={\i}t\=a (the shadow S\={\i}t\=a; the false S\={\i}t\=a). When the time of S\={\i}t\=a's ordeal by fire will take place I will give Her back to you. The Devas united have sent me to you. I am not really a Br\=ahmi\d{n}; but I am Ag\d{n}i Deva (eater of oblations).'' R\=amachandra heard Fire and gave his assent. But his heart shattered. He did not speak of this to Lak\d{s}ma\d{n}a. By the yogic power Ag\d{n}i (Fire) created a M\=ay\=a S\={\i}t\=a. This M\=ay\=a S\={\i}t\=a, O N\=arada, was perfectly equal to the real S\={\i}t\=a. Fire, then, handed this M\=ay\=a S\={\i}t\=a to the hands of R\=amachandra. H\=ut\=asana (fire) took the real S\={\i}t\=a and said, ``Never divulge this to any other body'' and went away. What to speak of divulging the secret to any other body, Lak\d{s}ma\d{n}a even could not know it. By this time R\=ama saw one deer, made of all gold. To bring that deer carefully to her, S\={\i}t\=a sent R\=amachandra with great eagerness. Putting S\={\i}t\=a under Lak\d{s}ma\d{n}a's care, in that forest, R\=ama went himself immediately and pierced the deer by one arrow. That M\=ay\=a mr\={\i}ga (the deer created by magic powers) on being pierced, cried out ``Ha Lak\d{s}ma\d{n}a!'' and seeing Hari before him and remembering the name of Hari, quitted

his life. The deer body then vanished; and a divine body made its appearance in its stead. This new body mounting on an aerial car made of jewels, ascended to Vaikuntha. That M\=ayik (magic) deer was in its previous birth, a servant, of the two gate-keepers of Vaikuntha; but, for the sake of some emergency, he had to take up this R\=akh\d{s}asa birth. He again became the servant of two door-keepers of Vaikuntha. On the other hand S\={\i}t\=a Dev\={\i}, hearing the cry ``Ha Laksmana!'' became very distressed and sent Lak\d{s}ma\d{n}a in search of R\=ama. No sooner did Lak\d{s}ma\d{n}a get out of the hermitage, the irresistible R\=ava\d{n}a took away S\={\i}t\=a gladly to the city of Lanka (Ceylon). Now R\=amachandra, seeing Lak\d{s}ma\d{n}a on the way in the forest, became merged in the ocean of sorrows and without losing any time came hurriedly to the hermitage where he could not find S\={\i}t\=a. Instantly he fell unconscious, on the ground; and, after a long time, when he regained his consciousness, he lamented and wandered here and there in search of Her. After some days on the banks of the river God\=avar\={\i}, getting the information of S\={\i}t\=a, he built a bridge across the ocean with the help of His monkey armies. Then he entered with his army into Lanka and slew R\=ava\d{n}a with arrows with all his friends. When S\={\i}t\=a's ordeal by fire came, Ag\d{n}i (Fire) handed over the real S\={\i}t\=a to R\=amachandra. The Shadow S\={\i}t\=a then humbly addressed Ag\d{n}i and R\=ama Chandra, ``O Lord! What am I to do now? Settle my case.''

49-53. Ag\d{n}i and R\=amachandra both of them then said to Chh\=ay\=a S\={\i}t\=a :-- ``O Dev\={\i}! Go to Pu\d{s}kara and practise tapasy\=a there; that place is the giver of religious merits and then you will be the Svarga Lak\d{s}m\={\i} (Lak\d{s}m\={\i} of Heaven).'' Hearing this, the Chh\=ay\=a S\={\i}t\=a went and practised tapasy\=a for the three divine lakh years and became Mah\=a Lak\d{s}m\={\i}. This Svarga Lak\d{s}m\={\i} appeared at one time from the sacrificial Kunda (pit). She was known as the daughter of Drupada and became the wife of the five P\=andavas. She was Veda Vat\={\i}, the daughter of Ku\'sadhvaja in the Satya Yuga; S\={\i}t\=a, the wife of R\=ama and the daughter of Janaka in Tret\=a Yuga; and Draupad\={\i}, the daughter of Drupada, in the Dv\=apara Yuga. As she existed in the Satya, Tret\=a, and Dv\=apara Yugas, the Three Yugas, hence She is Trih\=ayan\={\i}.

54. N\=arada said :-- ``O Chief of Munis! O Remover of doubts! Why had Draupad\={\i} five husbands? A great doubt has arisen in my mind on this point. Remove my doubt.''

55-63. N\=ar\=aya\d{n}a said :-- O Devar\d{s}i! When, in the city of Lanka, the real S\={\i}t\=a came before R\=ama, then Chh\=ay\=a S\={\i}t\=a, full of youth and beauty, became very anxious. Ag\d{n}i Deva and R\=amachandra both told

Her to go to Pu\d{s}kara and worship \'Samkara. While this Chh\=ay\=a S\={\i}t\=a was practising austerities in Pu\d{s}kara, She became very anxious to get a good husband and asked from Mah\=aDeva the boon ``Grant me a husband'' and repeated it five times. \'Siva, the chief among the humorous, witty persons, hearing this, said, ``O Dear! You will get five husband\'s' and thus granted her the boon. Therefore She became the dearest wife of the five P\=andavas. Now hear other facts. When the war at Lanka was over, \'Sr\={\i} R\=amachandra got his own dear wife S\={\i}t\=a, and installing Vibh\={\i}sana on the throne of Lanka, returned to Ayodhy\=a. He ruled for eleven thousand years in Bh\=arata and finally went to Vaikuntha with his all his subjects. Vedavat\={\i}, the incarnation in part of Lak\d{s}m\={\i} dissolved in the body of Kamal\=a. Thus I have described to you the pure anecdote of Vedavat\={\i}. Hearing this destroys sins and increases virtue. The four Vedas reigned incarnate, in their true forms, on the lips of Vedavat\={\i}; hence She was named Vedavat\={\i}. Thus I have told you the anecdote of the daughter of Ku\'sadhvaja. Now hear the story of Tulas\={\i}, the daughter of Dharmadhvaja.

Here ends the Sixteenth Chapter in the Ninth Book on the incarnation of Mah\=a Lak\d{s}m\={\i} in the house of Ku\'sadhvaja in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



