\chapter{On the anecdote of Gang\=a}

1. N\=arada said :-- O Lord of the Devas! Kindly say in what Loka did Gang\=a go after 5000 (five thousand) years of the Kali Yuga?

2-4. N\=ar\=aya\d{n}a said :-- The Bh\=ag\={\i}rath\={\i} Gang\=a came down to Bh\=arata under the curse of Bh\=arat\={\i}; and when the term expired, She went back, by the Will of God, to the region of Vaikuntha. Also at the end of the period of their curses, Bh\=arat\={\i} and, Lak\d{s}m\={\i}, too, left Bh\=arata and repaired to N\=ar\=aya\d{n}a. Gang\=a, Lak\d{s}m\={\i}, and Sarasvat\={\i}, these three and Tulas\={\i} all these four are so very dear to \'Sr\={\i} Hari.

5-6. N\=arada said :-- How did Gang\=a appear from the lotus feet of Vi\d{s}\d{n}u? Why did Brahm\=a put Her in His Kamandalu? I have heard that Gang\=a is the wife of \'Siva; how then, came She to be the wife of N\=ar\=aya\d{n}a? Kindly describe all these in detail to me.

7-8. N\=ar\=aya\d{n}a said :-- O Muni! In ancient times, in the region of Goloka, Gang\=a assumed the liquid appearance. She was born of the bodies of R\=adh\=a and Kri\d{s}\d{n}a. So She is of the nature of both of them and their parts. Gang\=a is the presiding deity of water. She is unequalled in Her beauty in this world. She is full of youth and adorned with all ornaments.

9-43. Her face was like the autumnal mid-day lotus and sweet smile was always reigning on Her lips; Her form was very beautiful; Her colour was as bright as melted gold and She looked brilliant like the Autumnal Moon. Eyes and mind get cool and become pleasant at Her beauty and radiance; She was of purely Suddha Sattva; Her loins were bulky and hard and She was covered with excellent clothings all over Her body. Her breasts were plump and prominent; they were raised, hard, and nicely round. Her eyes very fascinating, always casting sidelong glances. Her braids of hair situated a little oblique and the garland of M\=alat\={\i} flowers over it made Her look extremely handsome. The saudalpaste dot and the vermillion dot were seen on Her forehead. On Her cheeks the leaves of musk were drawn and Her lips were red like Bandh\=uka flowers and they looked enchanting. Her rows of teeth looked like rows of ripe pomegranates; the ends of Her cloth not inflammable by fire, worn in front in a knot round the waist. She sat by the side of Kri\d{s}\d{n}a, full of amorous desires, and abashed. She covered Her face with the end of Her cloth and was seeing, with a steadfast gaze the face of the Lord and She was drinking the nectar

of His face with great gladness. Her lotus face bloomed and became gladdened at the expectation of a first amorous embrace. She fainted on seeing the Form of Her Lord and a thrill of joy passed all over Her body. In the meanwhile R\=adhik\=a came up there. R\=adh\=a was attended by thirty kotis of Gop\={\i}s. She looked brilliant like tens of millions of Moons. Seeing Gang\=a by the side of \'Sr\={\i} Kri\d{s}\d{n}a, Her face and eyes became reddened with anger like a red lotus. Her colour was yellow like champaka and Her gait was like a maddened elephant. She was adorned with various invaluable ornaments made of jewels. Her pair of clothings were tied round Her waist. They were decked with invaluable jewels and not inflammable by fire (fire-proof). The Arghya offered by \'Sr\={\i} Kri\d{s}\d{n}a was on Her lotus-feet of the colour of a flowering shrub, Hibiscus mutabilis, and She was going slowly step by step. The \d{R}i\d{s}is began to fan Her with white Ch\=amaras no sooner She, descending from the excellent aeroplane decked with jewels, began to walk. Below the point where the parting of the hairs on the head is done, there was the dot of Sindura on Her forehead. It looked brilliant like a bright lamp flame. On both sides of this Sindurabindu, the dot of musk and the dot of Sandal-paste were seen. When She began to quiver with anger, Her braid, with P\=arij\=ata garland round it began to tremble also. Her lips adorned with beautiful colours, began to quiver also. She took Her seat angrily on a jewel throne by the side of \'Sr\={\i} Kri\d{s}\d{n}a. Her attendants took their seats in their allotted positions. Seeing R\=adh\=a, \'Sr\={\i} Kri\d{s}\d{n}a got up at once from His seat with reverence and addressed Her, smiling and began to converse with Her in sweet words. The Gop\={\i}s, very much afraid and with their heads bent low, began to chant hymns to Her with devotion. \'Sr\={\i} Kri\d{s}\d{n}a also began to praise Her with stotras. At this moment Gang\=a Dev\={\i} got up and praised Her with various hymns and asked Her welfare with fear and with humble words. Out of fear, Her throat, lips and palate were parched up. She took refuge humbly at \'Sr\={\i} Kri\d{s}\d{n}a's feet. \'Sr\={\i} Kri\d{s}\d{n}a then, took Gang\=a Dev\={\i} on His breast when She became calm and quiet. At this interval Sure\'svar\={\i} Gang\=a looked at R\=adh\=a, seated on a throne, lovely and sweet, as if She was burning with Brahm\=a Fire. Since the beginning of creation, She is the Sole Lady of innumerable Brahm\=as and She is Eternal. At the first sight, She looked young as if of twelve years old. Nowhere in any Universe can be seen a lady so beautiful and so qualified. She was peaceful, calm and quiet, lovely, infinite and having no beginning nor end. She was auspicious, well endowed with all auspicious signs, prosperous, and having the good fortune of having a best

husband. She was the foremost jewel amongst the ladies and appeared as if all the beauties were concentrated in Her. R\=adh\=a is the (left) half of \'Sr\={\i} Kri\d{s}\d{n}a's body; whether in age or in strength or in beauty she was in every way perfectly equal to \'Sr\={\i} Kri\d{s}\d{n}a. Lak\d{s}m\={\i} and the Lord of Lak\d{s}m\={\i} both worship R\=adh\=a. The excellent brilliance of Sr\={\i} Kri\d{s}\d{n}a was overpowered by the beauty of R\=adh\=a. Taking Her seat on the throne She began to chew betels offered by Her attendants (Sakh\={\i}s). She is the Mother of all the worlds; but no one is Her mother. She is fortunate, respected and proud. She is the Ruling Lady of \'Sr\={\i} Kri\d{s}\d{n}a's Life and Soul and ever dearer to Him than His Pr\=a\d{n}a (vital breath). O Devar\d{s}i! Gang\=a, the Governess of the Devas, looked at Her over and over again with a steadfast gaze; but Her eyes and mind were not at all satiated. At this moment, R\=adh\=a addressed smilingly to \'Sr\={\i} Kri\d{s}\d{n}a, the Lord of the world, humbly and in sweet words. O My Lord! Who is that Lady sitting by Thy side, looking askance, eager and with a smiling countenance. She is enchanted with Thy beauteous form and fainting away. Her whole body is excited with rapturous joy. Hiding Her face with cloth She is frequently looking at Thee. Thou also dost look on Her smilingly and with desires. What are all these? Even during My presence in this Goloka, all these bad practices are being rampant.

44-51. It is Thou that art doing all these bad things often and often! We are female sex; what shall we do? We are naturally, of a very pleasing temper, simple. I bore and forgave all these out of our love. O Licentious One! Take Thy Beloved and go away quickly from this Goloka. Otherwise these things will not bid fair to Thee. Firstly, one day I saw Thee, united with Viraj\=a Gop\={\i}, in Chandana (Sandal wood) forest. What to do? At the request of the \'Sakh\={\i}s, I did forgive Thee. Then, hearing My footsteps Thou didst fly away. Viraj\=a, out of shame, quitted Her body and assumed the form of a river. That is million Yojanas wide an four times as long. Even to this day that Viraj\=a is existing, testifying to Thy Glory (near Puri, Jagann\=atha)! When I went back to My home Thou didst go to Viraj\=a again and cried aloud ``O Viraje! O Viraje!'' Hearing Your cry, Viraj\=a, the Siddha Yogin\={\i} arose from the waters, out of Her Yogic power, and when She showed Thee Her divine appearance decked with ornaments, Thou didst draw Her to Thy side and cast Thy seed in Her. It is owing to the casting of that seed in the womb of Viraj\=a that the seven oceans have come into existence!

52-107. Secondly, one day I saw Thee in actual intercourse with the Gop\={\i} named \'Sobh\=a! Hearing My footsteps, Thou fled'st away that day also.

Out of shame \'Sobh\=a quitted Her body and departed to the sphere of Moon (Chandra Mandal). The cooling effect of the Moon is due to this \'Sobh\=a. When \'Sobh\=a was thus distressed, Thou didst divide Her and put some parts to gems and jewels, part to gold, partly to excellent pearls and gems, partly on the face of women, partly to the bodies of Kings, partly to the leaves of trees, partly to flowers, partly to ripe fruits, partly to corns, partly to palaces and temples, partly to purified materials, partly to young and tender shoots and foliage, and partly to milk. Thirdly, I saw Thee united with Prabh\=a Gop\={\i} in Brind\=aban. Thou fled'st away, hearing My footsteps. Out of shame, Prabh\=a quitted Her body and departed to the Solar atmosphere. This Prabh\=a (lustre) is fierce luminosity of the Solar atmosphere. Out of the pangs of separation Thou criedest and didst divide Prabh\=a and didst put some parts in Fire, partly amidst the Yak\d{s}as, partly into lions, among men, partly amongst the Devas, partly in Vai\d{s}\d{n}avas, partly in serpents, partly in Br\=ahma\d{n}as, partly in Munis, partly in ascetics, and partly in fortunate and prosperous ladies. Thou hadst to weep then after Thou hadst thus divided Prabh\=a, for Her separation and and fourthly I saw Thee in love union with the Gop\={\i} \'S\=anti in R\=asa Mandalam. On the coming of the spring season, one day Thou with garlands of flowers on Thy neck and with Thy body besmeared with sandal paste and decked with ornaments, wast sleeping on a bed of flowers with \'S\=anti Gop\={\i}, decked with gems, in a temple made of gems and pearls and illumined by a lamp of jewels and Thou wast chewing the betel, given by Thy beloved. Hearing My sound Thou fled'st away. \'S\=anti Gop\={\i}, too, out of fear and shame quitted Her body and disappeared in Thee. Therefore \'S\=anti is reckoned as one of the noblest qualities. Out of the pain of separation, Thou didst divide the body of \'S\=anti and distributed partly to forests, partly to Brahm\=a, partly to Me, partly to \'Suddha Sattv\=a Lak\d{s}m\={\i}, partly to Thy Mantra worshippers, partly to My Mantra worshippers, partly to the ascetics, partly to Dharma, and partly to the religious persons. Fifthly, dost Thou remember that one day anointing all over Thy body fully with the sandal paste and good scent and with garlands on Thy neck, well dressed, decked with jewels, Thou wast sleeping with K\d{s}am\=a (forgiveness) Gop\={\i} in ease and happiness, on a nice bedding interspersed with flowers and well scented. Thou wert so much over-powered by sleep after fresh intercourse that when I went and disturbed, then Thou two didst get up from the sweet sleep. I took away Thy yellow robes, the beautiful Mural\={\i} (flute), garlands made of forest flowers, Kaustubha gems, and invaluable earrings of pearls and gems. I gave it back to Thee at the earnest request of the \'Sakh\={\i}s. Thy body turned black with sin and dire shame, K\d{s}am\=a then quitted Her body out of shame and went down to the

earth. Therefore K\d{s}am\=a turned out to be the repository of best qualities. Out of affection to Her, Thou didst divide Her body and distributed them partly to Vi\d{s}\d{n}u, partly to the Vai\d{s}\d{n}avas, partly to Dharma, partly to the religious persons, partly to weak persons, partly to ascetics, partly to the Devas, and partly to the Pundits (literary persons). O Lord! Thus I have described Thy qualities as far as I know. What more dost thou want to hear? Thou hast many more qualities! But I am not aware of them. Having thus spoken, the red-lotus eyed R\=adh\=a began to rebuke Gang\=a sitting by the side of \'Sr\={\i} Kri\d{s}\d{n}a with Her head bent low out of shame. At this time Gang\=a, who was a Siddha Yogin\={\i} came to know all the mysteries, and instantly disappeared from the assembly in Her own water form.

The Siddha Yogin\={\i} R\=adh\=a came to know also, by Her Yogic power, the secrets of Gang\=a and became ready to drink the whole water in one sip. Gang\=a, knowing this intention of R\=adh\=a, by Her Yogic power, took refuge of Kri\d{s}\d{n}a and entered into His feet. Then R\=adh\=a began to look out for Gang\=a everywhere :-- First She searched in Goloka, then Vaikuntha, then Brahm\=a-loka; then She searched all the Lokas one by one but no where did She find Gang\=a. All the places in Goloka became void of of water; all turned out dried mud and all the aquatic animals died and fell to the ground. And Brahm\=a, Vi\d{s}\d{n}u, \'Siva, Ananta, Dharma, Indra, Moon, Sun, Manus, Munis, Siddhas, ascetics all became very thirsty and their throats became parched. They then went to Goloka, and bowed down with devotion to \'Sr\={\i} Kri\d{s}\d{n}a, Who was the Lord of all, beyond Prakriti, the Supreme, worthy to be worshipped, the Bestower of boons, the Best, and the Cause of boons; Who is the Lord of Gopas and Gop\={\i}s; Who is formless, without any desire, unattached, without refuge, attributeless, without any enthusiasm, changeless, and unstained; Who is All Will and who assumes forms to show favour to His devotees; Who is Sattva, the Lord of truth, the Witness and eternal Puru\d{s}a and Who is the Highest, the Supreme Lord, the Best and Excellent, the Highest Self and the Supreme God. They began to hymn Him. All were filled with intense feelings with devotion; tears of love were flowing from their eyes and the bodies of all were filled with ecstasy, the hairs standing in ends. He was Para Brahm\=a; His Substance was made of Transcendental Light, Who is the Cause of all Causes, who was seated in a wonderful throne, built of invaluable gems and jewels, who was being fanned by the Gopas with white chowries, who was seeing and hearing with great delight, and smiling countenance, the dancing and singing of the Gop\={\i}s, who was chewing the scented betel offered by R\=adh\=a and who

was residing in. the heart of His dearest \'Sr\={\i} R\=adh\=a, who was the Perfect, all pervading, and the Lord of the R\=asa Circle. The Manus, Munis, and the ascetics all bowed down to \'Sr\={\i} Kri\d{s}\d{n}a, no sooner they beheld Him. Joy and wonder at once caught hold of their hearts. They then looked at one another and gave over to Brahm\=a the task of communicating their feelings. The four faced Brahm\=a, with Vi\d{s}\d{n}u on His right and V\=ama Deva on His left, gradually came in front of \'Sr\={\i} Kri\d{s}\d{n}a. Wherever He cast His glance in the R\=asa Mandalam, He saw \'Sr\={\i} Kri\d{s}\d{n}a, full of the Highest Bliss, of the nature of the Highest Bliss, sitting. All have turned out Kri\d{s}\d{n}as; their seats were all uniform; all were two armed and with flutes in their hands; on every one's neck is the forest garland; peacock's tail was on the top of everyone's crest and Kaustubha jewels were on all their breasts. The Forms of all of them were very beautiful; very lovely and very peaceful. No difference at all between them whether in form, or in qualities, or in ornaments, or in radiance, in age, in lustre, in no respect no one was inferior to another. No one was imperfect; no one was deficient in lordliness. It was indeed very difficult to make out who was the master and who was the servant. Sometimes He is seen in His Teja form as the Great Light, and there is nothing else; sometimes there is that Clear Divine Form; sometimes He comes Formless; sometimes with form; and again sometimes both with and without form. Sometimes there is no R\=adh\=a; there is only Kri\d{s}\d{n}a; And sometimes again in every seat there is the Yugal Murti R\=adh\=a and Kri\d{s}\d{n}a combined. Sometimes R\=adh\=a assumes the form of Kri\d{s}\d{n}a. So the Creator Brahm\=a could not make out whether \'Sr\={\i} Kri\d{s}\d{n}a was a female or a male. At last He meditated on \'Sr\={\i} Kri\d{s}\d{n}a in his heart-lotus and began to chant hymns to Him with devotion and prayed for forgiveness for his misdoings. When \'Sr\={\i} Kri\d{s}\d{n}a got pleased, the Creator, opening His eyes, saw \'Sr\={\i} Kri\d{s}\d{n}a on the breast of \'Sr\={\i} R\=adh\=a. There were His attendants on all the sides and the Gop\={\i}s all around. Seeing this, Brahm\=a, Vi\d{s}\d{n}u, and Mahe\'svara bowed down to Him and sang His praises.

108-113. \'Sr\={\i} Kri\d{s}\d{n}a, the Lord of Lak\d{s}m\={\i}, the Omnipresent, Cause of all, the Lord of all, and the Internal Ruler of all, knew their intentions and, addressing them, separately said :-- ``O Brahm\=an! Is it all well with you? O Lord of Kamal\=a! Come here. O Mah\=adeva! Come here; let all be well to you. You all have come to me for Gang\=a. Gang\=a has taken refuge under My feet out of fear for R\=adh\=a. Seeing Gang\=a by My side, R\=adh\=a wanted to drink Her up, However I

will give over Gang\=a to the hands of you all; but you will have to pray to R\=adh\=a, so that Gang\=a becomes fearless of Her.'' The lotus born Brahm\=a smiled at \'Sr\={\i} Kri\d{s}\d{n}a's words and began to sing hymns to R\=adh\=a, Who is fit to be worshipped by all. The Creator Brahm\=a, the Compiler of the Four Vedas, the Four-faced One praised R\=adh\=a with His Four heads, bent low and addressed Her thus :--

114-125. Brahm\=a said :-- ``O R\=adhe! Gang\=a, appeared from Thee and the Lord \'Sr\={\i} Kri\d{s}\d{n}a. Both of you were transformed before into the liquid forms in the R\=a\d{s}\=a Mandalam, on hearing the music of \'Sankara. And That Lquid Form is Gang\=a. So She is born of Thee and \'Sr\={\i} Kri\d{s}\d{n}a. Hence She is like Thy daughter and to be loved as such. She will be initiated in Thy Mantra and She will worship Thee. The four armed Lord of Vaikuntha will be Her husband. And when She will appear in parts on earth, the Salt Ocean will be Her husband. O Mother! The Gang\=a that dwells in Goloka, is dwelling everywhere. O Governess of the Devas! Thou art Her mother; and She is always Thy Self born daughter.'' Hearing thus, the words of Brahm\=a, R\=adh\=a gave Her assent towards the protection of Gang\=a. And then Gang\=a appeared from the toe-tip of \'Sr\={\i} Kri\d{s}\d{n}a. The liquid Gang\=a, then, assumed Her own form and, getting up from water, was received with great honour by the Devas. Bhagav\=an Brahm\=a took a little of that Ganges water in His Kamandalu and Bhagav\=an Mah\=adeva kept some of it in His own head.

The lotus born Brahm\=a, then, initiated Gang\=a into the R\=adh\=a Mantra and gave Her instructions, R\=adh\=a Stotra (hymn of R\=adh\=a) according to the S\=ama Veda, R\=adh\=a Kavacha (protection mantras), R\=adh\=a Dhy\=an (meditation on R\=adh\=a), method of worship of R\=adh\=a, and R\=adh\=a's pura\d{s}chara\d{n}a. Gang\=a worshipped R\=adh\=a according to those instructions and went to Vaikuntha. O Muni! Lak\d{s}m\={\i}, Sarasvat\={\i}, Gang\=a, and the world purifying Tulas\={\i}, these four became the wives of N\=ar\=aya\d{n}a. Kri\d{s}\d{n}a, then, smiled and explained to Brahm\=a the history of Time, hardly to be comprehended by others. He then spoke :-- ``O Brahm\=a! O Vi\d{s}\d{n}u! O Mahe\'svara! Now you better take Gang\=a and I will now tell you what a change has been effected by this time. Hear.

126-136. You, the three Devas, the other Devas, Munis, Manus, Siddhas, and other Mah\=atm\=as that are present here, are living now. For this region of Goloka is not affected by K\=ala (Time). Now the Kalpa is going to expire. So in the other regions than Goloka and Vaikuntha, the Brahm\=as, etc., that were existing in all other Universes, have all now dissolved in My Body. O Lotus-born! Save Goloka and Vaikuntha, all are now under water, the pre-state of earth. Better go and create your own

Brahm\=andas and Gang\=a will go to that newly created Brahm\=anda. I will also create other worlds and the Brahm\=as thereof. Now you all better go with the Devas and do your own works respectively. You have waited here for a long interval. As many Brahm\=as that have fallen all appear again.'' Thus saying \'Sr\={\i} Kri\d{s}\d{n}a, the Lord of R\=adh\=a went to His Inner Chamber. The Devas also instantly retired from that spot and engaged themselves earnestly in the creation work. Gang\=a remained as before till then in the region of Goloka, Vaikuntha, \'Sivaloka, Brahm\=a-loka, and in other places, by the command of \'Sr\={\i} Kri\d{s}\d{n}a. She is named Vi\d{s}\d{n}upad\={\i}, because She appeared from the feet of Vi\d{s}\d{n}u. Thus I have described to you this pleasant, essential story of Gang\=a, leading to liberation. What more do you now want to hear? Say.

Here ends the Thirteenth Chapter on the anecdote of Gang\=a in the Ninth Book in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



