\chapter{On the story of Gang\=a becoming the wife of N\=ar\=aya\d{n}a}

1-2. N\=arada said :-- O Lord! Gang\=a, Lak\d{s}m\={\i}, Sarasvat\={\i}, and the world purifying Tulas\={\i}, these four, are dearest to N\=ar\=aya\d{n}a. Out of these, Gang\=a went from the region of Goloka to Vaikuntha. So I have heard. But how did She come to be the wife of N\=ar\=aya\d{n}a. I have not heard. Kindly describe this.

3. N\=ar\=aya\d{n}a said :-- Brahm\=a came from Goloka to the region of Vaikuntha accompanied by Gang\=a.

4-23. Brahm\=a said to N\=ar\=aya\d{n}a :-- O Lord! Gang\=a, born of the bodies of R\=adh\=a and Kri\d{s}\d{n}a, full of youth, modest, extraordinarily beautiful, of pure Suddha Sattva, and void of anger and egoism, does not like to marry anyone save Thee as She is born of Thee. But R\=adh\=a is of a very proud nature and very wrathful. She was even ready to drink up Gang\=a. But Gang\=a at once and intelligently took refuge into the feet of \'Sr\={\i} Kri\d{s}\d{n}a. So the whole Goloka became void of water. Seeing this, I have come here to know in particular the whole history of the case. Then \'Sr\={\i} Kri\d{s}\d{n}a, the Knower and the Ruler of the hearts of all, came to know my heart and instantly caused Gang\=a to issue from His toe and handed Her over to me. I bowed down to \'Sr\={\i} Kri\d{s}\d{n}a and now I have come with Gang\=a to Thee. Now Thou dost marry the Sure\'svar\={\i} Gang\=a according to the G\=andharva rule of marriage. As Thou art a Deva of taste and humorous in the assembly of the Devas, so Gang\=a is. As Thou art a gem amongst the males, so She is the gem amongst the females. And the union of a humorous man with a humorous woman is exceedingly pleasant. Now marry this Lady who has come of Her own

accord to Thee. \'Sr\={\i} Mah\=a Lak\d{s}m\={\i} becomes annoyed with one who does not marry a woman who has come spontaneously. There is no doubt in this. The wise men do never insult the Prakriti. All the Puru\d{s}as (males) are born of Prakriti and all the females are parts of Prakriti. So Prakriti and Puru\d{s}a are both inseparable and verily one and the same. So these two should never insult each other. (If Thou sayest that Gang\=a is attached to Kri\d{s}\d{n}a; how canst Thou marry Her. The reply is) As \'Sr\={\i} Kri\d{s}\d{n}a is beyond the attributes and beyond Prakriti, so Thou art also above Prakriti. The one-half of \'Sr\={\i} Kri\d{s}\d{n}a is two-armed; the other half of \'Sr\={\i} Kri\d{s}\d{n}a is four-armed. R\=adh\=a has appeared from the left side of \'Sr\={\i} Kri\d{s}\d{n}a. He Himself is the right half and Padm\=a is His left-half. As there is no difference between R\=adh\=a and Kamal\=a so there is no difference between \'Sr\={\i} Kri\d{s}\d{n}a and Thee. Therefore as Gang\=a is born of Thy body, she wants to marry Thee. As Prakriti and Puru\d{s}a are really one and the same without any difference, so the males and females are one. Thus speaking to N\=ar\=aya\d{n}a, Brahm\=a handed Gang\=a over to N\=ar\=aya\d{n}a and went away. N\=ar\=aya\d{n}a, then, married Gang\=a smeared with sandal paste and flowers, according to the rules of the Gandarbha marriage. The Lord of Lak\d{s}m\={\i} then spent his time happily in enjoyment with Gang\=a. Gang\=a had to go to the earth (under the Bh\=arati's curse) and afterwards returned to Vaikuntha. As Gang\=a appeared from the feet of Vi\d{s}\d{n}u, She is denominated Visnupad\={\i}. Gang\=a Dev\={\i} was very much overpowered with enjoyment in Her first intercourse with N\=ar\=aya\d{n}a; so much so that She remained motionless. Thus Gang\=a spent the days happily with N\=ar\=aya\d{n}a. Sarasvat\={\i}'s jealousy towards Gang\=a did not disappear, though She was advised by Lak\d{s}m\={\i} Dev\={\i} not to do so. Sarasvat\={\i} cherished incessantly the feeling of jealousy towards Gang\=a. But Gang\=a had not the least feeling of jealousy towards Sarasvat\={\i}.

At last, one day, when vexed too much, Gang\=a became angry and cursed Sarasvat\={\i} to take Her birth in Bh\=arata. So Lak\d{s}m\={\i}, Sarasvat\={\i} and Gang\=a were the wives of N\=ar\=aya\d{n}a. Lastly Tulas\={\i} became the wife of N\=ar\=aya\d{n}a. So the number of wives of N\=ar\=aya\d{n}a amounted to four.

Here ends the Fourteenth Chapter in the Ninth Book on the story of Gang\=a becoming the wife of N\=ar\=aya\d{n}a in the Mah\=apura\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



