\chapter{On the origin of the Ganges}

1-3. The Devar\d{s}i N\=arada said :-- O Thou, the foremost of the knowers of the Vedas! I have heard the excellent narration of Earth. Now I want to hear the anecdote of Gang\=a. I heard, ere long, that Gang\=a, of the nature of Vi\d{s}\d{n}u and appearing from the feet of Vi\d{s}\d{n}u, the \=I\'svar\={\i} of the Devas, appeared, due to the curse of Bh\=arat\={\i}, on Bh\=arata; why has she come to Bh\=arata; in which Yuga and asked by whom did she come to Bh\=arata? O Lord! Now describe to me this auspicious anecdote capable to destroy sins and yield religious merits.

4-38. N\=ar\=aya\d{n}a said :-- O Child! In ancient days there was born a prosperous Emperor King of Kings, in the Solar dynasty. He had two beautiful wives; one was named Vaidarbh\={\i}, and the other was named \'Saivy\=a. \'Saivy\=a delivered a very lovely son; his name was Asamanj\=a. On the other hand, the queen Vaidarbh\={\i} desirous of getting a son, worshipped \'Sankara, the Lord of Bh\=utas who became pleased and granted her request; and Vaidarbh\={\i} became pregnant. After one hundred years of pregnancy she gave birth to one mass of flesh. Seeing this, the queen became very afflicted and taking refuge of Mah\=adeva, began to cry loudly and very often. Bhagav\=an \'Sankara, then, appeared there in a Br\=ahmi\d{n} form and cut that mass of flesh into thousand pieces. Those thousand pieces turned out into thousand very powerful sons. Their bodies looked more brillant than the mid-day sun. But they were all burnt to ashes by the curse of Kapila Muni. And the King began to lament bitterly and he entered into the forest. Asamaj\=a practised tapasy\=a to bring the Gang\=a for one lakh years when he quitted his body in course of time. His son Am\'sum\=an practised tapasy\=a for one lakh years to bring Gang\=a unto Bh\=arata and he, too, died. Then the son of Amsum\=an, the intelligent Bhag\={\i}ratha, a great devotee of Vi\d{s}\d{n}u, free of old age and death and the store of many qualifications, practised tapasy\=a for one lakh years to bring Gang\=a on earth. At last he saw \'Sr\={\i} Kri\d{s}\d{n}a brilliant like ten millions of summer suns. He had two hands; there was a flute in his hand; be was full of youth in the dress of a cow-herd. A sight of His Gop\=ala Sundar\={\i} form, wearing a Sakh\={\i}'s dress, makes one think that He is ever ready to show grace on His devotees. He is Para Brahm\=a, whose Substance is Will; be has no deficiencies. Brahm\=a, Vi\d{s}\d{n}u and Mahe\'svara and the other Devas and Munis, etc. all praise Him, who pervades everywhere. He is not concerned with anything; yet He is the Witness of all. He is beyond

the three gu\d{n}as, higher than Prakriti. A sweet smile is always in his face, which makes it the more lovely. There is none equal to him in showing Grace to the Bhaktas. His raiment is purified (uninflammable) by fire and he is decorated with gems, jewels and ornaments. The King Bhag\={\i}ratha saw that unforeseen appearance, bowed down and began to praise over and over again. His whole body was filled with ecstasy. Then he clearly told what he wanted for the deliverance of his family. Bhagav\=an \'Sr\={\i} Kri\d{s}\d{n}a then, addressed Gang\=a and said :-- ``O Sure\'svar\={\i}! Go quickly and appear in Bh\=arata, under the curse of Bh\=arat\={\i}. By My command go quickly and purify the sons of Sagara. They will all be purified by the touch of the air in contact with the Ganges and rise up in divine aerial cars, assuming forms like Mine and they will come to My abode. They will there remain always as My attendants and they will not be involved in the sins that they committed in their previous births. O N\=arada! It is stated thus in the Vedas, that if the human souls, taking their births in Bh\=arata, commit sins for millions and millions of births, the sins will be completely destroyed if they touch once the air in contact with and carrying the particles of the Ganges. The sight of the Ganges and the touch of the Ganges water give religious merits ten times more than the touch with the air in contact with the Ganges water. People become freed of their sins then and there especially if they bathe in the Ganges. It is heard in the \'Srutis that the bathing in the Ganges, if done according to rules, destroys all the sins, e.g., the murder of a Br\=ahmi\d{n}, etc., acquired in one thousand millions of births done consciously or unconsciously. The merits that are acquired by the bathing in the Ganges on a day of religious merit, cannot be described even by the Vedas. Whatever is mentioned in the \=Agamas is but a mere trifle. Even Brahm\=a, Vi\d{s}\d{n}u and Mahe\'sa cannot describe fully the merits of the bathing in the Ganges. O Br\=ahmi\d{n}! Such is the glory of ordinary bathing. Now I will describe the effect of the Ganges bath done with a Sankalpa (resolve); hear. Ten times more the result is obtained when the Ganges bathing is done with a resolve (Sankalpa) than ordinary bath and if one bathes on the day when the sun passes from one sign to another (in the Zodiac), thirty times more religious merits accrue. On the new Moon (Am\=avyas) day, the Ganges bath gives the merits as above mentioned; but when the Sun is in his Southern course (Dak\d{s}i\d{n}\=aya\d{n}a) double the merits are obtained and when the sun is in his Northern course, ten times more religious merits are obtained. The Ganges bathing in the time of Ch\=aturm\=asya, full Moon day, Ak\d{s}ay\=a Navam\={\i} or Ak\d{s}ay\=a tritt\={\i}y\=a yields merits that cannot be measured. And if on the above Parva (particular periods of the year on which certain ceremonies are commanded) days both bathing, and making

over gifts are done, there is no limit to the religious merits acquired; hundred times more than ordinary bath, religious merits are obtained. Great religious merits accrue from the Ganges bath on Manvantar\=a tithi, Yug\=ady\=a, \'Sukl\=a seventh day of the month of M\=agha, Bh\={\i}sm\=astam\={\i} day, A\'sok\=astam\={\i} day, and \'Sr\={\i} R\=ama Navam\={\i} day. Again double the merits than those of the above arise from bathing in the Ganges during the Nand\=a ceremony. The Gauges bath in the Da\'sahar\=a tenth tithi gives merits equivalent to Yug\=ady\=a Sn\=anam (bath). And if the bathing be done on Mah\=anand\=a or in Mah\=av\=aru\d{n}\={\i} day, four times more religious merits accrue. Ten million times more religious merits accrue from the Ganges bathing on Mah\=a Mah\=a V\=aru\d{n}\={\i} day than ordinary bath. The Ganges bath in the Solar eclipse yields ten times more religious merits than in the Lunar eclipse. Again the Sn\=anam in Ardhodaya Yoga yields hundred times more religious merits than that of the (solar) eclipse. Thus saying to Gang\=a before Bhag\={\i}ratha, the Lord of the Devas remained silent. The Dev\={\i} Gang\=a with her head bowed down with devotion, said :--

39-42. Gang\=a said :-- If I am after all, to go to Bh\=arata as Thou commandest and under the curse given previously by Bh\=arat\={\i}, then tell me how I would be freed of the sins that the sinners will cast on me. How long will I have to remain there? When, O Lord! Shall I be able to return to the Highest place of Vi\d{s}\d{n}u? O Thou, the Inner Self of all! O All Knowing! O Lord! Whatever else I desire, Thou knowest them all. So be pleased to instruct me on all these points.

43-69. Bhagav\=an \'Sr\={\i} Hari said :-- O Sure\'svar\={\i}! I know all that you desire; when you will assume the liquid form, the Salt Ocean will be your husband. He is My part and you are of the nature of Lak\d{s}m\={\i}; so the union of the lover with the love stricken in the world will turn out a happy and qualified one. Of all the rivers Sarasvat\={\i} and others in Bh\=arata, that go to mix with the ocean, you will be the best and highly fortunate of them all. From today you will have to remain in Bh\=arata for a period of five thousand years, under the curse of Bh\=arat\={\i}. You will be able to enjoy daily and always the pleasures with the Ocean. O Dev\={\i}! As you are a clever lady, so He is also apt and expert. The inhabitants of Bh\=arata will praise Thee and worship Thee with great devotion by the stotra which Bhag\={\i}ratha has composed. He will derive the fruit of one horse-sacrifice who will meditate on Thee as per K\=a\d{n}va\'s\=akh\=a and worship, praise and bow down to Thee daily. Even if one utter ``Gang\=a,'' ``Gang\=a,'' though one is one hundred Yojanas away from the Ganges, one will be freed of all sins and go to Visnu-loka. Whatever sins will be cast in Thee by thousand sinners bathing in Thee will be destroyed by the touch of the devotees of Prakriti Dev\={\i}. Even if

thousands and thousands of sinners touch the dead bodies and bath in Thee, all those will be destroyed when the Dev\={\i} Bhaktas,the worshippers of Bhuvane\'svar\={\i} and M\=ay\=a V\={\i}ja, will come and touch Thee. O Auspicious One! Thou wilt wash away the sins of the sinners, by Thy stay in Bh\=arata with other best rivers Sarasvat\={\i} and others. That will be at once a sacred place of pilgrimage where Thy glories will be chanted. By the touch of the dust of Thine, the sinner will be at once purified and he will dwell in the Dev\={\i}-loka (Mani Dv\={\i}pa) for as many years as will be the dust there. All Hail to the Dev\={\i} Bhuvane\'svar\={\i}! He who will leave his body on Thy lap with full consciousness and remembering My name, will certainly go to My abode and will remain, as My chief attendant for an infinite period. He will see countless Pr\=akritik Layas (dissolutions of the Universe). Unless a man has collected hordes of religious merits, be cannot die in the Ganges; and if he dies on the Ganges he goes to Vaikuntha as long as the Sun rises in this world. I get many bodies for him where he can enjoy the fruits of his Karma and I then give him My Sv\=ar\=upya (Form resembling Mine) and make him My attendant. If any ignorant man, void of any J\~n\=anam, quits his body by touching merely Thy water, I give him S\=alokya (place in My region) and make him My attendant. Even if one quits one's body in a far off place, uttering Thy Name, I give him place in My region for one life time of Brahm\=a. And if he remembers Thee with devotion, and quits his mortal coil at another place, I give him S\=ar\=upya (Form resembling Mine) for a period of endless Pr\=akritik Layas. He instantly gets on an aerial car made of jewels and goes with My attendants to the region of Goloka and gets form like Me. Those that worship daily My Mantra, that pass their day eating the remnant of the food offered to Me, they need not have any distinction whether they die in T\={\i}rath or not. They themselves can easily purify the three worlds. Getting on the excellent and best aerial car built of jewels, they go to the region of Goloka. O Chaste One! Even if the friends of My devotees, be born in animal births, they also will be purified by the devotion shown towards Me and getting on a jewelled aerial car will be able to go to Goloka, so diffcult of access. Wherever the Bhaktas may be, if they simply remember Me with devotion, they will become liberated while living by the power of My Bhakti. Thus saying to Gang\=a, Bhagav\=an \'Sr\={\i} Hari addressed Bhag\={\i}ratha :-- O Child! Now worship Gang\=a Dev\={\i} with devotion and chant hymns to Her. The pure Bhag\={\i}rata meditated with devotion as per Kauthuma \'S\=akh\=a and worshipped the Dev\={\i} and praised Her repeatedly. Then Gang\=a and Bhag\={\i}ratha bowed down to \'Sr\={\i} Kri\d{s}\d{n}a and He disappeared at once from their sight.

70. The Devar\d{s}i N\=arada said :-- ``O Thou the foremost of the Knowers of the Vedas! How, by what Kuthuma \'S\=akh\=a, the noble King Bhagi-

ratha meditated on the Dev\={\i} Gang\=a; what stotra did he recite and what was the method with which he worshipped the Ganges.

71-75. N\=ar\=aya\d{n}a said :-- O N\=arada! One should first take one's bath, and putting on a clean washed clothing should perform one's daily duties. Then one should control oneself and with devotion worship the six Devat\=as Ga\d{n}e\'sa, Sun, Fire, Vi\d{s}\d{n}u, \'Siva and \'Siv\=an\={\i}. Thus one becomes entitled to worship. First worship is to be given to Ga\d{n}e\'sa for the destruction of obstacles; the Sun is to be next worshipped for health; Fire, for purification; Vi\d{s}\d{n}u is then worshipped for getting wealth and power; \'Siva is worshipped for knowledge and \'Siv\=an\={\i} is worshipped for Mukti. When these Devat\=as are worshipped, one is entitled to worship the Deity. Otherwise contrary become the effects. Now I am saying what Dhy\=a\d{n}am (meditation) did Bhag\={\i}ratha practise towards the Dev\={\i} Gang\=a.

Here ends the Eleventh Chapter of the Ninth Book on the origin of the Ganges in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



