\chapter{On the origin of Gang\=a}

1-15. N\=ar\=aya\d{n}a said :-- O N\=arada! Now about the meditation (Dhy\=an) of the Dev\={\i} Gang\=a as per K\=a\d{n}va \'S\=akh\=a, which destroys all the sins, O Gange! Of white colour like white lotuses! Thou destroyest all the sins of men. Thou hast appeared from the body of \'Sr\={\i} Kri\d{s}\d{n}a. Thou art powerful like Him. Thou art very chaste and pure. Thou hast worn the raiment, uninflammable and decorated all over with ornaments made of jewels. Thou art more brilliant than one hundred autumnal Moons. Thou art also well pleased with a smile on Thy lips. And Thou art always of steady youthful beauty (that never wanes). Thou art dear to N\=ar\=aya\d{n}a, calm and of peaceful temper, and proud of being His with His fortune. Thou bearest the braid of hair, decked with garlands of M\=alat\={\i} flowers; Thy cheeks are anointed with sandal dots, with Sind\=ura bindu (dots of red powder, vermilion) and well adorned with various artistic lines made of musk. Thy garment and Thy beautiful lips are more red than the ripe Bimba fruit (the red fruit of a cucurbitaceous plant); Thy teeth vie as it were, with the rows of pearls. How lovely are Thy eyes! How delightsome is Thy side-long glance! How close are Thy breasts like Bel fruits! Thy loins are thicker and more solid then the plantain trees. How do Thy feet look beautiful, defying the beauty of the Sthalapadma (ground Lotus)!

How do the red sandals look lovely with Kunkuma and alaktak (red powder)! What a red tinge Thy feet have shewn with the honey of

P\=arij\=ata flower that is seen on the head of Indra. The Devas, the Siddhas, the Munis, offer always Arghyas (offerings of rice with Durba grass) at Thy feet; the ascetics bow down at Thy feet, and it seems as though so many lines of bees are on Thy lotus feet. O Mother! Thy lotus feet give liberation to those that want Mukti and enjoyment to those that want Bhukti (enjoyments). O Mother! Thou art the boon; Thou art the chief excellent; Thou grantest boons and Thou showest Thy favour to Thy devotees; Thou bestowest the Vi\d{s}\d{n}upadam (the place of Vi\d{s}\d{n}u); but Thou hast come from the feet of Vi\d{s}\d{n}u. Thus meditating on the Dev\={\i} Gang\=a flowing by three routes (in Heaven, earth and infernal regions), the bestower of good things one should offer to the Dev\={\i} sixteen things :-- \=Asana, P\=adya, Arghya, water for bathing, ointment (an\=ulepana), Dh\=upa (scents), D\={\i}pa (lights), Naivedya (offerings of food), betel, cool water, clothings, ornaments garlands, sandal-paste, \=Achaman\={\i}ya (water for sipping), and beautiful beddings and worship Her with these. Then, with folded hands, one should perform stotra to Her and bow down to Her with devotion. Thus the worshipper gets the fruits of A'svamedha sacrifice.

16. N\=arada said :-- O Lord of the Devas! At present I am desirous to hear the sin-destroying and virtue-bestowing stotra (hymn) of Gang\=a Dev\={\i}, the Purifier of all those that are fallen from virtue, originated from the feet of Vi\d{s}\d{n}u, the Lord of world and the husband of Lak\d{s}m\={\i}. Kindly narrate all these in detail.

17-41. N\=ar\=aya\d{n}a said :-- O N\=arada! Now I am narrating the stotra of Gang\=a Dev\={\i}, that destroys all sins and bestows all religious merit. Hear. I bow down to the Ganges who appeared from the body of \'Sr\={\i} Kri\d{s}\d{n}a, enchanted by the music of \'Siva, and, who was bathed with the perspiration (water coming out of the body) of \'Sr\={\i} R\=adh\=a. I bow down to Gang\=a Dev\={\i} who first appeared in the circular dance (R\=asa M\=andalam) in the region of Goloka and who always remains with \'Sankara. My obeisance to the Dev\={\i} Gang\=a who remains in the auspicious grand utsab of R\=adh\=a (R\=asa M\=andalam), crowded with Gopas and Gop\={\i}s, in the Full Moon night of the month of K\=artik. She is one koti yojanas wide and one lakh times one koti yojanas long in the region of Goloka. My Obeisance to Her! In Vaikuntha, Gang\=a is sixty lakh yojanas in width and four times that in length. My Obeisance to Her! In Brahm\=a-loka, Gang\=a is thirty lakh yojanas wide and five times as long. I bow down to Her. In \'Siva-loka, She is thirty lakh yojanas wide and four times that in length. I bow down to Her. In Dhruva-loka, She is one lakh yojanas wide and seven times as long. I bow down to Her. In Chandra-

loka She is one lakh yojanas wide and five times as long. My obeisance to \'Sr\={\i} Gang\=a Dev\={\i}. I bow down to the Ganges who is sixty thousand yojanas wide in the S\=urya loka and ten times that in length. I bow down to Gang\=a in Tapo-loka who is one lakh yojanas wide and five times that in length. My obeisance to Gang\=a Dev\={\i} in Janar-loka, who is one thousand yojanas wide and ten times that as long. I bow down to Gang\=a in Mahar-loka who is ten lakh yojanas wide and five times that in length. My obeisance to Gang\=a Dev\={\i} in Kail\=a\'sa who is one thousand yojanas wide and one hundred times as long. I bow down to Gang\=a Dev\={\i} who is known as Mand\=akin\={\i} in Indra-loka, and who is one hundred yojanas wide and ten times than that in length. My obeisance to Gang\=a Dev\={\i}, known as Bhogavati in P\=at\=ala who is ten yojanas wide and five times as long. I bow down to Gang\=a Dev\={\i}, known as Alakanand\=a in this earth, who is two miles wide, in some places more wide and in some places less wide. I bow down to Gang\=a Dev\={\i} who was of the colour of milk in Satya yuga, of the colour of Moon in Tret\=a Yuga, of the colour of white sandal-paste in Dv\=apara yuga. I bow down to \'Sr\={\i} Gang\=a Dev\={\i} who is as water in Kali yuga in this earth and as milk in Kali yuga in Heaven. O Child! By the touch of one molecule of the water of the Ganges, all the horrible sins incurred in ten million births, the murder of a Br\=ahmi\d{n} and so forth, are burnt to ashes. Thus I have described in twenty-one verses the great stotra (human) of the sin-destroying and the virtue-increasing merits of Gang\=a. He reaps the fruit of the A'svamedha sacrifice (Horse sacrifice), who daily sings this praise of Gang\=a after worshipping Her with devotion. There is no doubt in this. The persons that are without any sons get sons hereby and those who have no wives get wives. The diseased get themselves free from their diseases, and the man who is under bondage, is liberated from that bondage. He who getting up early in the morning reads this stotra of Gang\=a, becomes widely known even if he be not known at all and he becomes illumined with wisdom even if he be quite ignorant. Even if he sees a bad dream, he acquires the merit of bathing in the Ganges and of seeing good dreams.

42-44. \'Sr\={\i} N\=ar\=aya\d{n}a spoke :-- O N\=arada! With this stotra (hymn) did Bhag\={\i}ratha praise the Gang\=a Dev\={\i}. Who then went with him to the spot where the Sagara's sons were burnt to ashes by the curse of Kapila. By the contact of the wind in touch with the particles of water of the Ganges, those sons of Sagara were instantly freed of their curses and they all repaired to Vaikuntha. She is named Bh\=agirath\={\i}, because Bhag\={\i}ratha brought Her to this earth. Thus I have described to you the story of the Ganges.

This anecdote is highly meritorious and the great step to liberation. What more do you now want to hear? Say.

45-46. N\=arada said :-- O Lord! How did Gang\=a come to flow through the three worlds by three routes, and thus purify them? How was she carried and to which places? How did the people of those localities accord respect to Her? Kindly describe all these in detail.

47-79. N\=ar\=aya\d{n}a said :-- O N\=arada! On the Full Moon night of the month of K\=artik in the R\=asa m\=andalam, at the great festivity in honour of R\=adh\=a, \'Sr\={\i} Kri\d{s}\d{n}a worshipped R\=adh\=a and remained there. Next R\=adh\=a, worshipped by \'Sr\={\i} Kri\d{s}\d{n}a, was worshipped by Brahm\=a and the other Devas, by Saunaka and the other \d{R}i\d{s}is, who also stopped there with much gladness. At this moment the Dev\={\i} Sarasvat\={\i}, the Presiding Deity of the Science of Music began to sing lovely songs regarding Kri\d{s}\d{n}a, in tune with vocal and instrumental music. Brahm\=a became glad and presented to Sarasvat\={\i} a necklace of jewels; Mah\=a Deva gave her gems and jewels rare in this universe; Kri\d{s}\d{n}a presented the best Kaustubha jewel; R\=adhik\=a offered excellent invaluable necklace of jewels; N\=ar\=aya\d{n}a presented to her the best and most excellent garland of jewels; Lak\d{s}m\={\i} gave her invaluable golden earrings decked with gems; Vi\d{s}\d{n}u-M\=ay\=a M\=ula Prakriti, Bhagvat\={\i} Durg\=a, who is N\=ar\=aya\d{n}\={\i}, \=I\'svar\={\i}, \=I\'s\=an\={\i}, presented Her devotion to Brahm\=a, so very rare; Dharma gave her devotion to Dharma and high fame; Agni (fire) gave her excellent raiments purified by fire and V\=ayu gave Her N\=upura (toe ornaments) made of gems and jewels. At this time, Mahe\'svara, the Lord of Bh\=utas (elements) began to sing, at the suggestion of Brahm\=a, songs relating to \'Sr\={\i} Kri\d{s}\d{n}a's grand R\=asa festival. Hearing this, the Devas became very much enchanted and remained motionless like statues. With great difficulty, they regained their consciousness. Then they saw that there was no R\=adh\=a nor Kri\d{s}\d{n}a in this R\=asa m\=andala; everything was deluged with water. The Gopas, Gop\={\i}s, Devas and Br\=ahma\d{n}as began to cry loudly. Brahm\=a in his meditation then came to know that R\=adh\=a and Kri\d{s}\d{n}a both have assumed this liquid appearance for the deliverance of the people of the world. Brahm\=a and others, all began to praise \'Sr\={\i} Kri\d{s}\d{n}a and said :-- ``O All prevading One! Now be pleased to show us Thy form and grant us our desired boons.'' At that instant a sweet incorporeal voice was clearly heard by all, as coming from air above, that, ``I am the Self of all, pervading all; and this my \'Sakti, R\=adh\=a, is also the Self of all, prevades all; so there is no separation of us from you all even for a moment. It is only to show our favour to the devotees that we assume special forms. For this reason only there is separation of us from you as regards this body! There is nothing else. Besides you have no necessity with our bodies. O Devas! Now if my Manus, men, Munis, Vai\d{s}\d{n}avas and you all, purified by Mantras desire very much to see My Form clearly, then I tell you to request Mahe\'svara to carry out My word. O Brahm\=a! O Creator! Better ask

Mah\=adeva, the World Teacher, that He would better compose the beautiful Tantra \'S\=astra, in accordance with the limbs of the Vedas. And that the above \'S\=astra be full of Mantras, capable to yield desired fruits, Stotras (hymns) and Kavachas (protection mantras) and rules of due worship in proper order. And that also My Mantra, My Stotra, and My Kavacha be also given there in a hidden form. So that those people that are sinners might not understand their real meanings and thus turn out against Me. It may be that one in a thousand or in a hundred may worship My Mantra. And My Mantra worshippers, the saints, become purified and come to My Abode. If My \'S\=astra be not well made (i. e., if everyone be able to understand its meaning) and if everyone be able to go from Bh\=urloka to Goloka, then Thy labour in this creation of the world will all be in vain. Therefore dost Thou better create different worlds according to the differences of S\=attvik, R\=ajasik, and Tamo Gu\d{n}as; then some will be the inhabitants of this Bh\=urloka, some will be the inhabitants of Dyuloka according to their Karmas. O Br\=ahma\d{n}! If Mah\=a Deva promises earnestly in this assembly of the Devas, I will then exhibit My True Form.'' O N\=arada! Thus speaking, the Eternal Puru\d{s}a \'Sr\={\i} Kri\d{s}\d{n}a remained silent (i. e., the aerial incorporeal voice stopped). Hearing this, Brahm\=a, the Creator of the world, gladly informed \'Siva of this. When the Lord of Knowledge, the Foremost of the J\~n\=anins, Bh\=utan\=atha heard the words of the Creator, He took the Ganges water in His hands and swore that ``I will complete the Tantra \'Sastra, full of R\=adh\=a mantras and not opposed to the Vedas. If one touches the Ganges water and speaks lies, one remains in the terrible K\=alasutra hell for a period of one Brahm\=a's life time.'' O Dvija! When Bhagav\=an \'Sankara said this before the assembly of the Devas in the region of Goloka, \'Sr\={\i} Kri\d{s}\d{n}a appeared there with R\=adh\=a. The Devas became exceedingly glad to see Him. They praised Him, the Best Puru\d{s}a and they were all filled with rapture and again engaged themselves in the grand R\=asa Festival. Some time after, M\=ah\=a Deva lighted the Torch of Mukti, i.e., the Tantra \'S\=astra was published by Him, as promised. O Child! Thus I have disclosed to you this anecdote, so very secret, and hard to be attained. Thus \'Sr\={\i} Kri\d{s}\d{n}a Himself, is verily the liquid Gang\=a sprung in the region of Goloka. This holy Gang\=a, born of the bodies of Kri\d{s}\d{n}a and R\=adh\=a inseparable from each other, grants enjoyment, lordship and liberation. \'Sr\={\i} Kri\d{s}\d{n}a, the Highest Self, has placed Her in various places; so Gang\=a is of the nature of \'Sr\={\i} Kri\d{s}\d{n}a and is everywhere, equally honoured everywhere in the Brahm\=anda (universe).

Here ends the Twelfth Chapter in the Ninth Book on the origin of Gang\=a in the Mah\=apur\=a\d{n}am \'Sr\={\i}mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



