\chapter{On the going of the Devas to Vaikuntha after Tulas\={\i}'s marriage with \'Sankhach\=uda}

1. N\=arada said :-- ``O Bh\=agav\=an! Wonderful is the story that has been now recited by you. My ears are not satisfied. So tell me what happened afterwards.''

2-94. N\=ar\=aya\d{n}a said :-- O N\=arada! The Creator Brahm\=a, blessing them, departed to His own abode. The D\=anava married Tulas\={\i} under the Gandharba method. The celestial drums sounded and the flowers were showered. In the beautiful lovely house the D\=anavendra, remained in perfect enjoyment. Tulas\={\i}, too, being busy with fresh intercourses, became almost mad after them. The chaste Tulas\={\i} and \'Sankhach\=uda both became deeply immersed in the ocean of bliss in their sexual union and began to enjoy sixty-four sorts of amorous sports. In the \'S\=astras on love affairs, all the connections of limbs with limbs that are described, as the lover and the loved desire, they both enjoyed those with perfect freedom and pleasure. The place was solitary; to add to it, the scenery was grand and lovely; so nothing remained untasted of the several tastes of amorous pleasures. On the banks of the river, in flower-gardens, they slept on the flower beds smeared with sandal-paste, and enjoyed the amorous pleasures. Both were adorned with

jewel ornaments; both were skilled in amorous practices; so no one desisted. The chaste Tulas\={\i} out of her nimbleness due to young age, easily stole into the heart of her husband. \'Sankhach\=uda, too, a great expert in knowing other's amorous sentiments, attracted the heart of Tulas\={\i}. Tulas\={\i} obliterated the sandal marks from the breast of the King and the sign of tilak from his nose. The King also wiped away the dot of Sindur and Alak\=a (vermillion) marks from Tulas\={\i}'s forehead and put marks of nails on her round plump breasts. Tulas\={\i} also hurt the King's left side by her bracelets. Then the King bit the lips of Tulas\={\i}. Thus each one embraced the other, kissed each other and each one began to champoo the thighs, legs, etc. When both of them thus spent their time in amorous sports, they got up and began to dress themselves as they desired. Tulas\={\i} smeared \'Sankhach\=uda's nose with red sandal-paste mixed with kumkum (saffron), smeared his body with sweet-scented sandal-paste, offered sweet-scented betels in his mouth, made him put on celestial garments (fireproof; brought from Fire) put unto his neck the wonderful garland of P\=arij\=ata flowers, destructive of disease and old age, invaluable jewel rings on his hand and offering him excellent gems, rare in the three worlds, said :-- ``O Lord! I am your maidservant'' and uttering this repeatedly bowed down to the feet of her husband with devotion. She then got up and with smiling countenance began to look on his face with a steadfast gaze. The king \'Sankhach\=uda then attracted his dear Tulas\={\i} to his breast and took off the veil fully from her face and began to look on that, next moment he kissed on her cheek and lips and gave her a pair of garments brought from the Varu\d{n}a's house, a necklace of jewels, hard to get in the three worlds, the tinklets of Sv\=ah\=a, the wife of Ag\d{n}i, the Keyura (armlets) of the Sun's wife Chh\=ay\=a, the two earrings of Rohi\d{n}\={\i}, the wife of the Moon, the finger rings of Rati, the wife of K\=amadeva, and the wonderfully beautiful conch, given by Vi\'sva Karm\=a, excellent bedding studded with pearls and jewels and various ornaments; and when he gave her all these things, he smiled. The king then put garlands on Tulas\={\i}'s braid of hair, nicely variegated Alak\=as on her cheek, three crescent lines of sweet-scented sandal paste within the Alak\=as, dots of saffron all around that, the brilliant Sindura mark looking like a flame, and red \=Alt\=a on the feet and toes; he then placed those feet on his breast and utterred repeatedly :-- ``I am your servant'' and then held her on his breast. They then left the hermitage, in that state and began to travel in various places. In the Malaya mountain, in mountains after mountains, in solitary flower gardens, in the mountain caves, in beautiful sea-beaches, on the banks of the Pu\d{s}pabhadr\=a river, cool with

watery breeze, in various rivers and riversides, in Vi\d{s}pandana forest echoed with sweet songs of the birds of the vernal season. They then went from Vi\d{s}pandana forest to the Surasana forest, from the Surasana, forest to the Nandana forest, from the Nandana forest to the nice Chandana forests, from Chandana forest to Champaka, Ketak\={\i}; M\=adhav\={\i} Kunda, M\=alat\={\i}, and Kumuda and lotus forests; thence they went to the forest of desire gratifying trees (Kalpavrik\d{s}a forest,) and P\=arij\=ata trees. They then went to the solitary place K\=anchan, thence to the K\=anch\={\i} (forest) they then went to the K\={\i}\~njalaka forest, thence to the K\=anchan\=akar (the gold mine), thence to Kanchuka and various other forests echoed with the sweet sounds of cuckoos. There, on beds strewn with flowers and scented with sandal paste they both enjoyed each other to their hearts content and with great pleasure. But none of them, whether \'Sankhach\=uda or Tulas\={\i}, got quenched with their thirst. Rather their passions were inflamed like the fire on which clarified butter is poured (in sacrifices). The King of the D\=anavas, then, brought Tulas\={\i} to his own kingdom and, there, in his own beautiful garden house, he incessantly enjoyed her. Thus the powerful king of the D\=anavas passed away one Manvantara in the enjoyment of his kingdom. He spread his sway over the Devat\=as, Asuras, D\=anavas, Gandharbas, Kinnaras, and R\=ak\d{s}asas. The Devas, dispossessed of their realms, wandered everywhere like beggars. At last they united in a body and went to the Brahm\=a's assembly and there they began to cry and then related the whole history how the D\=anava \'Sankhach\=uda oppressed them. Hearing all this, Brahm\=a took them to \'Sankara and informed Him of the whole history of the case. When Mah\=adeva heard all this, He took them all to the highest place, Vaikuntha devoid of old age and death. Going towards the first entrance of N\=ar\=aya\d{n}a's abode, they saw the gate-keepers watching the gate, taking their seats on jewel thrones. They all looked brilliant, clothed with the yellow garments, adorned with jewel ornaments, garlanded with forest flowers, all of \'Sy\=ama Sundara (dark blue, very beautiful) bodies. They were four-armed, holding on their hands, conch, mace, discus and lotus; sweet smile was on their faces and eyes beautiful like lotus leaves. On Brahm\=a asking them for entrance to the assembly, they nodded their assent. He, then, accompanied by the Devas, passed one by one, sixteen gates and at last came before N\=ar\=aya\d{n}a. On reaching there, He saw that the assembly was completely filled with Devar\d{s}is, and four-armed N\=ar\=aya\d{n}like P\=ari\d{s}adas (attendants), decked with Kaustubba jewels. The sight of the Sabh\=a (assembly) makes one think that the Moon has just arisen, shedding effulgent rays all round. By the will of \'Sr\={\i} Hari, excellent diamonds, invaluable gems and necklaces of gems and jewels were

placed at various places. At other places rows of pearls were shedding their splendour and brilliance like the garlands of gems and jewels. At others, the mirrors were placed in a circle; and at various other places, the endless wonderful artistic picture lines were drawn. Again at other places the jewels called Padmar\=agas were artistically arranged as if the lotuses were there spreading their lustrous beauty all around. At many other places rows of steps were made of wonderful Syamantak jewels. All around the assembly, there were the excellent pillars, built of Indran\={\i}lam jewels. Over those pillars, sandal leaves strung on strings from to pillar to pillar, were suspended. Golden jars, all brimful with water were located at various places. All around, the garlands of P\=arij\=ata flowers were seen. The hall was decorated with sweet scented sandal trees, red like saffron and musk. Sweet scents were being emitted all round. The Vidyadh\=aris were dancing at places. The assembly hall measured one thousand Yojanas. Countless servants were engaged all over on various works. Brahm\=a, \'Sankara, and the other Gods saw there \'Sr\={\i} Hari seated in the centre on an invaluable jewel throne, as a Moon looks surrounded by stars. There were the crown on His head, the ear-rings on His ears; garlands made of wild flowers were on his neck and His body was smeared all over with sandal paste and He was holding Kelipadma (a sort of lotus) in His hand. He was seeing, with a smiling countenance, the dancing and music before Him. He was full of peace, the Lord of Sarasvat\={\i}. Laksm\={\i} was holding gently His lotus feet and He was chewing the sweet scented betel offered by Her. Gang\=a also was fanning Him devotedly with a white Ch\=amara and the others were singing hymns to Him with their heads bent low with devotion. Brahm\=a and the other Gods all bowed down to Him; their bodies were all filled with Pulaka (excessive joy causing hair stand on end); tears flowed from their eyes and their
voices were choked out of emotion. The creator Brahm\=a, then, with clasped hands informed Him, with head bowed down, of the whole history of \'Sankhach\=uda. Hearing this, the omniscient Hari, knowing the minds of all, smiled and spoke to Brahm\=a all the interesting secrets :-- O Lotus born! I know all about \'Sankhach\=uda. He was in his previous birth My great devotee, an energetic Gopa. Now I speak to you the ancient history of Goloka; hear. This story about Goloka is sin destroying and highly meritorious. \'Sankhach\=uda, in his previous birth was the Gopa Sud\=am\=a, My chief P\=ari\d{s}ad (attendant). He has now become a D\=anava on account of the dire curse pronounced by \'Sr\={\i}
R\=adh\=a. One day when I went from My abode, accompanied by Viraj\=a Gop\={\i}, to the R\=asa Mandala, My beloved R\=adh\=a, hearing this news from a maid servant, came up at once with Her whole host of Sakh\={\i}s

wrathful, to the R\=asa Mandalam (ball dance in Goloka) and, not being able to see Me, saw Viraj\=a turned into a river, She thought that I had disappeared. So She went back to Her own abode with Her Sakh\={\i}s. But when I returned to the house with Sud\=am\=a, R\=adh\=a rebuked Me very much. I remained silent. But Sud\=am\=a could not bear and he rebuked R\=adh\=a in My presence, a thing quite intolerable to Her dignity! On hearing this rebuke, R\=adh\=a's eyes became red with anger and She immediately ordered Her Sakh\={\i}s to drive him away. Sud\=am\=a began to tremble with fear. Immediately on Her command lakhs and lakhs of Sakh\={\i}s got up immediately and drove that hot irresistible Sud\=am\=a away. Sud\=am\=a repeated his chafings and roarings. On hearing these, She cursed him :-- ``You better be born in the womb of a D\=anav\={\i}.'' Hearing the terrible curse, Sud\=am\=a bowed down to Me and went away crying; then R\=adh\=a, who was all-mercy, became melted with mercy. And She prevented him repeatedly, not to go away. R\=adh\=a wept and told him, ``O Child! Wait. Where are you going? No more you will have to go; return.'' Thus saying She became very distressed. The Gopas and Gop\={\i}s also began to weep. I then explained to them, ``In about half a moment Sud\=am\=a will come back, fulfilling the conditions of the curse. O Sud\=am\=a! Come here when the curse expires.'' Then he appeased R\=adh\=a also. ``Know that one moment (K\d{s}a\d{n}) in Goloka is equal to one Manvantara on earth. The Yogi \'Sankhach\=uda, expert in M\=ay\=a and very powerful will soon return from the earth. Take this My weapon \'S\=ula and go early to Bh\=arata. \'Siva will slay the D\=anava by this \'S\=ul\=astra. The D\=anava holds always on his neck My auspicious Kavacha and will therefore become the conqueror of the universe. No one will he able to kill him as long as be holds the above Kavacha. So, first of all, I will go to him in the form of a Br\=ahma\d{n}a and ask from him the Kavacha. O Creator! Thou also didst give him the boon that his death would occur when the chastity of of his wife would be destroyed. I will go and hold intercourse with his wife. Then his death will occur without fail. His wife after her death will come and become My dearest wife. Thus saying, N\=ar\=aya\d{n}a gave over to Mah\=adeva the \'S\=ul\=astra. Then He went gladly to His inner compartments. On the other hand, Brahm\=a and Rudra and the other Devas incarnated themselves in Bh\=arata.

Here ends the Nineteenth Chapter on the going of the Devas to Vaikuntha after Tulas\={\i}'s marriage with \'Sankhach\=uda in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



