\chapter{On the fight between the Devas and \'Sankhach\=uda}

1-75. \'Sr\={\i} N\=ar\=aya\d{n}a spoke :-- Then the King of the D\=anavas, very powerful, bowed down to Mah\=adeva and ascended on the chariot with his ministers. Mah\=adeva gave orders to His army to be ready at once. So \'Sankhach\=uda did. Terrible fight then ensued between Mahendra and Vri\d{s}aparv\=a, Bh\=askara and Viprachitti, Ni\'s\=akara and Dambha, between K\=ala and K\=ale\'svara, between Fire and Gokar\d{n}a, Kuvera and K\=alakeya, between Vi\'svakarm\=a and M\=ay\=a, between Mrityu and Bhayamkar, between Yama and Samh\=ara, between Varu\d{n}a and Vikamka, between Budha and Dhritapristha, between \'Sani and Rakt\=ak\d{s}a, Jayanta and Ratnas\=ara, between the Vasus and Varchasas, between the two A\'svin Kum\=aras and D\={\i}ptim\=an, between Nalak\=ubara and Dh\=umra, between Dharma and Dhurandhara, between Mangala and U\d{s}\=ak\d{s}a, Bh\=anu and \'Sov\=akara, between Kandarpa and P\={\i}thara, between the eleven \=Adityas and Godh\=amukha, Ch\=ur\d{n}a and Khadgadhvaja, Ka\~nch\={\i}mukha and Pinda, Dh\=umra and Nand\={\i}, between Vi\'sva and Pal\=a\'sa, between the eleven Rudras and the eleven Bhayamkaras, between Ugrachand\=a and the other Mah\=am\=ar\={\i}s and Nand\={\i}\'svara and the other D\=anavas. The battlefield, then, assumed a grim aspect, as if the time of Dissolution had come. Bhagav\=an Mah\=adeva sat under the Vata (peepul) tree with K\=artikeya and Bhadrak\=al\={\i}. \'Sankhach\=uda, decked with his jewel ornaments, sat on the jewel throne, surrounded by kotis and kotis of D\=anavas. The \'Sankara's army got defeated at the hands of the D\=anavas. The Devas, with cuts and wounds on

their bodies, fled from the battlefield, terrified. K\=artikeya gave words ``Don't fear'' to the Devas and excited them. Only Skanda resisted the D\=anava forces. In one moment he slew one hundred Ak\d{s}auhi\d{n}\={\i} D\=anava forces. The lotus eyed K\=al\={\i} also engaged in killing the Asuras. She became very angry and no sooner did She slay the Asura forces, than She began to drink their blood. She easily slew with Her one hand and at every time put into Her mouth ten lakhs, and hundred lakhs and Kotis and Kotis of elephants. Thousands and thousands of headless bodies (Kavandhas) came to be witnessed in the field. The bodies of the D\=anavas were all cut and wounded by the arrows of K\=artikeya. They were all terrified and fled away. Only Vri\d{s}aparv\=a, Viprachitti, Dambha, and Vikamka\d{n}ah remained fighting with Skanda with an heroic valour. Mah\=am\=ar\={\i}, too, did not shew his back and he fought out vigorously. By and by they all became very much confused and distressed; but they did not turn their backs. Seeing this terrible fight of Skanda, the Devas began to shower flowers. The killing of the D\=anavas looked like a Prakritik Dissolution. \'Sankhach\=uda, then, began to shoot arrows from his chariot.

The shooting of arrows by the king seemed as if rains were being poured in by the clouds. Everything became pitch dark. Fires only were seen emitting their golden tongues. The Devas, Nand\={\i}\'svara and others, fled away, terrified. Only K\=artikeya remained in the battlefield. Then \'Sankhach\=uda began to throw terribly showers and showers of mountains, snakes, stones, and trees. So much so, that K\=artikeya was covered by them as the Sun becomes obscured by fog. The Demon King cut off the weighty quiver and the pedestal of Skanda and broke His chariot. By the divine weapons of the D\=anava, the peacock (the vehicle) of K\=artikeya became exhausted. K\=artikeya threw one \'Sakti (weapon) on the breast of the D\=anava; but before it fell, the D\=anava cut off that, lustrous like the Sun and, in return, darted his \'Sakti. By that stroke, K\=artikeya became stunned for a moment; but he immediately regained his consciousness. He then took up the quiver that Bhagav\=an Vi\d{s}\d{n}u gave him before and many other weapons; and ascending on another chariot, built of jewels, began to fight out violently and valiantly. Getting angry, he resisted all those showers of snakes, mountains and trees by his divine weapons. He resisted fire by his watery (P\=aryannya) weapon. Then He cut off easily \'Sankhach\=uda's chariot, bow, armour, charioteer, and his bright crown and he threw on his breast one blazing \'Sakti of white colour. The D\=anavendra fell unconscious; but, at the next moment, he regained his consciousness quickly, mounted on another chariot and took a fresh quiver. The D\=anava was the foremost in his magic powers. He, by his power of M\=ay\=a, made a shower of arrows so much so that

K\=artikeya became completely covered by that multitude of arrows. Then the D\=anava took one invincible \'Sakti, lustrous like one hundred Suns. It seemed that flames of fire were licking high as if the Dissolution Time had come aright. Inflamed by anger, the D\=anava threw that \'Sakti on K\=artikeya. It seemed, then, that a burning mass of fire fell on him. The powerful K\=artikeya became senseless. Bhadrak\=al\={\i} immediately took Him on Her lap and carried him before \'Siva. \'Siva easily restored him to his life by his knowledge-power and gave him the indomitable strength. He then got up in full vigour. Bhadrak\=al\={\i} went to the field to see the K\=artikeya's forces. Nand\={\i}\'svara and other heroes, the Devas, Gandharbas, Yak\d{s}as, R\=ak\d{s}asas and Kinnaras followed Her. Hundreds of war drums were sounded and hundreds of persons carried Madhu (wine). Going to the battle-ground, She gave a war-cry. The D\=anava forces got fainted by that cry. Bhadrak\=al\={\i} shouted aloud inauspicious peals after peals of laughter. Then She drank Madhu and danced in the battlefield. Ugra Damstr\=a, Ugrachand\=a, Kotav\={\i}, the Yogin\={\i}s, D\=akin\={\i}s, and the Devas all drank Madhu (wine). Seeing K\=al\={\i} in the battlefield, \'Sankhach\=uda came up again and imparted the spirit of Fearlessness to the Daityas, trembling with fear. Bhadrak\=al\={\i} projected, then, the Fire weapon, flaming like the Great Dissolution Fire; but the king quickly put out that by the Watery weapon. K\=al\={\i} then projected the very violent and wonderful Varu\d{n}\=astra. The D\=anava cut off that easily with Gandharb\=astra. K\=al\={\i} then threw the flame-like Mahe\'svar\=astra. The king made it futile by the Vai\d{s}\d{n}av\=astra. Then the Dev\={\i} purifying the N\=ar\=aya\d{n}\=astra with the mantra, threw it on the king. At this the king instantly alighted from his chariot and bowed down to it. The N\=ar\=aya\d{n}\=astra rose high up like the Dissolution Fire. \'Sankhach\=uda fell prostrate on the ground with devotion. The Dev\={\i} threw, then, the Brahm\=astra, purifying it with Mantra. But it was rendered futile by the D\=anava's Brahm\=astra. The Dev\={\i} again shot the divine weapons purifying them with mantras; but they also were nullified by the divine weapons of the D\=anava. Then Bhadrak\=al\={\i} threw one \'Sakti extending to one Yojana. The Daitya cut it to pieces by his divine weapon. The Dev\={\i}, then, being very much enraged, became ready to throw P\=a\'supata \=Astra, when the Incorporeal Voice was heard from the Heavens, prohibiting Her, and saying, ``O Dev\={\i}! The high-souled D\=anava would not be killed by the P\=a\'supata weapon. For Brahm\=a granted him this boon that until the Vi\d{s}\d{n}u's Kavacha will remain on his neck and until his wife's chastity be not violated, old age and death will not be able to touch him.'' Hearing this Celestial Voice, the Dev\={\i} at once desisted. But She, out of hunger, devoured hundreds and lakhs of D\=anavas. The terrible Dev\={\i} K\=al\={\i}, then, went with great speed to devour \'Sankhach\=uda but the D\=anava resisted Her by his sharp

divine weapons. The Dev\={\i} then threw on him a powerful axe, lustrous like a summer Sun; but the D\=anava cut it to pieces by his divine weapon. The Dev\={\i} seeing this, became very angry and proceeded to devour him; but the D\=anava King, the Lord of all Siddhis, expanded his body. At this, K\=al\={\i} became violently angry and assuming a terrific appearance, went quickly and with the blow of one fist, broke his chariot and dropped down the charioteer. Then she hurled on the Asura one \'S\=ula weapon, blazing like a Pralaya Fire. \'Sankhach\=uda easily held that by his left hand. The Dev\={\i} became angry and struck the D\=anava with Her fist; the Daitya's head reeled, and, rolling, he fell unconscious for a moment. Next moment regaining his consciousness he got up. But he did not fight hand to hand with the Dev\={\i}. Rather he bowed down to Her. The weapons that the Dev\={\i} threw afterwards were partly cut down by the D\=anava and partly taken up by him and absorbed in him and thus rendered futile. Then Bhadrak\=al\={\i} caught bold of the D\=anava and whirling him round and round threw him aloft. Then the powerful \'Sankhach\=uda fell down on the ground from high with great force; he immediately
got up and bowed down to Her. He then gladly ascended on his beautiful chariot, built of excellent jewels. He did not feel any fatigue with the war and went on fighting. Then the Dev\={\i} Bhadrak\=al\={\i}, feeling hungry began to drink the blood of the D\=anavas and ate the fat and flesh. She came before Mah\=adeva and described to Him the whole history of the warfare from beginning to end. Hearing the killing of the D\=anavas, Mah\=adeva began to laugh. She went on saying, ``The D\=anavas that get out of my mouth while I was chewing them, are the only ones that are living. This number will be about one lakh. And when I took up the P\=a\'supata weapon to kill the D\=anava, the Incorporeal Celestial Voice spoke :-- He is invulnerable by you. But the very powerful D\=anava did no more fling any weapon on Me. He simply cut to pieces those that I threw on him.''

Here ends the Twenty-Second Chapter in the Ninth Book on the fight between the Devas and \'Sankhach\=uda in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



