\chapter{On the birth, etc., of S\=avitr\={\i}}

1-2. N\=ar\=aya\d{n}a said :-- O N\=arada! After having chanted the above hymn to the Goddess S\=avitr\={\i} and worshipped Her in accordance with due rites and ceremonies, the king A\'svapati saw the Dev\={\i}, effulgent like the lustre of thousand suns. She then smilingly told the king, as a mother to her son, whilst all the quarters were illumined with the lustre of Her body :--

3-14. S\=avitr\={\i} said :-- ``O King! I know your desire. Certainly I will give what you and your wife long for. Your chaste wife is anxious for a daughter, while you want a son. So, one after another,

the desires of both of you will be fulfilled.'' Thus saying, the Dev\={\i} went to the Brahma Loka. The King also returned to his house. First a daughter was born to him. As the daughter was born, as if a second Lak\d{s}m\={\i} was born after worshipping S\=avitr\={\i}, the King kept her name as S\=avitr\={\i}. As time rolled on, the daughter grew, day by day, like the phases of bright fortnight moon, into youth and beauty. There was a son of Dyumat Sena, named Satyav\=ana, always truthful, good natured and endowed with various other qualifications. The daughter chose him for her bridegroom. The King betrothed her with jewels and ornaments, to Satyav\=ana, who gladly took her home. After one year expired, the truthful vigorous Satyav\=ana gladly went out, by his father's command, to collect fruits and fuel. The chaste S\=avitr\={\i}, too, followed him. Unfortunately Satyav\=ana fell down from a tree and died. Yama, the God of Death, saw his soul as a Puru\d{s}a of the size of one's thumb and took it and went away. The chaste S\=avitr\={\i} began to follow Him. The high souled Yama, the Foremost of the Sadhus, seeing S\=avitr\={\i} following Him, addressed her sweetly :-- O S\=avitr\={\i}! Whither are you going in your this mortal coil? If you like to follow after all, then quit your this body.

15-25. The mortal man, with his transient coil of these five elements, is not able to go to My Abode. O Chaste One! The death time of your husband arrived; therefore Satyav\=ana is going to My Abode to reap the fruits of his Karma. Every living animal is born by his Karma. He dies again through his life long Karma. It is his Karma alone that ordains pleasure, pain, fear, sorrows, etc. By Karma, this embodied soul here becomes Indra; by Karma he can become a Brahm\=a's son. What more than this that J\={\i}va, by his Karma, can be in Hari's service and be free from birth and death! By one's own Karma all sorts of Siddhis and immortality can be obtained; the four blessed regions as Vi\d{s}\d{n}u's S\=alokya, etc., also can be obtained by Karma. What more than this that by Karma, a being becomes divine, human, or a King, or \'Siva or Gane\'sa! The state of Mun\={\i}ndra, asceticism, K\d{s}attriyahood, Vai\'syahood, Mlechhahood, moving things, stones, R\=ak\d{s}asahood, Kinnaras, Kingship, becoming trees, beasts, forest animals, inferior animals, worms, Daityas, D\=anavas, Asuras, all are fashioned and wrought by Karma and Karma alone. O N\=arada! Thus speaking, Yama remained silent.

Here ends the Twenty-seventh Chapter of the Ninth Book on the birth, etc., of S\=avitr\={\i} in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



