\chapter{On the enumeration of various hells for sinners}

1-28. N\=ar\=aya\d{n}a said :-- Then, initiating her with the Great Seed, the \=Adi Radical Mantra of the Mah\=a \'Sakti, \'Sr\={\i} Bhuvane\'svar\={\i} in accordance with due rules, the son of S\=urya began to recite the various effects of various Karmas, auspicious and inauspicious. Never do the persons go to hell when they perform good Karmas; it is only the bad works that lead men to hells. The different Pur\=a\d{n}as narrate various heavens. The J\={\i}vas go to those places as the effects of their various good Karmas. The good Karmas do not lead men to hells; but the bad Karmas do lead them veritably to various hideous hells. In different \'S\=astras, different hell-pits are ascertained. Different works lead men to different hells. O Child! Those hell-pits are very wide, deep, painful and tormenting, very horrible and ugly. Of these, eighty six pits or Kundas are prominent. Many other Kundas exist. Now listen to the names of the Kundas mentioned in the Vedas. Their names are :-- Vah\d{n}i Kunda, Tapta Kunda, K\d{s}\=ara Kunda, Bhay\=anaka Kunda, Vit Kunda, M\=utra Kunda, \'Sle\d{s}ma Kunda, Gara Kunda, D\=u\d{s}ik\=a Kunda, Vas\=a Kunda, \'Sukra Kunda, \'So\d{n}ita Kunda, A\'sr\=u Kunda, G\=atramala Kunda, Kar\d{n}amala Kunda, Majj\=a Kunda, M\=amsa Kunda, impassable Nakra Kunda, Loma Kunda, Ke\'sa Kunda, impassable Asthi Kunda, T\=amra Kunda, the exceedingly hot and painful Lauha Kunda (the pit of molten iron), Charma Kunda, the hot Sur\=a Kunda, sharp Thorny Kunda, Vi\d{s}a Kunda, the hot Taila Kunda, very heavy Astra Kunda, Krimi Kunda, P\=uya Kunda, terrible Sarpa

Kunda, Ma\'saka Kunda, Dam\'sa Kunda, dreadful Garala Kunda, Vajra Damstra Vri\'schika Kunda, \'Sara Kunda, S\=ula Kunda, awful Khadga Kunda, Gola Kunda, Nakra Kunda, sorrowful K\=aka Kunda, Manth\=ana Kunda, V\={\i}ja Kunda, painful Vajra Kunda, hot P\=at\'s\=a\d{n}a Kunda, sharp P\=a\'s\=a\d{n}a Kunda, L\=al\=a Kunda, Mas\={\i} Kunda, Chakra Kunda, Vakra Kunda, very terrible Kurma Kunda, Jv\=al\=a Kunda, Bhasma Kunda, Dagdha Kunda, and others. Besides these, there are the Taptas\=uch\={\i}, Asipatra, K\d{s}uradh\=ara, S\=uch\={\i}mukha, Gokh\=am\=ukha, K\=umbh\={\i}p\=aka, K\=alas\=utra, Matsyoda, Krimi, Kantuka, P\=am\'subhojya, P\=a\'savesta, S\=ulaprota, Prakampana, Ulk\=amakha, Andhak\=upa, Vedhana, T\=adana, J\=alarandhra, Dehach\=urna, Dalana, \'So\d{s}a\d{n}a, Ka\d{s}a, \'S\=urpa, Jv\=al\=am\=ukha, Dh\=um\=andha, N\=agavestana and various others. O Savitri! The Kundas give much pain and torment greatly the sinners; they are under the constant watch of innumerable servants. They hold rods in their hands; some of them have nooses; others hold clubs, \'Saktis, awful scimitars; they are fierce fanatics, maddened with vanity. All are filled with Tamogu\d{n}as, merciless, irresistible, energetic, fearless and tawny-eyed (like copper). Some of them are Yogis; some are Siddhas, they assume various forms. When the sinners are about to die, they see these servants of Yama. But those who do their own duties, who are \'S\=aktas, Sauras, or G\=a\d{n}apatyas or those who are virtuous Siddha Yogis, they never see the servants of Yama. Those who are engaged in their own Dharmas, who are possessed of wisdom, who are endowed with knowledge, who are mentally strong, who are untouched by fear, who are endowed with the feelings of the Devas, and those who are real Vai\d{s}\d{n}avas, they never see these servants of Yama. O Chaste One! Thus I have enumerated to you the Kundas. Now hear who live in the Kundas.

Here ends the Thirty-second Chapter of the Ninth Book on the enumeration of various hells for sinners in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



