\chapter{On the description of the various hells}

1-28. Dharma R\=aja Yama said :-- O Fair One! If, in this Bh\=arata, any murderer, merciless and fierce, kills any man, out of greed for money, he goes and miserably dwells in the Asipattra hell for fourteen Indra's life periods. And if that murderer kills a Br\=ahma\d{n}a, he lives in that hell for one hundred manvantaras. While in hell, his body becomes fiercely cut and wounded by the swords. There My messengers chastise him and beat him and he cries aloud and passes his time without any food. Then he becomes born for one hundred years as a Manth\=ana insect, for hundred births as a boar, for seven births as a cock; for seven births as a fox, for seven births as a tiger; for three births, as a wolf; for seven births, as a frog; then as a buffalo when he becomes freed of his sins of murders. If anybody sets fire to a city or a village, he will have to live in K\d{s}uradh\=ara Kunda for three yugas with his body severed. Then he becomes a Preta (disembodied spirit) and travels over the whole earth, being burnt up with fire. For seven births he eats unclean and unholy food and spends his time as a pigeon. Then for seven births he becomes diseased with a severe colic pain, for seven births as a leper; when ultimately he gets a pure human body. If anybody whispers in one's ear another's calumny and thus glorifies himself and abuses and vilifies the Devas and Br\=ahma\d{n}as, he goes and remains in S\=uch\={\i} Kunda for three Yugas, and he is pierced there by needles. Then he becomes a scorpion for seven births, a serpent for seven births, and an insect (Bha\d{s}ma K\={\i}ta) for seven births; then he gets a diseased human body when, at last, he becomes purified. If anybody breaks into another's house and steals away all the household articles, cows, goats or buffaloes, he goes to Gok\=a Mukha Kunda where faeces are

like cow's hoofs, there, beaten by My servants, for three Yugas. (Gok\=a is Gok\d{s}ura, hoof of a cow). Then, for seven births, he becomes a diseased cow; for three births, a sheep; for three births, a goat; and finally he becomes a man. But in this man-birth he is born first as diseased, poor, deprived of wife and friends, and a repenting person; when ultimately he is freed of his sin. If anybody steals any ordinary thing, he goes to Nakra Mukha Kunda and lives there for three years, greatly tormented by My messengers. Next for seven births, he becomes a diseased ox. Then he attains a very diseased man-birth, and ultimately he is freed of his sins. Such are the horrible results. If anybody kills a cow, elephant, horse, or cuts a tree, he goes to Gaja Dam\'sa Kunda for three yugas. There he is punished by My messengers freely by the teeth of elephants. Then he attains three elephant\'s births, three horse-births; then he becomes born as a cow and ultimately he is born a Mlechcha when he becomes pure. If anybody obstructs any thirsty cow from drinking water, he goes to Krimi Kunda and Gomukha Kunda filled with hot water and lives there for one manvantara. Next when he attains a human birth he owns not any cattle nor any wealth; rather he is born as a man, very much diseased, in low castes, for seven births when he becomes freed. If anybody, being born in Bh\=arata, kills cows, Br\=ahmi\d{n}s, women, beggars, causes abortions or goes to those not fit to be gone into, he lives in the Kumbh\={\i}p\=aka hell for fourteen Indra's life periods. There be is pulverised always by My messengers. He is made to fall sometimes in fire, sometimes over thorns, sometimes in hot oil, sometimes in hot water, sometimes in molten iron or copper. That great sinner gets thousand vulture births, hundred boar births, seven crow births and seven serpent births. He then becomes worms of faeces for sixty thousand years. Thus travelling frequently in ox births he at last becomes born as a very poor leper.

29-31. S\=avitr\={\i} said :-- ``O Bhagav\=an! What is, according to the \'S\=astras, Brahmahatty\=a (murdering a Br\=ahmin) and Gohaty\=a (killing a cow)? Who are called Agamy\=as (women unfit to be approached)? Who are designated as void of Sandhy\=a (daily worship of the twice born castes)? Who can be called uninitiated? Who are said to take Pratigrahas (gifts) in a T\={\i}rath? What are the characteristics of a real Gr\=amay\=aj\={\i} (village priests), Devala, (Br\=ahma\d{n}a of an inferior order who subsists upon the offerings made to the images which he attends), the cook of a \'S\=udra, of one who is infatuated (Pramatta) and the Vri\d{s}al\={\i}pati (one who has married an unmarried girl twelve years old in whom menstruation has commenced; a barren woman). Kindly describe all these to me.''

32-91. Dharmar\=ajan said :-- O Fair S\=avitr\={\i}! If anybody makes a distinction between Kri\d{s}\d{n}a and His Image or between any Deva and his image, between \'Siva and His phallic emblem, between the Sun and the stone S\=urya K\=anta (a precious stone of a bright and glittering colour) between Gane\'sa and Durg\=a, he is said to be guilty of the sin Brahmahatty\=a. If anybody makes any difference (superiority or inferiority) between his own I\d{s}ta Deva (his Deity), his Spiritual Teacher, his natural father, and mother, is certainly involved in the sin of Brahmahatty\=a. He who shews any difference (superiority or inferiority) between the devotees of Vi\d{s}\d{n}u and those of other Devas, is said to commit Brahmahatty\=a. He who makes any difference in matters of respect between the waters of the feet of any Br\=ahma\d{n}a and those of \'S\=alagr\=ama stone, is said to commit Brahmahatty\=a. The difference between the offerings to Hari and Hara leads to Brahmahatty\=a. He who shews any difference between Kri\d{s}\d{n}a, Who is verily the God of gods, the Cause of all causes, the Origin of all, Who is worshipped by all the Devas, Who is the Self of all, Who is attributeless and without a second yet Who by His Magic powers assumes many forms and who is I\'s\=ana, is said to commit, indeed, the Brahmahatty\=a. If any Vai\d{s}\d{n}ava (a devotee of Vi\d{s}\d{n}u) abuses and envies a \'Sakta (a devotee of \'Sakti), he commits Brahmahatty\=a. He who does not worship, according to the Vedas, the Pitris and the Devas or prohibits others in doing so, commits Brahmahatty\=a. He who abuses Hri\d{s}ike\'sa, Who is the Highest of the Holy things, Who is Knowledge and Bliss and Who is Eternal, Who is the only God to be served by the Devas and Vai\d{s}\d{n}avas, and those Who are worshippers of His Mantra, and those who do not worship themselves are said to commit Brahmahatty\=a. He who abuses and vilifies M\=ulaprakriti Mah\=a Dev\={\i}, Who is of the nature of Causal Brahm\=a (K\=ara\d{n}a Brahm\=a), Who is All Power and the Mother of all, Who is worshipped by all and who is of the nature of all the Devas and the Cause of all Causes, Who is \=Ady\=a \'Sakti Bhagavat\={\i}, is said to commit Brahmahatty\=a. He who does not observe the Holy \'Sr\={\i} Kri\d{s}\d{n}a Janm\=astami, \'Sr\={\i} R\=ama Navam\={\i}, \'Sivar\=atri, the Ek\=ada\'s\={\i} happening on Sunday, and five other holy P\=arva\d{n}as (festivals), commits Brahmahatty\=a; is considered more sinful than a Ch\=and\=ala. He who in this land of Bh\=arata, digs earth on the day of Ambuv\=ach\={\i} or makes water, etc., in the waters of the tanks, is involved in the sin of Brahmahatty\=a. He who does not support his spiritual teacher, mother, father, chaste wife, son and daughter, though they are faultless, commits Brahmahatty\=a. He whose marriage does not take place during his whole life-time, who does not see the face of his son, who does not cherish devotion to Hari, who eats things unoffered to \'Sr\={\i} Hari, who never worshipped throughout his life Vi\d{s}\d{n}u or an earthen symbol of \'Siva, verily commits Brahmahatty\=a. O Fair One! Now I will

recite the characteristics, according to the \'S\=astras, of Gohatty\=a (killing a cow). Listen. If anybody does not prohibit one, seeing one to beat a cow, or if he goes between a cow and a Br\=ahmi\d{n}, he is involved in the sin of Gohatty\=a. If any illiterate Br\=ahma\d{n}, carrying an ox, daily beats with a stick, the cows, certainly he commits the Gohatty\=a. If anybody gives the remains of another's meal to a cow to eat, or feeds a Br\=ahmi\d{n} who carries, rather moves or drives, cows and oxen; or eats himself the food of such a Br\=ahmi\d{n} driver, he commits Gohatty\=a. Those who do sacrifices of the husband of a barren woman (Vri\d{s}al\={\i}) or eat his food, commit sin equal to one hundred Gohatty\=as; there is no doubt in this. Those who touch fire with their feet, beat the cows or enter the temple bathing but not washing their feet, commit Gohatty\=a. Those who eat without washing their feet or those who sleep with their feet wetted with water and those who eat just after the Sun has risen, commit Gohatty\=a. Those who eat the food of women without husbands or sons or the food of pimps and pampers or those who do not perform their Sandhy\=as thrice, commit Gohatty\=a. If any woman makes any difference between her husband and the Devat\=a, or chastises and uses harsh words to her husband, she commits Gohatty\=a. If anybody destroys cow's pasture land, tanks, or land for forts and cultivates there grains, he commits Gohatty\=a. He who does not do Pr\=aya\'schitta (expiation, atonement) for the expiation of the sin of Gohatty\=a done by his son (for fear of his son's life), commits the sin himself. If any trouble arises in the state or from the Devas, and if any master does not protect then his own cows, rather torments them, he is said to commit Gohatty\=a. If any J\={\i}va oversteps the image of a Deva, fire, water, offerings to a god, flowers, or food, he commits the great sin Brahmahatty\=a. When a guest comes, if the master of the house always says, ``there is nothing, nothing with me; no, no,'' and if he be a liar, cheat and an abuser of the Devas, he commits the above sin. O fair One! Whoever seeing his spiritual teacher, and a Br\=ahma\d{n}, does not bow down and make respectful obeisance to them, commits Gohatty\=a. If any Br\=ahmi\d{n}, out of sheer anger, does not utter blessings to a man who bows down or does not impart knowledge to a student, he commits Gohatty\=a. O Fair One! Thus I have described to you the characteristics, approved by \'S\=astras, of cow-killing (Gohatty\=a) and murdering a Br\=ahmi\d{n} (Brahmahatty\=a). Now hear which women are (Agamy\=as) not fit to be approached and those which are fit to be approached (Gamy\=as). One's own wife is fit to be approached (Gamy\=a) and all other women are Agamy\=as, so the Pundits, versed in the Vedas, declare. This is a general remark; now hear everything in particular. O Chaste One! The Br\=ahmi\d{n} wives of \'S\=udras or the \'S\=udra wives of Br\=ahma\d{n}as are Aty\=agamy\=as (very

unfit to be approached) and blameable both in the Vedas and in the society. A \'S\=udra going to a Br\=ahma\d{n}\={\i} woman commits one hundred Brahmahatty\=as; so a Br\=ahma\d{n}a woman going to a \'S\=udra goes to the Kumbh\={\i}p\=aka hell. As a \'S\=udra should avoid a Brahm\=a\d{n}i, so a Br\=ahma\d{n}a should avoid a \'S\=udra woman. A Br\=ahma\d{n}a going to a \'S\=udra woman is recognised a Bri\d{s}alipati (one who has married an unmarried girl twelve years old in whom menstruation has commenced). So much so that that Br\=ahma\d{n}a is considered an outcast and the vilest of the Ch\=and\=alas. The offerings of Pindas by him are considered as faeces and water offered by him is considered as urine. Nowhere whether in the Devaloka or in the Pitriloka, his offered Pindas and water are accepted. Whatever religious merits he has acquired by worshipping the Devas, and practising austerities for Koti births, he loses all at once by the greed of enjoying the \'S\=udra woman. There is no doubt in this. A Br\=ahmi\d{n}, if he drinks wine, is considered as the husband of a Vri\d{s}al\={\i}, eating faeces. And if he be a Vai\d{s}\d{n}ava, a devotee of Vi\d{s}\d{n}u, his body must be branded with the marks of a Taptamudr\=a (hot seal); and if he be a \'Saiva, his body is to be branded with the Tapta \'S\=ula (hot trident). The wife of a spiritual teacher, the wife of a king, step-mother, daughter, son's wife, mother-in-law, sister of the same father and mother, the wife of one's brother (of the same father and mother), the wife of a maternal uncle, the father's mother, mother's mother, the mother's sister, sisters, the brother's daughter, the female disciple, the disciple's wife, the wife of the sister's son, the wife of the brother's son, these all are mentioned by Brahm\=a as Aty\=agamy\=as (very unfit to be approached). The people are hereby warned. If anybody, overpowered by passion, goes to these Aty\=agamy\=a women, he becomes the vilest of men. The Vedas consider him as if going to his mother and he commits one hundred Brahmahatty\=a sins. These have no right to do any actions. They are not to be touched by any. They are blamed in the Vedas, in the society everywhere. Ultimately they go to the dreadful Kumbh\={\i}p\=aka hells. O Fair One! He who performs Sandhy\=as wrongly or reads it wrongly or does not perform at all the three Sandhy\=as daily, is called as void of Sandhy\=a. He is said to remain uninitiated who does not, out of sheer vanity, receive any Mantra, whether he be a Vai\d{s}\d{n}avite, \'Saivite, or a Sun worshipper or the Gane\'sa worshipper. Where there is the running stream of the Ganges, lands on either side, four hands in width, are said to be the womb of the Ganges (Gang\=a Garbha) Bhagav\=an N\=ar\=aya\d{n}a incessantly dwells there. This is called the N\=ar\=aya\d{n}a (K\d{s}etra). One goes to Vi\d{s}\d{n}upada who dies in such a place. V\=ar\=a\d{n}as\={\i} (Benares), Vadar\={\i}, the Confluence of the Ganges with the ocean (Ganga-S\=agara), Pu\d{s}kara, Hari Hara K\d{s}ettra (in Behar near Ch\=apr\=a), Prabh\=asa,

K\=amar\=upa, Hardwar, Ked\=ara, M\=atripura, the banks of the river Sarasvat\={\i}, the holy land Bindr\=aban, God\=avar\={\i}, Kau\'sik\={\i}, Trive\d{n}\={\i} (Allahabad), and the Him\=alay\=as are all famous places of pilgrimages. Those who willingly accept gifts in these sacred places are said to be T\={\i}rthapratigr\=ah\={\i}s (the acceptors of the gifts in the T\={\i}rtha). These T\={\i}rthaprathigr\=ah\={\i}s go in the end to Kumbh\={\i}p\=aka hell. The Br\=ahma\d{n}a who acts as priest to the \'S\=udras is called \'S\=udray\=aj\={\i}; the village priests are called Gr\=amay\=aj\={\i}s. Those who subsist on the offerings made to the gods are called Devalas. The cooks of the \'S\=udras are called S\=upak\=aras. Those who are void of Sandhy\=a Bandanams are called Pramattas (mad). O Bhadre! These are the marks of the Vri\d{s}al\={\i}patis that I have (now) enumerated. These are the Great Sinners (Mah\=a P\=atakas). They go ultimately to the Kumbh\={\i}p\=aka hell. O Fair One! I now state by and by the other Kundas (hells) when other people go. Listen.

Here ends the Thirty fourth Chapter of the Ninth Book on the description of the various hells in the Mah\=a Pur\=a\d{n}am \'Sr\={\i}mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



