\chapter{On the glories of the Dev\={\i} and on the nature of Bhakti}

1-6. S\=avitr\={\i} said :-- ``O Lord! Give me the devotion to the Dev\={\i} to that \=Ady\=a \'Sakti Bhagavat\={\i} Mah\=a M\=aye, Parame\'svar\={\i} M\=ay\={\i} that is the Essence of all essences, the Door of final liberation to the human beings, and the Cause of delivering them from hells, that is the Root of all the Dharmas that lead to Mukti, that destroys all the inauspiciousness, that takes away the fear of all the Karmas, and that takes away always all the sins committed before. O Thou, the Foremost amongst the knowers of the Vedas! How many kinds of Muktis are there in this world? What is the True Bhakti? What are its characteristics? What is to be done by which the enjoyment of the karmas done can be desisted and nullified? O Bhagav\=an! The woman kind has been created by the Creator as devoid of any Tattvaj\~n\=ana or true knowledge; now tell me something about this True Knowledge. All the charities, sacrifices, bathing in the sacred places of pilgrimages, observing vows and austerities cannot be compared with one sixteenth of imparting knowledge to those who are ignorant (of true knowledge). Mother is hundred times superior to father; this is certain; but the Spiritual Teacher, the Giver of True Knowledge, is hundred times more to be reverenced and worshipped than the mother. O Lord!''

7-79. Dharmar\=aja said :-- O Child! What boons you desired of Me before, I granted them all to you. Now I grant this boon to you that, ``Let the devotion towards the \'Sakti now arise in your mind.'' O Auspicious One! You want to hear the reciting of the Glories of \'Sr\={\i} Dev\={\i}; by this, both he who puts forward the question and he who hears the answer, all their families are delivered. When the \'Sesa N\=aga Ananta Deva with His thousand mouths is unable to recite the glories of the Dev\={\i}, when Mah\=adeva cannot describe with His five mouths, when the Creator Brahm\=a is incapable to recite Her glories with His four mouths, when Vi\d{s}\d{n}u, the Omniscient, falls back, when K\=art\={\i}keya with His six mouths cannot sufficiently describe, when Gane\'sa, the Guru of the Gurus of the great yogis is incapable, when the Pundits, the knowers of the four Vedas, the Essence of all the \'S\=astras, cannot know even a bit of Her, when Sarasvat\={\i} becomes inert in going to describe Her glories; when Sanatkumara, Dharma, San\=atana, Sananda, Sanaka, Kapila, S\=urya and other sons of the Creator have fallen back, when the other Siddhas, Yog\={\i}ndras, Mun\={\i}ndras are quite incapable to glorify the deeds of Prakriti Dev\={\i}, then how can I recite fully the Glories of Her? Whose lotus-feet Brahm\=a,

Vi\d{s}\d{n}u, \'Siva and others meditate; and lo! when it becomes difficult for Her devotees even to think of Her, then what wonder is there that She will be so very rare to others! Brahm\=a, skilled in the knowledge of the Vedas knows more of Her than what other ordinary persons know so little of Her auspicious Glories. More than Brahm\=a, Gane\'sa, the Guru of the J\~n\=anins, knows; again \'Sambhu, the Omniscient, knows the best of all. For, in ancient times, the knowledge of the Prakriti Dev\={\i} was given to Him by Kri\d{s}\d{n}a, the Highest Spirit in a solitary place in the R\=asa Mandalam in the region of Goloka. Mah\=adeva, again, gave it to Dharma in the \'Sivaloka; Dharma again gave the Prakriti Mantra to My father. My father became successful (Siddha) in the worship of Prakriti Dev\={\i} when he practised austerities. Of old, the Devas wanted to offer to me the Government of the Yama Loka; but as I was very much dispassionate towards the world, I became unwilling and became ready to perform austerities. Then my Father told the Glories of Prakriti Dev\={\i}. Now I describe to you what I heard from my Father and what is stated in the Vedas, though very difficult to comprehend. Listen carefully. O Fair Faced One! As the eternal space does not know its own extent, so Prakriti Dev\={\i} Herself knows not Her own Glories; then what can be said of any other person on this! She is the Self of all, endowed with all powers and lordship, the Cause of all causes, the Lord of all, the Origin of all and the Preserver of all; She is Eternal, always with Her Cosmic Body, full of everlasting bliss, without any special form, unrestrained, having no fear, without any disease and decay, unattached, the Witness of all, the Refuge of all, and Higher than the Highest; She is with M\=ay\=a and She is M\=ula Prakriti; the objects created by Her being known as the Prakriti creation; Who remains as Prakriti and Puru\d{s}a inseparable from each other as Ag\d{n}i and Her burning force; the Mah\=a M\=ay\=a, of the nature of everlasting existence, intelligence and bliss. Though formless, She assumes forms for the gratification of the desires of Her Bhaktas. She created first the beautiful form of Gop\=ala Sundar\={\i}, i.e., the form of \'Sr\={\i} Kri\d{s}\d{n}a very lovely and beautiful, captivating the mind. His body is blue like the fresh rain cloud; He is young and dressed like that of a cow herd. Millions of Kandarpas (the Love deity) are, as it were, playing in his body. His eyes vie with the midday lotus of the autumn. The beauty of His face throws under shade the millions and millions of the Full Moon. His body is decorated with invaluable ornaments decked with jewels. Sweet smile reigns ever in His lips; it is adorned moreover with His yellow coloured invaluable robe. He is Parama Brahm\=a. His whole body is burning with the Brahm\=a Teja, the Fire of Brahm\=a.

His Body is Fiery. He is lovely, sweet to look at, of a peaceful temper, the Lord of R\=adh\=a and He is Infinite. (Note :-- The Universe; as we see, is unreal like what we see in the kaleidoscope; various apparent pictures of an endless variety of beautiful colours and forms.) He is sitting on a jewel throne in the R\=asa Mandalam, and is incessantly looked upon by the smiling lovely Gop\={\i}s at one and the same time. He is two-armed. A garland made of wild flowers is hanging from His neck. He is playing on His flute. His breast is resplendent with Kaustubha gem that He always wears. His body is anointed with Kunkuma (saffron), aguru (the fragrant and cooling paste of the Aguru wood), musk, and sandal-paste. The garland of beautiful Champaka and M\=alat\={\i} flowers is hanging from His neck. On His head, the beautiful crest is being seen, a little obliquely situated in the form of the beautiful Moon. Thus the Bhaktas, filled with Bhaktis, meditate on Him. O Child! It is through His fear that the Creator is doing His work of creation of this Universe; and is recording the Pr\=arabdha fruits of their Karmas. It is through His fear that Vi\d{s}\d{n}u is awarding the fruits of Tapas and preserving the Universe. By His command the K\=alagni Rudra Deva is destroying all. By Whose favour \'Siva has become Mrityumjaya, the Conqueror of Death and the Foremost of the J\~n\=anins; knowing whom \'Siva has become Himself endowed with knowledge and the Lord of the knowers of knowledge, full of the Highest Bliss, devotion and dispassion. Through Whose fear the wind becomes the foremost of runners and carries things, the Sun gives heat, Indra gives rain, Yama destroys, Ag\d{n}i burns, and Water cools all the things. By Whose command the Regents of the (ten) quarters of the sky are watching and preserving nice orders; through Whose fear the planets are describing their several orbits. Through Whose fear, trees flower and yield fruits; By Whose command the K\=ala destroys all. By Whose command all the beings whether on land or in water are quitting their lives in time; until the proper time comes no man does not die even if he be pierced whether in battle or in danger. By Whose command the wind supports the water; the water supports the tortoise; the tortoise supports the Ananta and the Ananta supports the earth; the earth supports the oceans, mountain and all the jewels. The earth is of the nature of forgiveness, i.e., endures all. For this reason all things, moving and non-moving, rest on Her and again melt away in Her. Seventy-one Divine Yugas constitute one Indra's life period. Twenty-eight Indra's life periods constitute Brahm\=a's one day and one night. Thus thirty days constitute Brahm\=a's one month; so two months constitute one Ritu (season); six Ritus make one year. Thus one hundred years constitute Brahm\=a's

life. When Brahm\=a dies, \'Sr\={\i} Hari's eye closes. That is the Pr\=akritik Pralaya. At this time, everything, moving and non-moving, from the Deva loka to Bh\=ur loka (earth) dies. The Creator Brahm\=a gets dissolved in the navel of \'Sr\={\i} Kri\d{s}\d{n}a. The four-armed Vi\d{s}\d{n}u, of Vaikuntha, sleeps on K\d{s}\={\i}ra Samudra, the ocean of milk, i.e., He dissolves on the left side of \'Sr\={\i} Kri\d{s}\d{n}a, the Highest Spirit. All the other \'Saktis (forces) dissolve in M\=ula Prakriti, the M\=ay\=a of Vi\d{s}\d{n}u. The M\=ula Prakriti Durg\=a, the Presiding Deity of Buddhi (reason) dissolves in the Buddhi of Kri\d{s}\d{n}a. Skanda, the part of N\=ar\=aya\d{n}a, dissolves in His breast. Gane'sa, the foremost of the Devas, born in part of Kri\d{s}\d{n}a, dissolves in the arm of \'Sr\={\i} Kri\d{s}\d{n}a. And those who are born in parts of Padm\=a, dissolve in Her body and Padm\=a dissolves in the body of R\=adh\=a. All the cow-herdesses and all the bodies of the Devas dissolve in R\=adh\=a's body. But R\=adh\=a, the Presiding Deity of the Pr\=a\d{n}a of \'Sr\={\i} Kri\d{s}\d{n}a, dissolves in the Pr\=a\d{n}a of \'Sr\={\i} Kri\d{s}\d{n}a. S\=avitr\={\i}, the four Vedas and all the \'S\=astras dissolve in Sarasvat\={\i}; and Sarasvat\={\i} gets dissolved in the tongue of \'Sr\={\i} Kri\d{s}\d{n}a, the Highest Self. The Gop\=as in the region of Goloka dissolve in the pores of His skin; the Pr\=a\d{n}a V\=ayu of all dissolve in His Pr\=a\d{n}a V\=ayu; the fire dissolves in the fire in His belly; water dissolves in the tip of His tongue, and the Vai\d{s}\d{n}avas, (devotees of Vi\d{s}\d{n}u), drinking the nectar of Bhakti, the Essence of all essences, dissolve in His lotus-feet. All smaller Vir\=ats dissolve in the Great Vir\=at and the Great Vir\=at dissolves in the Body of \'Sr\={\i} Kri\d{s}\d{n}a. O Child! He is Kri\d{s}\d{n}a, on the pores of Whose skin are situated endless Universes; at the closing of Whose eyes, the Pr\=akritic Pralaya comes and on the opening of Whose eyes, the creation takes place. The closing and opening of the eyes takes the same time. Brahm\=a's creation lasts one hundred years and the Pralaya lasts one hundred years. O One of good vows! There is no counting how many Brahm\=as or how many creations and dissolutions have taken place. As one cannot count the number of dusts, so one cannot count the creations and dissolutions. This is the Great Unspeakable Wonder! Again on Whose closing of the eyes the Pralaya takes place and on whose opening of the eyes the creation takes place, out of the will of God, That Kri\d{s}\d{n}a dissolves at the time of Pralaya in Prakriti. This Highest \'Sakti, the M\=ula Prakriti is the Only One without a second; it is the only one Nirgu\d{n}a and the Highest Puru\d{s}a. It is considered as ``Sat'' existing, by the Seers of the Vedas. Such a thing as M\=ula Prakriti is the unchanged state (Mukti). During the Pralaya, this only One M\=ula Prakriti appears as J\~n\=ana \'Sakti or the Knowledge Force. Who can in this universe recite Her glories? Mukti is of four kinds.

(1) S\=alokya, (2) S\=ar\=upya, (3) S\=am\={\i}pya and (4) Nirv\=a\d{n}a. So it is stated in the Vedas. Out of them Bhakti towards the Deva is the highest; so much so that the Deva Bhakti is superior to Mukti. Mukti gives S\=alokya, S\=ar\=upya, S\=am\={\i}pya, and Nirv\=a\d{n}a. But the Bhaktas do not want anything. They want service of the Lord. They do not want anything else. The state of becoming \'Siva, of becoming an Amara or an immortal, becoming a Brahm\=a, the birth, death, disease, old age, fear, sorrow, or wealth, or assuming a divine form, or Nirv\=a\d{n}a or Moksa all are looked on alike by the Bhaktas with disregard and contempt. Because Mukti is without any service while Bhakti increases this service. Thus I have told you the difference between Bhakti and Mukti. Now hear about the cutting off of the fruits of the past Karmas. O Chaste one! This service of the Highest Lord severs the ties of Karmas (past acts). This service is really the True Knowledge. So, O Child! I have now told you the Real Truth, leading to auspicious results. Now you can go freely as you desire. Thus saying to S\=avitr\={\i}, Yama, the son of S\=urya, gave life back to her husband and blessing her, became ready to go to His own abode. Seeing Dharmar\=aja ready to go away, S\=avitr\={\i} became sorry to have the bereavement of a good company, bowed down at His feet and began to cry. Yama, the Ocean of Mercy, hearing the crying of S\=avitr\={\i} began to weep and told the following words :--

80-96. Dharma said :-- O Child! You enjoy in this holy Bh\=arata happiness for one lakh years and you will in the end go to the Dev\={\i}loka or Ma\d{n}i Dv\={\i}pa. Now go back to your house and observe for fourteen years the vow called S\=avitr\={\i}-vrata for the mukti of women. This Vrata is to he observed on the fourteenth day of the white fortnight in the month of Jyaistha. Then observe the Mah\=a-Lak\d{s}m\={\i} Vrata. Its proper time is the eighth day of the bright fortnight of the month of Bh\=adra. For sixteen years consecutively without any break this vow is to be observed. The woman who practises with devotion this vow, goes to the abode of M\=ula Prakriti. You would worship on every Tuesday in every month the Dev\={\i} Mangala Chandik\=a, the giver of all good; on the eighth day in the bright fortnight you should worship Dev\={\i} S\=asth\={\i} (i.e., Devasen\=a); you should worship Manas\=a Dev\={\i}, the giver of all siddhis, on the Samkranti day (when the Sun enters another sign) in every year; you should worship R\=adh\=a, the Central Figure of R\=asa, more than the Pr\=a\d{n}a of Kri\d{s}\d{n}a on every Full Moon night in the month of K\=artik and you should observe fasting on the eighth day in the bright fortnight and worship the Vi\d{s}\d{n}u M\=ay\=a Bhagavat\={\i} Dev\={\i}, the Destructrix of all difficulties and dangers.

(Note :-- The R\=asa is the playing out of the Vedantic saying of Brahm\=a as ``Raso vai Sah.'' He is of the nature of R\=asa, the most sweet and lovely Divine Principle which unites the Rasika and the Rasik\=a.) The chaste woman having husband and sons who worships the World Mother M\=ula Prakriti, whether in Yantra, or in Mantra or in image, enjoys all pleasures in this world; and, in the end, goes to the Dev\={\i}loka or Ma\d{n}i Dv\={\i}pa. O Child! The worshipper S\=adhaka (one who is in one's way to success) must worship all the manifestations of the Dev\={\i}, day and night. At all times one must worship the omnipresent Durg\=a, the Highest \=Isvar\={\i}. There is no other way to attain blessedness than this. Thus saying, Dharmar\=aja went to His own abode. S\=avitr\={\i}, too, with her husband Satyav\=an went to her home. Both S\=avitr\={\i} and Satyav\=an, when they reached home, narrated all their stories to their friends and acquaintances. In time, by the blessing of Yama, S\=avitr\={\i}'s father got sons and father-in-law recovered his eye-sight and kingdom and S\=avitr\={\i} Herself got sons. For one lakh years, S\=avitr\={\i} enjoyed pleasures in this holy land of Bh\=arata, and ultimately went with her husband to the Dev\={\i}loka. S\=avitr\={\i} is the Presiding Deity of the S\=uryamandalam, the solar orb. The Sun is the central Para Brahm\=a. The G\=ayatr\={\i} Mantra, the Presiding Dev\={\i}, proves the existence of the highest Brahm\=a in the centre of the Sun. Therefore She is called S\=avitr\={\i}. Or Her name is S\=avitr\={\i} because all the Vedas have come out of Her. Thus I have narrated the excellent anecdote of S\=avitr\={\i}, and the fruitions of the Karmas of the several J\={\i}vas. Now what more do you want to hear? Say.

Here ends the Thirty-eighth Chapter in the Ninth Book on the glories of the Dev\={\i} and on the nature of Bhakti in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



