\chapter{On the conversation between S\=avitr\={\i} and Yama and on the fruition of Karmas}

1. S\=avitr\={\i} said :-- ``O Dharmar\=ajan! Kindly tell me in detail about those works that lead the meritorious persons to the Heavens and various other spheres.''

2-20. Dharmar\=aj said :-- O Child! He who gives rice and food to the Br\=ahma\d{n}as in India, goes to the \'Sivaloka where he dwells with great respect for years equivalent to the measure of that food. This ``Anna-d\=ana'' (the giving of boiled rice and other eatables) is a great d\=an (charity) and this can be done not only to the Br\=ahma\d{n}as but to the other castes also, where similar results also follow. There is, or will be, no other charity superior to this charity of anna (rice, boiled it may be and other eatables). For here no distinction is made as to what caste will get it or not, nor the discrimination of time, when to give such a charity. O Child! Seats (\=Asanas) given to the Devas and the Br\=ahma\d{n}as, carry the donor to the Vi\d{s}\d{n}u Loka, where he dwells for Ayuta years with great respect and love. Giving excellent cows yielding milk to Br\=ahma\d{n}as take the donor to Vi\d{s}\d{n}uloka, where he is glorified and remains for years equivalent to the number of pores in that cow or those cows. And if cows be given on a meritorious day, four times the merits accrue, and if given in a sacred place of pilgrimage, hundred times the result occurs; and if given in a t\={\i}rath, where N\=ar\=aya\d{n}a is worshipped, koti times the results accrue. He who gives with devotion, cows to the Br\=ahma\d{n}as in Bh\=arata, remains in the Chandraloka for one Ayuta years and is glorified. He who gives a two-mouthed cow to a Br\=ahma\d{n}a, goes to Vi\d{s}\d{n}uloka and remains there for as many years as there exist the numbers of hairs on the body of that cow and is glorified. A gift of a beautiful white umbrella to a Br\=ahma\d{n}a makes one go to Varu\d{n}aloka for Ayuta years where he remains with great pleasure. Giving garments to the diseased Br\=ahma\d{n}as makes one fit to remain with glory in Vayuloka for one ayuta years. Giving to a Br\=ahma\d{n}a the \'S\=alagr\=ama with garments makes one remain with glory in Vaikuntha as long as there exist the Sun and Moon. Giving a beautiful bedding to a Br\=ahma\d{n}a, glorifies a man in the Chandraloka as long as there exist the Sun and Moon. To give lights to the Devas and Br\=ahma\d{n}a glorifies a man in Ag\d{n}iloka (the region of Fire) for one Manvantara. To give elephants to the Br\=ahma\d{n}as in Bh\=arata, makes one sit in the same throne with Indra for his life period.

Giving horses to the Br\=ahma\d{n}as makes one remain in Varu\d{n}aloka for fourteen Indra's life periods. Giving a good palanquin to a Br\=ahma\d{n}a makes one remain in the Varu\d{n}aloka for fourteen Indra's life-periods. Giving a good site or a good orchard to a Br\=ahma\d{n}a leads one to the Vayuloka where he remains with glory for one Manvantara. Giving a white ch\=amara and fan to a Br\=ahma\d{n}a, leads the donor to the Vayuloka where he remains for one ayuta years. Giving grains and jewels make one long-lived and both the donors and receivers go certainly to Vaikuntha.

21-40. He who always recites the name of \'Sr\={\i} Hari, lives for ever and Death goes far far away from him. The intelligent man that celebrates the Swinging Festival (Dol J\=atr\=a) in the last quarter of the Full Moon night in this land of Bh\=arata, becomes liberated while living, enjoying pleasures in this world, goes in the end to Vi\d{s}\d{n}uloka, where he remains for one hundred Manvantaras; there is no doubt in this. If the Swinging Festival be performed under the influence of the asterism Uttara Ph\=algun\={\i} then the fruits become doubled; this is the saying of Brahm\=a Himself. The performer lives to the end of a Kalpa. To give til (Sesamum) to a Br\=ahma\d{n}, leads one to \'Siva Loka, where one enjoys for a number of years equal to the number of til. Then one is born in a good yoni and becomes longlived and happy. To give a copper plate yields double the effect. To give in India a chaste wife with garments and ornaments to a Br\=ahma\d{n} (and then to purchase her with an equivalent in gold) leads one to Chandra Loka where one remains for fourteen Indra's life periods and enjoys day and night the celestial Apsar\=as. Thence the donor goes to the Gandharba Loka for one ayuta years and day and night enjoys Urva\'s\={\i}. Then he gets for thousand births chaste, fortunate, wealthy, gentle and sweet-speaking, beautiful wives. He who gives nice and delicious fruits to the Br\=ahma\d{n}as, remains with glory in the Indra Loka for a number of years equivalent to the fruits. He gets again a good Yoni (birth) and gets excellent sons. To give thousand trees while there are fruits on them, or nice fruits only to the Br\=ahma\d{n}as, makes one enjoy the Heavens for a long, long time and he then comes back to Bh\=arata. To give various things and good edifices with grains, etc., to the Br\=ahma\d{n}as leads one to the regions of the Devas where he remains for one hundred Manvantaras. Then he gets a very good birth and becomes the master of abundant wealth. He who gives with devotion to the Br\=ahma\d{n}as lands certainly goes for one hundred Manvantaras and remains there in glory for one hundred Manvantras; and, coming again to be born in good wombs, they become Kings. The earth does not leave him for hundred births. He becomes prosperous, wealthy and possesses many sons and

becomes the lord of his subjects. He who gives a good village with pasture land and cows, dwells with glory in Vaikuntha for one lakh manvantaras. Then he gets a good birth (becomes born in a high caste family) and obtains a lakh villages. The earth quits him not even if he be born a lakh times. (This is very bad then, to one who does not like to be born again).

41-60. He who gives a village inhabited by good and obedient subjects with ripe excellent grains, various tanks, trees and adorned with fruits and leaves dwells in Kail\=a\d{s}a with great glory for ten lakh Indra's life periods. Getting again born in a high family, he becomes R\=aja Dhir\=aja in Bh\=arata and obtains Niyuta towns. There is no doubt in this. The earth quits him not, even if he be born ayuta times. Really he gets the highest prosperity in this earth. He who gives to a Br\=ahma\d{n}a one hundred towns and countries, inhabited by good or mediocre subjects, with wells, tanks, and various trees, remains with glory in Vaikuntha for one koti manvantaras. Then he becomes born in this earth in a high caste family, becomes the Lord of Jambudv\={\i}pa and attains in this earth great prosperity like Indra. The earth quits him not even if he comes here Koti times; in reality he is a Mah\=atm\=a (a great soul man), R\=ajr\=aje\'svara (the Lord of Kings) and lives upto the end of a Kalpa. He who gives his whole property to a Br\=ahmi\d{n}, gets in the end four times that; there is no doubt in this. He who gives Jambu Dv\={\i}pa to an ascetic Br\=ahma\d{n}a, gets undoubtedly in the end one hundred times the fruit. If you give away Jambu Dv\={\i}pa, the whole earth; if you travel all the T\={\i}rthas, if you perform all sorts of asceticisms, if you give shelter to all, if you make gifts of all sorts, know that you will have to come again to be reborn in this earth; but if you become a devotee of M\=ula Prakriti, then be sure that you won't have to come here and be reborn. The devotees of M\=ula Prakriti go to Ma\d{n}i Dv\={\i}pa, the highest place of \'Sr\={\i} Bhuvane\'sar\={\i} Dev\={\i} and remain there and they see the fall of innumerable Brahm\=as. The worshippers of the Dev\={\i} Mantra when they quit their mortal coils, assume divine appearances endowed with Bibh\=utis (manifestations of powers) and free from birth, death and old age, assume the S\=ar\=upya (the same form) of the Dev\={\i} and remain in Her Service. They reside in Ma\d{n}idv\={\i}pa and see the part Pralayas. The Devas die, the Siddhas die, the whole universe vanishes; but the Dev\={\i} Bhaktas never die and they remain free form birth, death, and old age. He who offers Tulas\={\i} leaf to Bhagav\=an Hari in the month of K\=artik resides for three yugas in the temple of Hari. Getting again a good birth, he acquires the devotion to \'Sr\={\i} Hari and becomes the Foremost of those who restrain their senses. He who bathes in the Ganges early before the rising of the Sun remains in enjoyment in the temple of Hari for sixty thousand yugas. Getting again a good birth, he gets

the Vi\d{s}\d{n}u Mantra, and, quitting his mortal coil, becomes united with the Feet of \'Sr\={\i} Hari.

61-77. He has not to come back from Vaikuntha to this earth. He remains in Hari's Service and gets the same form of Hari. He who bathes daily in the Ganges, becomes purified like the Sun and gets the result of performing the Horse-sacrifice at every step. The earth becomes purified by the dust of his feet and he enjoys in Vaikuntha as long as the Sun and Moon exist. Then again he becomes born in a good and beautiful womb, and is liberated by acquiring the devotion to Hari. He becomes very energetic and the foremost of the ascetics, pure, religious, learned, and self-restrained. When the Sun comes midway between Pisces and Cancer and heats intensely the earth, the man who in Bh\=arata gives cool water to drink to the people, resides in happiness in Kail\=a\'sa for fourteen Indra's life periods. Getting again a good birth here, he becomes beautiful, happy, devoted to \'Siva, energetic and expert in the Vedas, and the Vedangas. He who gives to a Br\=ahma\d{n}a the \'Saktu (sattu). in the month of Vai\'s\=akhe enjoys in the \'Siva temple for as many years as there are number of particles in that quantity of sattu (powders in parched oat). He who performs the Kri\d{s}\d{n}a Janm\=astam\={\i} vow in this Bh\=arata, is freed from the sins incurred in his hundred births; there is no doubt in this. The observer of the vow remains in great enjoyment in Vaikuntha for fourteen Indra's life periods, gets again a good birth here and acquires Hari Bhakti. He who performs the \'Sivar\=atri vow in this Bh\=arata Var\d{s}a, resides with great joy in \'Siva Loka for seven manvantaras. He who offers the Bel leaves to \'Siva in \'Sivar\=atri time, resides with great joy in \'Siva's Abode for as many yugas as there are number of leaves. Getting again a good birth here, he acquires the devotion to \'Siva and becomes learned, prosperous and possesses sons, subjects and lands. He who performs vow and worships \'Sankara in the month of Chaitra or M\=agha and who, with a branch of a tree in hand, dances day and night for one month, or half a month, for ten days or for seven days, dwells in \'Siva Loka for as many yugas as the number of days he dances. He who performs the vow of \'Sr\={\i} R\=ama Navam\={\i}, lives in the abode of Vi\d{s}\d{n}u for seven Manvantaras in great joy. Getting again a good birth, he becomes devoted to \'Sr\={\i} R\=ama, the foremost of those who have self restraint and he becomes very wealthy.

78-87. He who performs the S\=arad\={\i}y\=a P\=uj\=a (the great Durg\=a P\=uj\=a in the month of autumn) of the M\=ula Prakriti with incense, lights, offerings of food, and animal sacrifices of buffaloes, goats, sheep, rhinoceros, frogs or other animals, together with dancing, music, and various other aus-

picious things, resides in the \'Siva Loka for seven Manvantaras. Getting an excellent birth, and a pure understanding, he gets unbounded prosperity, sons and, no doubt, grandsons and he becomes a very powerful sovereign possessing many horses and elephants. There is no doubt in this. Again he who worships daily with devotion for a fortnight beginning from the eighth day of the bright fortnight the Mah\=a Dev\={\i} Lak\d{s}m\={\i}, remains in the region of Goloka for fourteen Indra's life periods. Then, obtaining an excellent birth, he becomes a sovereign. He who in the full moon night in the month of K\=artik prepares a R\=asa mandal with one hundred Gopas and Gopis and worships \'Sr\={\i} Kri\d{s}\d{n}a and R\=adh\=a in \'S\=alagr\=ama or in images with sixteen varieties of offerings remains in Goloka for Brahm\=a's life-time and coming again to Bh\=arata acquires an unflinching devotion to \'Sr\={\i} Kri\d{s}\d{n}a.

88-99. And when this Bhakti becomes greatly intensified, ho gets initiated into \'Sr\={\i} Hari mantra and after quitting his mortal coil, he goes to the Goloka. Then he gets the S\=ar\=upya (the same form) of Kri\d{s}\d{n}a and becomes the chief P\=ari\d{s}ad (attendant of Kri\d{s}\d{n}a) and, becoming free from old age, he has no fear, to fall again down to this earth. He who observes the Ek\=ada\'s\={\i} day, remains fasting and performing penances in the bright or dark eleventh day, remains in Vaikuntha in great enjoyment and comfort. Then, again coming into this Bh\=arata he becomes a devotee of Hari. And when that Bhakti is intensified he becomes solely devoted to Hari and quitting his mortal coil, goes again to the Goloka and gets the S\=ar\=upya of Kri\d{s}\d{n}a and becomes His P\=ari\d{s}ada (attendant). Then, freed of old age and death, he does not fall. He who worships Indra in the month of Bh\=adra in the twelfth day of the white fortnight is worshipped in the regions of Indra for sixty thousand years. He who performs in Bh\=arata the worship of the Sun on Sunday Sankr\=anti (when the Sun goes from one sign to another) and the bright seventh Tithi, according to due rules and ceremonies and eats the food called Havi\d{s}y\=anna (rice boiled in ghee), dwells in the S\=uryaloka for fourteen Indra's life periods. Then coming to Bh\=arata, he becomes free from all diseases and becomes prosperous. He who worships S\=avitr\={\i} on the fourteenth day of the black fortnight dwells in the region of Brahm\=a for seven Manvantaras with great ecl\=at and glory. Coming again to Bh\=arata he enjoys beauty, unequalled valour, long life, knowledge and prosperity. He who worships on the fifth day of the bright fortnight in the month of M\=agha, with his senses controlled and full of devotion, the Dev\={\i} Sarasvat\={\i} with sixteen articles of food, resides in Ma\d{n}i Dv\={\i}pa for one day and one night of Brahm\=a.

100-140. On getting re-birth, he becomes a poet and a learned man. He who daily gives with devotion for his whole life, cow and gold to a Br\=ahma\d{n}a dwells in Vi\d{s}\d{n}u Loka for twice as many years as there are

the numbers of hairs on the bodies of these cows and plays and jests with Vi\d{s}\d{n}u and doing auspicious things he finds pleasure. In the end he comes again to this Bh\=arata and becomes the King of Kings. He becomes fortunate, prosperous, possesses many sons, becomes learned, full of knowledge and happy in every way. He who feeds a Br\=ahma\d{n}a here with sweetmeats goes to Vi\d{s}\d{n}u Loka and enjoys there for as many years as there are hairs on the body of the Br\=ahmi\d{n}. In the end he comes again to Bh\=arata and becomes happy, wealthy, learned, long lived, fortunate and very powerful. He who utters the name of Hari or gives the name (i.e., the mantra) of Hari to others, is worshipped in Vi\d{s}\d{n}u loka for as many yugas as the number of times, the name or mantra was uttered. Coming again to Bh\=arata, he becomes happy and wealthy. And if such things be done in N\=ar\=aya\d{n}a K\d{s}ettra, koti times the above results ensue. He who repeats the name of Hari koti times in N\=ar\=aya\d{n}a K\d{s}ettra, becomes, no doubt, freed of all sins and liberated while living and he will not get rebirth. He lives always in Vaikuntha. He gets the S\=alokya (the same region of Vi\d{s}\d{n}u), is not liable to fall, becomes a Bhakta of Vi\d{s}\d{n}u. He who daily worships the earthen phallic symbol (after making it daily) for his whole life, goes to the \'Siva Loka and dwells there, for as many years as there are the number of particles of earth. Getting rebirth he becomes the King of Kings. He who worships daily the \'S\=alagr\=ama stone and eats the water (after bathing it) is glorified in Vaikuntha for one hundred Brahm\=a's lives and becomes born again. When he acquires the rare Hari Bhakti and quiting his mortal coil goes to Vi\d{s}\d{n}u Loka, whence he is not to return. He who performs all the Tapasy\=as (asceticims) and observes all the vratas (vows), dwells in Vaikuntha for fourteen Indra's life peroids. Getting rebirth in Bh\=arata he becomes the King of Kings and then he becomes liberated. He is not to return any more. He who bathes in all the T\={\i}rthas and makes a journey round the whole world, gets Nirv\=a\d{n}a. He is not reborn. He who performs the Horse-Sacrifice in this holy land Bh\=arata enjoys half the Indraship for as many years as there are hairs on the body of the horse. He who performs a R\=ajas\=uya Sacrifice, gets four times the above result. Of all the sacrifices, the Dev\={\i} Yaj\~n\=a, or the Sacrifice before the Dev\={\i} is the Best. O Fair One! Of old, Vi\d{s}\d{n}u, Brahm\=a, Indra and when Tripur\=asura was killed, Mah\=a Deva did such a sacrifice. O Beautiful One! This sacrifice before the \'Sakti is the highest and best of all the sacrifices. There is nothing like this in the three worlds. This Great Sacrifice was done of yore by Dak\d{s}a when he collected abundant sacrificial materials of all sorts. And a quarrel ensued on this account between Dak\d{s}a and \'Sankara. The Br\=ahmi\d{n}s

conducting the sacrifice cursed the Nand\={\i} and others. And Nand\={\i} cursed the Br\=ahma\d{n}as. Mah\=adeva, therefore, disallowed the going on of sacrifice and brought it to a dead stop. Of yore the Praj\=apati Dak\d{s}a did this Dev\={\i} Yaj\~n\=a; it was done also by Dharma, Ka\'syapa; Ananta, Kardama, Sv\=ayambhuva Manu, his son Priyavrata, \'Siva, Sanat Kum\=ara, Kapila and Dhruva. The performance of this sacrifice brings fruits equal to performing thousands and thousands of R\=ajas\=uya sacrifices. Therefore there is no other sacrifice greater than this Dev\={\i} Yaj\~n\=a. One becomes surely endowed with a long life of one hundred years and is liberated while living. He becomes equal to Vi\d{s}\d{n}u in knowledge, energy, strength, and asceticism. This is as true as anything. O Child! This Dev\={\i} Yaj\~n\=a is the best and highest of all the sacrifices as Vi\d{s}\d{n}u is the highest amongst the Devas; N\=arada, amongst the Vai\d{s}\d{n}avas; the Vedas, amongst all the \'S\=astras; the Br\=ahma\d{n}as amongst all the castes; the Ganges amongst the sacred places of pilgrimages, \'Siva amongst the Holy of Holies, the Ek\=ada\'s\={\i} vow amongst all the Vratas; Tulas\={\i}, amongst all the flowers; the Moon, amongst the asterisms; Garuda, amongst the birds; Prakriti, R\=adh\=a, Sarasvat\={\i} and Earth amongst the females; the mind, amongst the quick-going and restless senses; Brahm\=a, amongst the Praj\=apatis; Brahm\=a, amongst all the subjects; Vrindr\=aban, amongst all the forests; Bharat Var\d{s}a, amongst all the Var\d{s}as; Lak\d{s}m\={\i}, amongst the prosperous; Sarasvat\={\i}, amongst the learned; Durg\=a, amongst the chaste; Radhik\=a, amongst the fortunate. If one hundred horse sacrifices are performed, Indrahood is sure to be obtained. It is by the influence of bathing in all the T\={\i}rthas, performing all the sacrifices, observing all the Vratas, practising all the austerities, studying all the Vedas and circumambulating the whole earth, that this Highest \'Sakti's service is obtained and this service of \'Sakti is the direct cause of Mukti (liberation). To worship the lotus-feet of the Dev\={\i} is the best and highest, is stated in all the Pur\=a\d{n}as, in all the Vedas, and in all the Itih\=asas. To sing the glories of M\=ula Prakriti, to meditate on Her, to chant Her Name and attributes, to remember Her stotras, bow down before Her, to repeat Her Name, and to drink daily Her P\=adodoka (water after washing Her feet) and the offerings already offered to Her, these are approved of by all; and everyone desires this. So worship, worship this M\=ula Prakriti, Who is of the nature of Brahm\=a, and, lo! Who is again endowed with M\=ay\=a. O Child! Take your husband and live happily with him in your home. O Child! Thus I have described to you the fruition of the Karmas. This is auspicious to every human being, desired by all and approved of by all. The Real Knowledge springs from this. There is no doubt in this.

Here ends the Thirtieth Chapter of the Ninth Book on the conversation between S\=avitr\={\i} and Yama and on the fruition of Karmas in the Great Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



