\chapter{On the description of the destinies of different sinners in different hells}

1-19. Dharmar\=ajan said :-- Those that are in Hari's service, pure, the Siddhas in Yoga (those that have attained success in Yoga), the performers of Vratas (vows), the chaste, the ascetics, the Brahmach\=aris never go to hells. There is no doubt in this. Those persons that are proud of their strong positions and who use very harsh burning words to their friends, they go to Vah\d{n}i Kunda and live there for as many years as there are hairs on his body; next they attain animal births for three births and get themselves scorched under the strong heat of the Sun. He goes to the Tapta Kunda hell who does not

entertain any Br\=ahma\d{n} guest with any eatables who comes to his house hungry and thirsty. He lives there for as many years as there are hairs on his body and he has to sleep on a bed of fire, very tormenting. Then he will have to be born for seven births as birds. If anybody washes any clothing with any salt on Sunday, or on the day of Samkr\=anti (when the Sun enters another sign), or on any new-moon day or on any \'Sr\=addha day (when funeral ceremonies are performed), he will have to go to the K\d{s}\=ara Kunda hell where he remains for as many years as there are threads in that clothing and finally he becomes born for seven births as a veritable washerman. The wretch that abuses M\=ula Prakriti, the Vedas, the \'S\=astras, Pur\=a\d{n}as, Brahm\=a, Vi\d{s}\d{n}u, \'Siva and the other Devas, Gaur\={\i}, Lak\d{s}m\={\i}, Sarasvat\={\i} and the other Dev\={\i}s, goes to the hell named Bhay\=anaka Narakakunda. There is no other hell more tormenting than this. The sinners live here for many Kalpas and ultimately become serpents. There is no sin greater than the abuse of the Dev\={\i}. There is no expiation for it. So one ought never to abuse the Dev\={\i}. If one discontinues the allowances given by oneself or other persons to the Devas or Br\=ahma\d{n}as, one goes to Visth\=a Kunda and has to eat the faeces there for sixty thousand years and finally to be born in Bh\=arata as worms in faeces the same number of years. If any person without the owner's permission digs another's tank dried of water, or makes water in the water of any tank, he goes to M\=utra Kunda and drinks urine for as many years as there are the particles in that tank. Then he becomes born in this Bh\=arata its an ox for one hundred years. If any person eats good things himself without giving any portion thereof to the member of his family, he goes to \'Sle\d{s}ma Kunda where he eats phlegm, for full one hundred years. Then he becomes born as Preta (disembodied spirits) in this Bh\=arata for hundred years and drinks phlegm, urine and pus; then he becomes pure. He who does not support his father, mother, spiritual teacher, wife, sons, daughters and the helpless persons, goes to Gara Kunda where he eats poison for full one hundred years. Finally he becomes born and wanders as Bh\=utas (disembodied spirits). Then he becomes pure.

20-50. He who becomes angry and shrinks his eyes at the sight of a guest who has come to his house offends the Devas or Pitris, who do not accept the water offered to them by that villain. On the contrary, he earns all the sins of Brahmahaty\=a (murder of a Br\=ahmi\d{n} and so forth) and finally goes to D\=u\d{s}ik\=akunda where he remains for one hundred years and eats polluted things. Then wandering as Bh\=utas for one hundred years he becomes purified. If anybody makes a gift of any article to a Br\=ahmi\d{n} and then again gives that article to a different man, he goes to Vas\=a Kunda

where he eats marrows for one hundred years. Then he has to roam about in India for seven births as a Krikal\=asa (lizard) and finally he becomes born as a very poor man with a very short life. If any woman or any man makes another of a different sex eat semen, out of passion, he goes to \'Sukra Kunda where he drinks semen for one hundred years. Then he crawls about as worms for one hundred years. And then he gets purified. If anybody beats a Br\=ahma\d{n}a who is a family preceptor and causes his blood to come out, he will have to go to Rakta Kunda where he has to drink blood for one hundred years. Finally he has to roam about for seven births in India as tigers; then he becomes pure by degrees. If anybody mocks and laughs at any devotee of Kri\d{s}\d{n}a who sings with rapt consciousness and sheds tears of joy, he will have to go to A\'sru Kunda where he drinks tears for one hundred years. Then he has to roam as Ch\=and\=ala for three births and then he becomes pure. He who always cheats his friends, lives for one hundred years in G\=atramala Kunda. Then roaming about for three births as an ass and for three births as a fox concurrently, he becomes purified. Out of vanity, if anybody jests at a deaf person, he goes to Kar\d{n}amalakunda where he eats for one hundred years the wax of the ear. Next he comes to the earth as a deaf and very poor man for seven births, when at last he gets purified. If anybody commits murder out of greed to support his family, he goes to the hell Majj\=akunda where he eats marrow for one lakh years. Next he becomes a fish for seven births, for seven births he becomes a mosquito, for three births he becomes a boar, for seven births he becomes a cock, deer and other animals concurrently; at last he gets purified. If any stupid person sells the daughter whom he has supported, out of greed for money, he goes to M\=a\d{n}sakunda and lives there for as many years as there are hairs on her body. The Yama's servants beat him with their clubs. His head becomes overloaded with the burden of the flesh; and, out of hunger, he licks the blood coming out of his head. Next that sinner comes to Bh\=arata and for sixty years becomes a worm in any daughter's faeces, for seven births he becomes a hunter; for three births, a boar; for seven births, cock; for seven births, frog; for seven births, leech; and for seven births, crow; when he gets purified. One who shaves on the day of observing vows, fasting and funeral ceremony day, becomes impure and unfit to do any action, and, in the end, he goes to the Nakha Kunda where he receives blows of clubs and eats nails for one hundred Deva years. If anybody worships, out of carelessness, the earthen \'Siva phallic symbol with any hairs on it, he goes to the hell Ke\'sa Kunda where he remains for as many years as there are particles in that hair; then he gets to the yoni (womb) a Y\=avan\=an\={\i} (a Mlechcha woman) out of Hara's wrath. After one hundred

years he becomes freed from that and then he becomes a R\=ak\d{s}asa; there is no doubt in this. He who does not offer Pindas to the Vi\d{s}\d{n}up\=ada in honour of his Pitris at Gay\=a goes to the hell Asthikunda where he remains for as many years as there are dirts on his body. Then he becomes a man; but for seven births he becomes lame and poor. Then he gets purified. The stupid man who commits outrage and violence on his pregnant wife, resides for one hundred years in the hot T\=amra Kunda (where coppers are in a molten condition). He who takes the food of a childless widow and the same of any woman that has just bathed after menstruation goes for one hundred years to the hot Lauha Kunda (where iron is in a molten condition). For seven births he becomes then a crow and for seven births he becomes born of a washerwoman, full of sores and boils, and poor. Then he gets purified.

51-61. If one touches the things of the Devas after touching skins or impure hides, one remains in the Charma Kunda for full one hundred years. If any Br\=ahmi\d{n} eats a \'S\=udra's food, requested by him, he lives for one hundred years in the hot Sur\=a Kunda. Then for seven births he performs, the funeral rites for a \'S\=udra; at last he becomes pure. If any foul-mouthed person uses always harsh and filthy language to his master, he will have to go to T\={\i}k\d{s}\d{n}a Kantaka Kunda where he eats thorns. Besides, the Yama's servants give severe beatings to him with their clubs. For seven births he will have to become horses when he gets purified. If any man ministers poison to another and so takes away his life, he will have to remain for endless years in Vi\d{s}akunda, where he will have to eat poison. Then he will have to pass for one hundred years as a murderer Bhilla, full of sores and boils, and for seven births he will have to be a leper when at last he gets purified. Being born in this holy land Bh\=aratavar\d{s}a, if any man strikes a cow with a rod or any driver does so whether by himself or by his servant, he will have to dwell certainly in the hot Lauha Kunda for four yugas. He will have to pass as many years as a cow as there are hairs on that cow when ultimately he gets purified. If anybody strikes any other body with a red-hot iron dart (Kunta weapon), he will have to dwell in the Kunta Kunda for ayuta years. Then he will have to remain for one birth in a good womb, with a diseased constitution, when ultimately he will be purified.

62-85. If any Br\=ahmi\d{n} villain eats, out of greed, any flesh (not sacrificed before the goddess) or anything not offered to Hari, he will have to remain in the Krimi Kunda where he eats those things for as many years as there are hairs on his body. Then he will have to pass for three births as Mlechchas when ultimately he becomes born in a Br\=ahmi\d{n}

family. If any Br\=ahmi\d{n} performs the \'Sr\=adh of a \'S\=udra, eats the food pertaining to a \'Sr\=adh of a \'S\=udra or burns the dead body of a \'S\=udra, he will have to dwell certainly in P\=uya Kunda, where, being beaten by the rod of Yama, he eats the pus, etc., for as many years as there are hairs on his body. Then he becomes reborn in this Bh\=arata as one greatly diseased, poor, deaf and dumb and ultimately he will have to roam for seven births as a \'S\=udra. He who kills a black serpent on whose hood there is the lotus mark, lives in Sarpa Kunda for as many years as there are hairs on his body and he is bitten by serpents there and beaten by the servants of Yama and eats the excrescences of snakes and finally becomes born as a serpent. Then he becomes a man shortlived and having the cuticaneous disease and ringworm. And his death also comes out of snake-bite. He who kills mosquitoes and other small fanged-animals, that earn their substance rightly and pass so their lives, goes to Dam\'sa ma\'sa Kunda where he is eaten by mosquitoes and other fanged-creatures and lives there without food and crying, weeping, for as many years as the numbers of lives destroyed. Besides the Yama's servants tie his hands and feet and beat him. Then he becomes born as flies when ultimately he becomes purified. He who beats and chastises any man not fit to be chastised and beaten and as well as a Br\=ahma\d{n}a, goes to Vajra Damstra Kunda, full of worms, and lives there day and night for as many years as there are the number of hairs on the chastised person. When he is bitten by the worms and beaten by Yama's servants, he cries sometimes, weeps sometimes, and becomes very miserable. Next he is reborn as a crow for seven births when ultimately he gets purified. If any foolish king punishes and gives trouble to his subjects out of greed of money, he goes to Vri\d{s}chika Kunda where he lives for as many years as there are hairs on the bodies of his subjects. There is no doubt in this. Finally he becomes born in this Bh\=arata as a scorpion; then a man diseased and defective in limbs, when ultimately he becomes freed of his sins. If any Br\=ahmi\d{n} carries or raises weapons, washes the clothes of others who do not perform Sandhy\=as and abandons his devotion to Hari, he lives in Sar\=adi Kunda for as many years as there are hairs on his body; he is, then, pierced by arrows. Finally he becomes purified. If any king maddened by his own folly and fault, shuts his subjects in a dark cell and kills them, then he will have to go to a dreadful dark hell filled with worms having fanged teeth and covered with dirt. This hell is named Gola Kunda. He lives there bitten by insects for as many years as there are hairs on the bodies of his subjects. Finally he becomes a slave of those subjects, when he gets purified.

86-103. If anybody kills the sharks and crocodiles, etc., that rise out of the water spontaneously, he will have to remain, then, in Nakra

Kunda for as many years as there are thorns or edged points on those animals. Then he will have to be born as crocodiles, etc., for some time, when he will be purified. If any man, overpowered with lust, sees another's wife's uncovered breast, loins, and face, he will have to remain in K\=aka Kunda for as many years as there are hairs in his own body. Here the crows take out his eyes. Finally for three births he gets himself burned by Fire when he becomes pure. He who steals in India the gold of the Devas and the Br\=ahma\d{n}as, dwells certainly in Manth\=ana Kunda for as many years as there are hairs on his body. My servants give him good beatings, and cudgellings; his eyes are covered by Manth\=ana Danda insects (or animals) and he eats their dirty faeces. Then he is reborn as a man but for three births he becomes blind and for seven births he becomes very poor, cruel, and a sinful goldsmith and then he is born a Svar\d{n}ava\d{n}ik (Son\=ar bene). O Fair One! He who steals in India copper or iron, silver or gold, dwells in V\={\i}ja Kunda for as many years as there are hairs on his body. There the V\={\i}jas (a kind of insect) cover his eyes and he eats the excrescences of those insects. My messengers torment him. Finally he gets purified. If anybody steals in India any Devat\=a or the articles of a Devat\=a, he dwells in Vajra Kunda for as many years as there are hairs on his body. There his body gets burnt up. My messengers torment him and he a cries and weeps and remains without any food. Then he gets purified. If anybody steals the metal gold or silver, cows, or garments of any Deva or a Br\=ahma\d{n}a, certainly he dwells in hot P\=a\d{s}\=a\d{n}a Kunda for as many years as there are hairs on his body. Next for three births he becomes a tortoise and all sorts of white birds. Finally for three births he becomes a leper and for one birth be becomes a man with white marks on his body. Next for seven births he becomes diseased with a severe colic pain and bad blood and lives short. Then he gets purified. If anybody steals brass or K\=amsya properties of any Deva or a Br\=ahma\d{n}a, he will have to remain in the sharp P\=a\d{s}\=a\d{n}a Kunda for as many years as there are hairs on his body. Next he becomes born in Bh\=arata for seven births as horses; and ultimately his both the testicles get enlarged and he gets diseases in his legs when he gets purified. If anybody verily eats the food of an adulterate woman or lives on her alms, he will have to go to the L\=al\=a Kunda for as many years as there are hairs on his body. My messengers torment him there and he eats the saliva and thus lives miserably. Then he gets eye diseases and colic; when ultimately he gets purified.

104-126. If any Br\=ahma\d{n}a lives on writing only or on the service of Mlechchas, he lives in Masi Kunda very painfully, eating ink, tormented by My messengers for as many years as there are hairs

on his body. Then he becomes a black animal for three births and for another three births he becomes a black goat. Then he becomes a T\=al tree when he gets purified. If anybody steals a Deva's or
a Br\=ahma\d{n}a's grains, or any other good materials, betel, \=Asan (seat) or bedding, he lives in Ch\=ur\d{n}a Kunda for one hundred years, tormented by My D\=utas (messengers). Next for three births he gets
himself born as a goat, cock, and monkey. Finally he becomes born as a man with the heart disease, without any issue, poor, and short lived. When, at last, he gets purified. If anybody steals any Br\=ahmin's property and thereby does chakra p\=uj\=a (the famous chakra circle worship in Tantra), or prepares a potter's wheel or any other wheels, he will have to go to Chakra Kunda and remain there for one hundred years, tormented by My messengers. Then he will be born for three births as an oilman suffering from very severe diseases when he will ultimately be poor, without any issue and diseased. Finally he gets purified. If anybody casts a sinful eye on any Br\=ahma\d{n}a or on cows, he will have to remain in Vakra Kunda for one hundred Yugas. Next for three births he becomes a cat, for three births he becomes a vulture; for three births he becomes a boar; for three births he becomes a peacock; for seven births he becomes a man deformed and defective in limbs, his wife being dead, without any issue. Finally he becomes purified. If any person born in a Br\=ahmi\d{n} family eats the flesh of a tortoise that is prohibited, he lives in K\=urma Kunda, for one hundred years, eaten by tortoises. Then he becomes for three births a tortoise; for three births, a boar; for three births a cat; for three births, a peacock; till at last he gets purified. If anybody steals clarified butter or oil of any Devas or a Br\=ahma\d{n}a he will have to go to Jv\=al\=a Kunda or Bha\d{s}ma Kunda. That sinner remains in oil for one hundred years and gets soaked through and through. Then for seven births he becomes a fish and a mouse when he gets purified. If anybody, born here in this holy land Bh\=arata steals sweet scented oil of a Deva or of a Br\=ahma\d{n}a, the powdered myrobalan or any other scent, he goes to Dagdha Kunda where he lives, burnt day and night for as many years as there are hairs on his body. For seven births he becomes born emitting a nasty smell, for three births he becomes musk (mriga-n\=abhi); for seven births, as a Manth\=ana insect. Then he becomes born as a man. If, out of envy, a powerful man appropriates to his purpose another's ancestral property by cheating, by using force, he goes to the hot S\=uch\={\i} Kunda, being tormented there like a J\={\i}va dropped in the midst of a very hot oil tank, full of boiling oil. His body is, then, being burnt up severely as the

result of his own Karma; the wonder being that his body never gets completely destroyed nor reduced to ashes. For seven manvantaras he lives there without any food. My messengers give him good beatings and cudgellings and chastise him; he cries aloud. Next he gets himself born as worms of faeces for sixty thousand years. When he becomes born as a pauper without owning any land. Thus that villain, getting a fresh lease of human birth, begins again to do fresh good acts.

Here ends the Thirty-third Chapter of the Ninth Book on the description of the destinies of different sinners in different hells in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



