\chapter{On the description of the various hells for the various sinners}

1-44. Dharmar\=ajan said :-- O Chaste One! Without serving the Gods, the Karma ties can never be severed. The pure acts are the seeds of purities and the impure acts lead to impure seeds. If any Br\=ahma\d{n}a goes to any unchaste woman and eats her food, he will have to go ultimately to the K\=alas\=utra hell. There he lives for one hundred years when ultimately he gets a human birth when he passes his times as a diseased man and ultimately he gets purified. Those women who are addicted to their (one) husbands only are called Pativrat\=as. Those addicted to two persons are named Kulat\=as; to three, are called Dhar\d{s}i\d{n}\={\i}s; to four, called Pum\'schal\={\i}s; to five, six persons, called Ve\'sy\=as; to seven, eight, nine persons are called Pung\={\i}s; and to more than these, are called Mah\=ave\'sy\=as. The Mah\=ave\'sy\=as are unfit to be touched by all the classes. If any Br\=ahma\d{n}a goes to Kulat\=a, Dhar\d{s}i\d{n}\={\i}, Pum\'schal\={\i}, Pung\={\i}, Ve\'sy\=a and Mah\=ave\'sy\=as, he will have to go to the Matsyoda Kunda. Those who go to Kulat\=as remain there for one hundred years; those who go to Dhar\d{s}i\d{n}\={\i}s, remain for four hundred years, those who go to Pum\'schal\={\i}s for six hundred years; those who go to Ve\'sy\=as, for eight hundred years; those who go to Pung\={\i}s, remain for one thousand years and those who go to Mah\=ave\'sy\=as remain in the Matsyoda Kunda for ten thousand years. My messengers chastise and beat and torment them very severely. And when their terms expire, the Kulat\=a-goers become Tittiris (a bird), the Dhar\d{s}i\d{n}\={\i}-goers

become crows, the Pum\'schal\={\i}-mongers become cuckoos, the Ve\'sy\=a haunters become wolves; the Pung\={\i}-goers become for seven births boars. If any ignorant person eats food during the lunar and solar eclipses, he goes to Aruntuda Kunda for as many years as there are particles in that food. He then becomes born diseased with Gulma (a chronic enlargement of spleen) having no ears nor teeth, and after passing his time so, he becomes freed of his previous sin. If anybody makes a promise to give his daughter to one but he gives actually to a different person, he goes to P\=am\'su Kunda where he eats ashes for one hundred years. Again if anybody sells his daughter, he sleeps on a bed of arrows in Pam\'suvesta Kunda for one hundred years, chastised and beaten by My messengers. If any Br\=ahma\d{n}a does not worship with devotion the phallic emblem of \'Siva, he goes to the dreadful \'S\=ulaprota Kunda for that heinous sin. He remains there for one hundred years; then he becomes a quadruped animal for seven births and again he becomes born a Devala Br\=ahmi\d{n} for seven births when he becomes freed. If any Br\=ahma\d{n}a defeats another Br\=ahma\d{n}a in a bad useless argument and trifles him and makes him tremble, he goes to the Prakampana Kunda for as many years as there are hairs on his body. If any woman, being very furious with anger, chastises and uses harsh words to her husband, she goes to Ulk\=amukha Kunda for as many years as there are hairs on his body. My servants put fiery meteors or torches in her mouth and beat on her head. At the end of the term, she becomes a human being but she has to bear the torments of widowhood for seven births. Then she is again born as diseased; when at last she gets herself freed. The Br\=ahma\d{n}a woman, enjoyed by a \'S\=udra, goes to the terrible dark Andhak\=upa hell, where she remains, day and night, immersed in the impure water and eats that for fourteen Indra's life periods. Her pains are unbounded and My messengers beat her severely and incessantly.

At the expiry of the term in that hell, she becomes a female crow for thousand births, a female boar for one hundred births, a female fox for one hundred births, a hen for one hundred years, a female pigeon for seven births, and a female monkey for seven births. Then she becomes a Ch\=and\=al\={\i} in this Bh\=arata, enjoyed by all. Then she becomes an unchaste woman with the pthisis disease, a washerwoman, and then an oilwoman with leprosy when she becomes freed. O Fair One! The Ve\'sy\=as live in the Vedhana, and Jalarandhra hells; the Pung\={\i}s live in the Dandat\=adana hell; the Kulat\=as live in the Dehach\=urna hells; the Svairin\={\i}s live in the Dalana hells; the Dhar\d{s}i\d{n}\={\i}s live in \'So\d{s}ana hells. Their pains know no bounds at all those places. My messengers always beat and chastise them and they eat always the urine and faeces for

one Manvantara. Then, at the expiry of their hell period, they become worms of faeces for one lakh years when they become freed. If a Br\=ahma\d{n}a goes to another Br\=ahma\d{n}a's wife, if a K\d{s}attriya, Vai\'sya and \'S\=udra do so, they go to the Ka\d{s}\=aya hell. There they drink the hot Ka\d{s}\=aya water for twelve years when they become purified. The lotus-born Brahm\=a has said that the wives of Br\=ahmi\d{n}s, K\d{s}attriyas, etc., live in hells like Br\=ahmi\d{n}s, K\d{s}attriyas, etc., and they then get freed. If a K\d{s}attriya or a Vai\'sya goes to a Br\=ahmi\d{n}'s wife, he is involved in the sin of his going to his mother and goes and lives in the \'S\=urpa hell. There the worms of the size of a \'S\=urpa bite that K\d{s}attriya, that Vai\'sya and that Br\=ahma\d{n}a's wife. My messengers chastise them and they have to eat the hot urine. Thus they suffer pains for fourteen Indra's life periods. When they become boars for seven births and goats for seven births, when at last they are freed. Now if anybody makes a false promise or swears falsely, taking the T\=ulas\={\i} leaf in his hands, if anybody makes a false promise, taking the Ganges water, \'S\=alagr\=ama stone, or any other images of God in his hand; if anybody swears falsely, placing his right palm on the palm of another; if anybody swears falsely, being in a temple or touching a Br\=ahma\d{n}a or a cow; if anybody acts against his friends or others, if he be treacherous or if he gives a false evidence; then all these persons go to Jv\=al\=a Mukha hell, and remain there for fourteen Indra's life periods, chastised and beaten by My messengers and feeling pain as if one's body is being burnt by red hot coal. One who gives a false evidence, with the T\=ulas\={\i} (holy basil) in his hand becomes a Ch\=and\=ala for seven births; one who makes a false promise with the Ganges water in his hand, becomes a Mlechcha for five births; one who swears falsely while touching the \'S\=alagr\=ama stone, becomes a worm of the faeces for seven births; one who swears falsely, touching the image of the God, becomes a worm in a Br\=ahmi\d{n}'s house for seven births; one who gives a false evidence touching with the right hand, becomes a serpent for seven births; then he becomes born as a Br\=ahmi\d{n}, void of the knowledge of the Vedas, when he becomes freed. One who speaks falsely, while in a temple, is born as a Devala for seven births.

45-59. If one swears falsely, touching a Br\=ahma\d{n}a, one becomes a tiger. Then he becomes dumb for three births, then for three births he becomes deaf, without wife, without friends, and his family becomes extinct. Then he becomes pure. Those that rebel against their friends, become mongoose; the treacherous persons become rhinoceroses; the hypocrite and treacherous persons become tigers and those who give false evidences become frogs. So much so, that their seven generations

above and seven generations below go to hell. If any Br\=ahma\d{n}a does not perform his daily duties (Nitya Karma), he is reckoned as Jada (an inert matter). He has no faith in the Vedas. Rather he laughs at the Vedic customs. He does not observe vows and fastings; he blames others who give good advices. Such persons live in Dh\=umr\=andhak\=ara hell where they eat dark smoke only. Then he roams about as an aquatic animal for one hundred births successively. Then he becomes born as various fishes when he is freed. If anybody jests at the wealth of a Deva or a Br\=ahma\d{n}a, then he with his ten
generations above and below becomes fallen and he himself goes to the Dh\=umr\=andhak\=ara hell, terribly dark and filled with smoke. There his pains know no bounds and he lives there for four hundred years, eating smoke only. Then he becomes a mouse for seven births, and he becomes various birds and worms, various trees and various animals when ultimately he gets a human birth. If a Br\=ahmi\d{n} earns his livelihood by being an astrologer or if he be a physician and lives thereby or if he sells lac, iron, or oil, etc., he goes to the N\=agavstana Kunda hell where he lives for as many years as there are hairs on his body, tied up by snakes. Then he becomes born as various birds; ultimately he gets a human birth and becomes an astrologer for seven births and a physician for seven births. Then for sometime he becomes a cowherd (milkman), for sometimes a blacksmith; for sometimes a painter, when he becomes freed of his sin. O Chaste One! Thus I have described to you all the famous Kundas or hells. Besides there are innumerable small Kundas. The sinners go there and suffer the fruits of their own Karmas and travel through various wombs. O Fair One! What more do you now want to hear ? Say.

Here ends the Thirty-Fifth Chapter of the Ninth Book on the description of the various hells for the various sinners in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



