\chapter{On the eighty-six Kundas and their characteristics}

1-60. Dharmar\=aja said :-- All the Kundas (hells) are circular in form like the Full Moon. Of these, the Vah\d{n}ikunda has the fire lit at its bottom, by the help of various kinds of stones. This Kunda will not be destroyed till Mah\=apralaya comes. Here the sinners are tormented severely. It looks like a blazing coke. The flames are rising from it one hundred hands high. In circumference those flames are two miles. This is named Vah\d{n}ikunda. It is full of sinners crying loudly. It is constantly watched by My messengers who are chastising and punishing the sinners. Next comes the Tapta Kunda. It is filled with hot water and full of rapacious animals. The sinners there are severely beaten by My messengers and they are always crying out very loudly, which is being echoed and re-echoed all around terribly. It extends for one mile. This Kunda is filled with hot salt water and the abode of many crows. Then there is the Bhay\=anaka Kunda. It extends for two miles and it is filled with sinners. They are being punished by My messengers and they are incessantly crying, ``Save us, save us.''

Next comes the Visth\=a Kunda. It is filled with faeces and excrements where the sinners are moving without any food and with their palate and throats dry. Its size is two miles and it is very bad and ugly with foetid and nasty smell. It is always filled with sinners, who are being chastised by My D\=utas (messengers) and eat those faeces and excrements. The worms therein are constantly biting and stinging them and they are crying, ``deliver us, deliver us.'' Then comes the hot M\=uttra Kunda. It is filled with the hot urine and the worms thereof. The great sinners always dwell here. It measures four miles; and it is quite dark. My D\=utas always beat them and their throats, lips, palates are all dry. Then comes the \'Sle\d{s}ma Kunda. It is filled with phlegm and the insects thereof. The sinners dwell in phlegm and eat that phlegm. Then comes the Gara Kunda. It is filled with (factitious) poison. It measures one mile. The sinners eat this poison and dwell here. The worms thereof bite them. They tremble at the chastisement of My D\=utas and cry aloud. My messengers look like serpents, with teeth like thunderbolt and they are very furious and fierce, with their throats dry and their words very harsh. Then comes D\=usik\=a Kunda. It is filled with the rheum

and dirt of the eyes and it measures one mile. Innumerable worms are born therein. Numberless sinners live there, and as they move, the insects immediately bite and sting them. Next comes the Vas\=a Kunda. It is filled with the serum or marrow of the flesh and it measures one-half mile. The sinners dwell there, chastised and punished by My messengers. Then comes the \'Sukra Kunda. It measures two miles in diameter. The insects, born in the semen, bite the sinners, and they move on and on. Then comes the Rakta Kunda, with very offensive, foetid smell. It is deep like a well and filled with blood. The sinners dwell here, drinking blood. The insects therein are always biting them. Then follows the A\'sru Kunda. It measures in size one fourth the measure of the well (above-mentioned). It is always filled with hot tears of the eyes; and many sinners are seen there living weeping and crying and being bitten by the snakes. Then there is the G\=atra Mala Kunda. The sinners are chastised and punished there by My messengers and being bitten by the insects thereof, they eat the dirts of the body and dwell there. Then comes the Kar\d{n}a Mala Kunda. The sinners eat the wax of the ear and fill the place. The insects always bite them and they are crying aloud. It measures one fourth the measure of a V\=ap\={\i}. Then comes the Majj\=a Kunda. It is filled with fat and marrow, emitting foetid offensive odour. It measures one fourth the measure of a V\=ap\={\i}. The great sinners always dwell there. Then comes the M\=amsa Kunda. This is filled with the greasy flesh. It measures (one-fourth) that of a V\=ap\={\i}. Those who sell their daughters dwell here. My messengers always chastise and punish them and horrible insects bite and sting them and they cry, out of fear and agony, ``Save us, save us,'' and eat at times that flesh. Then come in succession the four Kundas Nakha, Loma and others. They also measure each one-fourth that of a V\=ap\={\i}. The sinners dwell there, always chastised by My messengers. Next comes the very hot T\=amra Kunda. Burning cokes exist on the top of very hot coppers. There are lakhs and lakhs of very hot copper figures in that Kunda. The sinners, being compelled by My messengers, are made to embrace each of these hot copper figures and they cry loudly and live there. It measures four miles. Then come the burning Ang\=ara Kunda and the hot Lauha Dh\=ara Kunda. Here the sinners are made to embrace the hot iron figures and, feeling themselves burnt, cry out of fear and agony. Whenever My messengers punish them, they immediately cry out, ``Save us, save us.'' It measures eight miles; and it is pitch dark and very awful. This is named the hot Lauha Kunda. Then come the Charma Kunda and Sur\=a Kunda. The sinners, beaten by My men, eat the skin and drink the hot urine and

dwell there. Then comes the \'S\=almal\={\i} Kunda; it is overspread with thorns and thorny trees, causing intense pain. It measures two miles. Millions and millions of great sinners are made by My men to fall from the tops of those trees down below where their bodies get pierced by very sharp thorns, six feet long; and thus they dwell there, beaten by My men. Out of thirst, their palates get dried up; and they cry out repeatedly, ``Water, water.'' Out of fear, they get very anxious and then their heads get broken by the clubs brought down on them by My men. So they move there like the beings burnt in very hot oil. Then comes the Vi\d{s}oda Kunda. It measures two miles in diameter and is filled with the poison of the serpent called Tak\d{s}akas. My men punish the sinners and they drink the poison thereof and dwell there. Then comes the hot Taila Kunda. There are no insects here. Only the great sinners dwell. All around burning coals are flaring and when My men beat the sinners, they run hither and thither. It is filled with horrible intense darkness and it is exceedingly painful. It looks dreadful and measures two miles. Then comes the Kunta Kunda. Sharp pointed iron weapons like tridents are placed in order all round. The sinners, pierced by those weapons, are seen encircling the Kunda. It measures one-half mile. Beaten by My men, their throats and lips get dried up. Then comes the Krimi Kunda. It is filled with terrible worms and insects, snake-like with sharp teeth, of the size of a \'Sanku (a S\=al tree) deformed and hideous looking; and it is filled with pitch darkness, terrible to look at. Beaten by My men, the great sinners dwell there. Then comes the P\=uya Kunda. It measures eight miles in diameter (or in circumference?). The sinners dwell and eat the pus thereof and, are beaten by My men. Then comes the Sarpa Kunda. Millions and millions of snakes of the length of a T\=al tree are existing there. These serpents encircle the sinners and as they bite them, My men also beat them at the same time. So there arises a general hue and cry, ``Save us, save us; we are done for.'' Then come in order the Dam\'sa Kunda, Ma\'saka Kunda, and the Garala Kunda. These are filled with gad-flies, mosquitoes, and poison respectively. Each of them measures one mile. The sinner\'s hands and feet are tied up. So when the gad-flies and mosquitoes fiercely sting them, and My men violently beat them simultaneously, they raise a loud uproar and are made to move on, in their tied states by My persons. Their bodies get thoroughly reddened and covered with blood by the stinging of the flies, etc. Then come the Vajra Kunda and the Vri\'schika Kunda filled respectively with Vajra insects and the scorpions. Each of them measures one-half that of the V\=ap\={\i}. The

sinners that dwell there, are incessantly bitten by insects (Vajras and scorpions). Then come in order the \'Sara Kunda, \'S\=ula Kunda, and the Khadga Kundas. They are filled respectively with arrows, spikes, and scimitars. Each of them measures one half that of the V\=ap\={\i}. The sinners are pierced by arrows, etc., and become covered over with blood and dwell there. Then comes the Gola Kunda. It is filled with boiling hot water and it is pitch dark. The sinners live there, bitten by the insects. This Kunda measures half that of the V\=ap\={\i}. The insects bite them and My men beat them so their fear knows no bounds; everyone of them is weeping and crying loudly. This Kunda is filled with hideously offensive smells. So the pains of the sinners are infinite. The Nakra Kunda comes next. It measures half the V\=ap\={\i}, is filled with millions and millions of crocodiles living in water. The horrible looking deformed sinners live there. The K\=aka Kunda then follows. The sinners here are being bitten by hundreds of deformed crows eating faeces, urines and phlegm. Then come the Manth\=ana Kunda and V\={\i}ja Kunda. These are filled respectively with insects called Manth\=ana and V\={\i}ja. Each of them measures one hundred Dhanus (one Dhanu - four hastas). Those insects are stinging the sinners and they cry out very loudly. Then follows the Vajra Kunda. It measures one hundred Dhanus. Many insects with their teeth as hard as thunderbolt live there and bite the sinners who cry out loudly. It is pitch dark. Then comes the hot P\=a\d{s}\=a\d{n}a Kunda. It measures twice that of the V\=ap\={\i}. It is so built of hot stones as it resembles a burning mass of coal. The sinners become restless with the heat and turn round and round in the middle. Then comes the P\=a\d{s}\=a\d{n}a Kunda and the L\=al\=a Kunda. The P\=a\d{s}\=a\d{n}a Kunda is made up of the sharp pointed stones, having sharp edges. Innumerable sinners dwell there. Many red beings live in the L\=al\=a Kunda. Then comes the Map\={\i} Kunda. Its size is one hundred Dhanus and its depth is two miles. It is made up of hot stones, each measuring the Anjana mountain. The sinners, beaten and driven by My persons, move on and on in the middle. Then comes the Ch\=ur\d{n}a Kunda. It measures two miles (in circumference) and is filled with (seven) ch\=ur\d{n}as (powders). The sinners, driven and beaten by My men, go on, restless hither and thither and eat the powders and get themselves burnt. Then comes the Chakra Kunda. Here a potter's wheel with sixteen sharp-edged spokes is constantly whirling round and round; the sinners are being crushed by this wheel.

61-80. Then comes the Vakra Kunda. Its depth is eight miles. lt is fashioned very much curved; and with and sharp slope it has gone down. It is built on the plan of a mountain cave, filled with hot water and it is enveloped with deep dense darkness. The aquatic animals there are biting

the sinners, who got very much restless and are crying out very loudly. Then comes the K\=urma Kunda. Here millions and millions of tortoises in the water awfully distorted, are biting the sinners. Then comes the Jv\=al\=a Kunda. It is built of fierce fiery flames. It measures two miles in circumference. The sinners here are always in great difficulty, with intense pain and crying out loudly. Next follows the Bha\d{s}ma Kunda. It measures two miles. The sinners get themselves well burnt in hot ashes and live there, eating the ashes. It is filled with hot stones and hot irons. The sinners here are always being burnt in hot irons and hot stones and their throats and palates are being parched up. Then comes the Dagdha Kunda. It is deep and horrible. It measures two miles in circumference. My messengers threaten always the sinners there. Then comes the S\=uch\={\i} Kunda. It is filled with salt water. Waves are always rising there. It is filled with various aquatic animals making all sorts of noises. It measures eight miles in circumference and it is deep and dark. The sinners here cannot see each other and are bitten by the animals. Pained very much, they cry out loudly. Then comes the Asipattra Kunda. On the top surface of the Kunda there is a very big T\=al tree very high. The edges of the leaves of this tree are sharp like the edge of a sword. One mile below this T\=al tree is situated the Kunda. The sharp-edged T\=al leaves, then, fall on the bodies of the sinners from the height of a mile and they get cut and wounded; blood comes out of them and the sinners, in great pain, cry out ``save, save.'' It is very deep, very dark and filled with Rakta K\={\i}ta a kind of blood like insects. This is the horrible Asipatra Kunda. Next comes the K\d{s}ura Dh\=ara Kunda, measuring one hundred Dhanus (one Dhanu - four hastas). It is filled with keen-edged weapons, as sharp as nice razors. The blood of the sinners is flowing here profusely. Then comes the S\=uch\={\i} Mukha Kunda, filled with sharp weapons of the form of long needles. It measures fifty Dhanus. The sinners get pierced by them end are constantly emitting blood. Their intense pain knows no bounds. Then comes the Gok\=amukha Kunda; inhabited by a sort of insect, called Gok\=a. They look like mouths; hence they are named Gok\=amukha. It is deep like a well and it measures twenty Dhanus. The great sinners suffer an intense amount of pain there. They have got to keep their mouths always downwards as the Gok\=a insects always bite and sting them. Then comes the Nakra Kunda. It resembles like the mouth of a crocodile and measures sixteen Dhanus. It is deep like a well and numbers of sinners dwell there. Then comes the Gaja Dam\'sa Kunda. It measures one hundred Dhanus. Next comes the Gomukha Kunda. It measures thirty Dhanus and resembles the mouth of a cow. It gives incessant pains and troubles to the sinners.

81-101. Then comes the Kumbh\={\i}p\=aka Kunda. It is like a wheel resembling that of this K\=alachakra, very horrible; and it is rotating incessantly. It looks like a water-jar, measuring eight miles, and it is quite dark. Its depth is one lakh Puru\d{s}as of the height of 100,000 persons. There are many other Kundas, Tapta Taila Kunda and Tapta Taila T\=amra Kunda, etc., within it. This Kunda is filled with almost unconscious great sinners and insects. They beat each other and cry out loudly. My messengers also threaten them with clubs and Mu\d{s}alas. So at times they fall dizzy-headed, at times they get unconscious, and sometimes they get up and cry. O Fair One! The numbers of sinners here equal to four times that of all the other sinners in all the other Kundas. They know no death, however much you beat them. Their lives persist. For the body is built up for sufferance, it is indestructible. This Kumbh\={\i}p\=aka Kunda is the chief of all the Kundas. This Kunda where the sinners are tied to a thread built by K\=ala, where My men lift the sinners on high at one time, and sink them down below at another time, where the sinners becoming suffocated for a long time, get unconscious, where their sufferings know no bounds, where it is filled with boiling oil, is named the K\=alasutra Kunda. Then comes the Matsyoda Kunda, hollow like a well. It is filled with boiling water and it measures twenty-four Dhanus. Next comes the Abatoda Kunda. It measures one hundred Dhanus. The sinners get their bodies burned and chastised by My persons, live there. No sooner they drop into the water of this Kunda, then they are attacked with all sorts of diseases. Then comes the Krimikantuka Kunda. The sinners are bitten by the Krimi Kantuka insects and cry out loudly, creating a general consternation and live there. Its another name is Aruntuda Kunda. Next comes the P\=am\'su Kunda. It measures one hundred Dhanus. It is overspread with burning rice husks. The sinners eat those hot husks and live there. Then comes the Pa\'save\d{s}tana Kunda. It measures two miles. No sooner the sinners fall in this Kunda than they are twined round by this rope or P\=a\'sa. Hence its name. Then comes the S\=ulaprota Kunda. It measures twenty Dhanus. No sooner the sinners fall here than they are encircled with the S\=ul\=astra (darts). Then comes the Prakampana Kunda. It measures one mile. It is filled with ice-cold water. The sinners, going there, shiver at once. Next follows the Ulka Kunda. It measures twenty Dhanus. It is filled with burning to ashes and meteors. My messengers thrust the torches and meteors into the mouths of the sinners living there. Next comes the Andha-K\=upa Kunda. It is pitch-dark, shaped like a well, circular and very horrible. The sinners beat each other and eat the insects thereof. Their bodies are burnt with hot water; they cannot see anything on account of dire darkness.

102-118. The Kunda where the sinners are pierced by various weapons is known as the Vedhana Kunda. It measures twenty Dhanus. Then comes the Dandat\=ada\d{n}a Kunda. It measures sixteen Dhanus. The sinners dwell here, threatened by My messengers. Then comes the J\=alarandhra Kunda. Here the sinners live encompassed by a great net as fishes, etc., are tied in a net. Next comes the Dehach\=ur\d{n}a Kunda. It is quite dark and its depth is that of the height of one koti persons; its circumference is twenty Dhanus. The sinners, here, encompassed by iron chains are made to fall below where their bodies are reduced to powders and they are inert and almost unconscious. The Kunda where the sinners are crushed and threatened by My messengers is known as the Dalana Kunda; it measures sixteen Dhanus in circumference. Next comes the \'So\d{s}ana Kunda. It is deep up to the height of one hundred persons and it is very dark. It measures thirty Dhanus. On falling on the hot sand, the throats and palates of sinners get dried up. Their pain knows no bounds. Hence it is called the \'So\d{s}ana Kunda. Then comes the Ka\d{s}a Kunda. It measures one hundred Dhanus. It is filled with the juices of skins and its smell is very offensive. The sinners eat those astringent waters and live there. Then comes \'S\=urpa Kunda. It measures twelve Dhanus and is extended like a winnowing basket. It is filled with hot iron dust and many sinners live there, eating those foetid iron dusts. Next comes the Jv\=al\=amukha Kunda. It is filled with red hot sand. From the (bottom) centre rises a flame, over-spreading the mouth of the Kunda. It measures twenty Dhanus. The sinners are burnt here by the flame and live awfully; they get fainted no sooner they are dropped in this Kunda. Then comes the Dhumr\=andha Kunda. It is dark, quite filled with smoke. Within that the hot bricks are placed. The sinners get suffocated with smoke; and their eye-sight becomes also obstructed. It measures one hundred Dhanus. Then comes the N\=agabestana Kunda. It is encircled and filled with the serpents. No sooner the sinners are let fall there, than they are surrounded by the snakes. O S\=avitr\={\i}! Thus I have spoken to you about the eighty-six Kundas and their characteristics. Now what more do you want to hear? Say.

Here ends the Thirty-seventh Chapter of the Ninth Book on the eighty-six Kundas and their characteristics in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



