\chapter{On the churning of the ocean and on the appearing of Lak\d{s}m\={\i}}

1-2. N\=ar\=aya\d{n}a said :-- O Twice-born! Indra then remembered Hari and took Brihaspati, the Guru of the Devas, to the assembly of Brahm\=a, accompanied by the other Devas. They soon reached the region of Brahm\=a and no sooner they saw Him, than Indra and the other Devas and Bihaspati all bowed down to Him.

3-25. Brihaspati, the \=Ach\=arya of the Devas, then communicated all to Brahm\=a. Hearing this, He smiled and spoke, addressing Indra :-- O Indra! You have been born in My race; nay, you are My grandson; the \=Arya Brihaspati is your Guru; you yourself are the ruler of the Devas and you are very wise and sagacious; the mother of your father is the powerful Dak\d{s}a, a great Vi\d{s}\d{n}u Bhakta. How is it, then that when the three sides of the family are pure, one would turn out so haughty and arrogant! Whose mother is so much devoted to her husband, whose father, mother's father and mother's brother are self-controlled and of pure Sattva Gu\d{n}a, he is not expected to be so very haughty! Every man may be guilty to Hari for the three faults :-- For some fault due to that of his father, or of his mother's father, or of Guru, the Spiritual Teacher. Bhagav\=an N\=ar\=aya\d{n}a, resides in this great holy temple of our this physical body controlling the hearts of all. At whatever moment, \'Sr\={\i} N\=ar\=aya\d{n}a quits this temple-body, at that moment this body becomes dead. I myself am the mind; \'Sankara is the Knowledge, Vi\d{s}\d{n}u is the vital

breaths, Bhagavat\={\i} Durg\=a is the intelligence (Buddhi), sleep, etc., the powers of Prakriti; when these are being reflected on by the \=Atman, J\={\i}va is formed with a body for enjoyment called Bhoga \'Sar\={\i}rabhrit. When a king departs, his attendants also follow him; so when this \=Atman departs from this temple of body, his attendants mind, buddhi, etc., instantly depart from this body and follow the \=Atman. O Indra! We all, are verily, the parts of \'Sr\={\i} Kri\d{s}\d{n}a. I myself, \'Siva, Ananta Deva, Vi\d{s}\d{n}u, Dharma, Mah\=a Vir\=at, you all are His parts and entirely devoted to Him. And you have shown contempt for His flower. Bhagav\=an \'Sankara, the Lord of the Bh\=utas, worshipped the lotus-feet of \'Sr\={\i} Kri\d{s}\d{n}a with that flower. The \d{R}i\d{s}i Durv\=as\=a gave you that flower. But you showed disrespect to it. The flower, P\=arij\=ata, after being offered at the lotus-feet of Kri\d{s}\d{n}a, should be placed on the head of an Immortal; His worship is to be done first; and it is the foremost amongst the Devas. So you are now being afflicted by the inevitable course of Fate; Fate is the most powerful of all. Who can save that unlucky man against whom Fate has turned? Seeing that you have rejected the flower offered to \'Sr\={\i} Kri\d{s}\d{n}a, \'Sr\={\i} Lak\d{s}m\={\i} Dev\={\i} has left you out of anger. Now come with Me and with your family priest Brihaspati to Vaikuntha and worship the Lord of Lak\d{s}m\={\i}; then by His grace you may get back your Heavens. Thus saying, Brahm\=a with Indra and all the other Devas, went to the Eternal Puru\d{s}a, Bhagav\=an N\=ar\=aya\d{n}a and saw that He was full of Fire and Energy like one hundred koti summer midday Suns, yet perfectly cool and calm. He has no beginning, and no end, nor any middle. He is Infinite. The four-armed P\=ari\d{s}adas, Sarasvat\={\i}, the four Vedas, and the Ganges, all were surrounding Him. Seeing Him, Brahm\=a and the other Devas bowed down with devotion and began to chant hymns to Him with tears in their eyes. Brahm\=a, then, informed Him of everything when all the Devas, dispossessed of their places, began to weep before Him. N\=ar\=aya\d{n}a saw that the Devas were very afraid and much distressed. They had no jewel ornaments as before, no vehicles (V\=ahanas), nor the Daivic splendour as before, not that brillance; always fearful. Then N\=ar\=aya\d{n}a, the Destroyer of fear, seeing the Devas in that state, addressed Brahm\=a and the other Devas :-- O Br\=ahma\d{n}! O Devas! Discard all fears. What fear can overcome you? I am here. I will give you again the immoveable prosperous R\=ajya Lak\d{s}m\={\i} (the Laksm\={\i} of the kings).

26-47. But for the present, I give you some advice proper for this moment. Listen. There are endless universes where exist innumerable persons. All of them are under Me. So know verily, that

I am under them also. My devotees regard Me as the Highest. They know no other than Me; they are fearless; so I do not remain in that house where My devotees are dissatisfied. I instantly quit that house with Lak\d{s}m\={\i}. Durv\=as\=a \d{R}i\d{s}i is born in part of \'Sankara. He is highly devoted to Me. He is a pakk\=a Vai\d{s}\d{n}ava. He cursed you and, as a matter of fact, I and Lak\d{s}m\={\i} instantly left your house.
Lak\d{s}m\={\i} does not reside in that house where conch shells are not blown, where there are no Tulas\={\i} trees, where there is no worship of \'Siva and \'Siv\=a, where the Br\=ahma\d{n}as are not fed. O Br\=ahma\d{n}! O Devas! Where I and My Bhaktas are blamed, Mah\=a Lak\d{s}m\={\i} becomes greatly displeased. She instantly goes away out of that house. Lak\d{s}m\={\i} does not stay even for a moment in that house where the stupid person, without any devotion for me, takes his food on the Hariv\=asara Ek\=ada\'s\={\i} day (the eleventh day of the moon's wane or increase) or on My anniversary birthday. If anybody sells My name or his own daughter, where the guests are not served, Lak\d{s}m\={\i} quits that house instantly and goes away.

(Note :-- That Guru is the Real Guru, who, being capable, imparts the name of God to worthy persons without taking any fee at all.)

If the son of an unchaste woman be a Br\=ahma\d{n}a, he and the husband of an unchaste woman are great sinners. If anybody goes to such a person's house or eats the food of a \'S\=udra during a Sr\=addha ceremony, Lak\d{s}m\={\i} becomes very angry and vacates that house. Being a Br\=ahmi\d{n}, if one burns a \'S\=udra's corpse, one becomes very wretched and the vilest of the Br\=ahmi\d{n}s. Lak\d{s}m\={\i} never stays for a moment more in that house. Being a Br\=ahmi\d{n}, if he be a \'S\=udra's cook and drives oxen, Lak\d{s}m\={\i} fears to drink water there and quits his house. Being a Br\=ahmi\d{n}, if his heart be unholy, if he be cruel, envies others and blames persons, if he officiates as a priest for the \'S\=udra, Lak\d{s}m\={\i} Dev\={\i} never stays in his house. The World-Mother never stays even for a moment in his house who eats at the house of one who marries an unmarried girl twelve years old in whom menstruation has commenced. He who cuts grass by his nails, or writes on the ground with his nails, or from whose house a Br\=ahma\d{n}a guest goes back disappointed, Lak\d{s}m\={\i} never stays in his house. If any Br\=ahma\d{n}a eats food at the early sunrise, sleeps during the day or engages in a sexual intercourse during the day, Lak\d{s}m\={\i} never stays in his house. Lak\d{s}m\={\i} slips away from that Br\=ahma\d{n}a who is devoid of \=Ach\=ara (rules of conduct), who accepts gifts from \'S\=udras, from him who remains uninitiated in his Mantram. The ignorant man who sleeps naked and with his feet wet, who laughs always, and always

talks at random on disconnected subjects like a mad man, is forsaken at once by Lak\d{s}m\={\i}. Lak\d{s}m\={\i} becomes angry and goes away from the house of that man who applies oil all over his body first and then touches the bodies of others and always makes some sounding noise on several parts of his body. If any Br\=ahma\d{n}a forsakes observing vows, fastings, the Sandhy\=a ceremony, purity and devotion to Vi\d{s}\d{n}u, Kamal\=a (Lak\d{s}m\={\i}) does not remain in his house any longer. If anybody blames always the Br\=ahma\d{n}as and shews his hatred always towards the Br\=ahma\d{n}as, if he does injury to the animals, and if he does not indulge in his heart anything of pity, kindness, Lak\d{s}m\={\i}, the Mother of the Worlds, quits him. O Lotus-born! But where Hari is worshipped and Hari's Name is chanted, Lak\d{s}m\={\i}, the Mother of all auspiciousness, remains there. Lak\d{s}m\={\i} remains where the glories of \'Sr\={\i} Kri\d{s}\d{n}a and His Bhaktas are sung.

48-59. Lak\d{s}m\={\i} always remains there with the greatest gladness where conch-shells are blown, where there are conchsells, the \'S\=alagr\=ama stone, the Tulas\={\i} leaves and the service and meditation of Lak\d{s}m\={\i} are daily done.

Where the phallic emblem of \'Siva is worshipped, and His glories sung, where \'Sr\={\i} Durg\=a is worshipped and Her glories are sung, Lak\d{s}m\={\i}, the Dweller in the Lotus, remains there. Where the Br\=ahma\d{n}as are honoured and they are gladly feasted, where all the Devas are worshiped; the chaste Lak\d{s}m\={\i}, the Lotus-faced, remains there. Thus saying to the Devas, the Lord of Lak\d{s}m\={\i} said :-- O Dev\={\i}! Go without any any delay to the K\d{s}iroda Ocean and incarnate there in part. He then addressed Brahm\=a and said :-- ``O Lotus-born! You also better go there and churn the K\d{s}iroda Ocean; when Lak\d{s}m\={\i} will arise, give her to the Devas.'' O Devar\d{s}i! Thus saying, the Lord Kamal\=a went to His inner compartment. On the other hand the Devas, after a long time, reached the shores of the K\d{s}iroda Ocean. The Devas and the Daityas then made the Golden Mountain (the Sumeru) the churning rod, the Deva Kurma (the tortoise), the churning pot and Ananta Deva (the thousand headed serpent) the churning cord and began to churn the ocean. While churning was going on, by and by arose Dhanvantar\={\i}, Amrita (the nectar), the horse Uchchaih\'srav\=a, various other invaluable jewels that were desired, the elephant Air\=avata and the beautiful eyed Lak\d{s}m\={\i}. Vi\d{s}\d{n}upriy\=a, \'Sr\={\i} Lak\d{s}m\={\i} Dev\={\i}, no sooner She got up from the ocean, she, the chaste woman, presented on the neck of N\=ar\=aya\d{n}a, of beautiful appearance, the Lord of all, Who slept on the K\d{s}iroda ocean the garland (indicative of accepting Him for her bridegroom). Then Brahm\=a and Mahe\'svara and the other Devas gladly worshipped Her and chanted hymns. At this time Lak\d{s}m\={\i} Dev\={\i} being pleased, cast a favourable glance towards the homes of the Devas,

in order to free them from their curse. Then, by the grace of Mah\=a Lak\d{s}m\={\i} and by the granting of the boon by Brahm\=a, the Devas got back their own possessions from the hands of the Daityas. O N\=arada! Thus I have described to you the story of Lak\d{s}m\={\i} Dev\={\i}, the Essence of all Essences, and very pleasant to hear. Now what more do you want to hear? Say.

Here ends the Forty-first Chapter of the Ninth Book on the churning of the ocean and on the appearing of Lak\d{s}m\={\i} in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



