\chapter{On the anecdote of Surabhi}

1. N\=arada said :-- ``O Bhagav\=an! Who was that Surabhi, who came down from the region of Goloka. I want to hear Her life. Kindly describe.''

2-23. N\=ar\=aya\d{n}a spoke :-- ``O Devar\d{s}i! The Dev\={\i} Surabhi sprang in the Goloka. She was the first in the creation of cows; and, from Her, all the other cows have come. She is the Presiding Deity of the cows. I will now speak Her history from the very beginning. Hear. Before, She appeared in the holy Vrind\=aban. One day the Lord of R\=adh\=a, surrounded by the Gop\={\i}s, was going gladly with R\=adh\=a to the Holy Vrind\=avan. There he began to enjoy in a solitary place with great pleasure. He is All Will and suddenly a desire arose in His mind that He would drink milk. Then He created easily the Dev\={\i} Surabhi, full of milk, with Her calf, from His own left side. The calf of Surabhi is nothing else but Her wish personified. Seeing Surabhi, \'Sr\={\i}d\=ama milked Her in a new earthen jar. The milk is more sweet than even the nectar and it prevents birth and death! The Lord of the Gop\={\i}s drank the milk. What milk dropped out of the jar, created a big tank! The tank measured one hundred Yojanas in length and in breadth and is known in Goloka by the name of K\d{s}\={\i}ras\=agara. The Gop\={\i}kas and R\=adh\=a play therein. At the will of \'Sr\={\i} Kri\d{s}\d{n}a, Whose Nature is All Will, that tank become full of excellent gems and jewels. Then, from every pore of Surabhi, there appeared suddenly one lakh koti K\=amadhenus (cows who yields according to one's desires). So much so that every Gopa who used to live there in Goloka had one K\=amadhenu and each house had one such. Their calves again became so many that no limit can be put to them. Thus, by degrees, the whole universe was filled with cows. This is the origin of the Cow Creation. O N\=arada! Surabhi was first worshipped by Bhagav\=an \'Sr\={\i} Kri\d{s}\d{n}a. Therefore She is so much honoured everywhere. On the day next the Div\=al\={\i} night (new moon in the month of October), Surabhi was worshipped by the command of \'Sr\={\i} Kri\d{s}\d{n}a. This is heard from the mouth of Dharma Deva. O Child! Now hear the Dhy\=anam, Stotra, and the method of worship of Surabhi as

mentioned in the Vedas. I will now speak on this. ``Om Surabhyai namah,'' is the principal six-lettered mantra of Surabhi. If anybody repeats this mantra one lakh times, he becomes Siddha in this mantra. This is like Kalpa Vrik\d{s}a (a tree yielding all desires) to the devotees. The Dhy\=anam of Surabhi is mentioned in the Yajur Vedas. Success, prosperity, increase and freedom come as the result of worshipping Surabhi. The Dhy\=anam runs as follows :-- ``O Dev\={\i} Surabhi! Thou art Lak\d{s}m\={\i}, Thou art best, Thou art R\=adh\=a; Thou art the chief companion of \'Sr\={\i} R\=adh\=a, Thou art the first and the source of the cow-creation, Thou art holy and Thou sanctifiest the persons; Thou fufillest the desires of the devotees and Thou purifiest the whole universe. Therefore I meditate on Thee.'' Reciting this Dhy\=anam, the Br\=ahma\d{n}as worship the Dev\={\i} Surabhi in jars, on the heads of cows, or on the pegs where cows are fastened or on \'S\=alagr\=ama stone or in water or in fire. O Muni! He who worships with devotion on the next day morning after Div\=al\={\i} night, becomes also worshipped in this world. Once a day in the V\=ar\=ahakalpa Surabhi did not yield milk, by the influence of Vi\d{s}\d{n}u M\=ay\=a. The Devas became very anxious. Then they went to the Brahmaloka and began to praise Brahm\=a. At His advice, Indra began to chant hymns (Stotra) to Surabhi :--

24-33. The Devendra said :-- ``O World-Mother! O Dev\={\i}! O Mah\=a Dev\={\i}! O Surabhi! Thou art the source of the cow creation. Obeisance to Thee! Thou art the dear companion of R\=adh\=a; Thou art the part of Kamal\=a; Thou art dear to \'Sr\={\i} Kri\d{s}\d{n}a; Thou art the mother of cows, I bow down to Thee. Thou art like the Kalpa Vrik\d{s}a (a tree yielding all desires), Thou art the Chief of all; Thou yieldest milk, wealth and prosperity and increase thereof. So I bow down to Thee. Thou art auspicious, Thou art good, Thou bestowest cows. Obeisance to Thee! Thou givest fame, name and Dharma. So I bow down to Thee.'' O N\=arada! Thus hearing the praise sung by Indra, the eternal Surabhi, the originator of the world, became very glad and appeared in the Brahmaloka. Granting boon to Mahendra, so very rare to others and desired by him, Subrabhi went to the Goloka. The Devas, also, went back to their own abodes. The whole world was now full of milk; clarified butter came out of the milk; and from clarified butter sacrifices began to be performed and the Devas were fed and they became pleased. O Child! He who recites this holy Stotra of Surabhi with devotion, gets cows, other wealth, name, fame and sons. The reciting of this Stotra qualifies one as if he had bathed in all the sacred places of pilgrimages and he had acquired the fruits of all the sacrifices. Enjoying happiness in this world, be goes in

the end to the Temple of \'Sr\={\i} Kri\d{s}\d{n}a. There living long in the service of Kri\d{s}\d{n}a, he becomes able to be a son of Brahm\=a.

Here ends the Forty-ninth Chapter of the Ninth Book on the anecdote of Surabhi in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



