\chapter{On the history of Sv\=ah\=a}

1-4. N\=arada said :-- ``O \d{R}i\d{s}i N\=ar\=aya\d{n}a! O Highly Fortunate One! O Lord! Thou art equal to N\=ar\=aya\d{n}a whether in beauty or in qualities, or in fame or in energy or in everything. Thou art equal to N\=ar\=aya\d{n}a. Thou art the foremost of the J\~n\=anins; there cannot be found a second like Thee as a Siddha Yog\={\i}, the ascetics and the Munis. And Thou art the crest of the knowers of the Vedas. I have heard the wonderful anecdote of Mah\=a Lak\d{s}m\={\i} that Thou hast told me. Now tell me any other thing that is unknown, very good in everyway, in accordance with Dharma, in the Vedas, and that which is not as yet written in the Pur\=a\d{n}as.''

5-6. N\=ar\=aya\d{n}a said :-- ``There are many wonderful hidden anecdotes that are not published as yet in the Pur\=a\d{n}as. What you have heard is very small in fact. Please mention to me what you like to hear, what you think as best amongst them and I will describe that to you.''

7-8. N\=arada said :-- ``When clarified butter is poured as libations in all the sacrificial ceremonies to the gods, Sv\=ah\=a is to be uttered everywhere as excellent, and commendable; so Svadh\=a is to be repeated in the offerings when the oblations are offered to the Pitris, the deceased ancestors. Then, again, Dak\d{s}i\d{n}\=a (the sacrificial fee) is always to be paid as right and excellent. So, O Knower of the Vedas! I like to hear the accounts of Svadh\=a, and Dak\d{s}i\d{n}\=a and their merits. Please now tell me about them.''

9. S\=uta said :-- ``Hearing the words of N\=arada, N\=ar\=aya\d{n}a \d{R}i\d{s}i smiled and began to speak the very ancient words of the Pur\=a\d{n}as.''

10-11. N\=ar\=aya\d{n}a said :-- Before the creation, the Devas assembled in the beautiful Brahm\=a's Council to decide on their food question.

They all brought to the notice of Brahm\=a the scarcity about their food. Brahm\=a promised to remove their food difficulties and began to chant hymns to Hari.

12. N\=arada said :-- ``O Lord! Bhagav\=an N\=ar\=aya\d{n}a Himself incarnated in part as Sacrifice. Are not the Devas satisfied when the Br\=ahma\d{n}as pour oblations of ghee to the Devas in those sacrifices?''

13-17. N\=ar\=aya\d{n}a said :-- O Muni! The Devas, in fact, did not get the offerings of the clarified butters that were poured in with devotion in sacrificial acts by the Br\=ahma\d{n}as and K\d{s}attriyas. So they were very much depressed and went again to the council of Brahm\=a and informed Him that they could not get any food for themselves. On hearing this, Brahm\=a at once meditated and took refuge of \'Sr\={\i} Kri\d{s}\d{n}a; Kri\d{s}\d{n}a advised them to worship M\=ula Prakriti. Brahm\=a then, by the command of \'Sr\={\i} Kri\d{s}\d{n}a, began to meditate on Prakriti, worship Her and chant hymns to Her. Then from the part of Prakriti, an all-powerful Dev\={\i} appeared. She was very beautiful, Shy\=am\=a (of a blue colour) and very lovely. This Dev\={\i} was Sv\=ah\=a. She looked always gracious with smile in Her face; it seemed that She was always ready to show favour towards the Bhaktas. She appeared before Brahm\=a and said :-- ``O Lotus born! Want any boon you desire.''

18-22. Hearing Her words, the Creator spoke reverentially :-- ``O Dev\={\i}! Let Thou be the burning power of Fire; without Thee Fire would not be able to burn anything. At the conclusion of any Mantra, whoever taking Thy name, will pour oblations in the Fire to the Gods, will cause those oblations to go to the Gods and reach them. And then they will be very glad. O Mother! Let Thou be the wealth of Fire, the beauty and housewife of Fire; let Thou be incessantly worshipped in the regions of the Gods and amongst men and other beings.'' Hearing these words of Brahm\=a, Sv\=ah\=a Dev\={\i} became very sad and expressed Her own intentions :-- ``I will get Kri\d{s}\d{n}a as my husband; let me perform Tapasy\=a as long as it takes. This is my object. All other things are false as dreams.

23-28. I always meditate with devotion on the lotus-feet of \'Sr\={\i} Kri\d{s}\d{n}a, serving Which You have become the Creator of this world, \'Sambhu has become the Conqueror of death, Ananta Deva is supporting this universe, Dharma is the Witness of the virtuous, Gane\'sa is getting, first of all, the first worship. Prakriti Dev\={\i} has become the adorable of all and the Munis and the \d{R}i\d{s}is respected by all.'' O Child! Having spoken thus to the Lotus-born, Padm\=a, with Her lotus-face, meditating incessantly on the Lotus-Feet of \'Sr\={\i} Kri\d{s}\d{n}a, free from any disease, started to perform

tapasy\=a for attaining Him. She, first of all, stood on one leg and practised austerities for one lakh years. Then She saw the Highest Puru\d{s}a \'Sr\={\i} Krisna, Who is beyond Prakriti and Her attributes. The beautiful amorous Sv\=ah\=a, seeing the Lovely Form of the Lord of Love, fainted.

29-43. The omniscient Bhagav\=an Kri\d{s}\d{n}a knowing Her intentions, took Her to His lap, reduced very much in body by long continued Tapasy\=a, and He said :-- ``O Dev\={\i}! Thou shalt be My wife in the next Var\=aha Kalpa. Then Thou wilt be the daughter of Nagnajit and wilt be known by the name of N\=agnajit\={\i}. O Beloved! At present let Thou be the Energy of Fire and be His wife. By My boon Thou wilt be worshipped by all. Fire will make Thee the Lady of His house and take the utmost care of Thee. Thou wilt be able to enjoy easily with Him.'' O N\=arada! Thus saying to Sv\=ah\=a, Bhagav\=an disappeared. On the other hand, Fire came in there by the command of Brahm\=a, with a doubtful mind and began to meditate on Her, the World Mother as per S\=ama Veda and worshipped Her. He then pleased and married Her with mantrams, etc. For one hundred divine years they enjoyed each other. In a very solitary place while they were enjoying each other, Sv\=ah\=a Dev\={\i} felt pregnant. For full twelve divine years She retained Her pregnancy. Then She delivered gradually three sons Daksi\d{n}\=agni, G\=arhyapaty\=agn\={\i}, and \=Ahavan\={\i}y\=agni. The \d{R}i\d{s}is, Munis, Br\=ahma\d{n}as, K\d{s}attriyas poured oblations of clarified butter pronouncing the terminal mantra ``Svah\=a.'' He who pronounces this excellent terminal Mantra ``Sv\=ah\=a'' gets immediate success in his actions. Then all the mantras without ``Sv\=ah\=a'' in the end became impotent as snakes become when void of poison, the Br\=ahma\d{n}as when they are devoid of the knowledge of the Vedas, the wife when she does not serve her husband, the men when they turn illiterate and the trees, when void of fruits and branches. O Child! The Br\=ahma\d{n}as then became satisfied. The Devas began to receive the oblations. With the ``Sv\=ah\=a'' mantra everything turned out fruitful. Thus I have described to you the anecdote of ``'Sv\=ah\=a.'' One who hears this essential anecdote gets his happiness enhanced and the Mok\d{s}a in his hands. What more do you want to hear ? Say.

44. N\=arada said :-- I like to hear how Fire worshipped Sv\=ah\=a and recited stotras (hymns of praise) to Her. Kindly tell me the method of worship, the Dhy\=anam and Stotra.

45-49. N\=ar\=aya\d{n}a said :-- O Best of Br\=ahma\d{n}as! I now tell you the meditation (Dhy\=anam) as per S\=ama Veda, the method of worship and stotra. Listen attentively. At the commencement of any sacrificial ceremony, one should first of all worship whether on the \'S\=alagr\=ama stone or in an earthen jar (ghata), the Dev\={\i} Sv\=ah\=a and then commence the

ceremony with the expectation of getting the desired fruit. The following is the Dhy\=anam (meditation) of Sv\=ah\=a Dev\={\i} :-- O Dev\={\i} Sv\=ah\=a! Thou art embodied of the Mantras; Thou art the success of the Mantras; Thou art Thyself a Siddh\=a; Thou givest success and the the fruits of actions to men; Thou dost good to all. Thus meditating, one should offer P\=adya (water for washing the feet), etc., uttering the basic Mantra; success then comes to him. Now hear about the Radical Seed Mantra. The said mantra (M\=ula mantra) is this :-- ``Om Hr\={\i}m \'Sr\={\i}m Vahnij\=ay\=ayai Devyai Sv\=ah\=a.'' If the Dev\={\i} be worshipped with this Mantra, all the desires come to a successful issue.

50-54. Fire recited the following stotra :-- ``Thou art Sv\=ah\=a, Thou art the Beloved of Fire, Thou art the wife of Fire; Thou pleasest all; Thou art the \'Sakti, Thou art the action, Thou art the bestower of K\=ala (time); Thou dost digest the food; Thou art the Dhruv\=a; Thou art the resort of men; Thou art the burning power; Thou canst burn everything, Thou art the essence of this world; Thou art the deliverer from the terrible world; Thou art the life of the gods and Thou nourishest the Gods.'' O N\=arada! He who reads with devotion these sixteen names, gets success both in this world as well as in the next. None of his works become deficient in any way; rather all the works are performed successfully and with a successful issue. Reading this stotra, one who has no wife, gets wife. So much so that the man who recites the stotra gets for his wife equal to Rambh\=a, the heavenly nymph, and passes his time in greatest bliss.

Here ends the Forty-third Chapter of the Ninth Book on the history of Sv\=ah\=a in \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, the Mah\=a Pur\=a\d{n}am, of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



