\chapter{On the story of Sv\=ayambhuva Manu}

1-6. N\=arada said :-- O N\=ar\=aya\d{n}a! O Thou, the Supporter of this whole world! The Preserver of all! Thou hast described the glorious characters of the Dev\={\i}, that take away all the sins. Kindly describe now to me the several forms that the Dev\={\i} assumed in every Manvantara in this world as well as Her Divine Greatness. O Thou, full of mercy! Describe also how and by whom She was worshipped and praised; how She, so kind to the devotees, having been thus pleased, fulfilled their desires. I am very eager to hear these, the very best and blissful characters of the Dev\={\i}. \'Sr\={\i} N\=ar\=aya\d{n}a said :-- Hear, O Mahar\d{s}i! The glories and greatness of the Dev\={\i} Bhagavat\={\i} leading to the devotion of the devotees, capable of giving all sorts of wealth and destroying all sins. From the navel lotus of Vi\d{s}\d{n}u, the holder of the Chakra (discus), was born Brahm\=a, the Creator of this universe, the great Energetic One, and the Grand Sire of all the worlds.

7-14. The four faced Brahm\=a, on being born, produced from His mind Sv\=ayambhuva Manu and his wife \'Satar\=up\=a, the embodiment of all virtues. For this very reason, Sv\=ayambbuva Manu has been known as the mind-born son of Brahm\=a. Sv\=ayambhuva Manu got from Brahm\=a the task to create and multiply; he made an earthen image of the Dev\={\i} Bhagavat\={\i}, the Bestower of all fortunes, on the beach of the sanctifying K\d{s}\={\i}ra Samudra (ocean of milk) and he engaged himself in worshipping Her and began to repeat the principal mystic mantra of V\=agbhava (the Deity of Speech). Thus engaged in worship, Sv\=ayambhuva Manu conquered by and by his breath and food and observed Yama, Niyama and other vows and became lean and thin. For one hundred years he remained standing always on one leg and became successful in controlling his six passions lust, anger, etc. He meditated on the feet of that \=Ady\=a \'Sakti (the Primordial \'Sakti) so much that he became inert like a vegetable or mineral matter. By his Tapas the Dev\={\i}, the World Mother appeared before him and said :-- ``O King! Ask divine boons from Me.'' Hearing these joyous words, the King wanted his long cherished and heart-felt boon, so very rare to the Devas.

15-22. Manu said :-- O Large eyed Dev\={\i}! Victory to Thee, residing in the hearts of all! O Thou honoured, worshipped! O Thou! the Upholdress of the world! O Thou, the Auspicious of all auspicious!

By Thy Gracious Look, it is that the Lotus born has been able to create the worlds; Vi\d{s}\d{n}u is perserving and Rudra Deva is destroying in a minute. By Thy command it is that Indra, the Lord of \'Sach\={\i}, has got the charge of controlling the three Lokas; and Yama, the Lord of the departed, is awarding fruits and punishing according, to their merits or demerits, the deceased ones. O Mother! By Thy Grace, Varu\d{n}a, the holder of the noose, has become the lord of all aquatic creatures and is preserving them; and Kuvera, the lord of the Yak\d{s}as, has become the lord of wealth. Agni (fire), Nairrit, V\=ayu (wind), \=I\'s\=ana and Ananta Deva are Thy parts and have grown by Thy power. Then, O Dev\={\i}! If Thou desirest to grant me my desired boon, then, O Thou! the Auspicious One! Let all the great obstacles to my work of procreating in this universe and increasing my dominions die away. And if anybody worships this great V\=agbhava Mantra or anybody hears with devotion this history or makes others hear this, they all shall be crowned with success and enjoyment and Mukti be easy to them.

23-24. Specially they would get the power to remember their past lives, acquire eloquence in speaking, all round beauty, success in obtaining knowledge, success in their deeds and especially in the increase of their posterity and children. O Bhagavat\={\i}! This is what I want most.

Here ends the First Chapter of the Tenth Book on the story of Sv\=ayambhuva Manu in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



