\chapter{On the narrative of Ch\=ak\d{s}u\d{s}a Manu}

1-7. N\=ar\=aya\d{n}a said :-- O N\=arada! I will now narrate the supreme glories of the Dev\={\i} and the anecdote how Manu, the son of Anga, obtained excellent kingdom by worshipping the Dev\={\i} Bhagavat\={\i}. The son of the king Anga, named Ch\=ak\d{s}u\d{s}a became the Sixth Manu. One day he went to the Brahm\=ar\d{s}i Pulaha \d{R}i\d{s}i and taking his refuge said :-- O Brahm\=ar\d{s}i! Thou removest all the sorrows and afflictions of those that come under Thy refuge; I now take Thy refuge. Kindly advise Thy servant how he may become the Lord of an endless amount of wealth. O Muni! What can I do so that I may get the sole undisputed sway over the world? How my arms can wield the weapons and manipulate them so that they may not be baffled? How my race and line be constant and my youth remain ever the same, undecayed? And how can I, in the end, attain Mukti? O Muni! Kindly dost Thou give instructions to me on these points and oblige. Hearing thus, the Muni wanted him to worship the Dev\={\i} and said :-- ``O King! Listen attentively to what I say you today. Worship today the all auspicious \'Sakti; by Her grace, all your desires will be fulfilled.''

8. Ch\=ak\d{s}u\d{s}a said :-- ``O Muni! What is that very holy worship of \'Sr\={\i} Bhagavat\={\i}? How to do it? Kindly describe all these in detail.''

9-20. The Muni said :-- O King! I will now disclose all about the excellent P\=uja of the Dev\={\i} Bhagavat\={\i}. Hear. You recite (mentally) always the seed mantra of V\=akbhava (Speech) (The Deity being Mah\=a Sarasvat\={\i}). If any one makes japam (recites slowly) of the V\=akbhava V\={\i}ja thrice a day, one gets both the highest enjoyment here and, in the end, release (Mukti). O Son of a K\d{s}attriya! There is no other V\={\i}ja Mantra (word) better than this of V\=ak (the Word). Through the Japam of this V\={\i}ja Mantra comes the increase of strength and prowess and all successes. By the Japam of this, Brahm\=a is so powerful and has become the Creator; Vi\d{s}\d{n}u preserves the Universe and Mahe\'svara has become the Destroyer of the Universe. The other Dikp\=alas (the Regents of the quarters) and the other Siddhas have become very powerful by the power of this Mantra, and are capable of favouring or disfavouring others. So, O King! You, too, worship the Dev\={\i} of the Devas, the World Mother and ere long you will become the Lord of unbounded wealth. There is no doubt in

this. O N\=arada! Thus advised by Pulaha \d{R}i\d{s}i, the son of the King Anga went to the banks of the Viraj\=a river to practise austerities. There the king Ch\=ak\d{s}u\d{s}a remained absorbed in making Japam of the V\=akbhava V\={\i}ja Mantra and took for his food the leaves of the trees that dropped on the ground and thus practised severe austerities.

The first year he ate leaves; the second year he drank water and in the third year he sustained his life by breathing air simply and thus remained steady like a pillar. Thus he remained without food for twelve years. He went on making Japam of the V\=akbhava Mantra and his heart and mind became purified. While he was sitting alone, absorbed in the meditation of the Dev\={\i} Mantra, there appeared before him suddenly the Parame\'svar\={\i}, the World Mother, the Incarnate of Lak\d{s}m\={\i}. The Highest Deity, full of dauntless fire and the Embodiment of all the Devas, spoke graciously in sweet words to Ch\=ak\d{s}u\d{s}a, the son of Anga.

21-29. O Regent of the earth! I am pleased with your Tapasy\=a. Now ask any boon that you want. I will give that to you. Ch\=ak\d{s}u\d{s}a said :-- ``O Thou, worshipped by the Devas! O Sovereign of the Deva of the Devas! Thou art the Controller Inside; Thou art the Controller Outside. Thou knowest everything what I desire in my mind. Still, O Dev\={\i}! When I am so fortunate as to see Thee, I say Thou grantest me the kingdom for the Manvantara period.'' The Dev\={\i} said :-- ``O Best of the K\d{s}attriyas! I grant unto you the kingdom of the whole world for one manvantara. You will have many sons, very powerful, indeed, and well qualified. Your kingdom will be free from any danger till at last you will certainly get Mukti.'' Thus granting the excellent boon to Manu, She disappeared then and there, after being praised by Manu, with deep devotion. The Sixth Manu, then favoured by the Dev\={\i}, enjoyed the sovereignty of the earth and other pleasures and became the best of the Manus. His sons became the devotees of the Dev\={\i}, very powerful and expert and became respected by all and enjoyed the pleasures of the kingdom. Thus getting the supremacy by the worship of the Dev\={\i}, the Ch\=ak\d{s}u\d{s}a Manu became merged in the end in the Holy Feet of the Dev\={\i}.

Here ends the Ninth Chapter of the Tenth Book on the narrative of Ch\=ak\d{s}u\d{s}a Manu in the Mah\=apur\=a\d{n}am \'Sr\={\i}mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



