\chapter{On the Devas going to Vi\d{s}\d{n}u}

1-5. S\=uta said :-- Then the Devas, on arriving at Vaikuntha, saw the Lord of Lak\d{s}m\={\i}, the Deva of the Devas, the World-Teacher, with his eyes beautiful like Padm\=a Pal\=asa (lotus-leaves), shining with brilliance and began to praise Him in a voice choked with intense feelings of devotion, thus :-- ``Victory to Vi\d{s}\d{n}u! O Lord of Ram\=a! Thou art prior to the Vir\=at Puru\d{s}a. O Enemy of the Daityas! O Thou, the Generator of desires in all and the Bestower of the fruits of those desires to all! O Gobinda! Thou art the Great Boar and Thou art of the nature of Great Sacrifices! O Mah\=a Vi\d{s}\d{n}u! O Lord of Dharma! Thou art the Cause of the origin of this world! Thou didst support the earth in Thy Fish Incarnation for the deliverance of the Vedas! O Thou Satyavrata of the form of a Fish! We bow down to Thee. O Thou! The Enemy of the Daityas! The Ocean of mercy! Thou dost do the actions of the Devas out of mercy. O Thou! the Tortoise Incarnation! That grantest Mukti to others! Obeisance to Thee!

6-18. O Thou! That didst assume the form of a Boar for the destruction of the Daityas Jaya and others and for raising the earth from the waters! Obeisance to Thee! Thou didst assume that form - Half man and half Lion of the Nrisimha M\=urti and tore asunder

Hira\d{n}ya Ka\'sipu, proud of his boons, by Thy nails. We bow down to Thee! Obeisance to Thee! That in Thy Dwarf Incarnation, didst deceive Bali, whose head got crazed by the acquisition of the kingdom over the three Lokas. We bow down to Thee, that in Thy Para\'su R\=ama Incarnation, didst slay K\=arta V\={\i}ry\=aryuna, the thousand handed, and the other wicked K\d{s}attriyas! Obeisance to Thee! That wert born of the womb of Re\d{n}uk\=a as the son of Jamadagni. Obeisance to Thee, of great prowess and valour, that in Thy R\=ama Incarnation as the son of Da\'saratha, didst cut off the heads of the wicked R\=ak\d{s}asa, the son of Pulastya! We bow down again and again to Thee, the Great Lord, that in Thy Kri\d{s}\d{n}a incarnation, didst deliver this earth from the clutches of the wicked King Duryodhana, Kam\d{s}a and others and didst establish the religion by removing the then prevailing vicious ideas and doctrines. We bow down to Thy Buddha Incarnation, that Great Deva who didst come down here to put a stop to the slaughtering of the innocent animals and to the performance of the wicked sacrificial ceremonies! Obeisance to the Deva! When almost all the persons in this world will turn out in future as Mlechchas and when the wicked Kings will oppress them, right and left, Thou wilt then incarnate Thyself again as Kalki and redress all the grievances! We bow down to Thy Kalki Form! O Deva! These are Thy Ten Incarnations, for the preservation of Thy devotees, for the killing of the wicked Daityas. Therefore Thou art called as the Great Reliever of all our troubles. O Thou! Victory to Thee! The Deva Who assumest the forms of women and water for destroying the ailings of the devotees! Who else can be so kind! O Thou, the Ocean of mercy!'' O \d{R}i\d{s}is! Thus praising the yellow robed Vi\d{s}\d{n}u, the Lord of all the Devas, the whole host of the Devas bowed down to Him and made Sh\=ast\=amgas. Then Vi\d{s}\d{n}u Gad\=adhara, hearing their hymns, gladdened them and spoke :--

19-27. \'Sr\={\i} Bhagav\=ana said :-- ``O Devas! I am pleased with your stotra. You need not be sorrowful. I will remove all your troubles that have become unbearable to you. O Devas! I am very glad to hear the praises that you have offered on Me. Better ask boons from Me. I will grant them though very rare even and obtained with difficulty. Any person who rises early in the morning and recites with devotion this stotra sung by you, will never experience any sorrow. O Devas! No poverty, no bad symptoms, no Vet\=alas nor planets nor Brahm\=a R\=ak\d{s}asas nor any misfortunes will overtake him. No disease, due to V\=ata (windy temperament), Pitta (bile) and Kapha (phlegm) nor untimely death will visit him. His family will not

be extinct and happiness will always reign there. O Devas! This stotra can give everything. Both the enjoyment and freedom will come within anyone's easy access. There is no doubt in this. Nov what is your difficulty? Give out. I will remove it at once. There is not a bit of doubt in this.'' Hearing these words of \'Sr\={\i} Bhagav\=an, the Devas became glad and spoke to Vi\d{s}\d{n}u.

Here ends the Fifth Chapter of the Tenth Book on the Deva\'s going to Vi\d{s}\d{n}u, in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



