\chapter{On the checking of the rise of the Bindhya Range}

1-21. S\=uta said :-- Hearing the words of the Devas, Agastya, the Best of the Br\=ahmi\d{n}s promised that he would carry out their works. O \d{R}i\d{s}is! All the Devas then became very glad when the Muni, born of the water jar, promised thus. They then bade good-bye to him and went back gladly to their own abodes. The Muni then spoke to his wife thus :-- ``O daughter of the King! The Bindhya Mountain has baffled the progress of the Sun's course and has thus caused a great mischief. What the Munis, the Seers of truths said before referring to K\=a\'s\={\i}, all are now coming to my mind when I am thinking why this disturbance has overtaken me. They said that various hindrances would come to him at every step, who is a S\=adhu intending to settle at K\=a\'s\={\i}. Let him who wants Mukti, never quit K\=a\'s\={\i}, the Avimukta place in any case. But, O Dear! Today I have got one hindrance during my stay at K\=a\'s\={\i}.'' Thus talking with much regret on various subjects with his wife, the Muni bathed in the Ma\d{n}ikar\d{n}ik\=a gh\=at, saw the Lord Vi\'sve\'svara worshipped Dandap\=an\={\i} and went to the K\=ala Bhairava. He said in the following terms :-- ``O Mighty armed K\=alabhairava! Thou destroyest the fear of the Bhaktas; Thou art the God of this K\=a\'s\={\i} City. Then why art Thou driving me away from this K\=a\'s\={\i}dh\=am. O Lord! Thou removest all the obstacles of the devotees and Thou preservest them. Then why, O Destroyer of the sorrows of the Bhaktas! art Thou removing me from here? Never I blamed others; nor did I practise any hypocrisy with any person nor did I lie; then under what sin, Thou art driving me away from K\=a\'s\={\i}.'' O \d{R}i\d{s}is! Thus praying to K\=ala Bhairava, the Muni Agastya, born of water jar and the husband of Lop\=amudr\=a, went to S\=ak\d{s}i Ga\d{n}e\'sa, the Destroyer of all evils and seeing and worshipping Him, went out of K\=a\'s\={\i} and proceeded to the south. The Muni, the ocean of great fortune, left K\=a\'s\={\i}; but he became very much distressed to leave it and he remembered it always. He began to march on with his wife. As if riding on his car of asceticism he arrived at the Bindhya mountain in the twinkling of an eye and saw that the Mountain had risen very high and obstructed the passage of the Sun in the Heavens. The Bindhya Mountain, seeing the Muni Agastya in front, began to tremble and as if desirous to speak something to the earth in a whisper became low and dwarfish and bowed down to the Muni and fell down with devotion in s\=ast\=angas with devotion just like a stick dropped flat on the ground before the Muni.

Seeing the Bindhya thus low, the Muni Agastya became pleased and spoke with a gracious look :-- ``O Child! Better remain in this state until I come back. For, O Child! I am quite unable to ascend to your lofty heights.'' Thus saying, the Muni became eager to go to the south; and, crossing the peaks of the Bindhya, alighted gradually again to the plains. He went on further to the south and saw the \'Sr\={\i} \'Saila Mountain and at last went to the Malay\=achala and there, building his \=A\'srama (hermitage), settled himself. O Saunaka! The Dev\={\i} Bhagavat\={\i}, worshipped by the Muni went to the Bindhya Mountain and settled there and became known, in the three worlds, by the name of Bindhyav\=asin\={\i}.

22-26. S\=uta said :-- Anybody who hears this highly pure narrative of the Muni Agastya and Bindhya, becomes freed of all his sins. All his enemies are destroyed in no time. This hearing gives knowledge to the Br\=ahma\d{n}as, victory to the K\d{s}attriyas, wealth and corn to the Vai\'syas and happiness to the \'S\=udras.

If anybody once hears this narrative, he gets Dharma if he want Dharma, gets unbounded wealth if he wants wealth and gets all desires if he wants his desires fulfilled. In ancient times Sv\=ayambhuva Manu worshipped this Dev\={\i} with devotion and got his kingdom for his own Manvantara period. O Saunaka! Thus I have described to you the holy character of the Dev\={\i} in this Manvantara. What more shall I say? Mention please.

Here ends the Seventh Chapter of the Tenth Book on the checking of the rise of the Bindhya Range in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



