\chapter{On the Devas going to Mah\=a Deva}

1-2. S\=uta said :-- O \d{R}i\d{s}is! Then Indra and all the other Devas taking Brahm\=a along with them and placing Him at the front, went to Mah\=adeva and took His refuge. They bowed down to Him and chanted sweet and great hymns to Him, Who holds Moon on His forehead, Deva of the Devas, thus :--

3-5. O Thou, the Leader of the host of Gods! Victory to Thee! O Thou, Whose lotus feet are served by Um\=a, Victory to Thee! O Thou, the Giver of the eight Siddhis and Vibh\=utis (extraordinary powers) to Thy devotees, Victory to Thee! O Thou, the Background of this Great Theatrical Dance of this Insurmountable M\=ay\=a! Thou art the Supreme Spirit in Thy True Nature! Thou ridest on Thy vehicle, the Bull, and residest in K\=ail\=a\'sa; yet Thou art the Lord of all the Devas. O Thou, Whose ornament is snakes, Who art the Honoured and the Giver of honours to persons! O Thou! the Unborn, yet comprising all forms, O Thou \'Sambhu! That findest pleasure in this Thy Own Self! Victory toThee!

6-9. O Thou, the Lord of Thy attendants! O Thou, Giri\'sa! The Giver of the great powers, praised by Mah\=a Vi\d{s}\d{n}u! O Thou, That livest in the heart lotus of Vi\d{s}\d{n}u, and deeply absorbed in Mah\=a Yoga! Obeisance to Thee! O Thou that can'st be known through Yoga, and nothing but the Yoga itself; Thou, the Lord of the Yoga! We bow down to Thee. Thou awardest the fruits of yoga to the Yogins. O Thou, the Lord of the helpless! The Incarnate of the ocean of mercy! The Relief of the diseased and the most powerful! O Thou, whose forms are the three gu\d{n}as, Sattva, Rajo, Tamas! O Thou! Whose Emblem (carrier) is the Bull (Dharma); Thou art verily the Great K\=ala; yet Thou art the Lord of K\=ala! Obeisance to Thee! (The Bull represents the Dharma or Speech).

10. Thus praised by the Devas, who take the offerings in sacrifices the Lord of the Devas, whose emblem is Bull, smilingly told the Devas in a deep voice :--

11. O Thou, the excellent Devas! The residents of the Heavens! I am pleased with the praises that you have sung of Me. I will fulfil the desires of you, all the Devas.

12-15. The Devas said :-- O Lord of all the Devas! O Giri\'sa! Thou whose forehead is adorned with Moon! O Thou, the Doer of good to the distressed. O Thou, the Powerful! Dost Thou do good to us. O Thou, the Sinless One! The Bindhya Mountain has become jealous of the Sumeru Mountain, and has risen very high up in the Heavens and he has obstructed the Sun's course, thereby causing great troubles to all. O

Thou, the Doer of good to all! O \=I\'s\=ana! Dost Thou check the mountain's abnormal rise. How can we fix time if the Sun's course be obstructed! And when there is no knowledge, what is now the time, the sacrifices to the Devas and the offerings to the Pitris are now almost dead and gone. O Deva! Who will now protect us? We see Thee as the Destroyer of the fear of us and of those who are terrified. O Deva! O Lord of Giri\'s\=a! Be pleased with us.

16-18. \'Sr\={\i} Bhagav\=an said :-- O Devas! I have no power to curb the Bindhya Mountain. Let us go to the Lord of Ram\=a and pay our respects to Him. He is our Lord, fit to be worshipped. He is Gobinda, Bhagav\=an Vi\d{s}\d{n}u, the Cause of all causes. We will go to Him and tell Him all our sorrows. He will remove them.

19. Hearing thus the words of Gir\={\i}\'sa, Indra and the other Devas with Brahm\=a placed Mah\=adeva at their front and went to the region of Vaikuntha, trembling with fear.

Here ends the Fourth Chapter of the Tenth Book on the going of the Devas to Mah\=adeva in the Mah\=a Pur\=a\d{n}am \'Sr\={\i}mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



