\chapter{On the Devas praying to the Muni Agastya}

1-6. S\=uta said :-- O \d{R}i\d{s}is! Hearing the words of the Lord of Lak\d{s}m\={\i}, all the Devas became pleased and they spoke. The Devas said :-- O Deva of the Devas! O Mah\=a Vi\d{s}\d{n}u! O Thou, the Creator, Preserver and the Destroyer of the Universe! O Vi\d{s}\d{n}u! The Bindhya mountain has risen very high and it has stopped the Sun's course. Therefore all the works on earth are suspended. We are not receiving our share of Yaj\~nas. Now where we will go, what we will do, we do not know. \'Sr\={\i} Bhagav\=an said :-- ``O Devas! There is now in Benares the Muni Agastya of indomitable power, in devoted service of that Primordial \'Sakti Bhagavat\={\i}, the Creatrix of this Universe. This Muni alone can put a stop to this abnormal Bindhya Range. Therefore it behoves you all to go to that fiery Dvija Agastya at Benares where the people get Nirv\=a\d{n}a; the Highest Place and pray to him (to kindly fulfil your object).''

7-19. S\=uta said: -- O \d{R}i\d{s}is! Thus ordered by Vi\d{s}\d{n}u, the gods felt themselves comforted and, saluting Him, went to the city of Benares.

In a moment they went to the Holy City of Benares, and bathing there at the Ma\d{n}ikar\d{n}ik\=a gh\=at, worshipped the Devas with devotion and offered Tarpa\d{n}as to the Pitris and duly made their charities. Then they went to the excellent \=A\'srama of the Muni Agastya. The hermitage was full of quiet quadruped animals; adorned with various trees, peacocks, herons, geese and Chakrav\=akas and various other birds, tigers, wolves, deer, the wild boars, rhinoceros, young elephants, Ruru deer and others. Though there were the ferocious animals, yet the place was free from fear and it looked exceedingly beautiful. On arriving before the Muni, the gods fell prostrate before him and bowed down again and again to him. They then chanted hymns to him and said :-- O Lord of Dvijas! O Thou honoured and most worshipful! Victory to Thee. Thou art sprung from a water jar. Thou art the destroyer of

V\=at\=ap\={\i}, the Asura. Obeisance to Thee! O Thou, full of \'Sr\={\i}, the son of Mitr\=avaru\d{n}a! Thou art the husband of Lop\=amudr\=a. Thou art the store house of all knowledge. Thou art the source of all the \'S\=astras. Obeisance to Thee! At Thy rise, the waters of the ocean become bright and clear; so obeisance to Thee! At Thy rise (Canopus) the K\=a\'sa flower blossoms. Thou art adorned with clots of matted hair on Thy head and Thou always livest with Thy disciples. \'Sr\={\i} R\=ama Chandra is one of Thy chief disciples. O great Muni! Thou art entitled to praise from all the Devas! O Best! The Store-house of all qualities! O great Muni! We now bow down to Thee and Thy wife Lop\=amudr\=a! O Lord! O very Energetic! We all are very much tormented by an unbearable pain inflicted on us by the Bindhya Range and we therefore take refuge of Thee. Be gracious unto us. Thus praised by the gods, the highly religious Muni Agastya, the twice born, smiled and graciously said :--

20-27. O Devas! You are the lords of the three worlds, superior to all, highsouled, and the preserver of the Lokas. If you wish, you can favour, disfavour, do anything. Especially He who is the Lord of heavens, whose weapon is the thunderbolt, and the eight Siddhis are ever at his service is your Indra, the Lord of the Devas. What is there that he cannot do? Then there is Agni, Who burns everything and always carries oblations to the gods and the Pitris, Who is the mouth piece of the Devas. Is there anything impracticable with him! O Devas! Then again Yama is there amongst you, the Lord of the R\=ak\d{s}asas, the Witness of all actions, and always quick in giving punishment to the offenders, that terrible looking Yama R\=aja. What is there that he cannot accomplish?

Still, O Devas! if there be anything required by you that awaits my co-operation, give out at once and I will do it undoubtedly. Hearing these words of the Muni, the Devas became very glad and joyfully began to say what they wanted. O Mah\=ar\d{s}i! The Bindhya mountain has risen very high and thwarted the Sun's course in the Heavens. A cry of universal distress and consternation has arisen and the three worlds are now verging to the ruins. O Muni! Now what we want is this that Thou, by Thy power of Tapas, curbest the rise of this Bindhya Mountain. O Agastya! Certainly, by Thy fire and austerities, that mountain will be brought down and humiliated. This is what we want.

Here ends the Sixth Chapter of the Tenth Book on the Deva\'s praying to the Muni Agastya for checking the abnormal rise of the Bindhya Range in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



