\chapter{On the killing of Madhu Kaitabha}

1-2. The king Suratha spoke :-- ``O Best of the twice born! Who is that Dev\={\i} that you spoke just now? Why the Dev\={\i} deludes all these beings? What for does She do so? Whence is the Dev\={\i} born? What is Her Form? and what are Her qualities? O Br\=ahmi\d{n}! Kindly describe all these to me.''

3-9. The Muni spoke :-- O King! I will now describe the nature of the Dev\={\i} Bhagavat\={\i} and why does She take Her Form in due time. Listen. In ancient days, when Bhagav\=an N\=ar\=aya\d{n}a, the king of the Yogis, was lying in deep sleep on the ocean on the bed of Ananta, after He had destroyed the Universe, there came out of the wax of his ear the two D\=anavas, Madhu and Kaitabha, of monstrous appearances. They wanted to kill Brahm\=a, who was lying on the lotus coming out of the navel of Bhagav\=an. Seeing the two Daityas Madhu and Kaitabha and seeing also Hari asleep the Lotus-born Brahm\=a became very anxious and thought :-- Now Bhagav\=an is asleep; and these two indomitable Daityas are ready

to kill me. Now what am I to do? Where to go? How shall I get ease? O Child! Thus thinking, the high souled Lotus born suddenly came to a practical conclusion. He said :-- ``Let me now take refuge to the Goddess Sleep, Nidr\=a, the Mother of all and under Whose power Bhagav\=an Hari is now asleep.''

10-24. Brahm\=a then began to praise Her thus :-- ``O Dev\={\i} of the Devas! O Upholdress of the world! Thou grantest desires of Thy devotees. O Thou auspicious! Thou art Para Brahm\=a! By Thy Command all are doing respectively their works in their proper spheres! Thou art the Night of Destruction (K\=ala R\=atri); Thou art the Great Night (Mah\=a R\=atri). Thou art the greatly terrible Night of Delusion (Moha R\=atri); Thou art omnipresent; omniscient; of the nature of the Supreme Bliss. Thou art regarded as the Great. Thou art highly worshipped; Thou art alone in this world as highly intoxicated; Thou art submissive to Bhakti only; Thou art the Best of all the things; Thou art sung as the Highest; Thou art modesty; Thou art Pusti (nourishment); Thou art forgiveness (K\d{s}am\=a); Thou art Beauty (K\=anti); Thou art the embodiment of mercy; Thou art liked by all; Thou art adorned by the whole world; Thou art of the nature of wakefulness, dream and deep sleep; Thou art the Highest; Thou art alone Highest Deity; Thou art highly attached to the Supreme Bliss. There is no other thing than Thee. There is One only and that is Thee. Hence Thou art denominated as One; Thou becomest again the two by contact with Thy M\=ay\=a. Thou art the refuge of Dharma, Artha and K\=ama; hence Thou Thou art Three; Thou art the Tur\={\i}ya (the fourth state of consciousness) hence Thou art Four. Thou art the God of the five elements; hence Thou art Pancham\={\i} (five); Thou presidest over the six passions K\=ama, anger, etc.; hence Thou art Sasth\={\i}; Thou presidest over the seven days of the week and Thou grantest boons seven by seven; hence Thou art Seven. Thou art the God of the eight Vasus; hence Thou art Astam\={\i}; Thou art full of the nine R\=agas and nine parts and Thou art the Goddess of nine planets; hence Thou art Navam\={\i}. Thou pervadest the ten quarters and Thou art worshipped by the ten quarters; hence Thou art named Da\'sam\={\i} (the tenth day of the fortnight); Thou art served by the Eleven Rudras, the Goddesses of eleven Ga\d{n}as and Thou art fond of Ek\=ada\'s\={\i} Tithi; hence Thou art denominated Ek\=ada\'s\={\i}; Thou art twelve armed and the Mother of the twelve \=Adityas; hence Thou art Dv\=ada\'s\={\i}; Thou art dear to the thirteen Ga\d{n}as; Thou art the presiding Deity of Visve Devas and Thou art the thirteen months including the Malas M\=asa (dirty month), hence Thou art Trayoda\'s\={\i}. Thou didst

grant boons to the fourteen Indras and Thou gavest birth to the fourteen Manus; hence Thou art Chaturda\'s\={\i}. Thou art knowable by the Pa\~nchada\'s\={\i}. Thou art sixteen armed and on Thy forehead the sixteen digits of the Moon are always shining; Thou art the sixteenth digit (ray) of the Moon named Am\=a; hence Thou art Soda\'s\={\i}. O Deve\'s\={\i}! Thou, though attributeless and formless, appearest in these forms and attributes. Thou hast now enveloped in Moha and Darkness the Lord of Ram\=a, the Bhagav\=an, the Deva of the Devas. These Daityas, Madhu and Kaitabha are indomitable and very powerful. So to kill them, Thou better dost awake the Lord of the Devas.''

25-34. The Muni said :-- Thus praised by the Lotus-born, the T\=amasi Bhagavat\={\i} (the Goddess of sleep and ignorance), the Beloved of Bhagav\=an, left Vi\d{s}\d{n}u and enchanted the two Daityas.

On being awakened, the Supreme Spirit Vi\d{s}\d{n}u, the Lord of the world, the Bhagav\=an, the Deva of the Devas, saw the two Daityas. Those two monstrous D\=anavas, beholding Madhu S\=udana, came up before Him, ready to fight. The hand-to-hand fight lasted amongst them for five thousand years. Then the two D\=anavas, maddened by their great strength, were enchanted by the M\=ay\=a of Bhagavat\={\i} and told the Supreme Deity, ``Ask boon from us.'' Hearing this, the Bhagav\=an \=Adi Puru\d{s}a, (the Prime Man) asked the boon that both of them would be killed that day by Him. Those two very powerful D\=anavas spoke to Hari again, ``Very well. Kill us on that part of the earth which is not under water.'' O King! Bhagav\=an Vi\d{s}\d{n}u, the Holder of the conch and club, spoke :-- ``All right. Indeed! Let that be so.''

Saying this He placed their heads on His thigh and severed them with His disc (chakra). O King! Thus Mah\=a K\=al\={\i}, the Queen of all the Yogas arose on this occasion when the praise was offered Her by Brahm\=a. O King! Now I will describe another account how this Mah\=a Lak\d{s}m\={\i} appeared on another occasion. Listen.

Here ends the Eleventh Chapter of the Tenth Book on the killing of Madhu Kaitabha in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



