\chapter{On the anecdote of S\=avar\d{n}i Manu}

1-6. The Muni said :-- O King! The powerful Asura Mahi\d{s}a, born of a she-buffalo, defeated all the Devas and became the Lord of the whole universe. That indomitable D\=anava seized forcibly all

the rights of the Devas and began to enjoy the pleasures of the kingdom over the three worlds. The Devas, thus defeated, were expelled from their abodes in Heavens. They took Brahm\=a as their Leader and went to the excellent regions where Mah\=a Deva and Vi\d{s}\d{n}u resided and informed them of all that had been done by that vicious Asura Mahi\d{s}a. They said :-- ``O Deva of the Devas! The insolent Mahi\d{s}\=asura has become unbearable and he has taken possessions of the rights and properties of the whole host of the Devas and he is now enjoying them. Both of you are quite capable to destroy the Asura. So why do you not devise means to annihilate him in no time!''

7-10. Hearing these pitiful words of the Devas, Bhagav\=an Vi\d{s}\d{n}u became quite indignant. \'Sankara, Brahm\=a and the other Devas all were inflamed with anger. O King! From the face of the angry Hari, then emanated an Unusual Fire, brilliant like thousand Suns. Then by and by emanated fires also from the bodies of all the Devas who were filled then with joy. From the mass of fire thus emanated there came out a beautiful Female Figure. The face of this figure was formed out of the fire that emanated from the body of Mah\=a Deva. Her hairs were formed out of the fire of Yama and Her arms were formed out of the fire that emanated from Vi\d{s}\d{n}u.

11-21. O King! From the fire of the Moon came out two breasts; from the fire of Indra came out Her middle portion; from the fire of Varu\d{n}a, appeared Her loins and thighs; from the fire of Earth, Her hips were formed; from the fire of Brahm\=a, Her feet were formed; from the fire of the Sun, Her toes were formed; from the fire of the Vasus, Her fingers were formed; from Kuvera's fire, Her nose came out; from the excellent tejas of Praj\=apati, teeth; from the fire of Agni, Her three eyes; from the fire of the twilights, Her eye-brows and from the fire of V\=ayu, Her ears appeared.

O Lord of men! Thus Bhagavat\={\i} Mahi\d{s}amardin\={\i} was born of the Tejas (fiery substances) of the Devas. Next \'Siva gave Her the \'S\=ula (weapon spear); Vi\d{s}\d{n}u gave Sudar\'sana (Chakra); Varu\d{n}a gave the conchshell; Fire gave \'Sakti (weapon); V\=ayu gave Her bows and arrows; Indra gave Her thunder bolt and the bell of the elephant Air\=avata; Yama gave Her the Destruction Staff (K\=ala Danda); Br\=am\=a gave Her the Rudr\=ak\d{s}a, rosary and Kamandalu; the Sun gave Her, in every pore the wonderful rays; the Time (K\=ala) gave Her sharp axe and shield; the oceans gave Her the beautiful necklace and new clothes (two in number); Vi\'svakarm\=a gladly gave Her the crown, ear-rings,

kataka, Angada, Chandr\=ardha, tinklets; and the Him\=alay\=as gave Her the Lion as Her Vehicle and various gems and jewels.

22-30. Kuvera, the Lord of wealth gave Her the cup filled with the drink; Bhagav\=an Ananta Deva gave Her a necklace of snakes (N\=agh\=ara). Thus the World Mother, the Dev\={\i}, became honoured by all the Devas. The Devas, very much oppressed by Mahi\d{s}\=asura, then, chanted various hymns of praise to the World Mother M\=ahe\'svar\={\i} Mah\=a Dev\={\i}.

Hearing their Stotras, the Deve\'s\={\i}, worshipped by the Devas, shouted aloud the War-Cry. O King! Mahi\d{s}\=asura, startled at that War-Cry, came to Bhagavat\={\i} with all his army corps. Then that great Asura Mahi\d{s}a hurled various weapons in the air and overcast the sky with them and began to fight with great skill. The several generals Chik\d{s}ura, Durdhara, Durmukha, V\=a\d{s}kala, T\=amraka, Vid\=al\=ak\d{s}a and various other innumerable generals as if Death incarnate, accompanied Mahi\d{s}a, the chief D\=anava. A fierce fight then ensued. Then the Dev\={\i} Who enchants all the beings, became red-eyed with anger and began to kill the generals of the against party. When the generals were killed one by one Mahi\d{s}\=asura, skilled in the science of magic, came up quickly to the front of the Dev\={\i}.

31-40. The Lord of the D\=anavas, then, by his magic power, began to assume various forms. Bhagavat\={\i}, too, began to destroy his those forms. Then the Daitya, the crusher of the Devas, assumed the form of a buffalo and began to fight. The Dev\={\i} then fastened the animal, the Asura, the Death of the Devas, tightly and cut off his head by Her axe. The remainder of his forces, then, fled away in terror and disorder with a loud cry. The Devas became very glad and began to chant hymns to the Dev\={\i}. O King! Thus the Lak\d{s}m\={\i} Dev\={\i} appeared to kill Mahi\d{s}\=asura. Now I will describe how Sarasvat\={\i} appeared. Listen. Once on a time the two very powerful Daityas \'Sumbha and Ni\'sumbha were born. They attacked the Devas, oppressed them and seized their houses and rights. The Devas became dispossessed of their kingdoms and went to the Him\=alay\=as and offered stotras to the Dev\={\i} with the greatest devotion :-- ``O Deve\'s\={\i}! O Thou, skilled in removing the difficulties of the Bhaktas! Victory to Thee! O Thou, the Sinless One! Old age and death cannot touch Thee. O Thou! Death incarnate to the D\=anavas! O Deve\'s\={\i}! O Thou, of mighty valour and prowess! O Thou, the embodiment of Brahm\=a, Vi\d{s}\d{n}u and Mahe\'sa! Unbounded is Thy might; Thou canst be easily reached by the power of devo-

tion. O Thou, the Creator, Preserver and Destroyer! O M\=adhav\={\i}! O Thou, the Giver of Bliss! Thou dancest with great joy at the time of the dissolution of all the things (Pralaya).

41-50. O Thou, full of mercy! O Deva Deve\'s\={\i}! Be gracious unto us. O Thou, the Remover of the sufferings of the refugees! We now come unto Thy protection. The terror of \'Sumbha and Ni\'sumbha is like an endless ocean unto us. Save, save us from their fast clutches. O Dev\={\i}! save us O King! verily.'' When the Devas praised thus, the daughter of the Him\=alay\=as, Bhagavat\={\i} became pleased and asked ``What is the matter?'' In the meanwhile, there emitted from the physical sheath of the Dev\={\i} another Dev\={\i} Kau\'sik\={\i} who gladly spoke to the Devas :-- ``O Suras! I am pleased with Thy Stotra. Now ask the boon that you desire.'' The Devas then asked for the following boon :-- ``O Dev\={\i}! The two famous Daityas \'Sumbha and Ni\'sumbha have attacked forcibly the three worlds. The wicked Lord of the D\=anavas, \'Sumbha, has overcome us by the power of his arm and is now tormenting us without any break. Kindly devise some means to kill him.'' The Dev\={\i} said :-- ``O Devas! Be patient. I will kill these two Daityas, \'Sumbha and Ni\'sumbha and thus remove the thorn on your way. At an early date I will do good to you.'' Thus saying to Indra and the other Devas the merciful Dev\={\i} disappeared at once before their eyes. The Deva with their hearts delighted went to the beautiful Sumeru Mountain and dwelt there in the caves thereof.

Here the servants of \'Sumbha and Ni\'sumbha Chanda and Munda, while they were making their circuits, saw the exquisitely beautiful Dev\={\i}, the Enchantress of the world, and came back to \'Sumbha their King and said :--

51-60. ``O Destroyer of enemies! O Giver of honour! O Greal King! You are the Lord of all the Daityas and are fit to enjoy all the gems and jewels. To-day we have seen an extraordinary beautiful woman jewel. She is fit to be enjoyed by you. So now you would better bring that perfectly beautiful woman and enjoy. No such enchanting women can be seen amongst the Asura women N\=aga Kany\=as Gandharbha women, D\=anav\={\i}s or men.'' Hearing thus the words of the servant, \'Sumbha, the tormentor of the foes, sent a Daitya named Sugr\={\i}va as a messenger to Her. The messenger went to the Dev\={\i} as early as possible and spoke to Her all that \'Sumbha had told him. ``O Dev\={\i}! The Asura \'Sumbha is now the conqueror of the three worlds and respected by the Devas. O Dev\={\i}!

He is now enjoying all that is best, the gems and jewels; O Dev\={\i}! I am his messenger sent here to convey to you his message as follows :-- O Dev\={\i}! I am the sole enjoyer of all the jewels. O Beautiful-eyed! You are a gem; so you would worship me. O Fair One! All the gems and jewels that are in the Deva loka, in the Daitya loka, or amongst the regions of men, are under my control. So you would lovingly worship me.'' The Dev\={\i} said :-- ``O Messenger! True that you are speaking for your King; but I made a promise before, how can I act against it? O Messenger! Hear what I promised.

61-70. Whoever in the three worlds will conquer Me by sheer force and thus crush My vanity, whoever will be as strong as Myself, He can enjoy Me. So the King of Daityas can prove My promise true and by sheer force can marry Me. What is there with him that he cannot do? So, Messenger! Go back to your master and tell him all this so that the powerful \'Sumbha may fulfil My promise.'' Hearing thus the words of the Great Dev\={\i}, the messenger went back to \'Sumbba and informed him everything regarding the Dev\={\i}'s sayings. The very powerful Lord of the Daityas, \'Sumbha became very angry at the unpleasant words of the messenger and commanded the Daitya named Dhumr\=ak\d{s}a :-- ``O Dhumr\=ak\d{s}a! Listen to my words with great attention. Go and catch hold of that wicked woman by her hairs and bring her to me. Go quickly; do not delay.'' Thus commanded, the very powerful and the best of the Daityas, Dhumr\=ak\d{s}a, went at once to the Dev\={\i} with sixty thousand Daityas and cried aloud to Her :--

``O Auspicious One! You would better worship quickly our Lord \'Sumbha, who is very powerful and mighty; you will then acquire all sorts of pleasures; else I will hold you by your hairs and take you to the Lord of the Daityas.''

71-80. Thus addressd by Dhumr\=ak\d{s}a Daitya, the enemy of the Devas, the Dev\={\i} said :-- ``O Powerful One! O Daitya! What you have spoken is perfectly right, but tell me first what you or your king \'Sumbha can do to Me?'' When the Dev\={\i} said thus, the Daitya Dh\=umralochana rushed on Her at once with arms and weapons. With one loud noise, M\=ahe\'svar\={\i} burnt him immediately to ashes. O King! The other forces were partly crushed by the Lion, the vehicle of the Dev\={\i} and partly fled away in disorder to all the quarters; some became senseless out of fear. \'Sumbba, the Lord of the Daityas, became very angry to hear this. His face assumed a terrible form with eyebrows contracted. Then he became impatient with anger and sent in order Chanda, Munda and Raktab\={\i}ja.

The three powerful Daityas went to the battle and tried their might to capture the Dev\={\i}. The Dev\={\i} Jagaddh\=atr\={\i}, of violent prowess, seeing that these three Daityas were coming to Her, killed them by Her trident and laid them prostrate on the ground. Hearing their death with all their army, \'Sumbha and Ni\'sumbha came in their own persons arrogantly to the battlefield. \'Sumbha and Ni\'sumbha fought for a time with the Dev\={\i} a terrible fight and became tired, when the Dev\={\i} killed them outright. When the Bhagavat\={\i}, Who is all this world, killed \'Sumbba and Ni\'sumbha, the Devas began to praise Bhagavat\={\i} the Supreme Deity of V\=ak (Word) incarnate.

81-93. O King! Thus I have spoken to you in due order the manner in which the beautiful K\=al\={\i}, Mah\=a Lak\d{s}m\={\i} and Sarasvat\={\i} incarnated themselves on the earth. That Supreme Deity, the Dev\={\i} Parame\'svar\={\i} thus creates, preserves, and destroys the Universe. You better take refuge of that highly adored Dev\={\i}, that causes the distinction and the delusion of this Universe. Then only you will attain success. N\=ar\=aya\d{n}a said :-- The king Suratha, hearing these beautiful words of the Muni, took refuge of the Dev\={\i}, that yields all desired objects. He built an earthen image of the Dev\={\i} and, with concentrated attention, thought wholly of the Dev\={\i} and began to worship Her with devotion. When the worship was over, he offered sacrifices of the blood of his body to the Dev\={\i}. Then the World-Mother, the Deity of the Devas, became pleased and appeared before him and asked him :-- Accept the boon that you desire. When the Dev\={\i} said thus, the king asked from the Mahe\'svar\={\i} that excellent knowledge whereby the ignorance is destroyed and as well the kingdom free from any dangers or difficulties. The Dev\={\i} said :-- ``O King! By My boon, you will get your foeless kingdom in this very birth as well as the J\~n\=anam that removes ignorance. O King! I will tell you also what you will be in the next birth. Hear. In your next birth, you will be the son of the Sun and be famous as S\=avar\d{n}i Manu. By My boon you will be the Lord of the Manvantara, become very powerful and you will get good many sons.'' Thus granting him this boon, the Dev\={\i} disappeared. By the Grace of the Dev\={\i}, Suratha became the Lord of the Manvantara. O S\=adhu! Thus I have described to you the birth and deeds of S\=avar\d{n}i. He who hears or reads this anecdote with devotion, will be a favourite of the Dev\={\i}.

Here ends the Twelfth Chapter of the Tenth Book on the anecdote of S\=avar\d{n}i Manu in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



