\chapter{On the greatness of the Rudr\=ak\d{s}am}

1-11. N\=arada said :-- O Sinless one! The greatness of the Rudr\=ak\d{s}a seed that you have described is verily such. Now I ask why is this Rudr\=ak\d{s}am so much entitled to worship by the people. Please speak clearly on this point. N\=ar\=aya\d{n}a spoke :-- O Child! This is the very question that was asked once by K\=artika, the sixfaced One, to Bhagav\=an Rudra, dwelling in Kail\=a\'sa. What He replied, I say now. Listen. Rudra Deva spoke :-- O Child Sad\=anana. I will dwell briefly on the secret cause of the greatness of the Rudr\=ak\d{s}a seed. Hear. In days of yore, there was a Daitya called Tripur\=a who could not be conquered by any body. Brahm\=a, Vi\d{s}\d{n}u and the other Devas were defeated by him. They then came to Me and requested Me to kill the Asura. At their request, I called in my mind the Divine Great weapon, named Aghora, beautiful and terrible and containing the strength of all the Devas, to kill him. It was inconceivable and it was blazing with fire.

For full divine one thousand years I remained awake with eyelids wide open in thinking of the Aghora weapon, the destroyer of all obstacles, whereby the killing of Tripur\=asura might be effected and the troubles of the Devas be removed. Not for a moment my eyelids dropped. There by my eyes were affected and drops of water came out of my eyes. Note here. How enemies are to be killed. It requires great thought, great concentration, great yoga and great powers. O Mah\=asena! From those drops of water coming out of my eyes, the great tree of Rudr\=ak\d{s}am did spring for the welfare of all. This Rudr\=ak\d{s}a seed is of thirty-eight varieties. From My S\=urya Netra, i.e., My right eye, symbolizing the Sun, twelve yellow coloured (Pingala colour) varieties have come; and from my left eye representing the Moon, the Soma Netra, sixteen varieties of white colour and from my third eye on the top, representing Fire, i.e., the Agni Netra, ten varieties of black colour have come out. Of these the white Rudr\=ak\d{s}ams are Br\=ahmi\d{n}s and they are used by the Br\=ahma\d{n}as; the red coloured ones are the K\d{s}attriyas and should be used by the K\d{s}attriyas and the black ones are \'S\=udras and should be used by the Vai\d{s}yas and the \'S\=udras.

12-19. One faced Rudr\=ak\d{s}a seed is the \'Siva Himself, made manifest and rendered vivid; even the sin incurred in killing a Br\=ahma\d{n}a is destroyed thereby. Two faced or two headed Rudr\=ak\d{s}am is like the Deva and the Dev\={\i}. Two sorts of sins are destroyed thereby. The three faced Rudr\=ak\d{s}am is like fire; the sin incurred in killing a woman is destroyed in a moment. The four faced Rudr\=ak\d{s}a seed is like Brahm\=a and removes the sin of killing persons. The five faced Rudr\=ak\d{s}am is verily an image of Rudra; all sorts of sins, e.g., eating prohibited food, going to the ungoables, etc., are destroyed thereby. The six faced Rudr\=ak\d{s}am is K\=artikeya. It is to be worn on the right hand. One becomes freed of the Brahm\=ahaty\=a sin. There is no manner of doubt in this. The seven faced Rudr\=ak\d{s}am is named Ananga. Holding this frees one from the sin of stealing gold, etc., O Mah\=asena! The eight faced Rudr\=ak\d{s}a is Vin\=ayaka. Holding this frees one from the sin of holding an illicit contact with a woman of a bad family and with the wife of one's Guru, etc., and other sins as well. It enables one to acquire heaps of food, cotton, and gold; and in the end the Highest Place is attained.

20-35. The fruit of holding the eight faced Rudr\=ak\d{s}a seed has been said. Now I will talk of the nine-faced Rudr\=ak\d{s}am. It is verily the Bhairava made manifest. On the left hand it should be worn. By this, the people get both Bhoga (enjoyment) and Mok\d{s}a (liberation)

and they become powerful like Me and get themselves freed at once, without the least delay, of the sins incurred by committing thousands of abortions, hundreds of Brahm\=ahatty\=as (killing the Br\=ahma\d{n}as). Holding the ten-faced Rudr\=ak\d{s}a is verily wearing Jan\=ardana, the Deva of the Devas. The holding of which pacifies the evils caused by planets, Pi\'s\=achas, Vet\=alas, Brahm\=a R\=ak\d{s}asas, and Pannagas. The eleven-faced Rudr\=ak\d{s}am is like the Eleven Rudras. The fruits, the efficacy of which I now describe. Hear. The fruits obtained through the performance of one thousand horse sacrifices, one hundred V\=ajapeya sacrifices, and making gifts of one hundred thousand cows are obtained thereby.

If one wears the twelve-headed Rudr\=ak\d{s}am on one's ear, the \=Adityas get satisfied. The fruits of performing Gomedha and A\'svamedha sacrifices are obtained thereby. No fear comes from burned buffaloes, armed enemies and wolves and tigers and other murderous animals. Also the several diseases of the body never come to him. The holder of the twelve-faced Rudr\=ak\d{s}a seed feels always happy and he is the master of some kingdoms He becomes freed of the sins incurred in killing elephants, horses, deer, cats, snakes, mice, frogs, asses, foxes and various other animals.

O Child! The thirteen faced Rudr\=ak\d{s}am is very rare; if anybody gets it, he becomes like K\=arti Keya and gets all desires fulfilled; and the eight siddhis are under his grasp. He learns how to make gold, silver and other metals; he attains all sorts of enjoyments. There is no manner of doubt in this. O Sad\=anana! If anybody holds the thirteen faced Rudr\=ak\d{s}am, he becomes freed from the sins incurred in killing mother, father and brothers.

O Son! If one holds on one's head the fourteen faced Rudr\=ak\d{s}am always, one becomes like \'Siva. O Muni! What more shall I speak to you! The Devas pay their respects to one holding the fourteen faced Rudr\=ak\d{s}as and he in the end attains the Highest Goal, the state of \'Siva. His body becomes verily the body of \'Siva.

36-40. The Devas always worship the Rudr\=ak\d{s}a seed; the highest goal is attained by wearing the Rudr\=ak\d{s}am. The Br\=ahma\d{n}as should hold on their heads at least one Rudr\=ak\d{s}am with devotion. A rosary of twenty-six Rudr\=ak\d{s}ams is to be made and tied on the head. Similarly a rosary of fifty seeds is to be worn and suspended on the breast; sixteen each on each of the two arms; twenty-four Rudr\=ak\d{s}ams to be worn on the wrists, twelve on each. O Sad\=anana! If a rosary be made of one hundred and eight, fifty or twenty-seven Rudr\=ak\d{s}ams and if japam be done with that, immeasurable merits are obtained. If anybody wears a rosary of one hundred and eight seeds, he gets at every moment

the fruit of performing the A\'svamedha sacrifices and uplifts his twenty-one generations and finally he resides in the \'Siva Loka.

Here ends the Fourth Chapter of the Eleventh Book on the Greatness of the Rudr\=ak\d{s}am in the Mah\=apura\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



