\chapter{On the Rudr\=ak\d{s}am rosaries}

1-14. \=I\'svara said :-- O K\=artikeya! Now I will speak how to count the Japam (repetition of the mantra) with the rosary. Hear. The face of Rudr\=ak\d{s}am is Brahm\=a; the upper summit point is \'Siva and the tail end of Rudr\=ak\d{s}am is Vi\d{s}\d{n}u. The Rudr\=ak\d{s}am has two-fold powers :-- It can give Bhoga (Enjoyment) as well as Mok\d{s}a (Liberation). Then string or tie together, like a cow's tail, and like the snake's coiling a body, twenty-five five faced Rudr\=ak\d{s}a seeds, thorny and of red, white, mixed colours bored through and through. The rosary is to taper as a cow's tail tapers down. In stringing the beads into a rosary, it should be seen that the flat face of one Rudr\=ak\d{s}am is in front of the flat face of another Rudr\=ak\d{s}am; so the tail, the pointed end of one, must come in front of the tail or the narrower end of another. The Meru or the topmost bead of the string must have its face turned upwards and the knot should be given over that. The rosary, thus strung, yields success of the Mantra (mantra-siddhi). When the rosary is strung, it is to be bathed with clear and scented water and afterwards with the Pa\~nchagavya (cow-dung, cow urine, curd, milk, and ghee); then wash it with clear water and sanctify it with the condensed electrical charge of the Mantra. Then recite the Mantra of \'Siva (Six limbed), with ``H\=um'' added and collect the rosaries. Then repeat over them the Mantra ``Sadyoj\=ata, etc.'', and sprinkle water over it one hundred and eight times. Then utter the principal mantra and place them on a holy ground and perform Ny\=asa over it, i.e., think that the Great Cause \'Siva and the World-Mother Bhagavat\={\i} have come on them. Thus make the Samsk\=ara of the rosary (i.e., purify it) and you will find then that your desired end will be attained successfully. Worship the rosary with the Mantra of that Devata for which it is intended. One is to wear the Rudr\=ak\d{s}a rosary on one's head, neck or ear and controlling oneself, one should make japam with the rosary. On the neck, head, breast, or the ears or on the arms, the rosary should he held with the greatest devotion. What is the use in saying about it so often? It is highly meritorious and commendable

to hold always the Rudr\=ak\d{s}am. Especially on such occasions as taking baths, making gifts, making japams, performing the Homas,or sacrifices to Vi\d{s}ve Devas, in performing the Pooj\=as of the Devas,in making Pr\=ayaschittams (penances), in the time of \'Sr\=adh and in the time of initiation, it is highly necessary to hold Rudr\=ak\d{s}am. A Br\=ahmi\d{n} is sure to go to hell if he performs any Vaidik act without wearing any Rudr\=ak\d{s}am. Note :-- It would be offering an insult to \'Siva!

15-29. It is advisable to use the true Rudr\=ak\d{s}am with gold and jewel, on the head, neck or on one's hand. Never use the Rudr\=ak\d{s}am worn by another. Use Rudr\=ak\d{s}am always with devotion; never use it while you are impure. Even the grass that grows with the air in contact with the Rudr\=ak\d{s}a tree, goes verily to a holy region for ever. J\=ab\=ala Muni says in the \'Sruti :-- lf a man wearing Rudr\=ak\d{s}am commits a sin, he gets deliverance from that sin. Even if animals hold Rudr\=ak\d{s}am, they become \'Siva; what of men! The devotees of \'Sr\={\i} Rudra should always use at least one Rudr\=ak\d{s}a on the head. Those great devotees, who with Rudr\=ak\d{s}am on take the name of the Highest Self \'Sambhu, get themselves freed of all sorts of sins and pains. Those who are ornamented with Rudr\=ak\d{s}am are the best devotees. It is highly incumbent on those who want their welfare to wear Rudr\=ak\d{s}am. Those who hold Rudr\=ak\d{s}am on their ears, crown hair, neck, hands, and breast, get Brahm\=a, Vi\d{s}\d{n}u, and Mahe\'svara under them as their Vibh\=utis (manifestations, powers). The Devas and all those \d{R}i\d{s}is that started the Gotra, the \=Adipuru\d{s}as (the first chief men in several families), held with reverence the Rudr\=ak\d{s}ams. All the other Munis, that descended from their families, the ardent followers of \'Srauta Dharma, the pure souled, held the Rudr\=ak\d{s}ams. It may be, that many might not like at first to hold this Rudr\=ak\d{s}am, the visible giver of liberation and so well written in the Vedas; but after many births, out of the Grace of Mah\=adeva many become eager to take the Rudr\=ak\d{s}ams. The Munis that are the J\=ab\=ala S\=akh\={\i}s are famous in expounding the inestimable greatness of Rudr\=ak\d{s}ams.

The effect of holding Rudr\=ak\d{s}ams is well known in the three worlds. Pu\d{n}yam (great merit) arises from the mere sight of Rudr\=ak\d{s}ams; ten million times that merit arises by its touch; and by wearing it, one hundred Koti times the fruit arises and if one makes Japam every day, then one lakh koti times the pu\d{n}yam arises. There is no manner of questionings in this.

30-36. He who holds in his hand, breast, neck, ears, head, the Rudr\=ak\d{s}ams, becomes an image of Rudra. There is no manner of doubt in this. By holding Rudr\=ak\d{s}ams, men become invulnerable of all the beings, become respected, like Mah\=a Deva, by the Devas and Asuras and they roam on the earth like Rudra. Even if a man be addicted to evil deeds and commits all sorts of sins, he becomes respected by all, on holding Rudr\=ak\d{s}ams. By this men are freed of the sin of taking Uchhista and of all the other sins. Even if you suspend a Rudr\=ak\d{s}am rosary on the neck of a dog and if that dog dies in that state, he gets liberation! Then what to speak of others! By holding Rudr\=ak\d{s}ams, men even if they be devoid of Japam and Dhy\=anam, become freed of all sins and attain the highest state. Even if one holds merely one Rudr\=ak\d{s}a seed purified and surcharged with Mantra \'Sakti, he uplifts his twentyone generations, gets to Heaven and resides there with respect. I am speaking now further of the Greatness of Rudr\=ak\d{s}am.

Here ends the Fifth Chapter of the Eleventh Book on the Rudr\=ak\d{s}am rosaries in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



