\chapter{On the greatness of one faced, etc., Rudr\=ak\d{s}am}

1-4. \'Sr\={\i} N\=ar\=aya\d{n}a said :-- O N\=arada! When Gir\={\i}\'sa thus explained to K\=artikeya the greatness of Rudr\=ak\d{s}am, he became satisfied. Now I have spoken to you of the glories of the Rudr\=ak\d{s}ams as far as I know. Now, as to our subject of right way of acting, I will now speak on other things that ought to be known. Listen. The seeing of Rudr\=ak\d{s}am brings in a lakh times of Pu\d{n}yam and koti times the merit arises from touching that; holding it brings in koti times merit; again if one makes the japam of a Mantra with that Rudr\=ak\d{s}am, one obtains merit one hundred lakh koti times and one thousand lakh koti times the merit. The merit in holding the Rudr\=ak\d{s}am is far superior to that in holding Bhadr\=ak\d{s}am. The Rudr\=ak\d{s}am seed that is of the size of an \=Amalak\={\i} is the best; which is the of the size of a plum, is middling; and which is of the size of a gram is the worst.

this is my word and promise. The Rudr\=ak\d{s}am tree is of four kinds :-- Br\=ahma\d{n}a, K\d{s}attriya, Vai\'sya, and \'S\=udra. The white colour is Br\=ahma\d{n}a; the red colour is K\d{s}attriya; the yellow colour is Vai\'sya and the black coloured Rudr\=ak\d{s}am seed is \'S\=udra. The Br\=ahma\d{n}as are to use the white coloured Rudr\=ak\d{s}ams; the K\d{s}attriyas, the red coloured ones, the Vai\'syas, the yellow coloured ones; and the \'S\=udras, the black ones. Those Rudr\=ak\d{s}a seeds that are nicely circular, smooth, hard, and whose thorns or points are distinctly visible, are the best. Those that are pierced by insects, broken in parts, whose thorns are not clearly visible, with swells and holes and those that are coated over, these six varieties of Rudr\=ak\d{s}ams are faulty. Those Rudr\=ak\d{s}ams that have their holes by nature running through and through are best; and those that base their holes pierced by men are middling. The Rudr\=ak\d{s}a seeds that are all of uniform shape, bright, hard, and beautifully circular should be strung together by a silken thread. How to test the Rudr\=ak\d{s}a seed? As gold is tested by a touch stone; so the Rudr\=ak\d{s}am is tested by drawing lines on it; those on which the lines are most uniform, bright and beautiful are the best and they should be worn by the \'Saivas. One should hold one Rudr\=ak\d{s}am on the crown hair, thirty on the head, thirty six on the neck; sixteen on each arm, twelve on each wrist, fifty on the shoulders, one hundred and eight Rudr\=ak\d{s}ams in the place of the sacrificial thread; and the devotee should have two or three rounds on the neck. On the earrings, on the crown of the head, the head, on bracelets, on armlets, on necklace, on the ornament worn on the loins one should hold Rudr\=ak\d{s}am always, whether one sleeps or eats. Holding three hundred Rudr\=ak\d{s}ams is the lowest; holding five hundred is middling; holding one thousand Rudr\=ak\d{s}ams is the best; so one ought to wear one thousand Rudr\=ak\d{s}ams. At the time of taking Rudr\=ak\d{s}am, on one's head, one should utter the Mantra of \=Is\=ana; the mantra of Tat Puru\d{s}a while holding on one's ears; Aghora mantra on one's forehead and heart; and the v\={\i}ja of Aghora mantra, i.e., ``hasau'' while holding on one's hands. One should wear the rosary of fifty Rudr\=ak\d{s}a seeds, suspended up to the belly, uttering the V\=amadeva mantra, i.e., Sadyoj\=at\=adi, etc., the five Brahm\=a mantras, and the six-limbed \'Siva mantra. One is to string every Rudr\=ak\d{s}a seed, uttering the root mantra and then hold it. One-faced Rudr\=ak\d{s}a reveals Paratattva (the highest Tattva); when worn, the knowledge of the highest Tattva arises; the Brahm\=a is seen then. The two-faced Rudr\=ak\d{s}am is Ardhan\=ar\={\i}\'svara, the Lord of the other half which represents women (in the same person); if worn, Ardhan\=ar\={\i}svara \'Siva is always pleased with that man who holds it. The three-faced Rudr\=ak\d{s}am is Fire made manifest; it destroys in a moment the sin of killing a woman.

The three-faced Rudr\=ak\d{s}am is the three Agnis, Dak\d{s}i\d{n}\=agni, G\=arhapatya, and \=Ahavaniya; Bhagav\=an Agni is always pleased with that man who wears the three-faced Rudr\=ak\d{s}am. The four-faced Rudr\=ak\d{s}am is Brahm\=a Himself. The wearer gets his prosperity enhanced, his diseases destroyed, the divine knowledge springs in him and his heart is always pleased. The five-faced Rudr\=ak\d{s}am is the five faced \'Siva Himself; Mah\=adeva gets pleased with him who holds it. The Presiding Deity of the six faced Rudr\=ak\d{s}am is K\=artikeya. Some Pundits take Ga\d{n}apati to be the Presiding Deity of the six-faced Rudr\=ak\d{s}am. The presiding Deity of the seven-faced Rudr\=ak\d{s}am is the seven M\=atrik\=as, the Sun and the seven \d{R}i\d{s}is. By putting on this, the prosperity is increased, health and the pure knowledge are established. It should be put on when one becomes pure. The Presiding Deity of the eight-faced Rudr\=ak\d{s}am is Br\=ahm\={\i}, the eight M\=atrik\=as. By holding this, the eight Vasus are pleased and the river Ganges is also pleased. The putting on of this makes the J\={\i}vas truthful and pleasant-minded. The Devat\=a of the nine-faced Rudr\=ak\d{s}am is Yama; holding this puts off the fears of Death. The Devat\=a of the eleven-faced Rudr\=ak\d{s}am is ten quarters, the ten quarters are pleased with him who wears the ten-faced Rudr\=ak\d{s}am. The Devata of the eleven mouthed Rudr\=ak\d{s}am is the eleven Rudr\=as and Indra. Holding this enhances happiness. The twelve-faced Rudr\=ak\d{s}am is Vi\d{s}\d{n}u made manifest; its Devat\=as are the twelve \=Adityas; the devotees of \'Siva should hold this. The thirteen-faced Rudr\=ak\d{s}am, if worn, enables one to secure one's desires; he does nowhere experience failures. The K\=ama Deva becomes pleased with him who wears this. The fourteen-faced Rudr\=ak\d{s}am destroys all diseases and gives eternal health. While holding this, one ought not to take wine, flesh, onion, garlic, Saj\~n\=a fruit, Ch\=alt\=a fruit and the flesh of the boar which eats excrements, etc. During the Lunar and Solar eclipses, during the Uttar\=aya\d{n}a \'Samkr\=anti or the Dak\d{s}in\=ayana \'Samkr\=anti, during the full Moon or the New Moon day, if Rudr\=ak\d{s}am be worn, one becomes instantly freed of all one's sins.

Here ends the Seventh Chapter of the Eleventh Book on the greatness of one faced, etc., Rudr\=ak\d{s}am in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



