\chapter{On the greatness of Rudr\=ak\d{s}ams}

1-21. \=I\'svara said :-- O K\=artikeya! Ku\'sagranthi, J\={\i}vapattr\={\i} and other rosaries cannot compare to one-sixteenth part of the Rudr\=ak\d{s}a rosary. As Vi\d{s}\d{n}u is the best of all the Puru\d{s}as, the Gang\=a is the best of all the rivers, Ka\'syapa, amongst the Munis, Uchchaihsrav\=a amongst the horses, Mah\=a Deva amongst the Devas, Bhagavat\={\i} amongst the Dev\={\i}s, so the Rudr\=ak\d{s}am rosary is the best of all the rosaries. All the fruits that occur by reading the stotras and holding all the Vratas, are obtained by wearing the Rudr\=ak\d{s}am bead. At the time of making the Ak\d{s}aya gift, the Rudr\=ak\d{s}am bead is capable of giving high merits. The merit that accrues by giving Rudr\=ak\d{s}am to a peaceful devotee of \'Siva, cannot be expressed in words. If anybody gives food to a man holding the Rudr\=ak\d{s}am rosary, his twenty one generations are uplifted and he ultimately becomes able to live in the Rudr\=a Loka. He who does not apply ashes on his forehead and who does not hold Rudr\=ak\d{s}am and is averse to the worship of \'Siva is inferior to a ch\=and\=ala. If Rudr\=ak\d{s}am be placed on the head then the flesh-eaters, drunkards, and the associates with the vicious become freed of their sins. Whatever fruits are obtained by performing various sacrifices, asceticism and the study of the Vedas are easily attained by simply holding the Rudr\=ak\d{s}am rosary. Whatever merits are obtained by read-

ing the four Vedas and all the Pur\=a\d{n}as and bathing in all the T\={\i}rthas and the results that are obtained by immense practice in learning all are obtained by wearing Rudr\=ak\d{s}am. If at the time of death, one wears Rudr\=ak\d{s}am and dies, one attains Rudrahood. One has not to take again one's birth. If anybody dies by holding Rudr\=ak\d{s}am on his neck or on his two arms, he uplifts his twenty-one generations and lives in the Rudra Loka. Be he a Br\=ahma\d{n} or a Ch\=and\=ala, be he with qualities or without qualities, if he applies ashes to his body and holds Rudr\=ak\d{s}am he surely attains \'Sivahood. Be he pure or impure; whether he eats uneatables or be he a Mlechha or a Ch\=and\=ala or a Great Sinner, any body if he holds Rudr\=ak\d{s}am is surely equal to Rudra. There is no doubt in this.

If any body holds Rudr\=ak\d{s}am on his head he gets Koti times the fruit; on his ears, ten Koti times the fruit, on his neck, one hundred Koti times the fruit; on his holy thread, ayuta times the fruit; on his arm, one l\=akh Koti times the fruit and if one wears Rudr\=ak\d{s}am on one's wrist, one attains Mok\d{s}a. Whatever acts, mentioned in the Vedas be performed with Rudr\=ak\d{s}am on, the fruits obtained are unbounded. Even if a man be without any Bhakti and if he wears on his neck the Rudr\=ak\d{s}a rosary though he does always vicious acts, he becomes freed of the bondage of this world. Even if a man does not hold Rudr\=ak\d{s}a but if he be always full of devotion towards the Rudr\=ak\d{s}am, he attains the fruit that is got by wearing the Rudr\=ak\d{s}am and he attains the \'Siva Loka and is honoured like \'Siva. As in the country of K\={\i}kata, an ass which used to carry Rudr\=ak\d{s}am seed got \'Sivahood after his death, so any man, whether he be a J\~nani (wise) or Ajn\=ani (unwise), gets \'Sivahood if he holds Rudr\=ak\d{s}am. There is no doubt in this.

22-28. Skanda said :-- ``O God! How is it that in the country of K\={\i}kata (Bihar), an ass had to carry Rudr\=ak\d{s}a; who gave him the Rudr\=ak\d{s}ams! And what for did he hold that?''

Bhagav\=an \=I\'svara said :-- O Son! Now hear the history of the case. In the Bindhya mountain one ass used to carry the load of Rudr\=ak\d{s}am of a traveller. Once the ass felt tired and became unable to carry the load and fell down on the road and died. After his death the ass came to Me by My Grace, becoming Mahe\'svara with trident in his hand and with three eyes. O K\=artikeya! As many faces as there are in the Rudr\=ak\d{s}am, for so many thousand Yugas the holder resides with honour in the \'Siva Loka. One should declare the greatness of Rudr\=ak\d{s}am to one's own disciple; never to disclose its glories to one who is not a disciple nor a devotee of Rudr\=ak\d{s}am nor to him who is an illiterate brute. Be he a Bhakta or not a

Bhakta, be be low or very low, if he holds Rudr\=ak\d{s}am then he is freed from all sins. No equal can be to the merit of him who holds the Rudr\=ak\d{s}ams.

29-39. The Munis, the Seers of truth, describe this holding on of Rudr\=ak\d{s}am as a very great vow. He who makes a vow to hold one thousand Rudr\=ak\d{s}ams, becomes like Rudra; the Devas bow down before him. If thousand Rudr\=ak\d{s}ams be not obtained, one should hold at least sixteen Rudr\=ak\d{s}ams on each arm, one Rudr\=ak\d{s}am on the crown hair; on the two hands, twelve on each; thirty-two on the neck; forty on the head; six on each ear and one hundred and eight Rudr\=ak\d{s}ams on the breast; and then he becomes entitled to worship like Rudra. If any body holds Rudr\=ak\d{s}am together with pearls, Prab\=ala, crystal, silver, gold and gem (lapis lazuli), he becomes a manifestation of \'Siva. If a body, through laziness even, holds Rudr\=ak\d{s}am, the sin cannot touch him as darkness cannot come near light. If any body makes japam of a mantram with a Rudr\=ak\d{s}a rosary, he gets unbounded results. Such a merit giving Rudr\=ak\d{s}am, if one such Rudr\=ak\d{s}am be not found in any one's body, his life becomes useless, like a man who is void of Tripundrak (three curved horizontal marks made on the forehead by the worshippers of \'Siva). If any body simply washes his head all over with Rudr\=ak\d{s}am on, he gets the fruit of bathing in the Ganges. There is no doubt in this. One faced Rudr\=ak\d{s}am, the five faced, eleven faced and fourteen faced Rudr\=ak\d{s}ams are highly meritorious and entitled to worship by all. The Rudr\=ak\d{s}am is \'Sankara made manifest; so it is always worshipped with devotion. The greatness of Rudr\=ak\d{s}am is such as it can make a king out of a poor man. On this point, I will tell you an excellent Pur\=anic anecdote.

40-49. There was a Br\=ahmi\d{n}, named Girin\=atha in the country of Kosala. He was proficient in the Vedas and Ved\=amgas, religious and very rich. He used to perform sacrifices. He had a beautiful son named Gu\d{n}anidhi. The son gradually entered into his youth and looked beautiful like Kandarpa, the God of Love. While he was studying at his Guru Sudhi\d{s}a\d{n}a's house, he, by his beauty and youth captivated the mind of his Guru's wife named Mukt\=aval\={\i}. The Guru's wife became so much enchanted by his extraordinary beauty that she, being unable to control herself, mixed with him and for some time remained with him in secret enjoyment. Then feeling inconveniences, due to the fear of. his Guru, to enjoy her freely, used poison to the Guru, killed him and then he began to live freely with her. Next when his father, mother came to know about this, he put to death instantly his father and mother, administering poison to them. He became

addicted to various pleasures and his wealth was exhausted gradually. He began to steal in Br\=ahma\d{n}\'s houses and became addicted very much to drinking. His relatives outcasted him from the society for his bad behaviour and banished him outside the town. He then went into a dense forest with Mukt\=aval\={\i}; and he began to kill the Br\=ahmi\d{n}s for their wealth. Thus a long time passed away; when at last he fell into the jaws of death.

50-54. Then to take him to the region of Death, thousands of the Yama's messengers came; at the same time the \'Siva's messengers came from \'Siva Loka. O K\=artikeya! A quarrel then ensued between both the parties of Yama and \'Siva. The Yama's messengers then said :-- ``O Servers of \'Sambhu! What are the merits of this man that you have come to take him? First speak to us of his merits.'' \'Siva's messengers spoke :-- ``Fifteen feet below the ground where this man died, there exists the Rudr\=ak\d{s}am. O Yama's messengers! By the influence of that Rudr\=ak\d{s}am, all his sins are destroyed; and we have come to take him to \'Siva.'' Then the Br\=ahmi\d{n} Gu\d{n}anidhi assumed a divine form and, getting on an aerial car went with \'Siva's messengers before \'Siva. O One of good vows! Thus I have described briefly to you the greatness of Rudr\=ak\d{s}am. This is capable to remove all sorts of sins and yield great merits.

Here ends the Sixth Chapter of the Eleventh Book on the Greatness of Rudr\=ak\d{s}ams in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



