\chapter{On the Greatness of the Dev\={\i} P\=uj\=a}

1. N\=arada spoke :-- ``O Bhagav\=an! I am now very eager to hear the special P\=uj\=a of \'Sr\={\i} Dev\={\i}. The people get their desires fulfilled if they worship Her.''

2-23. N\=ar\=aya\d{n}a said :-- O Devar\d{s}i! I shall now specially speak to you how the World Mother Bhagavat\={\i} is worshipped; by worshipping Whom one easily gets objects of enjoyments, liberation and the destruction of all evils. Controlling one's speech and making \=Achaman, one must make one's sankalpa and perform Bh\=uta\'suddhi, M\=atrik\=any\=asa, \d{s}adangany\=asa, placing conchshell and doing other necessary acts. Offering the ordinary Arghya, one should give special Arghya and with the mantra ``Astr\=aya Phat'' sprinkle over all the articles brought for worship. Taking the Guru's permission, he is to go on with his P\=uj\=a. First worship the p\={\i}tha or seat whereon the Dev\={\i} would be placed; then perform dhy\=an (meditation of the Dev\={\i}). Then with great devotion, offer to the Deva, the seats (\=Asana) and other articles of worship; then perform the bath of the Dev\={\i} by the water of the Panch\=amrita (the five nectars). If anybody performs the bath ceremony of \'Sr\={\i} Dev\={\i} with one hundred jars of sugarcane juice, he will not have to incur any future birth.

He who performs this bath, and recites the Veda Mantras, with mango juice or sugarcane juice gets for ever Lak\d{s}m\={\i} ever and ever and Sarasvat\={\i} bound at his doors. He who gets this sacred bath of the Dev\={\i} with grape juice, along with his relatives and acquaintences dwells in the Dev\={\i}-loka for as many years as there are atoms in the juice. He who bathes the Dev\={\i} with the Vedic mantras, and with water scented with camphor, the fragrant aloe wood (aguru), saffron, and musk, becomes freed at once of the sins acquired in his hundred births. He who bathes the Dev\={\i} with jars of milk, lives in the ocean of milk (k\d{s}\={\i}ra samudra) for one Kalpa. So he who does this bathing ceremony with jars of curd, becomes the lord of Dadhikunda (the reservoir of curd). He who performs the Sn\=anams of the Dev\={\i} with honey, ghee and sugar becomes the lord of these things. He who bathes the Dev\={\i} with one thousand jars, becomes happy in this world as in the next. Note :-- Make the liquid current flow pure in your body is the esoteric meaning of the bath. If you give Her a pair of silken clothes, you will go to the V\=ayu-Loka. If you give Her

the jewel ornaments, you will become the Lord of gems and jewels. (Make your mind like the gem.) If anybody gives saffron, sandalpaste, musk, Sind\=ura and \=Alaktak (red things), he will go to the Heavens and become there the Indra, the Lord of the Devas, in the next birth. Various flowers ought to be offered in \'Sr\={\i} Bhagavat\={\i}'s worship; or the flowers of the season offered to the Dev\={\i} will lead the devotee to Kail\=a\'sa. The devotee that offers the beautiful Bel leaves to the Dev\={\i} never experiences anywhere pains and difficulties. The devotee who writes the V\={\i}ja mantra of M\=ay\=a ``Hr\={\i}m Bhuvane\'svaryai Namah'' with red sandalpaste thrice on the tri-leaves of the Vilva tree leaf and offers this to the lotus feet of the Dev\={\i}, becomes Manu by the merit of this virtuous act! The devotee becomes the Lord of the whole universe who worships the Dev\={\i} Bhagavat\={\i} with ten millions of entire Vilva leaves, fresh, green and spotless.

24-40. If any devotee worships with ten millions of entire fresh green Kunda flowers, with eight scents, he gets surely the Praj\=apatihood. The worship of the Dev\={\i} with ten millions of Mallik\=a and M\=alat\={\i} flowers besmeared with eight scents makes a man the four faced (Brahm\=a); and one hundred millions of such flowers will make the devotee a Vi\d{s}\d{n}u. In days of yore, Vi\d{s}\d{n}u worshipped the Dev\={\i} in the aforesaid way and so got His Vi\d{s}\d{n}uhood. If any devotee worships the Dev\={\i} with one hundred Kotis of Mallik\=a or M\=alat\={\i} flowers, the man becomes certainly Sutr\=atm\=a Hira\d{n}yagarbha. In ancient days Hira\d{n}yagarbha worshipped thus the Dev\={\i} with great devotion and so he became Hira\d{n}yagarbha! (These Hira\d{n}yagarbha, Brahm\=a, Vi\d{s}\d{n}u and Mahe\'sa were mere ordinary men before. See the Brihad\=ara\d{n}yaka Upani\d{s}ada). Note :-- The eight scents refer to Jat\=a mams\={\i} Kapiyut\=a \'Saktergandh\=as takam! So will be the results if Jav\=a, Vandh\=uka and D\=adim\={\i} flowers be offered in the worship. Various other beautiful flowers can be offered duly to the Dev\={\i} by the devotee. The merits accruing from such offers are not known even to the God \=I\'svara. The flowers that spring in their proper seasons are to be offered every year to the Dev\={\i}, repeating Her thousand names enumerated in the Twelfth Book or in the K\=urma Pur\=a\d{n}a. If the above worship be offered to the Dev\={\i}, then that man, whether he be a sinner or a great sinner, will be freed from all the sins and after leaving his mortal coil, he will get, no doubt, the lotus feet of the \'Sr\={\i} Dev\={\i} Bhagavat\={\i}. Offer Dh\=upa made of black Aguru, camphor, sandalpaste, red sandalpaste, Sihlaka and Guggula, saturated with ghee in such a way as the whole room of \'Sr\={\i} Bhagavat\={\i} scents with pure

fragrant smell. The Dev\={\i} Bhagavat\={\i} becomes pleased with this and offers the lordship of the three Lokas to the devotee. The devotee, who offers daily the light of camphor to the Dev\={\i}, goes to the S\=urya Loka. There is no doubt in this. With one's whole heart, one should give one hundred or one thousand lights to the Dev\={\i}. The devotee should offer heaps of food consisting of six R\=asas, the plates and dishes for chewing, sucking, licking and drinking, that is, all kinds of food solid, and liquid, mountain-like high. Always give food on golden flat plates and cups and various delicious sweet juicy nice heavenly fruits, nicely arranged on trays, cups and saucers. When \'Sr\={\i} Mah\=adev\={\i} Bhuvane'svar\={\i} gets pleased, the whole universe gets pleased. For the whole Universe is all Dev\={\i}; as a rope is mistaken for a snake, so this Mah\=adev\={\i} is mistaken for the universe.

41-59. Offer a jar of drinking Ganges water, cool and nice, scented with camphor to the Dev\={\i}; then offer betels with camphor, cardamum, cloves, and various delicious scents. These all are to be offered with great devotion so that the Dev\={\i} may be pleased. Next have music with lovely mrida\d{n}gas, flutes, murajas, Dhahk\=as and dundubhis and so please Her. The Veda mantras are to be recited, the Pur\=a\d{n}as are to be read and the hymns to be chanted. With whole head and heart offer to the Dev\={\i} the umbrella and ch\=amara, the two kingly offerings. Then circumambulate round Her and prostrate before Her and ask Her kindness and pray to Her to forgive all faults and shortcomings. The Dev\={\i} is pleased with anybody who remembers Her even once! What wonder then that She will be pleased with all these offerings! The Mother is naturally merciful to her child. When She is loved with devotion, then She becomes very merciful. There is nothing strange here! On this point I will recite to you the history of Vrihadratha R\=ajar\d{s}i. Hearing which gives rise to Bhakti and Love.

Once in a certain region in the Him\=alay\=as there lived a bird called Chakrav\=ak. It flew over many countries and went once to K\=a\'s\={\i}dh\=am. As a fruit of his Pr\=arabdha Karma, that bird, desirous to find some rice beans, voluntarily went like an orphan round about the temple of \'Sr\={\i} Annap\=ur\d{n}\=a Dev\={\i}. There circumambulating round the Dev\={\i} Bhagavat\={\i} the bird left the city K\=a\'s\={\i}, that grants liberation and flew away to another country. In time the bird left his body and went to Heavens. There he assumed a heavenly form of a youth and began to enjoy various pleasures. Thus he enjoyed for two Kalpas. Then he got back to the earth and took his birth as the best in the K\d{s}attriya family. He became celebrated as the king Vrihadratha in

this world. That King was truthful, controlled his senses, and practised \'Samyama and deep concentration and knew everything of the past, the present and the future. He conquered all the enemies and performed various sacrifices and became the Emperor of the sea-girt earth and acquired the very rare faculty in the knowledge of everything of his previous births. The Munis came to know of this from various rumours and came to the King. The King Vrihadratha duly entertained those guests. The Munis took their seats and asked :-- ``O King! We hear that all the events of previous births are vividly reflected in your memory. On this point great doubts have come upon us. Kindly describe in detail. By what Pu\d{n}yam (merits) you have come to know all about previous births and the knowledge of the past, the present and the future. We have come to you to know how you got this wonderful supersensual knowledge. Kindly say to us sincerely everything about this and oblige.''

60-71. N\=ar\=aya\d{n}a said :-- O Br\=ahma\d{n}! The very religious King Vrihadratha heard them and began to speak out all the secret causes for his knowledge of the past, the present and the future, thus :-- O Munis! Hear how I acquired this knowledge. In my previous birth I was a very low bird chakrav\=ak. Once, out of my ignorance, I circumambulated round the temple of the Dev\={\i} Bhagavat\={\i} Annapur\d{n}\=a at K\=a\'s\={\i}. And, as the result of that, I lived in the Heavens for a period of two Kalpas and I have got this birth and I have got the knowledge of the past, the present and the future. O You of good vows! Who can ascertain what amount of merits accrues from remembering the Feet of the World-Mother. Remembering Her glories, I always shed tears of joy. Those who do not worship the adorable Deity Jagadamb\=a are the Great Sinners and they are treacherous. Fie on their births! The worship of \'Siva or Vi\d{s}\d{n}u is not eternal. Only the Jagadamb\=a's worship is eternal. Thus it is stated in the \'Srutis. What more shall I speak on this worship of the World-Mother, which is void of the best trace of any doubt. Everyone ought to serve devotedly the lotus feet of the Dev\={\i} Bhagavat\={\i}. There is no other act more glorious in this world than serving the feet of Jagadamb\=a. It is highly necessary to serve the Highest Deity, whether in Her Sagu\d{n}a or in Her Nirgu\d{n}a aspect. (Eat the sugarcandy, holding it in any way. It makes no difference.) N\=ar\=aya\d{n}a said :-- Hearing the aforesaid words of the virtuous R\=ajar\d{s}i Vrihadratha, the Munis went back to their respective abodes. Such is the power of the Dev\={\i} Jagadambik\=a! So who can question about the certainty of the high merits arising from the Jagadambik\=a's worship and who will not reply, when so questioned? Their births are really fruitful who possess faith in the Dev\={\i} worship;

but of those who have no such faith, there is some wrong mixture, no doubt; in their births.

Here ends the Eighteenth Chapter of the Eleventh Book on the Greatness of the Dev\={\i} P\=uj\=a, in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



