\chapter{On the description of the greatness of the three kinds of Bha\d{s}mas}

N\=arada asked :-- O Bhagav\=an! How is the above Bha\d{s}ma of three kinds? I am eager to hear this. Kindly describe this to me. N\=ar\=aya\d{n}a said :-- O N\=arada! I am now talking of the three kinds of ashes; hearing this even destroys one's sins and brings in good fame. When a cow evacuates her dung, just as the cow dung leaves her and is far from reaching the ground, one should catch it with one's hand and this cow-dung burnt with ``Sadyoj\=at\=adi,'' i.e., Brahm\=a Mantra, becomes ashes which are called ``\'S\=antika Bha\d{s}ma,'' i.e., ashes producing peace. Before the cowdung is about to reach the ground, the devotee should take it with his hand and uttering the six lettered Mantra, he is to burn the cowdung. The ashes from this are called Paustik Bha\d{s}ma, i.e., ashes leading to nourishment.

If the cowdung be burned with the Mantra ``Haum,'' the ashes of this are called ``K\=amada Bha\d{s}ma,'' i.e., ashes leading to the granting of desires. O N\=arada! On the full moon day, new moon day or on the eighth lunar day, a man is to get up from his bed early in the morning and be pure and go the cow enclosure. He is to salute the cows and take the cowdung, uttering the Mantra Haum. If he be a Br\=ahmi\d{n}, he is to collect the white cow dung; if he be a K\d{s}attriya, he would take the red cow

dung; a Vai\'sya, yellow cow dung and if he be a \'S\=udra, he would take the black cow dung. Then by the mantra ``Namah'' he is to form that into a ball and cover it with the husk of rice or some other grain and dry it in a sacred place, repeating the mantra ``Haum.'' Bring fire from a forest or from the house of a Veda-knowing Br\=ahmi\d{n} and reduce the cowdung to ashes by this mantra, uttering the mantra Haum. Next take out the ashes carefully from the fire place (Agni Kunda) and place it in a new jar or pot, again remembering the mantra ``Haum''. Mix with the ashes the Ketak\={\i} dust, the P\=atala flower dust, the root of the fragrant grass called khas khas, saffron and other sweet scented things with the mantra ``Satyoj\=atam prapady\=ami,'' etc. First perform the water bath, then the bath of the ashes. In case one cannot have the water bath, one is to have the ash bath. Washing the hands, feet and head with the mantra ``I\'s\=anah Sarvavidy\=an\=am,'' etc., and uttering ``Tatpuru\d{s}a'' one is to besmear one's face with ashes and by the mantra ``Aghora'' apply ashes on one's chest; with the mantra V\=amadeva. He is to use ashes on his navel; and with the mantra ``Sadyo J\=ata,'' etc., all over his body; he is to quit his former cloth and put on another fresh cloth. Wash your hands and feet and sip (do \=Achaman). It will serve the purpose if one simply uses Tripundrak and if one does not besmear the whole body with ashes. Before the midday one is to use Bha\d{s}ma with water; but after the midday with dry ashes one is to draw the Tripundra lines of ashes with the forefinger, middlefinger and ringfinger. The head, forehead, ears, neck, heart, and the arms are the places whereon the Tripundras are used. On the head the ashes are applied with five fingers and with the mantra ``Haum''; on the forehead, the Tripundra is applied with mantra Sv\=ah\=a by the forefinger, mindlefinger and ringfinger; on the right ear, it is applied with ``Sadyoj\=ata'' mantra; on the left ear, with ``Vamadeva'' Mantra; on the neck with Aghora mantra by the middlefinger; on the chest with ``Namah'' mantra by the forefing middlefinger and ringfinger; on the right arm with va\d{s}at mantra by the three fingers; on the left arm with ``Ham'' mantra by the three fingers; and on the navel, the ashes are to be applied with the mantra \=I\'s\=anah sarva dev\=an\=am by the middlefinger. The first line in every Tripundra is Brahm\=a; the second line is Vi\d{s}\d{n}u; and the third, the topmost line is Mah\=a Deva. The line of ashes that is marked by one finger is \=I\'svara. The head is the place of Brahm\=a; the forehead is the site of \=I\'svara; the two ears are the seats of the two A\'svins and the neck is where Ga\d{n}e\'sa resides. The K\d{s}attriyas, Vai\'syas, and \'S\=udras are to use Tripundras without any mantra; they are also not to use the ashes on

the whole of the body. The lowest classes (e.g. the ch\=and\=alas, etc.,) and the uninitiated persons are to use the Tripundraks without any mantra.

Here ends the Eleventh Chapter of the Eleventh Book on the description of the greatness of the three kinds of Bha\d{s}mas in the Mah\=apur\=anam \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mah\=ar\d{s}i Veda Vy\=asa.



