\chapter{On the greatness in holding the Tripundra and Bha\d{s}ma}

1-20. \'Sr\={\i} N\=ar\=aya\d{n}a said :-- O Devar\d{s}i N\=arada! Hear now the great secret and the fruits of besmearing one's body with ashes, yieding all desires. The pure cowdung of the Kapila (brown) cow is to be taken up by the hand before it reaches the ground. It should not be like mire, i.e., not like a liquid; it should not be also very hard nor should it emit a bad stench. And in case if the cowdung that has already fallen on the earth, has to be taken, it should be scraped off from the top and bottom; make it into a ball and then burn it in a pure fire, repeating the principal mantra. Take the ash and tie it in a piece of cloth and keep it in a pot. The pot in which the ashes are to be kept should be nice and good, hard, clean and sprinkled over for purification. Uttering the principal mantra, one is to keep the ashes in the pot. The pot may be of metal, wood, earth, or cloth; or it can be kept in any other nice pot. The ashes can be kept in a silken bag where the mohurs are kept. In going to a distant land, the devotee can take the ashes himself or kept with his accompanying servant. When it is to be given to somebody, it is to be given with both the hands; never with one hand. Never keep it in an unholy place. Never apply feet to the ashes, nor throw it in an ordinary place nor ever cross it by your legs. Use always the ashes after purifying it with mantra. These rules of holding the Bha\d{s}ma are according to the Smritis. By holding Bha\d{s}ma in this way, the devotee becomes, no doubt, like \'Siva. The ashes, that the Vaidik devotees of \'Siva prepare are to be taken with devotion. All can ask for that. But the ashes that the followers of the Tantra cult prepare, are taken by the T\=antriks only; it is prohibited to the Vaidiks. The \'S\=udras, K\=ap\=alikas, and other heretics (e.g., Jains, Buddhists) can use the Tripundras. Never do they conceive in their minds that they would not take the Tripundra. The holding on of Bha\d{s}ma (ashes) is

according to the Vedas. Therefore one who does not apply it falls down. The Br\=ahma\d{n}as must use the Tripundras, repeating the mantra; and they are to besmear their whole body with ashes; if they don't do so, they are surely fallen. He can never expect to get liberation even after koti births who does not besmear his body with ashes devotedly and who does not hold the Tripundras. O N\=arada! The vile man who does not hold Bha\d{s}ma duly, know the birth of that man as futile as is the birth of a hog. Consider that body as a burning ground which does not bear the Tripundra marks. The virtuous man should not cast a glance at him even. Fie on that forehead which does not carry the Tripundra! Fie on that village which has not a single temple of \'Siva! Fie on that birth which is void of the worship of \'Siva! Fie on that knowledge which is void of the knowledge of \'Siva. Know them to be the slanderers of \'Siva who mock at Tripundra. Those that put on the Tripundras, bear \'Siva in their forehead. The Br\=ahmi\d{n} who is Niragnik (without the holy fire) is not nice in every way. So if the worship of \'Siva be not done with any Tripundra is not praiseworthy, even it be attended with abundance of other offerings. Those who do not besmear their bodies with ashes or who do not use the Tripundras, get their previous good deeds converted into bad ones.

21-42. Unless the Tripundra mark is taken up according to the \'S\=astras, the Vaidik Karmas (works) or those performed according to the Smritis prove injurious; the good works whatsoever done by any man count for nothing; the holy words heard seem as if unheard and the study of the Vedas counted as if not studied.

The study of the Vedas, Sacrifices, Charities, asceticisms, vows and fastings of that man, who does not use the Tripundra, all become fruitless. Without using Bha\d{s}ma (ashes) if one wants liberation, then that desire is equivalent to live after taking poison. There is no doubt in this. The Creator has not made the forehead vertically high nor round; but he has made it slightly slanting and curved fit to have the Tripundra. Making thus the forehead, the Creator wants, as it were, to inform everyone that every one ought to use Tripundra marks; the curved lines also are made visible for this purpose. Still the igorant illiterate man does not put up the Tripundra. Unless the Br\=ahma\d{n}as use the curved Tripundras, their meditation won't be successful; they will not have liberation, knowledge, nor their asceticism would bear any fruit. As the \'S\=udras have no right in the study of the Vedas, so the Br\=ahma\d{n}s have not any right to perform the worship of \'Siva, etc., unless they use the Tripundras. First of all, facing eastward, and washing hands and feet, he ought to make a resolve

and then he takes a bath of the ashes mentally, controlling his breath. Then taking the ashes of the Agnihotra sacrifice he is to put some ashes on his own head, uttering ``\=I\'s\=ana'' mantra. Then he is to recite the Puru\d{s}a \'S\=ukta Mantra and apply ashes on his face; with the Aghora mantra on his chest; with the V\=amadeva mantra, on his anus; with Sadyoj\=ata mantra on his legs; and with the mantra Om, he is to besmear his whole body with ashes. This is called the bath of fire by the Munis. So bring all the actions to a successful issue one is to take first of all this bath of fire. Washing his hands, then, he is to make \=Achaman duly; and, according to the above-mentioned rules, he is to apply ashes on his forehead, heart, and all round the neck with the five mantras above-mentioned; or with each mantra he is to apply the Tripundras. Thus all works are fructified and he gets the right to do all the Vaidik actions. The \'S\=udras, even, are not to use the ashes touched by the lowest classes. All the actions ordained by the \'S\=astras are to be done after being besmeared with ashes of the Agnihotra sacrifice; otherwise no action will bear any fruit. All his truth, purity, Japam, offering, oblations to the sacrifice, bathing in the holy place of pilgrimage, and worshipping the gods become useless, who does not hold Tripundra. No fear of disease, sins, famine, or robbers comes to the Br\=ahmi\d{n}s who use Tripundra and rosary of Rudr\=ak\d{s}a and thus remain always pure. In the end, they get the Nirv\=a\d{n}a liberation. During the time of Sr\=addhas (solemn obsequies performed in honour of the manes of deceased ancestors) the Br\=ahmi\d{n}s purify the rows where persons are fed; so much so that the Devas glorify them. One must use the Tripundra marks before one performs any Sr\=addha, Japam sacrifice, offering oblations or worshipping the Visvedev\=as; then one gets deliverance from the jaws of death. O N\=arada! I am now speaking further of the greatness in holding the Bha\d{s}ma; listen.

Here ends the Twelfth Chapter of the Eleventh Book on the greatness in holding the Tripundra and Bha\d{s}ma in the Mah\=apur\=anam \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



