\chapter{On Sad\=ach\=ara}

1. N\=arada said :-- O Bhagav\=an! Thou art the ocean of mercy; kindly speak out to me in brief all the things and the duties to be observed and applied in the \'S\=anti Karmas (the peace bringing acts) of G\=ayatr\={\i}.

2-20. N\=ar\=aya\d{n}a said :-- O N\=arada! The question asked by you is esoteric. Never divulge this to a wicked person or any hypocrite. It is to be kept secret. While doing this \'S\=anti Karma, the Br\=ahma\d{n}as are to perform the Homas with fuel soaked in milk (Payah). If the fuel of the \'Sam\={\i} tree be offered in Homa ceremony, then diseases caused by planets are cured. If the Homa be performed with wet A\'svaththa or Udumbara or other k\d{s}\={\i}ra trees, the diseases caused by demons and hob-goblins are cured. If one offers Tarpa\d{n}am with one palmful of water, repeating the mantra ``S\=uryam Tarpay\=ami namah'', the pending evils are also averted and troubles are ceased. The repetition of the G\=ayatr\={\i} mantra with knees immersed in water averts all evils. The repetition of G\=ayatr\={\i} with body immersed in water upto the throat, averts the danger of life; and the same with whole body immersed grants all success. This is the best of all the \'S\=anti karmas, the acts that bring health, wealth, happiness and peace. While performing Homa, light with the fuel of K\d{s}\={\i}ravrik\d{s}a (the trees that emit milky juice); place Pa\~nchagavya in a vessel made of gold, silver, copper or wood of k\d{s}\={\i}ra trees, or in the vessels made of earth, without any knot or crack; utter the G\=ayatr\={\i} mantra and offer one thousand homas. Sprinkle with water at every offering, touching the Pa\~nchagavya with Ku\'sa grass, thousand times. Then offer the sacrifices there where calamities or nuisances are seen and meditate on the Highest Devat\=a. Thus all the magic spells used by other persons for a malevolent purpose will be rendered nugatory. Bring under your control any Deva yoni, Bh\=ute Yoni, or Pi\'s\=acha Yoni that causes you troubles; then they will quit the house, village, city, way, even that kingdom. Now hear how they are brought under control. Prepare a sacrificial altar with sand, draw a square on it, place a S\=ula (spear) in its centre and plaster with Astagandha. For the sake of converting all the evils, repeat G\=ayatr\={\i} mantra thousand times and impregnate it with the mantra, dig the ground and place or bury the S\=ula under it. Place on the level ground or sthandila a jar or Navaphala, a fruit made of gold, silver, or copper or a newly made earthen jar and enclose it with thread. Then have the sacred waters brought from the several T\={\i}rthas by the Br\=ahmi\d{n}s and fill the jar, repeating the G\=ayatr\={\i} Mantra. Put within this jar the twigs,

then of Cardamom tree, sandal tree, karp\=ura tree, j\=ati, apar\=ajit\=a. Saha Dev\={\i}, P\=atala, Mallik\=a flower, Bel leaves, rice, barley, Til, mustard, a\'svattha, and udumbara trees and throw them within the jar. (The K\d{s}\={\i}ra trees are A\'svaththa, Udumbara, Plak\d{s}a, and Nyagrodha.) Doing all these, prepare one Ku\'sa Kurcha made of twenty seven Ku\'sa grass in the form of a braid of hair, round one end of a straight rod and tied in a knot and place it there. Then take your bath and repeat the G\=ayatr\={\i} mantra over it one thousand times with your whole mind fully concentrated. Then the Veda knowing Br\=ahma\d{n}as would recite the Saura mantra and sprinkle the man attacked by the demon with water and make him drink also the water saturated with the mantra and bathe him also with that. Then that person will be dispossessed of the devil and be happy. Even when the man (possessed by a devil), is going to die, he gets his life again if he be made to drink this water saturated with this mantra and have his bath with that. So a wise King must do this, with a desire to have a long life; and after he is sprinkled with this mantra-charged water he is to give one hundred cows to the Br\=ahma\d{n}as as the Dak\d{s}i\d{n}\=a.

21-44. The Dak\d{s}i\d{n}\=a is to be given according to one's might; and specially what gives satisfaction to the Br\=ahma\d{n}as. If one be terrified by a Devil or so or by the mischievous magic spells of others, one is to sit on Saturday under an A\'svattha tree and repeat one hundred G\=ayatr\={\i}s. For the cure of all diseases, if one has to perform the Mrityunjaya Homa, one will have to do Homa with the Gulancha creeper, soaked in milk and deducting the knots thereof. For the pacification of fever, mango leaves soaked with milk ought to be used in the Homa ceremony. The wasting diseases are cured if Homa be performed with the leaves of the Vacha soaked in milk; phthisis or consumption is cured if Homa be performed with curd, milk and ghee. Again, if offerings be given to the Solar Deity and if P\=ayas\=annam be given to him and if this be given to the consumptive patient, his disease will be cured. Again, on the Am\=avasy\=a tithi (new moon day) if Homa be performed with Soma creeper (excluding the knot joints) soaked in milk, then consumption will be cured. If Homa be made with the flowers of the \'Samkhya tree, then leprosy is cured; if the Homa be done with the seed of Ap\=am\=arga, the Mrigis or the hysteric and epileptic fits are cured. So if Homa be performed with the fuel of K\d{s}\={\i}ra trees, lunacy is cured; if, with the fuel of udumbara, meha (spermatorrhea) is cured; if with sugarcane juice, gonorrhea is cured; if with curd, milk and ghee or with the ghee of Kapil\=a cow, the homa be performed, the Mas\=urik\=a disease or smallpox will be cured, and if Homa be performed with the fuels of Udumbara, Vata, and A\'svatha be performed, then the diseases of cows, elephants

and horses are cured. If the trouble be caused by many ants and ant-hills (Madhu Valm\={\i}ka) then perform the Homa ceremony with the fuel of Sam\={\i} tree one hundred times and with the food prepared of ghee one hundred times and offer sacrifices with the rest of the food; then the above troubles will cease. If there be a earth quake or if there be seen flashes of lightnings, then homa is to be performed with the fuel of Vana Vetasa and the whole kingdom will be happy. If you surcharge any piece of iron with G\=ayatr\={\i} mantra repeated hundred times and if you throw it in any direction then no fear will arise from that quarter out of fire, air, or any other enemy. If one be imprisoned, and if he repeats the G\=ayatr\={\i} mentally, he will be liberated from the prison. If you touch the man possessed by a devil, disease or mortification and sorrow with the Ku\'sa grass and repeat the G\=ayatr\={\i} mantra, and thus charge him with Divine electricity, that man will be liberated from the fear caused thereby. If you make the man possessed by devils, etc., drink the water charged with the G\=ayatr\={\i} mantra or if you cast on his body the ashes charged with hundred G\=ayatr\={\i}s or tie those ashes on his head repeating the G\=ayatr\={\i} mantra, he will be instantly freed of all diseases and will live for one hundred years in happiness. In case a man is unable to do fully all these himself, he can get all these done by other Br\=ahma\d{n}as and pay Dak\d{s}i\d{n}\=a (fees) to them duly for the same.

O N\=arada! Now I will tell you how nourishment and wealth are attained. Wealth is attained if the Homa be performed with red lotus or fresh J\=at\={\i} flower or with the \'S\=ali rice or with the fuel of Bel tress, leaves, flowers, fruits or roots or with any portions thereof. If for one week the oblations be offered with fuel of Bel tree mixed with P\=ayasa or with ghee one hundred times, then Lak\d{s}m\={\i} Dev\={\i} will surely be attained. If the Homa be performed with L\=aja (fried rice) mixed with curd, milk, and ghee, the daughter will be obtained. If for one week, the Homa be performed with red lotus, then gold is obtained. If the Tarpa\d{n}am (peace offering) be offered to the Sun, then the treasures, gold hidden under the water, are obtained. If the Homa be performed with food (Anna) then Anna is obtained; if Homa be performed with rice, then rice is obtained.

45-51. If Homa be performed with calfdung, dried and powdered, then animals are obtained. If Homa be performed with Priyangu, P\=ayasa or ghee then the progeny is obtained. If the oblations of P\=ayas\=anna be offered to the Solar Deity and if the Pras\=adam (remnant) be given to one's wife under menstruation to eat, then

excellent sons will be obtained. If the Homa be performed with the fuel of wet pointed K\d{s}\={\i}ra trees, then longevity is attained. If Homa be performed with the fuel of the Pal\=a\'sa tree, pointed and wet and mixed with curd, milk, and ghee for one hundred times consecutively, then longevity and gold are attained. If the Homa be performed with Durb\=a grass, milk, honey or ghee, one hundred times, then longevity and golden lotus are obtained. If for one week the Homa ceremony be performed with the fuel of \'Sam\={\i} tree mixed with food, milk or ghee one hundred times of each or if for one week the Homa he performed with the fuel of Nyagrodha tree and afterwards one hundred Homas be made with P\=ayas\=anna, the fear of unnatural death is removed.

52-60. That man can conquer death who can remain for one week living on milk only and who performs during that time hundreds, and hundreds of Homas and repeats the G\=ayatr\={\i}, controlling his speech. If anybody can fast three nights and control his speech and repeat G\=ayatr\={\i} he gets himself freed from the hands of Death; or totally immersed in water if he repeats G\=ayatr\={\i}, he will be saved from the impending danger of death. If anybody repeats the G\=ayatr\={\i} mantra for one month, taking his seat under a Bel tree or performs Homa with Bel fruit, root or leaves, he gets kingdoms. (Know all the Mantrams are electric in their effects). Similarly if anybody performs Homa with one hundred lotuses, he gets a foeless kingdom. So one becomes the lord of a village if one performs Homa with Yav\=agu (barley gruel) and S\=alidh\=anya. If the Homa be performed with the fuel of Asvaththa tree, victory in battle is ensured and if the Homa be performed with the fuel of \=Akanda tree, then victory everywhere is ensured. If one hundred Homas be performed extending a week with Vetasa tree's leaves or fruits, dipped in milk and mixed with P\=ayasa, the rainfall is ensured. Similarly if anybody repeats G\=ayatr\={\i} for one week with his body upto navel immersed in water, the rainfall is ensured; on the contrary if the Homa be performed with ashes in water, then the cessation of heavy rainfall is ensured. The Homa with the fuel of Pal\=asa gives Brahmateja; Homa with the flowers of Pal\=asa gives everything desired. Homa with milk or drinking Br\=ahmarasa, charged with mantra, increases the intellect; and the Homa with ghee gives Buddhi (medh\=a) (intelligence).

61-69. Homa with flowers gives good smell; Homa with thread gives cloth; Homa with salt and honey mixed or Bel flowers gives one power to control anything and everything that is desired. If anybody bathes everyday immersed completely within water and sprinkles water on his body, he becomes cured of diseases and

he becomes very healthy. If any Br\=ahma\d{n}a does these things for others, he becomes also no doubt healthy. If anybody wants to increase his life period he should practise good deeds and repeat G\=ayatr\={\i} thousand times daily for one month. Thus his longevity will be increased. Two month\'s such practice gives long life and perpetual health; three month's such practice will give life, health, and wealth, four months such practice gives longevity, wealth, fame, women, sons, etc., five months such practice gives longevity, health, wealth, wife, sons and learning. So one should repeat this as many months in proportion to the number of his desires and he would get them. Again any Br\=ahma\d{n}a who stands on one leg without holding any other thing and raises both his hands and daily repeats three hundred G\=ayatr\={\i}s for one month, gets all his desires fulfilled. And if he repeats one thousand one hundred G\=ayatr\={\i}s, there is nothing in this world that is not met and attained with success. Controlling the Pr\=a\d{n}a (inhaling) and Ap\=ana (exhaling) V\=ayu (breath), he who repeats daily three hundred G\=ayatr\={\i}s to the Dev\={\i}, his highest desires are satisfied.

70-77. Vi\'sv\=amitra \d{R}i\d{s}i says :-- Standing on one leg, with both hands raised and controlling V\=ayu he who repeats daily one hundred G\=ayatr\={\i}s for one month, gets all his desires fulfilled. Similarly with three hundred or thousand repeatings, all things are attained. Submerged under water, if one repeats G\=ayatr\={\i}s as many times as mentioned above, he gets everything. If, for one year, with hands uplifted and without holding anything, anyone stands on one leg, controls one's breath and repeats G\=ayatr\={\i} mantra three hundred times or thousand times, eating Havisy\=annam only in the night time, he becomes a \d{R}i\d{s}i (Seer). This thing repeated two years gives infallible speech; three years gives knowledge of the present, past and future; four years will enable one to see face to face the Solar God; five years will give the eight Siddhis, lightness, etc., six years will enable one to assume forms as he desires; seven years gives immortality; nine years gives Manuhood; ten years gives Indrahood; eleven years gives Brahm\=a-hood; and twelve years gives the state of Parama Brahma.

78-90. O N\=arada! By these practices of Tapasy\=as you and other \d{R}i\d{s}is have been able to conquer the three Lokas (regions). Some ate only vegetables; some fruits; some, roots; some simply water; some, ghee; some, Somarasa; whereas some others ate only charu and did tapasy\=a. Some \d{R}i\d{s}is practised this great Tapasy\=a by eating very little for a fortnight only. Some ate food, only what they got by begging during the day; and some ate only Havi\d{s}y\=anna. O N\=arada! Now hear the rules for the purification and expiation of sins. For the expiation of the sin incurred in stealing gold, one is to repeat three thousand G\=ayatr\={\i}s (daily) for one

month; then the sin will be destroyed. By this act also the sins incurred by drinking or by going to one's Guru's wife are destroyed. Vi\'sv\=amitra \d{R}i\d{s}i says :-- The sin incurred by killing a Br\=ahma\d{n}a (Brahmahaty\=a) is destroyed if one erects a shed in a forest and, living there, repeats three thousand G\=ayatr\={\i}s daily for one month. Those Br\=ahma\d{n}as that have committed the Great Sins (i.e., Mah\=ap\=atakas), become free, if they repeat one thousand G\=ayatr\={\i}s daily, submerged under water, for twelve successive days. By controlling speech and by practising Pr\=a\d{n}\=ay\=ama, if one repeats three thousand G\=ayatr\={\i}s daily for one month, one will be free from the Mah\=ap\=atakas. If one practises one thousand Pr\=a\d{n}\=ay\=amas repeating the G\=ayatr\={\i}, one becomes freed also of Brahmahaty\=a. If one draws upwards the Pr\=a\d{n}a and Ap\=ana V\=ayus six times, repeating the G\=ayatr\={\i} with collected mind, this destroys all the sins and it is called all-sin destroying Pr\=a\d{n}\=ay\=ama. If one practises this Pr\=a\d{n}\=ay\=ama one thousand times for one month, the lord of the earth becomes freed of all sins. If any Br\=ahma\d{n}a incurs the sin of killing a cow, for twelve days he is to repeat three thousand G\=ayatr\={\i}s daily for expiation. Similarly the repetition of ten thousand G\=ayatr\={\i}s removes the sin of going to those not fit to be gone into, eating the uneatables, stealing and killing and this act brings in peace. All sins are destroyed by performing one hundred Pr\=a\d{n}\=ay\=amas with G\=ayatr\={\i}. Again if there be a mixture of various sins, one will have to live in the forest for one month and repeat one thousand G\=ayatr\={\i}s or practise fasting and repeat three thousand G\=ayatr\={\i}s; thus all sins will be destroyed.

91-100. To repeat G\=ayatr\={\i} twenty-four thousand times is equal to performing the Krichchhra vrata and to repeat sixty four thousand G\=ayatr\={\i}s duly is equal to performing the Ch\=andr\=ayana. If anybody repeats, in the morning and evening Sandhy\=a times, the merit giving G\=ayatr\={\i} one hundred times, with Pr\=a\d{n}\=ay\=ama, all his sins are destroyed. So, submerged under water, if one repeats the G\=ayatr\={\i} Dev\={\i}, meditating Her in the Sun, one hundred times daily, one's all the sins are fully destroyed. O N\=arada! Thus I have described to you all about averting or destroying the evils and the purification of various sins. All this is secret. Keep it carefully concealed. Never divulge this. Whoever divulges this will bring his own ruin. I have spoken to you, in brief, all about Sad\=ach\=ara (right way of living). If anybody practises this duly, according to rules, \'Sr\={\i} Mah\=am\=ay\=a Durg\=a Dev\={\i} becomes pleased with him. If anybody wants to have both enjoyment and liberation, he is to practise all these daily, as well as the occasional, and K\=amya (desired) duties duly according to rules. It is stated in all the \'S\=astras, that this \=Ach\=ara (right way of living) is the foremost and the chief Dharma, the Deity of which is

the Supreme Mother Herself. O N\=arada! That man who practises duly this \=Ach\=ara is, in this world, holy, happy and blessed. This I speak to you truly. If anybody desires to get the Dev\={\i} Bhagavat\={\i}'s Grace, he should first of all set himself at once to practise this Sad\=ach\=ara. He who hears this gets wealth and great happiness. There is no doubt in this. Now speak what more you want to hear.

Here ends the Twenty fourth Chapter of the Eleventh Book on Sad\=ach\=ara in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa. The Eleventh Book Completed.

[Here ends the Eleventh Book.]



