\chapter{On the cause of \'Sr\=addha in other Devas than the Dev\={\i} G\=ayatr\={\i}}

1-20. Vy\=asa said :-- O King Janamejaya! Once on a time, on account of an evil turn of Fate (Karma) of the human beings, Indra did not rain on this earth for fifteen years. Owing to want of rain, the famine appeared horribly; and almost all the beings lost their lives. No one could count in every house the number of the dead persons. Out of hunger the people began to eat horses; some began to eat bears and pigs, some began to eat the dead bodies while some others carried on any how their lives. The people were so much distressed with hunger that the mother did not refrain from eating her baby child and the husband did not refrain from eating his wife. O King! The Br\=ahma\d{n}as then united and after due discussion, came to the conclusion that they would go to the hermit Gautama who would be able to remove their distress. So all of them wanted to go quickly to the hermitage of the Muni Gautama. They began to say :-- ``We hear that there is no famine in the hermitage of Gautama. Various persons are running there from various quarters.'' Thus coming to a conclusion, the Br\=ahma\d{n}as went to the Gautama's \=A\'srama with their cows, servants and relations. Some went from the east; some from the south; some from the west, and some from the north. Thus from various

quarters the people flocked there. Seeing the Br\=ahma\d{n}as coming there, the \d{R}i\d{s}i Gautama bowed down to them and gave them a cordial welcome and served them with seats, etc. When all took their seats and became calm and quiet, Gautama enquired about their welfare and the cause of their arrival. They described everything about the dire famine and their own states and expressed their deep regret. Seeing them very much distressed, the Muni gave them word not to have any fear; he said :-- ``I am today become blessed by the arrival of the great ascetics and honourable persons like you. I am your servant. You consider all my houses as yours. Be quite comfortable. Bear no uneasiness. When your servant is alive, what fear do your entertain and whom do you fear? When the demerits are transformed into good merits by your mere sight, and when you have blessed my house with the dust of your holy feet, then who is more blessed than me? O Vipras! Kindly perform your Sandhy\=as, and Japams and rest here at ease.'' Vy\=asa said :-- O King Janamejaya! Thus consoling the Br\=ahma\d{n}as, the \d{R}i\d{s}i Gautama began to worship the G\=ayatr\={\i} Dev\={\i} with rapt devotional trance. ``O Dev\={\i} G\=ayatr\={\i}! Obeisance to Thee! Thou art the Great Vidy\=a, the Mother of the Vedas, Higher than the Highest; Thou art Vy\=arhiti represented by the Mantra `Om Bhur Bhuvah Svah,' O Mother! Thou art the state of equilibrium, i.e., the Tur\={\i}ya; Thou art of of the Form of Hr\={\i}m; Thou art Sv\=ah\=a and Svadh\=a; Thou grantest the desires of the Bhaktas. Thou art the Witness of the three states, J\=agrat (waking), Svapna (dreaming) and Su\d{s}upti (deep sleep). Thou art the Tur\={\i}ya and Sachchid\=ananda Brahm\=a. O Dev\={\i}! Thou resident in the Solar Orb and appearest as a ruddy girl in the morning, an youthful maiden at noon and a black old woman in the evening. O Dev\={\i}! Obeisance to Thee! Now shew favour on us at this severe famine time when all the beings are well nigh on the way to destruction.''

21-40. Thus praised and worshipped, the World-Mother appeared and gave to the \d{R}i\d{s}i one vessel (cup) full to the brim by which every one can he fed and nourished. The Mother told the Muni :-- ``This full vessel, given by me to you will yield whatever you wish.'' Thus saying, the Dev\={\i} G\=ayatr\={\i}, Higher than the Highest, vanished. Then, according to the wish of the Muni, came out from that cup, mountains of cooked rice, various curries and sweetmeats, lots of grass and fodder, silken clothings, various ornaments and various articles and vessels for sacrificial purposes. In fact whatever the Muni Gautama wished, that came out of the brimful cup, given by the Dev\={\i} G\=ayatr\={\i}. Then the Muni Gautama called the other Munis that came there and gave them wealth, grains, clothing, ornament, and the sacrificial ladles and spoons and cows and buffaloes for the sacrificial purposes. The

Munis then assembled and performed various yaj\~n\=as. The place all round, then flourished and became so much prosperous that it looked like a second heaven. In fact whatever fair and beautiful there exist in the Trilokas, all came from the brimful cup given by the Dev\={\i} G\=ayatr\={\i}. At this time the Munis, with sandalpaste all over on their bodies, and decorated with very bright ornaments looked like the gods and their wives looked like goddesses. Daily utsabs began to be held in in the \=A\'srama of Gautama. Nowhere were seen any diseases or dacoities and there was no fear from any such things. Gradually the \=A\'srama's boundary extended to one hundred Yojanas (4 hundred miles). Hearing this greatness of Gautama, many persons came there from various quarters. And the Muni Gautama, too, gave them words, ``cast away fear'' and fed them. The Devas, on the other hand, became very much satisfied by the various Yaj\~n\=as and extolled the Muni's greatness. So much so, that the famous Indra, the Lord of the Devas, came in the midst of the assembly and extolled his greatness, thus :-- ``This Gautama has fulfilled all our wishes and has verily become a Kalpa Vrik\d{s}a (celestial tree yielding all desires). If this man had not done such things, in this hard famine time, we would not have got the Havih offered in sacrifices and the prospect of our lives would have been at stake.'' O King Janamejaya! Thus the Muni Gautama fed and nourished for twelve years all the Munis, like his sons and that place came to be recognised as the chief centre (the Head Quarters) of the G\=ayatr\={\i} Dev\={\i}. Even today, all the Munis perform with devotion the Pura\d{s}acharanams and worship thrice the Bhagavat\={\i} G\=ayatr\={\i} Dev\={\i}. Even today the Dev\={\i} is there seen as a girl in the morning, as a youthful maiden at noon, as an old woman in the evening. Then, once on a time, N\=arada, of best conduct, came there playing on his great lute and singing in tune the highest glories of G\=ayatr\={\i} and took his seat in the assembly of the Munis.

41-62. Seeing the tranquil hearted N\=arada coming there, Gautama and the other Munis received him duly and worshipped him with the P\=adya and Arghya. In course of conversation he began to describe the glories of Gautama and said :-- ``O Best of Munis! I have heard from the mouth of Indra, in the assembly of the Devas, your glories as to your supporting and feeding the pure-hearted Munis and I have come to see you. By the Grace of \'Sr\={\i} Bhagavat\={\i} G\=ayatr\={\i} Dev\={\i}, you have now become blessed. There is no doubt in this.'' Thus saying, the Devar\d{s}i N\=arada entered into the temple of the Dev\={\i} G\=ayatr\={\i} and with eyes gladdened by love, saw the Dev\={\i} there and offered due hymns in praise of Her and then ascended to the Heavens. Here, on the other hand, the Br\=ahma\d{n}as that were fed by Gautama, became jealous at so much honour offered

to Gautama and tried their best so that no further honour be paid to him. They further settled not to stay any longer in his \=A\'srama, when the next good harvest season comes. (Thus his glories will wane). O King! Some days passed when good rains fell and there was an abundance of crops everywhere and the famine ended. Hearing this, all the Br\=ahmi\d{n}s united, Alas! O King! to curse the Gautama. Oh! Their fathers and mothers are blessed in whom do not arise such feelings of jealousy! This all is the wonderful play of the powerful Time; it cannot be expressed by any person. O King! These Br\=ahmi\d{n}s created, by M\=ay\=a, an aged cow, who was to die and pushed her in the sacrificial hall of the Muni Gautama at the time of the Homa ceremony. Seeing that cow entering into the enclosure where the Homa was being performed, Gautama cried out ``Hoom Hoom'' when the cow fell there and died. And the other Br\=ahma\d{n}as instantly cried out ``Look! Look! The wicked Gautama has killed the cow.'' Seeing this inconceivable event, Gautama was greatly nonplussed and, completing his Homa ceremony, entered into Sam\=adhi and began to think the cause of it. Then, coming to know that this has been concocted by the M\=ay\=a of the Br\=ahmi\d{n}s, he became angry like Rudra at the time of dissolution; his eyes were reddened and he cursed the \d{R}i\d{s}is, thus :-- Oh vile Br\=ahmi\d{n}s! When you are ready to cause mischief to me unjustly then let you be averse to meditate and do the japam of the Dev\={\i} G\=ayatr\={\i}, the Mother of the Vedas. For your this act, never you will be eager to perform any Vedic sacrificial acts or any action concerning thereof. There is no doubt in this. You will be always averse to the mantra of \'Siva or the Tantra of \'Siva. You will be always averse to M\=ula Prakriti \'Sr\={\i} Dev\={\i}, to Her Dhy\=anam, mantra, to any conversation regarding Her; to the visiting of Her place or Temple, to do worship and other ceremonies to Her, to see the Grand Festivals of the Dev\={\i}, to singing the names and glories of the Dev\={\i}, to sit before the Dev\={\i} and to adore Her.

63-81. O vile Br\=ahma\d{n}as! You will be always averse to see the festivals of \'Siva, to worship \'Siva, to Rudr\=ak\d{s}a, to the Bel leaves, and to the holy Bha\d{s}ma (ashes). You will be wholly indifferent to practise the right ways of living as presented in the Vedas and Smritis, to preserve your conduct good and to observe the path of knowledge to Advaita J\~n\=anam, to practise restraint of senses and continence, to the daily practices of Sandhy\=a Bandanam, to performing the Agnihotra ceremonies, to the study of the Vedas according to one's own \'S\=akh\=a or to the daily studies thereof as to teach those things or to give, as gifts, cows, etc., or to perform the \'Sr\=addhas of the fathers, etc., or to perform Krichchra Ch\=andr\=aya\d{n}a and other penances. O Vile Br\=ahma\d{n}as! As you are ready to do these mean things, you will have to suffer for this that you will desist from worshipping the Most

Adorable \'Sr\={\i} Bhagavat\={\i} Dev\={\i} and that you will worship the other Devas with faith and devotion and hold on your bodies \'Samkha, Chakra and other signs. You will follow the K\=ap\=alikas, Bauddha \'S\=astras and other heretics. You will sell your father, mother, brothers, sisters, sons and daughters and even your wives too!

You will sell the Vedas, T\={\i}rthas, and your Dharma. You will not feel ashamed in any way to sell all these. You will certainly have faith in K\=ap\=alika and Bauddha opinions, P\=a\~nchar\=atras and K\=ama \'S\=astras. O vile Br\=ahmanas! You will not hesitate to go to your mother, daughters or sisters and you will always be licentious with other\'s wives and spend your time in that. This is not to you only but to the women and men, all that will come in your families. Let the G\=ayatr\={\i} Dev\={\i} be always indignant with you and let you all go in the end to the Andha K\=upa hells, etc. Vy\=asa said :-- O Janamejaya! Thus taking the water symbolising the true rules and laws of creation, and cursing the Br\=ahma\d{n}as, the Muni Gautama went hastily to see the G\=ayatr\={\i} Dev\={\i} and, on arriving at the temple there bowed down to Her. The Dev\={\i}, too, became surprised to see their actions. O King! Even to day Her Lotus Face looks similarly astonished!

82-90. Then the G\=ayatr\={\i} Dev\={\i} told Gautama with amazement :-- ``O Gautama! The venom of the snake does not become less if you feed the serpents with milk; so never mind all these things; the Karmas take their peculiar turns; it is hard to say when will happen what things. Now be peaceful. Do not be sorry.'' Hearing these words of the Dev\={\i}, Gautama bowed down to Her and went thence to his own \=A\'srama. Here, on the other hand, the Br\=ahma\d{n}as forgot everything due to the curse of Gautama, of the Vedas and the G\=ayatr\={\i} Mantra. They then began to look at this event with wonder as unique and extraordinary. All united they afterwards repented and going before Gautama, fell prostrate at his feet. But they could not speak any word out of shame. Only they said frequently :-- ``Be pleased, be pleased with us.'' When all the assembly of the Br\=ahma\d{n}as prayed to him for favour and grace, the Muni Gautama took pity on them and replied. My word will never turn out false. You will have to remain in Kumbh\={\i}p\=aka hell upto the time when \'Sr\={\i} Kri\d{s}\d{n}a will take his incarnation. Then you will be born in the earth in the K\=al\={\i} age and whatever I have uttered will exactly come unto you. And if you are in earnest to avert my curse then go and worship the Lotus Feet of \'Sr\={\i} G\=ayatr\={\i} Dev\={\i}. The is no other remedy.

91-100. Vy\=asa said :-- Thus dismissing the Br\=ahma\d{n}as, Gautama Muni thought that all these occurred as a result of Pr\=arabdha Karma and he became calm and quiet. For this reason, after \'Sr\={\i} Kri\d{s}\d{n}a Mah\=ar\=aja ascended to the Heavens, when the K\=al\={\i} age came, those cursed Br\=ahma\d{n}as got out of the Kumbh\={\i}p\=aka hell and took their births in this earth as Br\=ahmi\d{n}s, devoid of the three Sandhy\=as, devoid of the devotion to G\=ayatr\={\i}, devoid of faith in the Vedas, advocating the heretic\'s opinion and unwilling to perform Agnihotra and other religious sacrifices and duties and they were devoid of Svadh\=a and Sv\=ah\=a. They forgot entirely the Unmanifested M\=ula Prakriti Bhagavat\={\i}. Some of them began to mark on their bodies various heretical signs, e.g., Taptam\=udr\=a, etc.; some became K\=ap\=alikas; some became Kaulas; some Bauddhas and some Jainas. Many of them, though learned, became lewd and addicted to other's wives and engaged themselves in vain and bad disputations. For these, they will have to go again surely to the Kumbh\={\i}p\=aka hell. So O King! Worship with your heart and soul \'Sr\={\i} Bhagavat\={\i} Parame\'svar\={\i} Dev\={\i}. The worship of Vi\d{s}\d{n}u or \'Siva is not constant (to be done everyday); only the worship of \'Sakti is to be constantly performed. For this reason whoever does not worship \'Sakti is sure to fall. Thus I have answered all your questions. Now I shall describe the highest and most beautiful place, Ma\d{n}idv\={\i}pa of the Primal Force Bhagavat\={\i}, the Deliverer from this bondage of world. Listen.

Here ends the Ninth Chapter of the Twelfth Book on the cause of \'Sr\=addha in other Devas than the Dev\={\i} G\=ayatr\={\i} in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



