\chapter{On the description of the Kavacha of \'Sr\={\i} G\=ayatr\={\i} Dev\={\i}}

1-3. N\=arada spoke :-- O Bhagav\=an! Thou art the Lord of this world; Thou canst shew favour and disfavour both; Thou art specially versed in the sixty-four Kal\=as (arts of learning); Thou art the chief of the Yogis. I therefore ask Thee to solve a doubt of mine. By what Pu\d{n}yam a man can become free from all his sins and limitations and he can realise and become of the nature of Brahm\=a. O Lord! And what are duly the \d{R}i\d{s}is, Chha\d{n}das, Devat\=as, Dhy\=an, and Ny\=asa, etc., of this meritorious act? I want to hear.

4-25. N\=ar\=aya\d{n}a said :-- O N\=arada! There is but one and the only one way to this and though that is very secret, I will disclose that to you. It is the G\=ayatr\={\i}-Kavacha. It can destroy all sins. Therefore to recite or to hold it on one's body enables the man to become free from all his sins and to get all his desires fulfilled and he gets the S\=ayuya Mukti with the Dev\={\i} (be merged in the Dev\={\i}-Body). Now hear the \d{R}i\d{s}is, Chha\d{n}das, etc., of this Kavacha :-- Brahm\=a, Vi\d{s}\d{n}u and Mahe\'svara are the \d{R}i\d{s}is; the Rik, Yajus, S\=ama and Atharva Vedas are the Chha\d{n}das; the Param\=a Kal\=a G\=ayatr\={\i} of the nature of Brahm\=a is the Devat\=a; ``Tat'' in G\=ayatr\={\i} is the V\={\i}ja; ``Bharga'' is the \'Sakti; and ``Dh\={\i}yah'' is the K\={\i}laka; and its viniyoga (application) is in getting the Mok\d{s}a (liberation.). With the first four syllables touch the heart; with

the next three letters touch the head; with the next four letters touch the tuft on the crown of the head; with the next three letters on the Kavacha; with the next four letters on the eyes and with the last four letters make the Ny\=asa, all over the body repeating ``Astr\=aya Phat,'' O N\=arada! Hear now the Dhy\=anam of G\=ayatr\={\i}, that grants all desires. The G\=ayatr\={\i} Dev\={\i} has five faces; one of which is of white colour; and the other four is of pearl, Vidruma, golden, and N\={\i}lak\=antama\d{n}i colour respectively. Each face has got three eyes; on the head there is a crown of jewels and the digit of the Moon is shining there. Her body is composed of the twenty-four tattvas. She has ten hands :-- On the top right and left hands there are two lotuses; lower down, there are disc and conch shell; lower down, there are rope and skull; lower down, there are noose and goad; and on the bottom hands right and left she is making signs of ``No fear'' and ``ready to grant boons.'' Thus meditating on \'Sr\={\i} G\=ayatr\={\i}, one is to recite the Kavacha thus :-- Let the G\=ayatr\={\i} Dev\={\i} protect my front; S\=avitr\={\i} Dev\={\i} protect my right; the Sandhy\=a Dev\={\i}, my back and the Dev\={\i} Sarasvat\={\i}, my left. Let my Mother P\=arvat\={\i} Dev\={\i} protect my quarters. Let Jala\'s\=ayin\={\i} protect the southeast; Y\=atudh\=ana Bhayankar\={\i} protect my South-west; Pavam\=anavil\=asin\={\i} my north-west; Rudrar\=upi\d{n}\={\i} Rudr\=a\d{n}\={\i} protect my north-east. Let Brahm\=a\d{n}\={\i} protect my top and Vai\d{s}\d{n}av\={\i} protect my nether regions. Let the word ``Tat'' in the G\=ayatr\={\i} protect my legs; ``Savituh'' protect my Knees; ``Vare\d{n}yam,'' protect my loins; ``Bhargah,'' my navel. Let ``Devasya'' protect my heart; ``Dh\={\i}mah\={\i}'' protect my neck; ``Dh\={\i}yah,'' protect my eyes; ``Yah,'' protect my forehead; ``Nah'' protect my head; and ``Prachoday\=at'' protect the tuft on the crown of my head.

Again let the ``Tat'' of the twenty-four syllabled G\=ayatr\={\i} protect my head; ``Sa,'' protect my forehead; ``Vi'' protect my eyes; ``Tu'' my cheeks; ``Va,'' protect my nostrils; ``Re,'' my mouth; ``\d{n}i'' protect my upper lip; ``Yah'' protect my lower lip; ``Bha'' within my face; ``rgo,'' protect my cheeks; ``De,'' my throat; ``Va'' my shoulders; ``Sya'' my right hand; ``Dhi'' my navel; ``ma,'' my heart, ``Hi,'' my belly; ``Dh\={\i},'' my navel; ``Yo'' my loins; ``Yo'', my anus; ``nah,'' my thighs, ``Pra,'' my Knees; ``Cho'' my shanks ``Da'' my heels; ``Y\=a'' my legs; and let ``at'' protect all my sides. O N\=arada! This divine Kavacha of the Dev\={\i} G\=ayatr\={\i} can baffle hundreds and thousands of obstructions and evils; can grant sixty-four Kal\=as and liberation. By the glory of this Kavacha, man can become free from all evils and can attain the state of Brahm\=a. Moreover whoever reads or hears this acquires the fruits of making a gift of a thousand cows.

Here ends the Third Chapter of the Twelfth Book on the description of the Kavacha of \'Sr\={\i} G\=ayatr\={\i} Dev\={\i} in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



