\chapter{On the description of Ma\d{n}i Dv\={\i}pa}

1-17. Vy\=asa said :-- O King Janamejaya! The Ratnagriha, above mentioned, is the Central, the Chief and the Crowning Place of M\=ula Prakriti. [The nine jewels are :-- (1) Mukt\=a, (2) M\=a\d{n}ikya, (3) Vaid\=urya, (4) Gomeda, (5) Vajra, (6) Vidruma, (7) Padmar\=aga, (8) Marakata, and (9) N\={\i}la.] This is situated in the centre of all the enclosures. Within this there are the four Mandapas, i.e., halls built of one thousand (i.e., innumerable) pillars. These are the \'Sring\=ara Mandapa, Mukti Mandapa, J\~n\=ana Mandapa and Ek\=anta Mandapa; on the top there are canopies of various colours; within are many scented articles scented by the Dh\=upas, etc. The brilliance of each of these is like that of one Koti Suns. On all sides of these four Mandapas there are nice groups of gardens of Ka\'sm\={\i}ra, Mallik\=a, and Kunda flowers. Various scents, and scented articles, for example, of musk, etc., are fully arranged in due order. There is a very big lotus tank here; the steps leading to it are built of jewels. Its water is nectar, on it are innumerable full-blown lotuses and the bees are humming always over them. Many birds, swans, K\=arandavas, etc., are swimming to and fro. The sweet scents of lotuses are playing all round. In fact, the whole Ma\d{n}idv\={\i}pa is perfumed with various scented things. Within the \'Sring\=ara Mandapa, the Dev\={\i} Bhagavat\={\i} is situated in the centre on an \=Asana (seat) and She hears the songs sung in tune by the other Dev\={\i}s along with the other Devas. Similarly sitting on the Mukti Mandapa, She frees the J\={\i}vas from the bondages of the world. Sitting on the J\~n\=ana Mandapa, She gives instructions on J\~n\=ana, and sitting on the fourth Ek\=anta Mandapa, She consults with Her ministers, the Sakh\={\i}s, Ananga Kusuma, etc., on the creation, preservation, etc., of the universe. O King! Now I shall describe about the main, Kh\=as, room of \'Sr\={\i} Dev\={\i}. Listen. The Kh\=as Mah\=al palace of the Dev\={\i} Bhagavat\={\i} is named \'Sr\={\i} Chint\=ama\d{n}i Griha. Within this is placed the raised

platform, the dais and sofa whereon the Dev\={\i} taketh Her honourable seat. The ten \'Sakti-tattvas form the staircases. The four legs are (1) Brahm\=a, (2) Vi\d{s}\d{n}u, (3) Rudra, and (4) Mahe\'svara. Sad\=a\'siva forms the upper covering plank. Over this \'Sr\={\i} Bhuvane\'svara Mah\=a Deva or the Supreme Architect of the Universe is reigning. Now hear something about this Bhuvane\'svara. Before creation while intending to sport, the Dev\={\i} Bhagavat\={\i} divided Her Body into two parts and from the right part created Bhuvane\'svara. He has five faces and each face has three eyes. He has four hands and He is holding in each hand speer, signs indicating do not fear, axe, and signs granting boons. He looks sixteen years old. The lustre of of His Body is more beautiful then Koti Kandarpas and more fiery than thousand Suns; and at the same time cool like Koti Suns. His colour is crystal white, and on His left lap \'Sr\={\i} Bhuvane\'svar\={\i} Dev\={\i} is always sitting.

18-29. On the hip of \'Sr\={\i} Bhuvane\'svar\={\i}, is shining the girdle with small tinkling bells, built of various jewels; the ornaments on the arms are made of burnished gold studded with Vaid\=uryama\d{n}is; the T\=atanka ornaments on Her ears are very beautiful like \'Sr\={\i}chakra and they enhance very much the beauty of Her lotus face. The beauty of Her forehead vies with, or defies the Moon of the eighth bright lunar day. Her lips challenge the fully ripened Bimba fruits. Her face is shining with the Tilaka mark made of musk and saffron. The divine crown on Her head is beautified with the Sun and Moon made of jewels; the nose ornaments are like the star Venus and built of transparent gems, looking exceedingly beautiful and shedding charming lustre all around. The neck is decorated with necklaces built of gems and jewels. Her breasts are nicely decorated with camphor and saffron. Her neck is shining like a conchshell decorated with artistic designs. Her teeth look like fully ripe pomegranate fruits. On Her head is shining the jewel crown. Her lotus face is beautified with alak\=a as if these are mad bees. Her navel is beautiful like the whirls in the river Bh\=agirath\={\i}; Her fingers are decorated with jewel rings; She has three eyes like lotus leaves; the lustre of Her body is bright like Padmar\=agama\d{n}i cut and carved and sharpened on stone. The bracelets are adorned with jewel tinkling bells; Her neck ornaments and medals are studded with gems and jewels. Her hands are resplendent with the lustre of the jewels on the fingers; the braid of hair on Her head is wreathed with a garland of Mallik\=a flowers; Her bodice (short jacket) is studded with various jewels.

30-45. O King! \'Sr\={\i} Dev\={\i} is slightly bent down with the weight of Her very high hard breasts. She has four hands and She is holding noose, goad and signs granting boons and ``do not fear.'' The all beautiful all merciful Dev\={\i} is full of love gestures and beauties. Her voice is sweeter than that of lute; the lustre of Her body is like Kotis and Kotis of Suns and

Moons if they rise simultaneously on the sky. The Sakh\={\i}s, attendants, the Devas and the Dev\={\i}s surround Her on all sides. Ichch\=a \'Sakti, J\~n\=ana \'Sakti, and Kriya \'Sakti all are present always before the Dev\={\i}. Lajj\=a, Tusti, Pusti, K\={\i}rti, K\=anti, K\d{s}am\=a, Day\=a, Buddhi, Medh\=a, Smriti, and Lak\d{s}m\={\i} are always seen here incarnate in their due Forms. The nine P\={\i}tha \'Saktis, Jay\=a, Vijay\=a, Ajit\=a, Apar\=ajit\=a, Nity\=a, Vil\=asin\={\i}, Dogdhr\={\i}, Aghor\=a, and Mangal\=a reside here always and are in the service of the Dev\={\i} Bhuvane\'svar\={\i}. On the side of the Dev\={\i} are the two oceans of treasures; from these streams of Navaratna, gold, and seven Dh\=atus (elements) go out and assume the forms of rivers and fall into the ocean Sudh\=a Sindhu. Because such a Dev\={\i} Bhuvane\'svar\={\i}, resplendent with all powers and prosperities, sits on the left lap of Bhuvane\'svara, that He has, no doubt acquired His omnipotence. O King! Now I will describe the dimensions of the Chint\=ama\d{n}i Griha. Listen. It is one thousand Yojanas wide; its centre is very big; the rooms situated further and further are twice those preceding them. It lies in Antar\={\i}k\d{s}a (the intervening space) without any support. At the times of dissolution and creation it contracts and expands like a cloth. The lustre of this Chint\=ama\d{n}i Griha is comparatively far more bright and beautiful than that of other enclosure walls. \'Sr\={\i} Dev\={\i} Bhagavat\={\i} dwells always in this place. O King! All the great Bhaktas of the Dev\={\i} in every Brahm\=anda, in the Devaloka, in N\=agaloka, in the world of men or in any other loka, all those that were engaged in the meditation of the Dev\={\i} in the sacred places of the Dev\={\i} and died there, they all come here and reside with the Dev\={\i} in great joy and festivity.

46-59. On all sides rivers are flowing; some of ghee, some of milk, curd, honey, nectar, pomegranate juice, jambu juice, and some of mango juice, sugarcane juices are flowing on all sides. The trees here yield fruits according to one's desires and the wells and tanks yield water also as people desire. Never is there any want felt here of anything. Never are seen here diseases, sorrow, old age, decrepitude, anxiety, anger, jealousy, and envy and other lower ideas. All the inhabitants of this place are full of youth and look like one thousand Suns. All enjoy with their wives and they worship \'Sr\={\i} Bhuvane\'svar\={\i}. Some have attained S\=alokya, some S\=am\={\i}pya, some S\=ar\=upya and some have attained S\=arsti and pass their days in highest comfort. The Devas that are in every Brahm\=anda all live here and worship \'Sr\={\i} Dev\={\i}. The seven Koti Mah\=a Mantras and Mah\=a Vidy\=as here assume forms and worship the Mah\=a M\=ay\=a \'Sr\={\i} Bhagavat\={\i}, Who is of the nature of Brahm\=a. O King! Thus I have described to you all about this Ma\d{n}idv\={\i}pa. The lustre of Sun, Moon and Kotis and Kotis of lightnings cannot be one Kotieth of one Koti part of Its lustre. At some places the lustre is like Vidrumama\d{n}i; some places as are illumined like the lustre of Marakata Ma\d{n}i; some, like S\=urya K\=anta ma\d{n}i and some

places are rendered brilliant like Kotis and Kotis of lightnings. The light at some places is like Sind\=ura; at some places like Indran\={\i}lama\d{n}i; at some places, like M\=a\d{n}ikya, and at some places like diamond. Some places are blazing like the conflagration of fire; and some places look like molten gold; some places seem filled with the lustre of Chandrak\=antama\d{n}i, and some places look brilliant like S\=uryak\=antama\d{n}i.

60-73. The mountains here are all built of gems and jewels; the entrance gates and enclosures are built of gems and jewels; the trees and their leaves all are of gems; in fact all that exist here are all of gems and jewels. At some places numbers of peacocks are dancing; at some places cuckoos are captivating the minds of persons by cooing in the fifth tune and at others doves and pigeons and parrots are making sweet cackling sounds. Lakhs and lakhs of tanks are there with their pure crystal-like waters. The Red lotuses have blown fully and enhanced the beauty of the place. The captivating scents of these lotuses extend to a distance one hundred Yojanas all round and gladden the minds of people. The leaves are rustling with gentle breeze. The whole sky overhead is radiant with the lustre of Chint\=ama\d{n}i gems and jewels. All the sides are illuminated with the brilliancy of the gems and jewels. O King! These jewels act like lamps. And the sweet scented trees emit their flagrance and it is transmitted by breeze all around. Thus these trees serve the purpose of dh\=up (scent). The rays of these gems pierce through the openings of the jewel screens on the houses and fall on the mirrors inside, thus causing a nice brilliant appearance that captivates the mind and causes confusion. O King! And what shall I say of this place, more than this, that all the powers, and wealth, all the love sentiments, all the dress suited to amorous interviews, all the splendours, fire, energy, beauty and brilliance, the omniscience, the indomitable strength, all the excellent qualities and all mercy and kindness are present here! The All Comprehending Bliss and the Brahm\=ananda can always be witnessed here! O King! Thus I have described to you about the Ma\d{n}idv\={\i}pa, the most exalted place of the Dev\={\i} Bhagavat\={\i}. At Her remembrance all the sins are instantly destroyed. The more so, if a man remembers the Dev\={\i} and about this place at the time of death, He surely goes there. O King! He who daily reads the five Chapters, i.e., from the eighth to this twelfth chapter, is surely untouched by any obstacles due to the Bh\=utas, Pretas and Pi\'s\=achas. Especially the recitation of this at the time of building a new house and at the time of V\=astuy\=aga ensures all good and auspiciousness.

Here ends the Twelfth Chapter of the Twelfth Book on the description of Ma\d{n}i Dv\={\i}pa in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



