\chapter{On the description of Janamejaya's Dev\={\i} Yaj\~n\=a}

1-4. Vy\=asa said :-- O King Janamejaya! Thus I have answered all your nice queries; also what N\=ar\=aya\d{n}a spoke to the highsouled N\=arada is also said by me. He who hears this greatly wonderful Pur\=a\d{n}am \'Sr\={\i} Dev\={\i} Bh\=agavatam certainly becomes dear to the Dev\={\i} and all his actions become fructified with success. Now as regards your mental distress, how you prevent any evil falling to your late father in his future life, I advise you to do the Yaj\~n\=a in the name of Bhagavat\={\i}; and certainly your father will be saved. And you also better take the Most Excellent Mantra of the Mah\=a Dev\={\i} duly, according to rules; and your human life will then be crowned with success; (your life will be saved; thus you as well as your father will be saved).

5-12. S\=uta said :-- O \d{R}i\d{s}is! Hearing thus, the King asked Vy\=asa Deva to initiate in the Great Dev\={\i} Mantra and thus to become his Guru. He was then initiated duly according to rules with the Great Mantra of Bhagavat\={\i} united with Pra\d{n}ava. When the Navar\=atra period arrived, he called Dhaumya and other Br\=ahma\d{n}as and performed the Navar\=atra Vrata so very dear to the Dev\={\i}, according to his state. At this time for the satisfaction of the Dev\={\i}, he caused this Dev\={\i} Bh\=agavata Pur\=a\d{n}a to be read by the Br\=ahma\d{n}as and fed innumerable Br\=ahma\d{n}as and Kum\=ar\={\i}s (virgins) and gave in charity lots of things to the poor, orphans, and the Br\=ahmi\d{n} boys and thus finished the Vrata. O \d{R}i\d{s}is! Thus completing the Dev\={\i}yaj\~n\=a, while the King was sitting on his seat, the fiery Devar\d{s}i N\=arada came there from above playing with his lute. Seeing him there, all on a sudden, the King got up, and paid due respects to him by asking him to take his seat, with other necessary things. When the Devar\d{s}i became relieved of his labour of journey, the King asked him about his welfare and then enquired into the cause of his coming there.

13-19. O Devar\d{s}i! Whence and what for are you now coming? By your arrival here I am become blessed and feel that my Lord has come to me; now what can I serve to you; kindly command and oblige. Hearing this, the Devar\d{s}i N\=arada said :-- ``O King! Today I saw in the Devaloka a very wonderful event. I wanted eagerly to inform that to you. Hence I have come here. Your father met with a bad turn of fate for his bad action. I saw today he assumed a divine form and he was going on a chariot. The Devas were praising him and the Apsar\=as were encircling him. It seemed he

was going in that dress to the Ma\d{n}i Dv\={\i}pa. O King! You performed the Navar\=atra Vrata and read the Dev\={\i} Bh\=agavata; it seems, as a result of that, your father has now been rewarded with such a noble and good turn of fate. Now you have become blessed and your actions have borne fruits. You have delivered your father from the hell and so you have become an ornament in your family. Today your name and fame have extended to the Devaloka.''

20-30. S\=uta said :-- O \d{R}i\d{s}is! Hearing these words from the mouth of N\=arada, the King Janamejaya became very much happy and delighted and fell prostrate at the feet of Vy\=asa Deva of glorious deeds and said :-- O Best of Munis! By Thy Grace, today I have become blessed. Now what return can I pay to Thee save bowing down to Thee. I pray that Thou dost shew such favours to me ever and again. O \d{R}i\d{s}is! Hearing these words of the King Janamejaya, V\=adar\=aya\d{n}a Veda Vy\=asa blessed him and spoke to him in sweet words :-- ``O King! Now leave all other actions. Read always the Dev\={\i} Bh\=agavata and worship the Lotus Feet of \'Sr\={\i} Dev\={\i}. Leave off all laziness and now perform the Dev\={\i} Yaj\~n\=a with great ecl\=at. And you will surely be able to cross this bondage of the world. True there are various Pur\=a\d{n}as, the Vi\d{s}\d{n}u Pur\=a\d{n}a, the \'Siva Pur\=a\d{n}a, but those cannot compare with one sixteenth of this Dev\={\i} Bh\=agavatam. In fact, this Pur\=a\d{n}a is the Essence of all the Pur\=a\d{n}as. How can the other Pur\=a\d{n}as be compared with this, wherein is established the Dev\={\i} M\=ula Prakriti? Reading this Pur\=a\d{n}a from the beginning to the end yields the result of reading the Vedas. So the wise persons should try their best to study it always.'' Thus saying to Janamejaya, Veda Vy\=asa departed. Then the pure minded Dhaumya and the other Br\=ahma\d{n}as highly praised the Dev\={\i} Bh\=agavatam and went to their desired places. And the King Janamejaya, on the other hand, began to read and hear always the Dev\={\i} Bh\=agavatam and spent his days happily in governing his kingdom.

Here end the Thirteenth Chapter of the Twelfth Book on the description of Janamejaya's Dev\={\i} Yaj\~n\=a in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18000 verses by Mahar\d{s}i Veda Vy\=asa.



