\chapter{On the description of the enclosure walls built of Padmar\=aga ma\d{n}i, etc., of the Ma\d{n}i Dv\={\i}pa}

1-30. Vy\=asa said :-- O King Janamejaya! Next to this Pu\d{s}par\=aga ma\d{n}i enclosure wall comes the tenth enclosure wall, made of Padmar\=aga ma\d{n}i, red like the red Kunkuma and the Rising Sun. It is ten yojanas high. All its ground, entrance gates and temples and arbours are all made of Padmar\=aga ma\d{n}i. Within this reside the sixty four Kal\=as or Sub-\'Saktis adorned with various ornaments and holding weapons in their hands. Each of them has a separate Loka (region) allotted and within this Loka he has get his own formidable weapons, V\=ahanas, families and their

leaders or Governors. O King! Now hear the names of the sixty four Kal\=as. They are :-- Pingal\=ak\d{s}\={\i}, Vi\'s\=al\=ak\d{s}\={\i}, Samriddhi, Vriddhi, \'Sraddh\=a, Sv\=ah\=a, Svadh\=a, M\=ay\=a, Sa\~ng\~n\=a, Vasundhar\=a, Tr\={\i}lokadh\=atr\={\i}, S\=avitr\={\i}, G\=ayatr\={\i}, Trida\'se\'svsr\={\i}, Sur\=up\=a, Bahur\=up\=a, Skandam\=at\=a, Achyutapriy\=a, Vimal\=a, Amal\=a, Aru\d{n}\={\i}, \=Arun\={\i}, Prakriti, Vikriti, \'Sr\={\i}sti, Sthiti, Samrhiti, Sandhy\=a, M\=at\=a, Sat\={\i}, Hams\={\i}, Mardik\=a, Vajrik\=a, Par\=a, Devam\=at\=a, Bhagavat\={\i}, Devak\={\i}, Kamal\=asan\=a, Trimukh\={\i}, Saptamukh\={\i}, Sur\=asura vimardin\={\i}, Lambosth\={\i}, \=Urdhake\'s\={\i}, Bahus\={\i}r\d{s}\=a, Vrikodar\={\i} Ratharekh\=ahvay\=a, \'Sa\'sirek\=a, Gaganaveg\=a, Pavanaveg\=a, Bhuvanap\=al\=a, Madan\=atur\=a, Anang\=a, Anangamathan\=a, Anangamekhal\=a, Anangakusum\=a, Visvar\=up\=a, Sur\=adik\=a, K\d{s}ayamkar\={\i}, Ak\d{s}yobhy\=a, Satyav\=adin\={\i}, Bahur\=up\=a, \'Suchivrat\=a, Ud\=ar\=a and V\=agi\d{s}'\={\i}. These are the sixty four Kal\=as. All of them have got luminous faces and long lolling tongues. Fire is always coming out from the faces of all of them. The eyes of all of them are red with anger. They are uttering :-- We will drink all the water and thus dry up the oceans; we will annihilate fire, we will stop the flow of air and control it. Today we will devour the whole universe and so forth. All of them have got bows and arrows in their hands; all are eager to fight. The four quarters are being reverberated with the clashing of their teeth. The hairs on their heads are all tawny and they stand upwards. Each of them has one hundred Ak\d{s}auhi\d{n}\={\i} forces under them. O King! What more to say than this that each of them has got power to destroy one lakh Brahm\=andas; and their one hundred Ak\d{s}auhi\d{n}\={\i} forces also can do the same. There is nothing that is not impracticable with them. What they cannot do cannot be conceived by mind nor can be uttered in speech. All the war materials exist within their enclosures. Chariots, horses, elephants, weapons, and forces all are unlimited. All the war materials are ready at all times and in abundance. Next comes the eleventh enclosure wall built of Gomedama\d{n}i. It is ten Yojanas high. Its colour is like the newly blown Jav\=a flower. All the ground, trees, tanks, houses, pillars, birds and all other things are all red and built of Gomedama\d{n}i. Here dwell the thirty-two Mah\=a \'Saktis adorned with various ornament made of Gomedama\d{n}i and furnished with various weapons. They are always eager to fight. Their eyes are always red with anger; their bees are like Pi\'s\=achas and their hands are like chakras (discs). ``Pierce him,'' ``Beat him,'' ``Cut him,'' ``Tear him asunder,'' ``Burn him down,'' are the words constantly uttered by them. The inhabitants of the place always worship them. Each of them has ten Ak\d{s}auhi\d{n}\={\i} forces. These are inordinately powerful. It is impossible to describe that. It seems that each \'Sakti can easily destroy one lakh Brahm\=andas.

Innumerable chariots, elephants, hordes, etc., and other v\=ahanas are here. Verily all the war materials of the Dev\={\i} Bhagavat\={\i} are seen in this Gomedama\d{n}i enclosure.

31-51. Now I am mentioning the auspicious, sin destroying names of these \'Saktis :-- Vidy\=a, Hr\={\i}, Pusti, Praj\~n\=a, Sin\={\i} v\=al\={\i}, Kuh\=u, Rudr\=a, Viry\=a, Prabh\=a, Nand\=a, Po\d{s}a\d{n}\={\i}, Riddhid\=a, \'Subh\=a, K\=alar\=atri, Mah\=ar\=atri, Bhadra K\=aI\={\i}, Kaparddin\={\i}, Vikriti, Dandi, Mundin\={\i}, Sendukhand\=a, \'Sikhandin\={\i}, Ni\'sumbha \'sumbha mathan\={\i}, Mahi\d{s}\=asura marddin\={\i}, Indr\=a\d{n}\={\i}, Rudr\=a\d{n}\={\i}, \'Sankar\=ardha sar\={\i}ri\d{n}\={\i}, N\=ar\={\i}, Nir\=aya\d{n}\={\i} Tri\'s\=ulin\={\i}, P\=alin\={\i}, Ambik\=a, and Hl\=adin\={\i}. (See the Dak\d{s}i\d{n}\=a M\=urti Samhit\=a and other Tantras.)

Never there is any chance that they will be defeated anywhere. Hence if all those \'Saktis get angry at any time, this Brahm\=anda ceases to exist. Next to this Gomeda enclosure comes the enclosure made of diamonds. It is ten yojanas high; on all sides there are the entrance gates; the doors are hinged there with nice mechanisms. Nice new diamond trees exist here. All the roads, royal roads, trees, and the spaces for watering their roots, tanks, wells, reservoirs, S\=aranga and other musical instruments are all made of diamonds. Here dwells \'Sr\={\i} Bhuvane\'svar\={\i} Dev\={\i} with Her attendants. O King! Each of them has a lakh attendants. All of them are proud of their beauty. Some of them are holding fans in their hands; some are holding cups for drinking water; some, betelnuts; some are holding umbrellas; some chowries; some are holding various clothings; some flowers; some, looking glasses; some, saffrons; some collyrium, whereas some others are holding Sind\=ura (red lead). Some are ready to do the painting works; some are anxious to champoo the feet; some are eager to make Her wear ornaments; some are anxious to put garlands of flowers on Her neck. All of them are skilled in various arts of enjoyments and they are all young. To gain the Grace of the Dev\={\i}, they consider the whole universe as trifling. Now I shall mention to you the names of the attendants of the Dev\={\i}, proud of their possessing lots of amorous gestures and postures. Listen. They are :-- Anangar\=up\=a, Anangamadan\=a, Madan\=atur\=a, Bhuvanaveg\=a, Bhuvanap\=alik\=a, Sarva\'si\'sira, Anangavedan\=a, Anangamekhal\=a, these are the Eight Sakh\={\i}s. Each of them is as fair as Vidyullat\=a. Each is adorned with various ornaments and skilled in all actions. When they walk to and fro with canes and rods in their hands in the service of the Dev\={\i}, they look as if the lightning flashes glimmer on all sides.

52-71 On the outer portion of the enclosure wall, on the eight sides are situated the dwelling houses of these eight Sakh\={\i}s and they are always full of various v\=ahanas and weapons. Next to this enclosure of diamond

comes the thirteenth enclosure wall made of Vaid\=urya ma\d{n}i. Its height is ten yojanas. There are entrance gates and doorways on the four sides. The court inside, the houses, the big roads, wells, tanks, ponds, rivers and even the sands are all made of Vaid\=urya ma\d{n}i. On the eight sides reside the eight M\=atrik\=as Br\=ahm\={\i}, etc., with their hosts. These M\=atrik\=as represent the sum-total of the individual M\=atrik\=as in every Brahm\=anda. Now hear their names :-- (1) Br\=ahm\={\i}, (2) M\=ahe\'svar\={\i}, (3) Kaum\=ar\={\i}, (4) Vai\d{s}\d{n}av\={\i}, (5) V\=ar\=ah\={\i}, (6) Indr\=a\d{n}\={\i}, (7) Ch\=amund\=a, and (8) Mah\=a Lak\d{s}m\={\i}. Their forms are like those of Brahm\=a and Rudra and others. They are always engaged in doing good to the Universe and reside here with their own V\=ahanas and weapons.

At the four gates, the various V\=ahanas of Bhagavat\={\i} remain always fully equipped. Somewhere there are Kotis and Kotis of elephants. At some places there are Kotis and Kotis of horses; at others there are camps, houses, at others there are swans, lions; at others there are Garudas; at other places there are peacocks, bulls and various other beings all fully equipped and arranged in due order. Similarly the above mentioned animals are yoked to Kotis and Kotis of chariots; there are coachmen (syces); at some places flags are fluttering high on them so as to reach the heavens and thus they are adding beauty. At other places the aerial cars are arranged in rows, countless, with various sounding instruments in them, with flags soaring high in the Heavens and endowed with various ensigns and emblems. O King! Next to this Vaid\=urya enclosure, comes the fourteenth enclosure wall built of Indran\={\i}lama\d{n}i; its height is ten Yojanas. The court inside, houses, roads, wells, tanks and reservoirs, etc., all are built of Indran\={\i}lama\d{n}i. There is here a lotus consisting of sixteen petals extending to many Yojanas in width and shining like a second Sudar\'sana Chakra. On these sixteen petals reside the sixteen \'Saktis of Bhagavat\={\i}, with their hosts. Now I am mentioning the names of these. Hear :-- Kar\=al\={\i}, Vik\=ar\=al\={\i}, Um\=a, Sarasvat\={\i}, \'Sr\={\i}, Durg\=a, \=U\d{s}\=a, Lak\d{s}m\={\i}, \'Sruti, Smriti, Dhriti, \'Sraddh\=a, Medh\=a, Mati, K\=anti, and \=Ary\=a. These are the 16 \'Saktis. They all are dark blue, of the colour of the fresh rain-cloud; they wield in their hands axes and shields. It seems they are ever eager to fight. O King! These \'Saktis are the Rulers of all the separate \'Saktis of the other Brahm\=andas. These are the forces of \'Sr\={\i} Dev\={\i}.

72-90. Being strengthened by the Dev\={\i}'s strength, these are always surrounded by various chariots and forces, various other \'Saktis follow them. If they like, they can cause great agitation in the whole universe. Had I thousand faces, I would not have been able to describe what an

amount of strength they wield. Now I describe the fifteenth enclosure wall. Listen. Next to this Indran\={\i}lama\d{n}i enclosure, comes the enclosure made of pearls (mukt\=a), very wide and ten Yojanas high. The court inside, its space, trees, all are built of pearls. Within this enclosure there is a lotus with eight petals, all of pearls. On these petals reside the eight \'Saktis, the advisers and ministers of the Dev\={\i}. Their appearances, weapons, dresses, enjoyments, everything is like those of \'Sr\={\i} Dev\={\i}. Their duty is to inform the Dev\={\i} of what is going on in the Brahm\=andas. They are skilled in all sciences and arts and clever in all actions. They are very clever, skillful and clever in knowing beforehand the desires and intentions of \'Sr\={\i} Dev\={\i} and they perform those things accordingly. Each one of them has many other \'Saktis who also live here. By their J\~n\=ana \'Sakti they know all the news concerning the J\={\i}vas in every Brahm\=anda. Now I mention the names of those eight Sakh\={\i}s. Listen. Anangakusum\=a, Anangakusum\=a-tur\=a, Anangamadan\=a, Ananga madan\=atur\=a, Bhuva\d{n}ap\=ala, Gaganaveg\=a, \'Sa\'sirekh\=a, and Gaganarekh\=a. These are the eight Sakh\={\i}s. They look red like the Rising Sun; and in their four hands they hold noose, goad, and signs of granting boons and ``no fear.'' At every instant they inform \'Sr\={\i} Dev\={\i} of all the events of the Brahm\=anda. Next to this comes the sixteenth enclosure wall made of emerald (marakata); it is ten Yojanas high; the court inside, its space, and houses and everything are built of emeralds (marakata ma\d{n}i). Here exist all the good objects of enjoyments. This is hexagonal, of the Yantra shape. And at every corner reside the Devas. On the eastern corner resides the four-faced Brahm\=a; he lives with G\=ayatr\={\i} Dev\={\i}; he holds Kamandalu, rosary, signs indicating ``no fear'' and Danda (rod). The Dev\={\i} G\=ayatr\={\i} is also decorated with these. Here all the Vedas, Smritis, the Pur\=a\d{n}as, and various weapons exist incarnate in their respective forms. All the Avat\=aras of Brahm\=a, G\=ayatr\={\i}, and Vy\=ahritis that exist in this Brahm\=anda, all live here. On the south-west corner Mah\=a Vi\d{s}\d{n}u lives with S\=avitr\={\i}; He holds conch shell, disc, club, and lotus. S\=avitr\={\i} has got also all these. The Avat\=aras of Vi\d{s}\d{n}u that exist in every Brahm\=anda Matsya, K\=urma, etc., and all the Avat\=aras of S\=avitr\={\i} that exist in every universe, all dwell in this place. On the north western corner exists Mah\=a Rudra with Sarasvat\={\i}. Both of them hold in their hands Para\'su, rosary, signs granting boons and ``no fear.''

91-110. All the Avat\=aras of Rudra and P\=arvat\={\i} (Gaur\={\i}, etc.) facing south that exist in all the Brahm\=andas, dwell here. All the chief \=Agamas, sixty four in number and all the other Tantras reside here, incarnate in their due forms. On the south-eastern corner, the Lord of wealth, Kuvera, of Bhagavat\={\i}, surrounded by roads

and shops resides here with Mah\=a Lak\d{s}m\={\i} and his hosts holding the jar of jewels (Ma\d{n}i Karandik\=a). On the western corner exists always Madana with Rati, holding noose, goad, bow and arrow. All his amorous attendants reside here, incarnate in their forms. On the north-eastern corner resides always the great hero Ga\d{n}e\'sa, the Remover of obstacles, holding noose and goad and with his Pusti Dev\={\i}. O King! All the Vibh\=utis (manifestations) of Ga\d{n}e\'sa that exist in all the universes reside here. What more to say than this, that Brahm\=a and the other Devas and Dev\={\i}s here represent the sum-total of all the Brahm\=as and the Devas and the Dev\={\i}s that exist in all the Brahm\=andas. These all worship \'Sr\={\i} Bhagavat\={\i}, remaining in their own spheres respectively. O King! Next come the seventeenth enclosure wall made of Prab\=ala. It is red like saffron and it is one hundred Yojanas high. As before, the court inside, the ground and the houses all are made of Prab\=ala. The goddesses of the five elements, Hrillekh\=a, Gagan\=a, Rakt\=a, Kar\=alik\=a, and Mahochchhu\d{s}m\=a reside here. The colours and lustres of the bodies of the goddessses resemble those of the elements over which they preside respectively. All of them are proud of their youth and hold in their four hands noose, goad and signs granting boons and ``no fear.'' They are dressed like \'Sr\={\i} Dev\={\i} and reside here always. Next to this comes the eighteenth enclosure wall built of Navaratna (the nine jewels). It is many yojanas wide. This enclosure wall is superior to all others and it is higher also. On the four sides there exist innumerable houses, tanks, reservoirs, all built of Navaratna; these belong to the Dev\={\i}s, the presiding Deities of \=Amn\=ayas (that which is to be studied or learnt by heart; the Vedas). The ten Mah\=a Vidy\=as, K\=al\={\i}, T\=ar\=a, etc., of \'Sr\={\i} Dev\={\i} and the Mah\=abhed\=as, that is, their all the Avat\=aras all dwell here with their respective \=Avara\d{n}as, V\=ahanas and ornaments. All the Avat\=aras of \'Sr\={\i} Dev\={\i} for the killing of the Daityas and for showing favour to the devotees live here. They are Pa\'samku\'se\'svar\={\i}, Bhuvane\'svar\={\i}, Bhairav\={\i}, Kap\=ala Bhuvane\'svar\={\i}, Amku\'sa Bhuvane\'svar\={\i}, Pram\=ada bhuvane\'svar\={\i}, \'Sr\={\i} Krodha Bhuvane\'svar\={\i}, Triput\=a\'sv\=ar\=udh\=a, Nityaklinn\=a, Annapurn\=a, Tvarit\=a, and the other avat\=aras of Bhuvane\'svar\={\i}, and K\=al\={\i}, T\=ar\=a and the other Mah\=avidy\=as are known as Mah\=avidy\=as. They live here with their \=Avara\d{n}a Devat\=as, V\=ahanas, and ornaments respectively. (Note :-- The \=Avara\d{n}a Deities are the attendant Deities.) Here live also the seven Kotis of Dev\={\i}s presiding over the Mah\=a Mantras, all brilliant and fair like the Koti Suns. O King! Next to this enclosure wall comes the chief and crowning palace of \'Sr\={\i} Dev\={\i}, built of

Chint\=ama\d{n}i gems. All the articles within this are built of Chint\=ama\d{n}i gems. Within this palace are seen hundreds and thousands of pillars. Some of these pillars are built of S\=uryak\=antama\d{n}i, some are built of Chandrak\=anta ma\d{n}i, and some are built of Vidyutk\=anta ma\d{n}i. O King! The lustre and brilliance of these pillars is so strong that no articles within this palace are visible to the eye. (Note :-- The face of the Goddess K\=al\={\i} is so bright that it appears like a shadow, i.e., black.)

Here ends the Eleventh Chapter on the description of the enclosure walls built of Padmar\=aga ma\d{n}i, etc., of the Ma\d{n}i Dv\={\i}pa in the Mah\=apur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.



