\chapter{On the recitation of the fruits of this Pur\=a\d{n}am}

1-17. S\=uta said :-- ``O \d{R}i\d{s}is! In days of yore, from the Lotus Face of the Dev\={\i} Bh\=agavata came out \'Sr\={\i} Mad Bh\=agavatam in the form of half a \'Sloka, as the decided conclusion of the Vedas. About what She gave instructions to Vi\d{s}\d{n}u, sleeping on a leaf of a Banyan tree, that same thing, the seed of the \'Sr\={\i} Mad Bh\=agavata, Brahm\=a Himself expanded into one hundred Koti \'slokas. Then, Veda Vy\=asa, in order to teach his own son \'Suka Deva, condensed them into eighteen thousand \'slokas, in Twelve Books and named it \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam, the present volume. That voluminous book comprising one hundred Koti \'slokas compiled by Brahm\=a are still extant in the Deva loka. There is no Pur\=a\d{n}a like the Dev\={\i} Bh\=agavatam, so merit-giving, holy and capable to destroy all the sins. The reading of every line yields the fruits of performing many A\'svamedha sacrifices. Human beings addicted to worldly affairs will get the merit of giving lands to the Br\=ahma\d{n}as and they will enjoy also all the pleasures of the world and in the end will go to the region of the Dev\={\i}, if they can hear, after they have fasted and controlled their passions, the recitation of this Pur\=a\d{n}am from the mouth of a Paur\=a\d{n}ik Br\=ahma\d{n}a, who has been worshipped and given clothings and ornaments and is considered as a second Veda Vy\=asa. Or, if anybody writes the whole of the Dev\={\i} Bh\=agavatam with his own hand or gets it written by a writer from the beginning to the end and gives to a Paur\=a\d{n}ik Br\=ahmi\d{n} the book placed in a box of the form of a lion made up of gold and a cow yielding milk with her calf with gold as his sacrificial fee; or if he feeds as many Br\=ahma\d{n}as as there are the number of chapters of the Dev\={\i} Bh\=agavatam and worship as many Kum\=ar\={\i}s (virgin girls) with saffron, sandalpaste and ornaments and feeds them with P\=ays\=anna, he gets the merits of giving lands and enjoys all the pleasures of the world and goes in the end to the region of the Dev\={\i}. He has no want of anything who daily hears with rapt devotion this Dev\={\i} Bh\=agavatam. One who has no wealth gets abundance of wealth, those who are students get knowledge, one who has no sons, gets sons if one hears this Dev\={\i} Bh\=agavatam with true devotion. A barren woman, or one who bears still-born children or whose offsprings never live long or who bears only a single child, gets all her defects removed, if she hears this Dev\={\i} Bh\=agavatam with a steadfast devotion. The house where this Pur\=a\d{n}a is worshipped, Lak\d{s}m\={\i} and Sarasvat\={\i} dwell there, leaving their animosities towards each other. By the influence of this Dev\={\i} Bh\=agavatam the D\=akin\={\i}s, Vet\=alas, R\=ak\d{s}asas, and other ghosts cannot cast a glance even on its devotee. If anybody gets fever and if the \'Sr\={\i} Dev\={\i} Bh\=agavatam be read touching him with a concentrated attention, all the complaints disappear. By reading this Bh\=agavatam, one hundred times even more difficult than the severe disease pthisis is cured.

18-20. If after performing the Sandhy\=a, one reads only one chapter of this Bh\=agavatam with a collected mind, he soon acquires the Real Knowledge. O Muni \'Saunaka! While going to read this Bh\=agavatam, first examine omens and then read. I have spoken already on this subject. If during the \'S\=arad\={\i}ya P\=uj\=a (the autumnal Durg\=a P\=uj\=a), at the

Navar\=atra period, one reads with devotion this Bh\=agavatam, the Dev\={\i} Bhagavat\={\i} becomes greatly pleased and awards him results more than his desires.

21-31. During the Navar\=atri period all can read well this book for the satisfaction of his \=Ista Deva (his own deity) whether he be a Vai\d{s}\d{n}ava, \'Saiva, Saura, G\=a\d{n}apatya or a \'S\=akta. All can read this for the satisfaction of Lak\d{s}m\={\i}, Um\=a and other \'Saktis. The Vaidik Br\=ahma\d{n}as are to recite this daily for the satisfaction of the Dev\={\i} G\=ayatr\={\i}. This Pur\=a\d{n}am is not contradictory to any sectarian belief. The reason of this being that to whatever deity he pays his worship, he must worship some \'Sakti or other, this is stated everywhere. So for the satisfaction of one's own \'Sakti, all can read this, without contradicting each other. Never any woman nor any \'S\=udra, is to read this herself or himself, even out of ignorance; rather they should hear this from the mouth of a Br\=ahma\d{n}a. This is the rule of the \'S\=astras. (The vibrations and the consequent results would be truer then.) O \d{R}i\d{s}is! What more to say on this book than this, that this Pur\=a\d{n}am is the most excellent of all and yields great merits. It is the essence of the Vedas. This I tell you with great certainty. There is not the least doubt in this. Reading or hearing this yields results equivalent to reading or hearing the Vedas. I now bow to the Dev\={\i} of the nature of Hr\={\i}m and established by G\=ayatr\={\i}, of the nature of Everlasting Existence, Intelligence and Bliss, Who stimulates our activities to the understanding of various subjects. Thus hearing the excellent words of S\=uta, the great Paur\=a\d{n}ik, all the Munis of Naimi\d{s}\=ara\d{n}ya worshipped him specially and as the result of hearing this Pur\=a\d{n}am glady became the servants of the Lotus Feet of the Dev\={\i} and they attained the Highest Rest. The Munis expressed their humility and gratitude to S\=uta frequently and bowed down to him again and again. And they said :-- ``O S\=uta! It is you that have saved us from this ocean of world.'' Thus (the great Bh\=agavata) S\=uta, the bee drinking the honey of the Lotus Feet of the Dev\={\i}, recited before the assemblage of the best of the Munis this Pur\=a\d{n}am from the beginning to the end, the Secret of all the Nigamas and full of the Glories of the Dev\={\i} Bhagavat\={\i}. After this the \d{R}i\d{s}is bowed down to him and he blessed and honoured them.

Then he went away to his desired place. Here the Dev\={\i} Bh\=agavatam ends and is fully completed.

THE END.

Here ends the Fourteenth Chapter of the Twelfth Book on the recitation of the fruits of this Pur\=a\d{n}am in the Mah\=a Pur\=a\d{n}am \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam of 18,000 verses by Mahar\d{s}i Veda Vy\=asa.

Here ends as well the Full Treatise, \'Sr\={\i} Mad Dev\={\i} Bh\=agavatam.
Om. Om. Om. Om Tat Sat. Om Hari Om.

